### 2021-06-29 01-preliminary.Rmd
getting wordbank data

---
title: "Looking for CDI Bias"
author: "Nathan & George"
date: "6/29/2021"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=F}
knitr::opts_chunk$set(echo = FALSE)

library(wordbankr)
library(tidyverse)
library(ggrepel)
library(mirt) # requires package loading, not just referencing. 
library(mirtCAT)
library(here)
library(kableExtra)
library(gridExtra)
library(printr)
#library(parallel)
#library(doParallel)
#library(mclust)
#library(Rtsne)
#library(factoextra)
```

First, let's get some wordbank data for American English.

```{r get-wordbank-data, echo=T}
get_instruments()

d_demo <- 
  get_administration_data(language = "English (American)", form = "WG") #%>%
  #bind_rows(get_administration_data(language = "English (American)", form = "WS"))

table(d_demo$age)
table(d_demo$mom_ed)

sum(is.na(d_demo$mom_ed)) # many missing

```

We have data from `r nrow(d_demo)` children for the WG form.
<!-- writing reminder -->

```{r get-wordbank-items}
d_long_wg <- get_instrument_data(language="English (American)", form = "WG") %>%
  mutate(produces = as.numeric(value == "produces")) # we'll start with production
# yes = understands; no = does not understand/produce, produces = produces (and understands)

d_mat <- d_long_wg %>% select(data_id, num_item_id, produces) %>% # ToDo: filter out non-word items here
  pivot_wider(id_cols = data_id, names_from = num_item_id, values_from = produces)

en_wg <- get_item_data(language="English (American)", form="WG") %>%
  filter(type=="word")

table(en_wg$category)
```

# DIF Analysis with mirt

```{r}
# https://rdrr.io/cran/mirt/man/DIF.html

mirt(d_mat, 1, "2PL")

# eventually: DIF(multigroupmodel based on sex or mom_ed)
```

### 2021-07-01 01-preliminary.Rmd
Update 01-preliminary.Rmd
minor changes

@@
 knitr::opts_chunk$set(echo = FALSE)
-
+# new libraries GK
+# new line
 library(wordbankr)
@@
 ```
+
+Test

### 2021-07-02 01-preliminary.Rmd
fitting basic 2PL model to WG production

--- 2021-06-29 01-preliminary.Rmd
@@
 title: "Looking for CDI Bias"
-author: "Nathan & George"
-date: "6/29/2021"
+author: "Nathan, George, and Mike"
+date: "`Sys.Date()`"
 output:
@@
 library(ggrepel)
-library(mirt) # requires package loading, not just referencing. 
-library(mirtCAT)
+library(mirt) 
 library(here)
@@
 
-First, let's get some wordbank data for American English.
+First, let's get some wordbank data for American English. We'll start with production data from the CDI: Words & Gestures (WG) form.
+We'll look for DIF based on 1) sex and 2) SES (high vs. low). 
+Questions: How do we choose a dozen anchor items (that we expect to be unbiased)?
 
-```{r get-wordbank-data, echo=T}
-get_instruments()
+```{r get-wordbank-data, eval=F, echo=F}
+# Run Once
+#get_instruments()
 
@@
 
+# 8-18 month-olds
 table(d_demo$age)
+
+# mother's years of education = proxy for SES
 table(d_demo$mom_ed)
+# skewed: only 72 with Primary/Some Secondary (most have Some College +)
 
-sum(is.na(d_demo$mom_ed)) # many missing
+#sum(is.na(d_demo$mom_ed)) # many mom_ed missing
 
@@
 
-We have data from `r nrow(d_demo)` children for the WG form.
-<!-- writing reminder -->
-
-```{r get-wordbank-items}
+```{r get-wordbank-items, eval=F, echo=F}
 d_long_wg <- get_instrument_data(language="English (American)", form = "WG") %>%
@@
 
-d_mat <- d_long_wg %>% select(data_id, num_item_id, produces) %>% # ToDo: filter out non-word items here
-  pivot_wider(id_cols = data_id, names_from = num_item_id, values_from = produces)
+en_wg <- get_item_data(language="English (American)", form="WG") %>%
+  filter(type=="word") %>% select(-complexity_category, -item_id)
 
-en_wg <- get_item_data(language="English (American)", form="WG") %>%
-  filter(type=="word")
+# subjects not producing any words yet (can't fit model with them)
+bad_Ss = d_demo[which(d_demo$production==0),]$data_id
+
+d_long_wg <- d_long_wg %>% left_join(en_wg %>% select(num_item_id, definition)) %>%
+  filter(!is.na(definition))
+
+d_mat <- d_long_wg %>% select(data_id, definition, produces) %>% 
+  pivot_wider(id_cols = data_id, names_from = definition, values_from = produces)
+
+save(d_mat, d_demo, en_wg, file=here("data/en_wg.Rdata"))
+```
+
+```{r load-data}
+# load saved CDI:WG data
+load(here("data/en_wg.Rdata")) # d_demo, d_mat, en_wg (items)
+
+d_demo <- d_demo %>% filter(production!=0) %>%
+  arrange(data_id)
+
+# there are more IDs in d_mat than in d_demo; remove those
+d_mat <- d_mat %>% filter(is.element(data_id, d_demo$data_id)) %>% arrange(data_id)
+
+sids = d_mat$data_id
+d_mat$data_id = NULL
+d_mat = data.matrix(d_mat)
+row.names(d_mat) = sids
 
@@
 
-# DIF Analysis with mirt
+We have data from `r nrow(d_demo)` children for the WG form.
+
+## Fit baseline 2PL model
+
+```{r}
+m1 <- mirt(d_mat, 1, "2PL", 
+           technical=list(NCYCLES=1000))
+
+coefs_2pl <- as_tibble(coef(m1, simplify = TRUE)$items) %>%
+  mutate(definition = rownames(coef(m1, simplify = TRUE)$items))
+fscores_2pl <- tibble(data_id = rownames(d_mat), 
+                             ability = fscores(m1, method = "MAP")[,1])
+
+save(file = here("data/eng_wg_mod_2pl.Rds"), "m1", "coefs_2pl") # "fscores_2pl",
+```
+
+# DIF Model: Sex
 
@@
 # https://rdrr.io/cran/mirt/man/DIF.html
+# https://rdrr.io/cran/mirt/man/multipleGroup.html
 
-mirt(d_mat, 1, "2PL")
+sex_group = d_demo$sex
 
-# eventually: DIF(multigroupmodel based on sex or mom_ed)
+mg1 <- multipleGroup(dat, model = 1, group = sex_group, verbose = FALSE) 
+mg2 <- multipleGroup(dat, model = 1, group = sex_group, 
+                     invariance = 'slopes', verbose = FALSE)
+
+anova(mg2, mg1)
 ```

### 2021-07-02 01-preliminary.Rmd
figuring out multigroup

@@
 
-d_demo <- d_demo %>% filter(production!=0) %>%
+d_demo <- d_demo %>% 
+  filter(production!=0, !is.na(sex)) %>% # can't fit children not producing words, or with NA sex in group model
   arrange(data_id)
@@
 
-sex_group = d_demo$sex
+sex_group = as.character(d_demo$sex)
 
-mg1 <- multipleGroup(dat, model = 1, group = sex_group, verbose = FALSE) 
-mg2 <- multipleGroup(dat, model = 1, group = sex_group, 
+# mg1 <- multipleGroup(d_mat, model = 1, group = sex_group, verbose = FALSE) 
+# Error: Multiple Group model will not be identified without proper constraints (groups contain missing data patterns where item responses have been completely omitted or, alternatively, the number of categories within each group is not equal to the total number of categories)
+mg2 <- multipleGroup(d_mat, model = 1, group = sex_group, 
                      invariance = 'slopes', verbose = FALSE)

### 2021-07-02 01-preliminary.Rmd
trying out multipleGroups

--- 2021-07-02 01-preliminary.Rmd
@@
 
-sex_group = d_demo$sex
+sex_group = as.character(d_demo$sex)
 
-mg1 <- multipleGroup(dat, model = 1, group = sex_group, verbose = FALSE) 
-mg2 <- multipleGroup(dat, model = 1, group = sex_group, 
-                     invariance = 'slopes', verbose = FALSE)
+# mg1 <- multipleGroup(d_mat, model = 1, group = sex_group, verbose = FALSE) 
+# Error: Multiple Group model will not be identified without proper constraints (groups contain missing data patterns where item responses have been completely omitted or, alternatively, the number of categories within each group is not equal to the total number of categories)
+mg2 <- multipleGroup(d_mat, model = 1, group = sex_group, technical=list(NCYCLES=1000), 
+                     invariance = 'slopes', verbose = T)
 
-anova(mg2, mg1)
+m_sc2 <- multipleGroup(d_mat, 1, group = sex_group,
+                             invariance=c('slopes', 'intercepts', 'free_var','free_means'))
+mod_sc1 <- multipleGroup(d_mat, 1, group = sex_group,  #fixed means
+                             invariance=c('slopes', 'intercepts', 'free_var'))
+mod_fullconstrain <- multipleGroup(d_mat, 1, group = sex_group,
+                             invariance=c('slopes', 'intercepts'))
+
+extract.mirt(mod_fullconstrain, 'time') #time of estimation components
+
+save(file = here("data/eng_wg_sex_slopes_invariant.Rds"), mg2)
+
+anova(m1, mg2) # AIC prefers group model, BIC prefers 2PL
+mg2s <- summary(mg2)
+fem = mg2s[[1]]$rotF
+mal = mg2s[[2]]$rotF
+
+plot(x=fem, y=mal) # slopes are constrained to be the same...
+# where are the difficulties? 
 ```

### 2021-07-06 01-preliminary.Rmd
looking at item DIF

@@
 m1 <- mirt(d_mat, 1, "2PL", 
-           technical=list(NCYCLES=1000))
+           technical=list(NCYCLES=500))
 
@@
 
-# mg1 <- multipleGroup(d_mat, model = 1, group = sex_group, verbose = FALSE) 
+# ToDo: define ses_group ("high" for mom_ed > X, "low") (choose X ..high school?)
+
+mg1 <- multipleGroup(d_mat, 1, group = sex_group, SE = T) 
 # Error: Multiple Group model will not be identified without proper constraints (groups contain missing data patterns where item responses have been completely omitted or, alternatively, the number of categories within each group is not equal to the total number of categories)
@@
                      invariance = 'slopes', verbose = T)
+
+DIF(mg2, c('d')) # No hyper-parameters were estimated in the DIF model. For effective DIF testing, freeing the focal group hyper-parameters is recommended.
 
@@
 
+residuals(mod_sc1)
+
 save(file = here("data/eng_wg_sex_slopes_invariant.Rds"), mg2)
+
+Mit = as_tibble(coef(mg2, simplify=T)$Male$items) %>%
+  mutate(definition = rownames(coef(mg2, simplify=T)$Male$items)) %>%
+  select(-g, -u) %>% 
+  rename(d_m = d)
+Fit = as_tibble(coef(mg2, simplify=T)$Female$items) %>%
+  mutate(definition = rownames(coef(mg2, simplify=T)$Female$items)) %>%
+  select(-g, -u) %>%
+  rename(d_f = d)
+
+combo <- Mit %>% left_join(Fit) %>%
+  mutate(d_diff = d_f - d_m,
+         d_diff_abs = abs(d_diff))
+
+thresh = median(combo$d_diff_abs) + sd(combo$d_diff_abs)
 

### 2021-07-07 01-preliminary.Rmd
fitted more sex DIF models

@@
 author: "Nathan, George, and Mike"
-date: "`Sys.Date()`"
+date: "`r Sys.Date()`"
 output:
+  html_document: default
   pdf_document: default
-  html_document: default
 ---
@@
 library(here)
-library(kableExtra)
+#library(kableExtra)
 library(gridExtra)
@@
 
-```{r}
+```{r, eval=F}
 m1 <- mirt(d_mat, 1, "2PL", 
@@
 
-```{r}
+```{r, eval=F}
 # https://rdrr.io/cran/mirt/man/DIF.html
@@
 
-save(file = here("data/eng_wg_sex_slopes_invariant.Rds"), mg2)
+save(file = here("data/eng_wg_sex_slopes_invariant.Rds"), 
+     mg2, m_sc2, mod_sc1, mod_fullconstrain)
+```
 
-Mit = as_tibble(coef(mg2, simplify=T)$Male$items) %>%
-  mutate(definition = rownames(coef(mg2, simplify=T)$Male$items)) %>%
+```{r}
+load(here("data/eng_wg_sex_slopes_invariant.Rds"))
+load(here("data/eng_wg_mod_2pl.Rds"))
+
+extract_group_df <- function(group_model) {
+  Mit = as_tibble(coef(group_model, simplify=T)$Male$items) %>%
+    mutate(definition = rownames(coef(group_model, simplify=T)$Male$items)) %>%
     select(-g, -u) %>% 
     rename(d_m = d)
-Fit = as_tibble(coef(mg2, simplify=T)$Female$items) %>%
-  mutate(definition = rownames(coef(mg2, simplify=T)$Female$items)) %>%
+  Fit = as_tibble(coef(group_model, simplify=T)$Female$items) %>%
+    mutate(definition = rownames(coef(group_model, simplify=T)$Female$items)) %>%
     select(-g, -u) %>%
@@
            d_diff_abs = abs(d_diff)) 
+  return(combo)
+}
 
-thresh = median(combo$d_diff_abs) + sd(combo$d_diff_abs)
+mg2_df = extract_group_df(mg2)
 
-anova(m1, mg2) # AIC prefers group model, BIC prefers 2PL
-mg2s <- summary(mg2)
-fem = mg2s[[1]]$rotF
-mal = mg2s[[2]]$rotF
+# same difficulties:
+mod_sc1_df = extract_group_df(mod_sc1)
 
-plot(x=fem, y=mal) # slopes are constrained to be the same...
-# where are the difficulties? 
+#anova(m1, mg2) # AIC prefers group model, BIC prefers 2PL
 ```
+
+### Items with a large difference in difficulty across sex
+
+```{r}
+# items showing > median + 1SD of absolute difference in difficulty have a lot of DIF
+thresh = median(mg2_df$d_diff_abs) + sd(mg2_df$d_diff_abs)
+big_dif = mg2_df[which(mg2_df$d_diff_abs > thresh),]
+
+#mg2s <- summary(mg2)
+#fem = mg2s[[1]]$rotF
+#mal = mg2s[[2]]$rotF
+big_dif %>% arrange(desc(d_diff_abs)) #%>% kable()
+```
+
+### Plot item ease for males vs. females
+
+We label the `r nrow(big_dif)` items with absolute ease difference of more than median + 1SD = `r round(thresh,2)`.
+
+```{r, fig.width=8, fig.height=6}
+ggplot(mg2_df, 
+       aes(x = d_m, y = d_f)) + 
+  geom_point(alpha=.7) + 
+  ggrepel::geom_label_repel(data = filter(mg2_df, d_diff_abs > thresh), 
+                            aes(label = definition), max.overlaps = 20) + theme_bw() +
+  xlab("Item Ease for Males") + 
+  ylab("Item Ease for Females") + 
+  geom_abline(intercept=0, slope=1, linetype='dashed')
+```

### 2021-07-07 01-preliminary-WGprod.Rmd
fitting DIF sex to WG comprehension

--- 2021-07-07 01-preliminary.Rmd
@@
 ---
-title: "Looking for CDI Bias"
+title: "Looking for CDI Bias: WG Production"
 author: "Nathan, George, and Mike"

### 2021-07-07 02-preliminary-WGcomp.Rmd
fitting DIF sex to WG comprehension

--- 2021-07-07 01-preliminary.Rmd
@@
 ---
-title: "Looking for CDI Bias"
+title: "Looking for CDI Bias: WG Comprehension"
 author: "Nathan, George, and Mike"
@@
 
+Let's do comprehension data.
+
 ```{r get-wordbank-items, eval=F, echo=F}
 d_long_wg <- get_instrument_data(language="English (American)", form = "WG") %>%
-  mutate(produces = as.numeric(value == "produces")) # we'll start with production
+  mutate(produces = as.numeric(value == "yes" | value == "produces")) # we'll start with production
 # yes = understands; no = does not understand/produce, produces = produces (and understands)
@@
 
-save(d_mat, d_demo, en_wg, file=here("data/en_wg.Rdata"))
+save(d_mat, d_demo, en_wg, file=here("data/en_wg_comp.Rdata"))
 ```
@@
 # load saved CDI:WG data
-load(here("data/en_wg.Rdata")) # d_demo, d_mat, en_wg (items)
+load(here("data/en_wg_comp.Rdata")) # d_demo, d_mat, en_wg (items)
 
@@
 
-save(file = here("data/eng_wg_mod_2pl.Rds"), "m1", "coefs_2pl") # "fscores_2pl",
+save(file = here("data/eng_wg_comp_mod_2pl.Rds"), "m1", "coefs_2pl") # "fscores_2pl",
 ```
@@
 
-save(file = here("data/eng_wg_sex_slopes_invariant.Rds"), 
+save(file = here("data/eng_wg_comp_sex_sexDIF.Rds"), 
      mg2, m_sc2, mod_sc1, mod_fullconstrain)
@@
 ```{r}
-load(here("data/eng_wg_sex_slopes_invariant.Rds"))
-load(here("data/eng_wg_mod_2pl.Rds"))
+load(here("data/eng_wg_comp_sex_sexDIF.Rds"))
+load(here("data/eng_wg_comp_mod_2pl.Rds"))
 

### 2021-07-08 02-preliminary-WGcomp.Rmd
ran basic WG comprehension DIF models

@@
 # subjects not producing any words yet (can't fit model with them)
-bad_Ss = d_demo[which(d_demo$production==0),]$data_id
+bad_Ss = d_demo[which(d_demo$comprehension==0),]$data_id # 8 bad comprehension Ss
 
@@
 d_demo <- d_demo %>% 
-  filter(production!=0, !is.na(sex)) %>% # can't fit children not producing words, or with NA sex in group model
+  filter(comprehension!=0, !is.na(sex)) %>% # can't fit children not producing words, or with NA sex in group model
   arrange(data_id)
@@
 
-We have data from `r nrow(d_demo)` children for the WG form.
+We will fit comprehension data from `r nrow(d_demo)` children for the WG form.
 
@@
 m1 <- mirt(d_mat, 1, "2PL", 
-           technical=list(NCYCLES=500))
+           technical=list(NCYCLES=1000))
 
@@
 # Error: Multiple Group model will not be identified without proper constraints (groups contain missing data patterns where item responses have been completely omitted or, alternatively, the number of categories within each group is not equal to the total number of categories)
-mg2 <- multipleGroup(d_mat, model = 1, group = sex_group, technical=list(NCYCLES=1000), 
+mg2 <- multipleGroup(d_mat, model = 1, group = sex_group, technical=list(NCYCLES=4000), 
                      invariance = 'slopes', verbose = T)
 
-DIF(mg2, c('d')) # No hyper-parameters were estimated in the DIF model. For effective DIF testing, freeing the focal group hyper-parameters is recommended.
+#DIF(mg2, c('d')) # No hyper-parameters were estimated in the DIF model. For effective DIF testing, freeing the focal group hyper-parameters is recommended.
 
-m_sc2 <- multipleGroup(d_mat, 1, group = sex_group,
+m_sc2 <- multipleGroup(d_mat, 1, group = sex_group, technical=list(NCYCLES=2000),
                        invariance=c('slopes', 'intercepts', 'free_var','free_means'))
-mod_sc1 <- multipleGroup(d_mat, 1, group = sex_group,  #fixed means
+mod_sc1 <- multipleGroup(d_mat, 1, group = sex_group, technical=list(NCYCLES=2000), #fixed means
                              invariance=c('slopes', 'intercepts', 'free_var'))
-mod_fullconstrain <- multipleGroup(d_mat, 1, group = sex_group,
+mod_fullconstrain <- multipleGroup(d_mat, 1, group = sex_group, technical=list(NCYCLES=2000),
                              invariance=c('slopes', 'intercepts'))
 
-extract.mirt(mod_fullconstrain, 'time') #time of estimation components
+#extract.group(m_sc2, "Male")
+#extract.group(m_sc2, "Female")
 

### 2021-07-10 02-preliminary-WGcomp.Rmd
looking at ability distributions

@@
 library(here)
-#library(kableExtra)
+library(kableExtra)
 library(gridExtra)
 library(printr)
+library(GGally)
 #library(parallel)
@@
 
-save(file = here("data/eng_wg_comp_mod_2pl.Rds"), "m1", "coefs_2pl") # "fscores_2pl",
+save(file = here("data/eng_wg_comp_mod_2pl.Rds"), "m1", "coefs_2pl", "fscores_2pl") # ,
 ```
@@
 #mal = mg2s[[2]]$rotF
-big_dif %>% arrange(desc(d_diff_abs)) #%>% kable()
+big_dif %>% arrange(desc(d_diff_abs)) %>% kable(digits = 2)
 ```
@@
 ```
+
+## Ability distribution comparison
+
+What are the ability distributions by sex in the 2PL model?
+
+```{r}
+d_demo <- d_demo %>% left_join(fscores_2pl %>% mutate(data_id = as.numeric(data_id)))
+
+d_demo %>% ggplot(aes(x=ability, fill=sex)) + 
+  geom_density(alpha=.3) + 
+  theme_classic()
+```
+
+```{r}
+# get ability scores from sex group model with varying difficulty
+fscores_mg2 <- tibble(data_id = rownames(d_mat), 
+                             ability_sex = fscores(mg2, method = "MAP")[,1])
+
+d_demo <- d_demo %>% left_join(fscores_mg2 %>% 
+                                 mutate(data_id = as.numeric(data_id)))
+
+ggpairs(d_demo, columns = c("age","comprehension","ability","ability_sex"),
+        ggplot2::aes(colour=sex, alpha=.5)) + theme_classic()
+```
+
+```{r}
+d_demo %>% group_by(sex) %>% 
+  summarise(ability_2pl = mean(ability),
+            ability_sex = mean(ability_sex)) %>%
+  kable(digits=2)
+```
+
+Average male ability is slightly higher in the sex group model (i.e., there is a larger female advantage in the standard 2PL model), showing that the 36 items that are easier for females boost female ability estimates.

### 2021-07-16 02-preliminary-WGcomp.Rmd
fitted SES DIF model

@@
 row.names(d_mat) = sids
+
+table(d_demo$mom_ed)
+# high SES = mom_ed >= Some College (739 high, 324 low-SES)
+d_demo <- d_demo %>% mutate(ses_group = ifelse(as.numeric(mom_ed)>4, "high", "low"))
 
@@
 Average male ability is slightly higher in the sex group model (i.e., there is a larger female advantage in the standard 2PL model), showing that the 36 items that are easier for females boost female ability estimates.
+
+## DIF Model: SES
+
+```{r, eval=F}
+ses_group = as.character(d_demo$ses_group)
+
+#mg1 <- multipleGroup(d_mat, 1, group = ses_group, SE = T) 
+mg2 <- multipleGroup(d_mat, model = 1, group = ses_group, technical=list(NCYCLES=4000), 
+                     invariance = 'slopes', verbose = T)
+
+#DIF(mg2, c('d')) # No hyper-parameters were estimated in the DIF model. For effective DIF testing, freeing the focal group hyper-parameters is recommended.
+
+m_sc2 <- multipleGroup(d_mat, 1, group = ses_group, technical=list(NCYCLES=2000),
+                       invariance=c('slopes', 'intercepts', 'free_var','free_means'))
+mod_sc1 <- multipleGroup(d_mat, 1, group = ses_group, technical=list(NCYCLES=2000), #fixed means
+                             invariance=c('slopes', 'intercepts', 'free_var'))
+mod_fullconstrain <- multipleGroup(d_mat, 1, group = ses_group, technical=list(NCYCLES=2000),
+                             invariance=c('slopes', 'intercepts'))
+
+#extract.group(m_sc2, "high")
+#extract.group(m_sc2, "low")
+
+residuals(mod_sc1)
+
+save(file = here("data/eng_wg_comp_SES_DIF.Rds"), 
+     mg2, m_sc2, mod_sc1, mod_fullconstrain)
+```

### 2021-07-16 02-preliminary-WGcomp.Rmd
plotting high vs. low-SES DIF

@@
 #mal = mg2s[[2]]$rotF
-big_dif %>% arrange(desc(d_diff_abs)) %>% kable(digits = 2)
+big_dif %>% relocate(definition) %>%
+  arrange(desc(d_diff_abs)) %>% kable(digits = 2, format = "html", table.attr = "style='width:40%;'")
 ```
@@
 ```
+
+```{r}
+load(here("data/eng_wg_comp_SES_DIF.Rds"))
+
+extract_ses_group_df <- function(group_model) {
+  Mit = as_tibble(coef(group_model, simplify=T)$high$items) %>%
+    mutate(definition = rownames(coef(group_model, simplify=T)$high$items)) %>%
+    select(-g, -u) %>% 
+    rename(d_hi = d)
+  Fit = as_tibble(coef(group_model, simplify=T)$low$items) %>%
+    mutate(definition = rownames(coef(group_model, simplify=T)$low$items)) %>%
+    select(-g, -u) %>%
+    rename(d_lo = d)
+
+  combo <- Mit %>% left_join(Fit) %>%
+    mutate(d_diff = d_hi - d_lo,
+           d_diff_abs = abs(d_diff)) 
+  return(combo)
+}
+
+mg2_df = extract_ses_group_df(mg2)
+```
+
+### Items with a large difference in difficulty across SES
+
+```{r}
+# items showing > median + 1SD of absolute difference in difficulty have a lot of DIF
+thresh = median(mg2_df$d_diff_abs) + sd(mg2_df$d_diff_abs)
+big_dif = mg2_df[which(mg2_df$d_diff_abs > thresh),]
+
+#mg2s <- summary(mg2)
+big_dif %>% relocate(definition) %>% 
+  arrange(desc(d_diff_abs)) %>% kable(digits = 2, format = "html", table.attr = "style='width:40%;'")
+```
+
+### Plot item ease for high vs. low SES
+
+We label the `r nrow(big_dif)` items with absolute ease difference of more than median + 1SD = `r round(thresh,2)`.
+
+```{r, fig.width=8, fig.height=6}
+ggplot(mg2_df, 
+       aes(x = d_lo, y = d_hi)) + 
+  geom_point(alpha=.7) + 
+  ggrepel::geom_label_repel(data = filter(mg2_df, d_diff_abs > thresh), 
+                            aes(label = definition), max.overlaps = 20) + theme_bw() +
+  xlab("Item Ease for low-SES") + 
+  ylab("Item Ease for high-SES") + 
+  geom_abline(intercept=0, slope=1, linetype='dashed')
+```

### 2021-07-19 ScriptRice.R
RScript for remote computers

@@ -1,0 +1,64 @@
+library(wordbankr)
+library(tidyverse)
+library(ggrepel)
+library(dbplyr)
+library(mirt) 
+library(here)
+library(kableExtra)
+library(gridExtra)
+library(printr)
+library(parallel)
+library(tidyr)
+
+
+
+d_demo <- 
+  get_administration_data(language = "English (American)", form = "WG")
+
+d_long_wg <- get_instrument_data(language="English (American)", form = "WG") %>%
+  mutate(produces = as.numeric(value == "produces")) # we'll start with production
+# yes = understands; no = does not understand/produce, produces = produces (and understands)
+
+en_wg <- get_item_data(language="English (American)", form="WG") %>%
+  filter(type=="word") %>% select(-complexity_category, -item_id)
+
+# subjects not producing any words yet (can't fit model with them)
+bad_Ss = d_demo[which(d_demo$production==0),]$data_id
+
+d_long_wg <- d_long_wg %>% left_join(en_wg %>% select(num_item_id, definition)) %>%
+  filter(!is.na(definition))
+
+d_mat <- d_long_wg %>% select(data_id, definition, produces) %>% 
+  pivot_wider(id_cols = data_id, names_from = definition, values_from = produces)
+
+d_demo <- d_demo %>% 
+  filter(production!=0, !is.na(sex)) %>% # can't fit children not producing words, or with NA sex in group model
+  arrange(data_id)
+
+# there are more IDs in d_mat than in d_demo; remove those
+d_mat <- d_mat %>% filter(is.element(data_id, d_demo$data_id)) %>% arrange(data_id)
+
+sids = d_mat$data_id
+d_mat$data_id = NULL
+d_mat = data.matrix(d_mat)
+row.names(d_mat) = sids
+
+table(en_wg$category)
+
+sex_group = as.character(d_demo$sex)
+
+# ToDo: define ses_group ("high" for mom_ed > X, "low") (choose X ..high school?)
+
+#mg1 <- multipleGroup(d_mat, 1, group = sex_group, SE = T) 
+# Error: Multiple Group model will not be identified without proper constraints (groups contain missing data patterns where item responses have been completely omitted or, alternatively, the number of categories within each group is not equal to the total number of categories)
+#mg2 <- multipleGroup(d_mat, model = 1, group = sex_group, technical=list(NCYCLES=1000), 
+#invariance = 'slopes', verbose = T)
+
[…]

### 2021-07-30 01-preliminary-WGprod.Rmd
multipleGroup model comparisons

@@
 
+# Get DIF Statistics
+
+```{r, eval=F, include=F}
+# which.par = c('a1','d')
+# items2test = 1  # specify which items to test -- can split up 
+# scheme = 'add' # No hyper-parameters were estimated in the DIF model. For effective DIF testing, freeing the focal group hyper-parameters is recommended.
+# mirtCluster() # allows parallel running of models
+start_time = Sys.time()
+d_s <- DIF(mg2, which.par='d', items2test = 1:2, scheme='drop')
+end_time = Sys.time()
+```
+
 ### Items with a large difference in difficulty across sex

### 2021-07-30 02-preliminary-WGcomp.Rmd
multipleGroup model comparisons

@@
 # mother's years of education = proxy for SES
-table(d_demo$mom_ed)
+table(d_demo$mom_ed) %>% kable()
 # skewed: only 72 with Primary/Some Secondary (most have Some College +)
@@
 
-table(en_wg$category)
+table(en_wg$category) %>% kable()
 ```
@@
 
-mg1 <- multipleGroup(d_mat, 1, group = sex_group, SE = T) 
+mg1 <- multipleGroup(d_mat, 1, group = sex_group) # this example should work, according to docs
+mg1 <- multipleGroup(d_mat, 1, group = sex_group, 
+                     invariance = c("free_var", "free_means")) 
 # Error: Multiple Group model will not be identified without proper constraints (groups contain missing data patterns where item responses have been completely omitted or, alternatively, the number of categories within each group is not equal to the total number of categories)
+
+# unless there are possibly items
+
 mg2 <- multipleGroup(d_mat, model = 1, group = sex_group, technical=list(NCYCLES=4000), 
                      invariance = 'slopes', verbose = T)
-
 #DIF(mg2, c('d')) # No hyper-parameters were estimated in the DIF model. For effective DIF testing, freeing the focal group hyper-parameters is recommended.
@@
                        invariance=c('slopes', 'intercepts', 'free_var','free_means'))
-mod_sc1 <- multipleGroup(d_mat, 1, group = sex_group, technical=list(NCYCLES=2000), #fixed means
+
+# constrain slopes, freely estimate latent group means (reference group constrained to a vector of 0)
+mg3 <- multipleGroup(d_mat, model = 1, group = sex_group, technical=list(NCYCLES=4000), 
+                     invariance =c('slopes', 'free_means'), verbose = T)
+
+# constrain slopes, allow varying group means and variances -- very slow!
+#mg4 <- multipleGroup(d_mat, model = 1, group = sex_group, technical=list(NCYCLES=4000), 
+#                     invariance =c('slopes','free_means', 'free_var'), verbose = T)
+
+# fixed means, varying group variance
+mod_sc1 <- multipleGroup(d_mat, 1, group = sex_group, technical=list(NCYCLES=2000),
                              invariance=c('slopes', 'intercepts', 'free_var'))
+# fully-constrained
 mod_fullconstrain <- multipleGroup(d_mat, 1, group = sex_group, technical=list(NCYCLES=2000),
                              invariance=c('slopes', 'intercepts'))
+
+# constrain intercepts to be equal across groups, but allow group slopes to vary
+m_int <- multipleGroup(d_mat, 1, group = sex_group, technical=list(NCYCLES=2000),
+                     invariance = 'intercepts')
+
+anova(mod_sc1, mod_fullconstrain) # mod_fullconstrain preferred by AIC/BIC
+anova(m_sc2, mod_fullconstrain) # mod_fullconstrain preferred by AIC/BIC
+
+anova(mg2, mod_fullconstrain) # AIC prefers slope-invariant model; BIC prefers fully-constrained model
+
+anova(m_int, mod_fullconstrain) # AIC prefers intercept-constrained model; BIC prefers fully-constrained
+
+anova(mg3, mg2) # AIC//BIC prefer model with varying group means
+#anova(mg3, mg4)
+
+coef(mg3)[[2]]$GroupPars # Female intercept = 0.22 (Male = 0)
 
@@
 save(file = here("data/eng_wg_comp_sex_sexDIF.Rds"), 
-     mg2, m_sc2, mod_sc1, mod_fullconstrain)
+     mg2, mg3,  m_sc2, mod_sc1, mod_fullconstrain) # not run: mg4
 ```
@@
 
+# slope-invariant, varying difficulty:
 mg2_df = extract_group_df(mg2)
+mg3_df = extract_group_df(mg3)
+
+# same item estimates with/without free_means? 
+cor(mg2_df$d_f, mg3_df$d_f) # .9999
+cor(mg2_df$d_m, mg3_df$d_m) # .9998
 
@@
 mod_sc1_df = extract_group_df(mod_sc1)
+
+# fully-constrained is just the same as ordinary 2PL model, no? but we can use this model for DIF
+mod_fullconstrain_df = extract_group_df(mod_fullconstrain)
+# DIF(mod_fullconstrain, which.par = 'd')
+
+#mod_int_df = extract_group_df(m_int)
+
+mod_sc2_df = extract_group_df(m_sc2)
 
@@
 # items showing > median + 1SD of absolute difference in difficulty have a lot of DIF
-thresh = median(mg2_df$d_diff_abs) + sd(mg2_df$d_diff_abs)
-big_dif = mg2_df[which(mg2_df$d_diff_abs > thresh),]
+thresh = median(mg3_df$d_diff_abs) + sd(mg3_df$d_diff_abs)
+big_dif = mg3_df[which(mg3_df$d_diff_abs > thresh),]
 
@@
 ```{r, fig.width=8, fig.height=6}
-ggplot(mg2_df, 
+ggplot(mg3_df, 
        aes(x = d_m, y = d_f)) + 
@@
 # get ability scores from sex group model with varying difficulty
+fscores_mg3 <- tibble(data_id = rownames(d_mat), 
+                             ability_sex = fscores(mg3, method = "MAP")[,1])
+
 fscores_mg2 <- tibble(data_id = rownames(d_mat), 
                              ability_sex = fscores(mg2, method = "MAP")[,1])
+
+# very similar
+#plot(fscores_mg3$ability_sex, fscores_mg2$ability_sex, pch='.')
 
@@
             ability_sex = mean(ability_sex)) %>%
-  kable(digits=2)
+  kable(digits = 2, format = "html", table.attr = "style='width:30%;'")
 ```
@@
                      invariance = 'slopes', verbose = T)
+
+mg3 <- multipleGroup(d_mat, model = 1, group = ses_group, technical=list(NCYCLES=4000), 
+                     invariance = c('slopes', 'free_means'), verbose = T)
+
+anova(mg2, mg3) # AIC/BIC prefer model with varying group mean
 
@@
 save(file = here("data/eng_wg_comp_SES_DIF.Rds"), 
-     mg2, m_sc2, mod_sc1, mod_fullconstrain)
+     mg2, mg3, m_sc2, mod_sc1, mod_fullconstrain)
 ```
@@
 load(here("data/eng_wg_comp_SES_DIF.Rds"))
+load(here("data/eng_wg_comp_mod_2pl.Rds"))
 
@@
 mg2_df = extract_ses_group_df(mg2)
+mg3_df = extract_ses_group_df(mg3)
+
+coef(mg3)[[2]]$GroupPars # low-SES group (high is ref) intercept = .24
 ```
@@
 ```{r, fig.width=8, fig.height=6}
-ggplot(mg2_df, 
+ggplot(mg3_df, 
        aes(x = d_lo, y = d_hi)) + 
   geom_point(alpha=.7) + 
-  ggrepel::geom_label_repel(data = filter(mg2_df, d_diff_abs > thresh), 
+  ggrepel::geom_label_repel(data = filter(mg3_df, d_diff_abs > thresh), 
                             aes(label = definition), max.overlaps = 20) + theme_bw() +
@@
 ```
+
+## Ability distribution comparison
+
+What are the ability distributions by SES in the 2PL model?
+
+```{r}
+d_demo <- d_demo %>% left_join(fscores_2pl %>% mutate(data_id = as.numeric(data_id)))
+
+d_demo %>% ggplot(aes(x=ability, fill=ses_group)) + 
+  geom_density(alpha=.3) + 
+  theme_classic()
+```
+
+```{r}
+# get ability scores from sex group model with varying difficulty
+fscores_mg2 <- tibble(data_id = rownames(d_mat[which(!is.na(d_demo$ses_group)),]), 
+                             ability_ses = fscores(mg2, method = "MAP")[,1])
+
+d_demo_ses <- d_demo %>% 
+  filter(!is.na(ses_group)) %>% 
+  left_join(fscores_mg2 %>% mutate(data_id = as.numeric(data_id)))
+
+ggpairs(d_demo_ses, columns = c("age","comprehension","ability","ability_ses"),
+        ggplot2::aes(colour=ses_group, alpha=.5)) + theme_classic()
+```
+
+```{r, echo=F}
+d_demo_ses %>% group_by(ses_group) %>% 
+  summarise(ability_2pl = mean(ability),
+            ability_ses = mean(ability_ses)) %>%
+  kable(digits = 2, format = "html", table.attr = "style='width:30%;'")
+```

### 2021-07-31 02-preliminary-WGcomp.Rmd
new comprehension fits, multigroup model comparisons

@@
 library(GGally)
+library(ggpubr)
 #library(parallel)
@@
 
-# mother's years of education = proxy for SES
-table(d_demo$mom_ed) %>% kable()
-# skewed: only 72 with Primary/Some Secondary (most have Some College +)
-
 #sum(is.na(d_demo$mom_ed)) # many mom_ed missing
@@
 d_long_wg <- get_instrument_data(language="English (American)", form = "WG") %>%
-  mutate(produces = as.numeric(value == "yes" | value == "produces")) # we'll start with production
+  mutate(understands = as.numeric(value == "yes" | value == "produces" | value == "understands")) # we'll start with production
 # yes = understands; no = does not understand/produce, produces = produces (and understands)
@@
 
-d_mat <- d_long_wg %>% select(data_id, definition, produces) %>% 
-  pivot_wider(id_cols = data_id, names_from = definition, values_from = produces)
+d_mat <- d_long_wg %>% select(data_id, definition, understands) %>% 
+  pivot_wider(id_cols = data_id, names_from = definition, values_from = understands)
 
-save(d_mat, d_demo, en_wg, file=here("data/en_wg_comp.Rdata"))
+d_long_wg <- d_long_wg %>% select(-value, -num_item_id)
+
+save(d_mat, d_demo, en_wg, d_long_wg, file=here("data/en_wg_comp.Rdata"))
 ```
@@
 
-table(en_wg$category) %>% kable()
+table(en_wg$category) %>% kable(format = "html", table.attr = "style='width:30%;'")
 ```
@@
 
+sex_group = as.character(d_demo$sex)
+
 save(file = here("data/eng_wg_comp_mod_2pl.Rds"), "m1", "coefs_2pl", "fscores_2pl") # ,
+```
+
+## GLIMMER Plots
+
+```{r fit-rasch-group-models, echo=F, eval=F}
+# first fit Rasch model to each group, and then compare resulting parameters
+
+m1pl_m <- mirt(d_mat[which(d_demo$sex=="Male"),], 1, "Rasch", SE=T, SE.type='Oakes', verbose=T)
+m1pl_f <- mirt(d_mat[which(d_demo$sex=="Female"),], 1, "Rasch", SE=T, SE.type='Oakes', verbose=T)
+
+m1pl_h <- mirt(d_mat[which(d_demo$ses_group=="high"),], 1, "Rasch", SE.type='Oakes', verbose=T)
+m1pl_l <- mirt(d_mat[which(d_demo$ses_group=="low"),], 1, "Rasch", SE.type='Oakes', verbose=T)
+
+save(file = here("data/eng_wg_rasch_group_models.Rds"),
+     "m1pl_m", "m1pl_f", "m1pl_h", "m1pl_l")
+
+res_m = residuals(m1pl_m)
+#m1pl_m_ci = boot.mirt(m1pl_m)
+#itemplot(m1pl_m, 1, CE=T)
+```
+
+```{r, echo=F}
+load(here("data/eng_wg_rasch_group_models.Rds"))
+
+# coef(m1pl_m) shows CI_2.5 and CI_97.5 for d...iterate over and add to dataframes?
+#ci_lo = coef(m1pl_m)$`baa baa`['CI_2.5','d']
+#ci_hi = coef(m1pl_m)$`baa baa`['CI_97.5','d']
+
+par_m = as_tibble(coef(m1pl_m, simplify = TRUE)$items) %>%
+      mutate(definition = rownames(coef(m1pl_m, simplify = TRUE)$items))
+par_f = as_tibble(coef(m1pl_f, simplify = TRUE)$items) %>%
+      mutate(definition = rownames(coef(m1pl_f, simplify = TRUE)$items))
+
+par_h = as_tibble(coef(m1pl_h, simplify = TRUE)$items) %>%
+      mutate(definition = rownames(coef(m1pl_h, simplify = TRUE)$items))
+par_l = as_tibble(coef(m1pl_l, simplify = TRUE)$items) %>%
+      mutate(definition = rownames(coef(m1pl_l, simplify = TRUE)$items))
+
+rasch_pars = tibble(definition = par_m$definition,
+                    d_Male = par_m$d,
+                    d_Female = par_f$d,
+                    d_LowSES = par_l$d,
+                    d_HighSES = par_h$d) %>%
+  mutate(d_sex_diff = abs(d_Female - d_Male),
+         d_ses_diff = abs(d_HighSES - d_LowSES))
+
+sex_thresh = median(rasch_pars$d_sex_diff) + sd(rasch_pars$d_sex_diff)
+ses_thresh = median(rasch_pars$d_ses_diff) + sd(rasch_pars$d_ses_diff)
+
+p1 <- ggplot(rasch_pars, 
+       aes(x = d_Male, y = d_Female)) + 
+  geom_point(alpha=.7) + 
+  ggrepel::geom_label_repel(data = filter(rasch_pars, d_sex_diff > sex_thresh), 
+                            aes(label = definition), max.overlaps = 20) + theme_bw() +
+  xlab("Item Ease for Males") + 
+  ylab("Item Ease for Females") + 
+  geom_abline(intercept=0, slope=1, linetype='dashed')
+
+p2 <- ggplot(rasch_pars, 
+       aes(x = d_LowSES, y = d_HighSES)) + 
+  geom_point(alpha=.7) + 
+  ggrepel::geom_label_repel(data = filter(rasch_pars, d_ses_diff > ses_thresh), 
+                            aes(label = definition), max.overlaps = 20) + theme_bw() +
+  xlab("Item Ease for Low SES") + 
+  ylab("Item Ease for High SES") + 
+  geom_abline(intercept=0, slope=1, linetype='dashed')
+```
+
+```{r, echo=F, fig.width=10, fig.height=5}
+ggarrange(p1, p2)
 ```
@@
 
-# ToDo: define ses_group ("high" for mom_ed > X, "low") (choose X ..high school?)
-
-mg1 <- multipleGroup(d_mat, 1, group = sex_group) # this example should work, according to docs
+# won't work if some items are all 0 (or 1?) for one sex
 mg1 <- multipleGroup(d_mat, 1, group = sex_group,
-                     invariance = c("free_var", "free_means")) 
-# Error: Multiple Group model will not be identified without proper constraints (groups contain missing data patterns where item responses have been completely omitted or, alternatively, the number of categories within each group is not equal to the total number of categories)
-
-# unless there are possibly items
+                     technical=list(NCYCLES=4000)) 
+#mg1 <- multipleGroup(d_mat, 1, group = sex_group, 
+#                     invariance = c("free_var", "free_means"))
 
@@
 # constrain slopes, allow varying group means and variances -- very slow!
-#mg4 <- multipleGroup(d_mat, model = 1, group = sex_group, technical=list(NCYCLES=4000), 
-#                     invariance =c('slopes','free_means', 'free_var'), verbose = T)
+mg4 <- multipleGroup(d_mat, model = 1, group = sex_group, technical=list(NCYCLES=4000), 
+                     invariance =c('slopes','free_means', 'free_var'), verbose = T)
 
@@
 
-anova(mod_sc1, mod_fullconstrain) # mod_fullconstrain preferred by AIC/BIC
-anova(m_sc2, mod_fullconstrain) # mod_fullconstrain preferred by AIC/BIC
+anova(mg1, mg2) # AIC and BIC prefer slope-invariant model (mg2)
+anova(mg2, mg3) # AIC/BIC prefer mg3, with varying group means
+anova(mg3, mg4) # AIC/BIC prefer mg3, with varying group means but equal variance
 
-anova(mg2, mod_fullconstrain) # AIC prefers slope-invariant model; BIC prefers fully-constrained model
+anova(mod_sc1, mod_fullconstrain) # mod_fullconstrain preferred by BIC, sc1 by AIC
+anova(m_sc2, mod_fullconstrain) # mod_fullconstrain preferred by BIC, sc2 by AIC
+
+anova(mod_fullconstrain, mg2) # AIC and BIC both prefer slope-invariant model
 
@@
 
-anova(mg3, mg2) # AIC//BIC prefer model with varying group means
-#anova(mg3, mg4)
-
-coef(mg3)[[2]]$GroupPars # Female intercept = 0.22 (Male = 0)
+coef(mg3)[[2]]$GroupPars # Female intercept = 0.047 (Male = 0)
 
@@
 save(file = here("data/eng_wg_comp_sex_sexDIF.Rds"), 
-     mg2, mg3,  m_sc2, mod_sc1, mod_fullconstrain) # not run: mg4
+     mg1, mg2, mg3, mg4, m_sc2, mod_sc1, mod_fullconstrain) 
 ```
@@
 # same item estimates with/without free_means? 
-cor(mg2_df$d_f, mg3_df$d_f) # .9999
-cor(mg2_df$d_m, mg3_df$d_m) # .9998
+cor(mg2_df$d_f, mg3_df$d_f) # .999
+cor(mg2_df$d_m, mg3_df$d_m) # .999
 
-# same difficulties:
-mod_sc1_df = extract_group_df(mod_sc1)
-
-# fully-constrained is just the same as ordinary 2PL model, no? but we can use this model for DIF
-mod_fullconstrain_df = extract_group_df(mod_fullconstrain)
-# DIF(mod_fullconstrain, which.par = 'd')
-
-#mod_int_df = extract_group_df(m_int)
-
-mod_sc2_df = extract_group_df(m_sc2)
-
-#anova(m1, mg2) # AIC prefers group model, BIC prefers 2PL
+anova(mg1, mg2) # AIC and BIC prefer slope-invariant model (mg2)
+anova(mg2, mg3) # AIC/BIC prefer mg3, with varying group means
 ```
@@
 
+### Anchor items
+
+For our DIF analysis, let's pick 10 anchor items that show the smallest difficulty difference in the slope-invariant free group means model (mg3).
+
+```{r}
+sex_anchors = mg3_df %>% arrange(d_diff_abs) %>% head(10) %>% 
+  relocate(a1, .after=definition) %>%
+  relocate(d_m, .after=d_f)
+
+sex_anchors %>% kable(digits = 3, format = "html", table.attr = "style='width:40%;'")
+```
+
 ## Ability distribution comparison
@@
 
-d_demo <- d_demo %>% left_join(fscores_mg2 %>% 
+d_demo <- d_demo %>% left_join(fscores_mg3 %>% 
                                  mutate(data_id = as.numeric(data_id)))
@@
 
-## DIF Model: SES
+## SES-related DIF
+
+Now we will look for SES-based DIF, using mother's education as a proxy for SES. 
+Here's the distribution of mom_ed in our data:
+
+```{r, mom-ed-tab, echo=F}
+table(d_demo$mom_ed) %>% kable(format = "html", table.attr = "style='width:30%;'")
+```
+
+We (arbitrarily) define high SES as their mother having at least some college (N=`r length(which(d_demo$ses_group=="high"))`), and low SES as having at most secondary education (N=`r length(which(d_demo$ses_group=="low"))`).
 
@@
 
-#mg1 <- multipleGroup(d_mat, 1, group = ses_group, SE = T) 
+mg1 <- multipleGroup(d_mat, 1, group = ses_group, SE = T, technical=list(NCYCLES=4000)) 
 mg2 <- multipleGroup(d_mat, model = 1, group = ses_group, technical=list(NCYCLES=4000), 
                      invariance = 'slopes', verbose = T)
-
 mg3 <- multipleGroup(d_mat, model = 1, group = ses_group, technical=list(NCYCLES=4000), 
@@
 
-anova(mg2, mg3) # AIC/BIC prefer model with varying group mean
+anova(mg1, mg2) # AIC/BIC prefer slope-invariant model (mg2)
+anova(mg2, mg3) # AIC/BIC prefer slope-invariant model without varying group mean
 
@@
 save(file = here("data/eng_wg_comp_SES_DIF.Rds"), 
-     mg2, mg3, m_sc2, mod_sc1, mod_fullconstrain)
+     mg1, mg2, mg3, m_sc2, mod_sc1, mod_fullconstrain)
 ```
@@
 
-coef(mg3)[[2]]$GroupPars # low-SES group (high is ref) intercept = .24
+coef(mg3)[[2]]$GroupPars # low-SES group (high is ref) intercept = .25
 ```
@@
 # items showing > median + 1SD of absolute difference in difficulty have a lot of DIF
-thresh = median(mg3_df$d_diff_abs) + sd(mg3_df$d_diff_abs)
-big_dif = mg3_df[which(mg3_df$d_diff_abs > thresh),]
+thresh = median(mg2_df$d_diff_abs) + sd(mg2_df$d_diff_abs)
+big_dif = mg2_df[which(mg2_df$d_diff_abs > thresh),]
 
@@
 ```{r, fig.width=8, fig.height=6}
-ggplot(mg3_df, 
+ggplot(mg2_df, 
        aes(x = d_lo, y = d_hi)) + 
   geom_point(alpha=.7) + 
-  ggrepel::geom_label_repel(data = filter(mg3_df, d_diff_abs > thresh), 
+  ggrepel::geom_label_repel(data = filter(mg2_df, d_diff_abs > thresh), 
                             aes(label = definition), max.overlaps = 20) + theme_bw() +
@@
   geom_abline(intercept=0, slope=1, linetype='dashed')
+```
+
+### Anchor items
+
+For our DIF analysis, let's pick 10 anchor items that show the smallest difficulty difference in the slope-invariant model (mg2).
+
+```{r ses-anchors, echo=F}
+ses_anchors = mg2_df %>% arrange(d_diff_abs) %>% head(10) %>% 
+  relocate(a1, .after=definition) %>%
+  relocate(d_hi, .after=a1)
+
+ses_anchors %>% kable(digits = 3, format = "html", table.attr = "style='width:40%;'")
 ```

### 2021-08-02 02-preliminary-WGcomp.Rmd
finished running some comprehension DIF stats

@@
 
+mirtCluster()
+mg3_dif = DIF(mg3, c('d'))
+
 save(file = here("data/eng_wg_comp_sex_sexDIF.Rds"), 
-     mg1, mg2, mg3, mg4, m_sc2, mod_sc1, mod_fullconstrain) 
+     mg3_dif, mg1, mg2, mg3, mg4, m_sc2, mod_sc1, mod_fullconstrain) 
 ```
@@
 # same item estimates with/without free_means? 
-cor(mg2_df$d_f, mg3_df$d_f) # .999
-cor(mg2_df$d_m, mg3_df$d_m) # .999
+#cor(mg2_df$d_f, mg3_df$d_f) # .999
+#cor(mg2_df$d_m, mg3_df$d_m) # .999
 
-anova(mg1, mg2) # AIC and BIC prefer slope-invariant model (mg2)
-anova(mg2, mg3) # AIC/BIC prefer mg3, with varying group means
+#anova(mg1, mg2) # AIC and BIC prefer slope-invariant model (mg2)
+#anova(mg2, mg3) # AIC/BIC prefer mg3, with varying group means
+```
+
+### Items with DIF in difficulty
+
+`r nrow(subset(mg3_dif, p<.05))` items show DIF based on a $\chi^2$ test yielding $p<.05$.
+Below we show the `r nrow(subset(mg3_dif, p<.001))` items with $p<.001$.
+
+```{r}
+mg3_dif %>% arrange(BIC) %>% filter(p < .001) %>%
+  select(-AICc, -SABIC, -HQ, -df) %>%
+  kable(digits = 2, format = "html", table.attr = "style='width:40%;'")
 ```
@@
 
-#DIF(mg2, c('d')) # No hyper-parameters were estimated in the DIF model. For effective DIF testing, freeing the focal group hyper-parameters is recommended.
+#DIF(mg1, c('d')) # No hyper-parameters were estimated in the DIF model. For effective DIF testing, freeing the focal group hyper-parameters is recommended.
 
@@
 
+mg2_dif = DIF(mg2, c('d')) # No hyper-parameters were estimated in the DIF model. For effective DIF testing, freeing the focal group hyper-parameters is recommended.
+mg3_dif = DIF(mg3, c('d')) # (though varying group mean not justified according to AIC/BIC)
+
 save(file = here("data/eng_wg_comp_SES_DIF.Rds"), 
-     mg1, mg2, mg3, m_sc2, mod_sc1, mod_fullconstrain)
+     mg1, mg2, mg3, m_sc2, mod_sc1, mod_fullconstrain,
+     mg2_dif, mg3_dif)
 ```
@@
 coef(mg3)[[2]]$GroupPars # low-SES group (high is ref) intercept = .25
+```
+
+### Items with DIF in difficulty
+
+`r nrow(subset(mg3_dif, p<.05))` items show DIF based on a $\chi^2$ test yielding $p<.05$.
+We show these items below.
+
+```{r}
+mg3_dif %>% arrange(BIC) %>% filter(p < .05) %>%
+  select(-AICc, -SABIC, -HQ, -df) %>%
+  kable(digits = 2, format = "html", table.attr = "style='width:40%;'")
 ```

### 2021-08-04 GLIMMER_WGcomp.R
GLIMMER plots for SES and sex DIF

@@ -1,0 +1,85 @@
+# LOAD PACKAGES -----------------------------------------------------------
+library(tidyverse)
+library(difR)
+library(mirt)
+library(janitor)
+library(here)
+#data("verbal")
+
+# load saved CDI:WG data  -------------------------------------------------------
+load(here("data/en_wg_comp.Rdata")) # d_demo, d_mat, en_wg (items)
+
+d_demo <- d_demo %>% 
+  filter(comprehension!=0, !is.na(sex)) %>% # can't fit children not producing words, or with NA sex in group model
+  arrange(data_id)
+
+# there are more IDs in d_mat than in d_demo; remove those
+d_mat <- d_mat %>% filter(is.element(data_id, d_demo$data_id)) %>% arrange(data_id)
+
+sids = d_mat$data_id
+d_mat$data_id = NULL
+d_mat = data.matrix(d_mat)
+row.names(d_mat) = sids
+
+sex_group = as.character(d_demo$sex)
+ses_group = as.character(d_demo$ses_group)
+
+# MIRT FUNCTION HELPERS ---------------------------------------------------
+fit_mod_intuitive <- function(data, groups){
+  multipleGroup(data, 1, itemtype = "Rasch", groups, invariance = "free_var", SE = TRUE, verbose = FALSE)
+}
+
+mod_intuitive_to_draws_df <- function(mod){
+  par_draws <- MASS::mvrnorm(n = 10000, mu = extract.mirt(mod, "parvec"), Sigma = extract.mirt(mod, "vcov"))
+  par_draws <- par_draws[ , str_detect(colnames(par_draws), "d")]
+  draws_df <- tibble(run = 1:nrow(par_draws))
+  stopifnot(ncol(par_draws) %% 2 == 0)
+  n_items <- ncol(par_draws) / 2
+  for (i in 1:n_items) {
+    draws_df[[paste0("item", i)]] <- par_draws[ , i] - par_draws[ , i + n_items]
+  }
+  draws_df
+}
+
+draws_df_to_logit_plot <- function(draws_df){
+  draws_df %>%
+    select(-run) %>%
+    gather(var, val) %>%
+    mutate(var = paste0("Item ", add_zero(parse_number(as.character(var))))) %>%
+    ggplot(aes(x = val, y = var)) +
+    ggridges::geom_density_ridges() +
+    labs(x = "", y = "")
+}
+
[…]

### 2021-08-04 GLIMMER_WGcomp.R
adding glimmer plots

@@ -56,19 +56,18 @@ plot_glimmer <- function(mod_intuitive, item_names, plotName='') {
   draws_df <- mod_intuitive %>% mod_intuitive_to_draws_df()
   names(draws_df) = c("run", item_names)
   
-  if(plotName!='') ggsave(paste0(plotName,'.pdf'), 
-                          width=8, height=50, # .2 * length(item_names)
-                          limitsize = F) 
+  
   p <- draws_df %>%
     clean_names() %>%
     gather(var, val, -run) %>%
     mutate(var = fct_reorder(var, val)) %>%
     ggplot(aes(x = val, y = var)) +
     ggridges::geom_density_ridges() +
-    geom_vline(aes(xintercept=0, linetype='dashed'), alpha=.5) +
+    geom_vline(aes(xintercept=0), linetype='dashed', alpha=.5) +
     labs(x = "", y = "") + theme_classic()
-  dev.off()
-  
+  if(plotName!='') ggsave(paste0(plotName,'.pdf'), 
+                          width=8, height=50, # .2 * length(item_names)
+                          limitsize = F) 
   return(p)
 }
 

### 2021-08-07 02-preliminary-WGcomp.Rmd
minor edits

@@
 
+# Takehome: by model comparison, we find that equal slopes (across sex) and equal variance, but varying group means (and item difficulties) are justified.
+
 anova(mod_sc1, mod_fullconstrain) # mod_fullconstrain preferred by BIC, sc1 by AIC
@@
 Average male ability is slightly higher in the sex group model (i.e., there is a larger female advantage in the standard 2PL model), showing that the 36 items that are easier for females boost female ability estimates.
+
+Extracting the estimated ability intercept for males from mg3: `r round(coef(mg3)[["Male"]]$GroupPars[1],2)`. 
+(Why is this positive? Females should have higher ability (see plot).)
 

### 2021-08-10 02-preliminary-WGcomp.Rmd
added ethnicity GLIMMER plot

@@
 First, let's get some wordbank data for American English. We'll start with production data from the CDI: Words & Gestures (WG) form.
-We'll look for DIF based on 1) sex and 2) SES (high vs. low). 
+We'll look for DIF based on 1) sex, 2) SES (high vs. low), and 3) race/ethnicity. 
 Questions: How do we choose a dozen anchor items (that we expect to be unbiased)?
@@
 
-```{r, echo=F}
+```{r, echo=F, include=F}
 load(here("data/eng_wg_rasch_group_models.Rds"))
@@
 
-```{r, echo=F, fig.width=10, fig.height=5}
+```{r, echo=F, fig.width=10, fig.height=5, include=F}
 ggarrange(p1, p2)
 ```
+
+(See external PDFs and GLIMMER_WGcomp.R)
 

### 2021-08-10 GLIMMER_WGcomp.R

@@ -11,10 +11,13 @@ load(here("data/en_wg_comp.Rdata")) # d_demo, d_mat, en_wg (items)
 
 d_demo <- d_demo %>% 
   filter(comprehension!=0, !is.na(sex)) %>% # can't fit children not producing words, or with NA sex in group model
-  arrange(data_id)
+  arrange(data_id) %>%
+  mutate(eth_group = ifelse(ethnicity=="White", "White", "Nonwhite"))
 
 # there are more IDs in d_mat than in d_demo; remove those
 d_mat <- d_mat %>% filter(is.element(data_id, d_demo$data_id)) %>% arrange(data_id)
+
+
 
 sids = d_mat$data_id
 d_mat$data_id = NULL
@@ -23,6 +26,7 @@ row.names(d_mat) = sids
 
 sex_group = as.character(d_demo$sex)
 ses_group = as.character(d_demo$ses_group)
+eth_group = as.character(d_demo$eth_group)
 
 # MIRT FUNCTION HELPERS ---------------------------------------------------
 fit_mod_intuitive <- function(data, groups){
@@ -75,10 +79,11 @@ plot_glimmer <- function(mod_intuitive, item_names, plotName='') {
 # RUN MODELS
 mod_intuitive_sex <- fit_mod_intuitive(d_mat, sex_group)
 mod_intuitive_ses <- fit_mod_intuitive(d_mat, ses_group)
-
+mod_intuitive_eth <- fit_mod_intuitive(d_mat, eth_group)
 
 # GLIMMER TIME
 plot_glimmer(mod_intuitive_sex, colnames(d_mat), plotName="GLIMMER_sex_compWG")
 plot_glimmer(mod_intuitive_ses, colnames(d_mat), plotName="GLIMMER_ses_compWG")
+plot_glimmer(mod_intuitive_eth, colnames(d_mat), plotName="GLIMMER_eth_compWG")
 
 

### 2021-08-11 02-preliminary-WGcomp.Rmd
production (WS only) glimmer plots

@@
 d_long_wg <- get_instrument_data(language="English (American)", form = "WG") %>%
-  mutate(understands = as.numeric(value == "yes" | value == "produces" | value == "understands")) # we'll start with production
+  mutate(understands = as.numeric(value == "yes" | value == "produces" | value == "understands")) # we'll start with comprehension
 # yes = understands; no = does not understand/produce, produces = produces (and understands)

### 2021-08-11 GLIMMER_production.R

@@ -1,0 +1,128 @@
+# LOAD PACKAGES -----------------------------------------------------------
+library(tidyverse)
+library(difR)
+library(mirt)
+library(janitor)
+library(here)
+#data("verbal")
+
+# gets WS production data and demo variables (ToDo: stitch with WG production data)
+get_production_data <- function() {
+  require(wordbankr)
+  
+  d_long_ws <- get_instrument_data(language="English (American)", form = "WS") %>%
+    mutate(produces = as.numeric(value == "produces")) # we'll start with production
+  
+  en_ws <- get_item_data(language="English (American)", form="WS") %>%
+    filter(type=="word") %>% select(-complexity_category, -item_id)
+  
+  d_demo <- get_administration_data(language = "English (American)", form = "WS")
+  
+  # subjects not producing any words yet (can't fit model with them)
+  bad_Ss = d_demo[which(d_demo$comprehension==0),]$data_id # 9 bad production Ss
+  
+  d_long_ws <- d_long_ws %>% left_join(en_ws %>% select(num_item_id, definition)) %>%
+    filter(!is.na(definition))
+  
+  d_mat <- d_long_ws %>% select(data_id, definition, produces) %>% 
+    pivot_wider(id_cols = data_id, names_from = definition, values_from = produces)
+  
+  d_long_ws <- d_long_ws %>% select(-value, -num_item_id)
+  
+  save(d_mat, d_demo, en_ws, d_long_ws, file=here("data/en_ws_production.Rdata"))
+}
+
+# run once
+#get_production_data()
+
+# load saved CDI:WS data  -------------------------------------------------------
+load(here("data/en_ws_production.Rdata")) # d_demo, d_mat, en_wg (items)
+
+d_demo <- d_demo %>% 
+  filter(comprehension!=0, !is.na(sex)) %>% # can't fit children not producing words, or with NA sex in group model
+  arrange(data_id) %>%
+  mutate(eth_group = ifelse(ethnicity=="White", "White", "Nonwhite"),
+         ses_group = 
+           ifelse(is.element(mom_ed, c("None", "Primary", "Some Secondary", "Secondary")), "low", "high"))
+  
+
[…]

### 2021-08-12 GLIMMER_production.R
add DIF histogram

@@ -126,3 +126,44 @@ plot_glimmer(mod_intuitive_ses, colnames(d_mat), plotName="GLIMMER_ses_prodWS")
 plot_glimmer(mod_intuitive_eth, colnames(d_mat), plotName="GLIMMER_eth_prodWS")
 
 
+# histograms of item difficulty differences
+
+#mod_p <- summary(mod_intuitive_sex)
+
+extract_group_df <- function(group_model, groups=c("Male","Female")) {
+  Mit = as_tibble(coef(group_model, simplify=T)[[groups[1]]]$items) %>%
+    mutate(definition = rownames(coef(group_model, simplify=T)[[groups[1]]]$items),
+           group1 = groups[1],
+           group2 = groups[2]) %>%
+    select(-g, -u) %>% 
+    rename(d_g1 = d)
+  Fit = as_tibble(coef(group_model, simplify=T)$Female$items) %>%
+    mutate(definition = rownames(coef(group_model, simplify=T)[[groups[2]]]$items)) %>%
+    select(-g, -u) %>%
+    rename(d_g2 = d)
+  
+  combo <- Mit %>% left_join(Fit) %>%
+    mutate(d_diff = d_g2 - d_g1,
+           d_diff_abs = abs(d_diff)) 
+  return(combo)
+}
+
+item_difficulty_difference_histogram <- function(mm) {
+  mm %>% ggplot(aes(x=d_g2 - d_g1)) +
+    geom_histogram() + theme_classic() +
+    geom_vline(aes(xintercept=median(d_diff)), linetype="dashed") +
+    xlab(paste(mm$group2, "-", mm$group1, "Item Difficulty"))
+}
+
+mm_sex <- extract_group_df(mod_intuitive_sex, groups=c("Male","Female"))
+sex_hist <- item_difficulty_difference_histogram(mm_sex)
+
+mm_ses <- extract_group_df(mod_intuitive_ses, groups=c("high","low"))
+ses_hist <- item_difficulty_difference_histogram(mm_ses)
+
+mm_eth <- extract_group_df(mod_intuitive_eth, groups=c("Nonwhite","White"))
+eth_hist <- item_difficulty_difference_histogram(mm_eth)
+
+
+require(ggpubr)
+ggarrange(sex_hist, ses_hist, eth_hist, nrow=1)

### 2021-08-12 GLIMMER_production.R
fixed histogram function

@@ -137,7 +137,7 @@ extract_group_df <- function(group_model, groups=c("Male","Female")) {
            group2 = groups[2]) %>%
     select(-g, -u) %>% 
     rename(d_g1 = d)
-  Fit = as_tibble(coef(group_model, simplify=T)$Female$items) %>%
+  Fit = as_tibble(coef(group_model, simplify=T)[[groups[2]]]$items) %>%
     mutate(definition = rownames(coef(group_model, simplify=T)[[groups[2]]]$items)) %>%
     select(-g, -u) %>%
     rename(d_g2 = d)
@@ -148,11 +148,19 @@ extract_group_df <- function(group_model, groups=c("Male","Female")) {
   return(combo)
 }
 
-item_difficulty_difference_histogram <- function(mm) {
-  mm %>% ggplot(aes(x=d_g2 - d_g1)) +
-    geom_histogram() + theme_classic() +
+item_difficulty_difference_histogram <- function(mm, withNormal=F) {
+  #yfit <- dnorm(mm$d_diff, mean = mean(mm$d_diff), sd = sd(mm$d_diff)) 
+
+  p <- mm %>% ggplot(aes(x=d_g2 - d_g1)) +
+    geom_histogram(aes(y =..density..), alpha=.7) + theme_classic() +
     geom_vline(aes(xintercept=median(d_diff)), linetype="dashed") +
     xlab(paste(mm$group2, "-", mm$group1, "Item Difficulty"))
+  
+  if(withNormal) {
+  p <- p + stat_function(fun = dnorm, alpha=.6, 
+                         args = list(mean = mean(mm$d_diff), sd = sd(mm$d_diff)))
+  }
+  return(p)
 }
 
 mm_sex <- extract_group_df(mod_intuitive_sex, groups=c("Male","Female"))
@@ -164,6 +172,13 @@ ses_hist <- item_difficulty_difference_histogram(mm_ses)
 mm_eth <- extract_group_df(mod_intuitive_eth, groups=c("Nonwhite","White"))
 eth_hist <- item_difficulty_difference_histogram(mm_eth)
 
-
 require(ggpubr)
 ggarrange(sex_hist, ses_hist, eth_hist, nrow=1)
+ggsave(file="item_DIF_hist_sex.pdf", width=8, height=3.5)
+
+sex_histn <- item_difficulty_difference_histogram(mm_sex, withNormal = T)
[…]

### 2021-08-13 GLIMMER_production.R
better histograms

@@ -150,10 +150,16 @@ extract_group_df <- function(group_model, groups=c("Male","Female")) {
 
 item_difficulty_difference_histogram <- function(mm, withNormal=F) {
   #yfit <- dnorm(mm$d_diff, mean = mean(mm$d_diff), sd = sd(mm$d_diff)) 
-
-  p <- mm %>% ggplot(aes(x=d_g2 - d_g1)) +
+  mm <- mm %>% mutate(d_diff = d_g2 - d_g1)
+  max_dif = mean(mm$d_diff) + 2*sd(mm$d_diff)
+  min_dif = mean(mm$d_diff) - 2*sd(mm$d_diff)
+  mm <- mm %>% mutate(extreme = ifelse((d_diff > max_dif) | (d_diff < min_dif), T, F))
+  
+  p <- mm %>% ggplot(aes(x=d_diff)) + # , fill=extrem
     geom_histogram(aes(y =..density..), alpha=.7) + theme_classic() +
     geom_vline(aes(xintercept=median(d_diff)), linetype="dashed") +
+    geom_vline(aes(xintercept=max_dif), linetype="dashed", color="red") +
+    geom_vline(aes(xintercept=min_dif), linetype="dashed", color="red") +
     xlab(paste(mm$group2, "-", mm$group1, "Item Difficulty"))
   
   if(withNormal) {
@@ -174,11 +180,11 @@ eth_hist <- item_difficulty_difference_histogram(mm_eth)
 
 require(ggpubr)
 ggarrange(sex_hist, ses_hist, eth_hist, nrow=1)
-ggsave(file="item_DIF_hist_sex.pdf", width=8, height=3.5)
+ggsave(file="item_DIF_hist_thresh.pdf", width=8, height=3.5)
 
 sex_histn <- item_difficulty_difference_histogram(mm_sex, withNormal = T)
 ses_histn <- item_difficulty_difference_histogram(mm_ses, withNormal = T)
 eth_histn <- item_difficulty_difference_histogram(mm_eth, withNormal = T)
 
 ggarrange(sex_histn, ses_histn, eth_histn, nrow=1)
-ggsave(file="item_DIF_hist_sex_withNormals.pdf", width=8, height=3.5)
+ggsave(file="item_DIF_hists_withNormals.pdf", width=8, height=3.5)

### 2021-08-17 ws_prod_report.Rmd
outlined report

---
title: "CDI:WS Production Bias Report"
author: "Nathan, George, & Mike"
date: "8/17/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)

library(tidyverse)
library(difR)
library(mirt)
library(janitor)
library(here)
require(kableExtra)

# created in GLIMMER_production.R
load(here("data/en_ws_production.Rdata")) # d_demo, d_mat, en_wg (items)
source("report_helper_functions.R")
```

# Introduction

## Item-Response Theory

## Differential Item Function

### Dataset

We analyze Wordbank CDI Words & Sentences form data from `r nrow(d_demo)` children aged `r min(d_demo$age)` to `r max(d_demo$age)`.

```{r, sex-table}
#d_demo <- d_demo %>% mutate(relevel() # remove empty "Other" level?

table(d_demo$sex) %>% kable(format = "html", table.attr = "style='width:30%;'")
```

## Goals

For each axis (sex, SES, race/ethnicity)
1) show GLIMMER plots (small sample of words)
2) histograms of DIF: fit normal distro, exclude >2SD (list those items)
  - overall statistic measuring DIF of the test? sd(normal), but that doesn't account for overall group difference (subtract mean?) (DTF definition/statistic)
3) re-fit IRT model, show new GLIMMER (discuss: does the test look much more fair, or are there new outliers?)
4) suggested new words from CHILDES (equal frequency for boys/girls) [but we can't do this for SES/race unless CHILDES has demo data]

## Sex DIF

### GLIMMER plots

```{r show-sex-glimmer, fig.width=5, fig.height=5.5, caption="GLIMMER plot for sample of items showing most DIF in both directions, as well as a random sample from the middle."}
load("data/glimmer_prodWS_models.Rds")

plot_glimmer(mod_intuitive_sex, colnames(d_mat)[2:ncol(d_mat)])

```

### Removing Biased Items

```{r}
mm_sex <- extract_group_df(mod_intuitive_sex, groups=c("Male","Female"))
sex_hist <- item_difficulty_difference_histogram(mm_sex)

orig_sex <- get_extreme_item_difficulty_differences(mm_sex) # min: -.53 max: 1.06 (mean + 2SD)
sex_hist
```

```{r sex-diff-table}
items_to_remove <- orig_sex %>% filter(extreme==T)

items_to_remove %>% filter(extreme==T) %>%
  arrange(d_diff) %>% 
  select(-d_diff_abs, -extreme, -a1) %>%
  kable(format = "html", digits = 2, table.attr = "style='width:30%;'")
```

ToDo: filter items_to_remove out of d_mat, re-run 1PL IRT model

see what the new distribution of item difficulty differences looks like (are there any more outliers?)

```{r, eval=F}
# chunk to run once:
mod_intuitive_sex <- fit_mod_intuitive(d_mat, sex_group)

save()
```

```{r}
load()
```

## SES DIF

```{r ses-table}

```

```{r}
plot_glimmer(mod_intuitive_ses, colnames(d_mat), plotName="GLIMMER_ses_prodWS")

```

## Race/Ethnicity DIF

```{r, race-table}
# breakdown white/non-white
```

# Discussion

Where to go from here -- other languages, regional variation, 
- using adult word vectors to find unbiased items?

### 2021-08-17 GLIMMER_production.R

@@ -148,6 +148,16 @@ extract_group_df <- function(group_model, groups=c("Male","Female")) {
   return(combo)
 }
 
+get_extreme_item_difficulty_differences <- function(mm) {
+  #yfit <- dnorm(mm$d_diff, mean = mean(mm$d_diff), sd = sd(mm$d_diff)) 
+  mm <- mm %>% mutate(d_diff = d_g2 - d_g1)
+  max_dif = mean(mm$d_diff) + 2*sd(mm$d_diff)
+  min_dif = mean(mm$d_diff) - 2*sd(mm$d_diff)
+  mm <- mm %>% mutate(extreme = ifelse((d_diff > max_dif) | (d_diff < min_dif), T, F))
+  print(paste("mininum difference:",min_dif, "maximum difference:",max_dif))
+  return(mm)
+}
+
 item_difficulty_difference_histogram <- function(mm, withNormal=F) {
   #yfit <- dnorm(mm$d_diff, mean = mean(mm$d_diff), sd = sd(mm$d_diff)) 
   mm <- mm %>% mutate(d_diff = d_g2 - d_g1)

### 2021-08-17 report_helper_functions.R

@@ -1,0 +1,89 @@
+extract_group_df <- function(group_model, groups=c("Male","Female")) {
+  Mit = as_tibble(coef(group_model, simplify=T)[[groups[1]]]$items) %>%
+    mutate(definition = rownames(coef(group_model, simplify=T)[[groups[1]]]$items),
+           group1 = groups[1],
+           group2 = groups[2]) %>%
+    select(-g, -u) %>% 
+    rename(d_g1 = d)
+  Fit = as_tibble(coef(group_model, simplify=T)[[groups[2]]]$items) %>%
+    mutate(definition = rownames(coef(group_model, simplify=T)[[groups[2]]]$items)) %>%
+    select(-g, -u) %>%
+    rename(d_g2 = d)
+  
+  combo <- Mit %>% left_join(Fit) %>%
+    mutate(d_diff = d_g2 - d_g1,
+           d_diff_abs = abs(d_diff)) 
+  return(combo)
+}
+
+get_extreme_item_difficulty_differences <- function(mm) {
+  #yfit <- dnorm(mm$d_diff, mean = mean(mm$d_diff), sd = sd(mm$d_diff)) 
+  mm <- mm %>% mutate(d_diff = d_g2 - d_g1)
+  max_dif = mean(mm$d_diff) + 2*sd(mm$d_diff)
+  min_dif = mean(mm$d_diff) - 2*sd(mm$d_diff)
+  mm <- mm %>% mutate(extreme = ifelse((d_diff > max_dif) | (d_diff < min_dif), T, F))
+  print(paste("mininum difference:",min_dif, "maximum difference:",max_dif))
+  return(mm)
+}
+
+item_difficulty_difference_histogram <- function(mm, withNormal=F) {
+  #yfit <- dnorm(mm$d_diff, mean = mean(mm$d_diff), sd = sd(mm$d_diff)) 
+  mm <- mm %>% mutate(d_diff = d_g2 - d_g1)
+  max_dif = mean(mm$d_diff) + 2*sd(mm$d_diff)
+  min_dif = mean(mm$d_diff) - 2*sd(mm$d_diff)
+  mm <- mm %>% mutate(extreme = ifelse((d_diff > max_dif) | (d_diff < min_dif), T, F))
+  
+  p <- mm %>% ggplot(aes(x=d_diff)) + # , fill=extrem
+    geom_histogram(aes(y =..density..), alpha=.7) + theme_classic() +
+    geom_vline(aes(xintercept=median(d_diff)), linetype="dashed") +
[…]

### 2021-08-25 03-preliminary-WSprod.Rmd
added WS production fits

--- 2021-08-10 02-preliminary-WGcomp.Rmd
@@
 ---
-title: "Looking for CDI Bias: WG Comprehension"
+title: "Looking for CDI Bias: WS Production"
 author: "Nathan, George, and Mike"
@@
 
-First, let's get some wordbank data for American English. We'll start with production data from the CDI: Words & Gestures (WG) form.
+First, let's get some wordbank data for American English. We'll now turn to production data from the CDI: Words & Sentences (WS) form.
 We'll look for DIF based on 1) sex, 2) SES (high vs. low), and 3) race/ethnicity.
-Questions: How do we choose a dozen anchor items (that we expect to be unbiased)?
-
-```{r get-wordbank-data, eval=F, echo=F}
-# Run Once
-#get_instruments()
-
-d_demo <- 
-  get_administration_data(language = "English (American)", form = "WG") #%>%
-  #bind_rows(get_administration_data(language = "English (American)", form = "WS"))
-
-# 8-18 month-olds
-table(d_demo$age)
-
-#sum(is.na(d_demo$mom_ed)) # many mom_ed missing
-
-```
-
-Let's do comprehension data.
-
-```{r get-wordbank-items, eval=F, echo=F}
-d_long_wg <- get_instrument_data(language="English (American)", form = "WG") %>%
-  mutate(understands = as.numeric(value == "yes" | value == "produces" | value == "understands")) # we'll start with production
-# yes = understands; no = does not understand/produce, produces = produces (and understands)
-
-en_wg <- get_item_data(language="English (American)", form="WG") %>%
-  filter(type=="word") %>% select(-complexity_category, -item_id)
-
-# subjects not producing any words yet (can't fit model with them)
-bad_Ss = d_demo[which(d_demo$comprehension==0),]$data_id # 8 bad comprehension Ss
-
-d_long_wg <- d_long_wg %>% left_join(en_wg %>% select(num_item_id, definition)) %>%
-  filter(!is.na(definition))
-
-d_mat <- d_long_wg %>% select(data_id, definition, understands) %>% 
-  pivot_wider(id_cols = data_id, names_from = definition, values_from = understands)
-
-d_long_wg <- d_long_wg %>% select(-value, -num_item_id)
-
-save(d_mat, d_demo, en_wg, d_long_wg, file=here("data/en_wg_comp.Rdata"))
-```
 
 ```{r load-data}
-# load saved CDI:WG data
-load(here("data/en_wg_comp.Rdata")) # d_demo, d_mat, en_wg (items)
+# see GLIMMER_production.R
+load(here("data/en_ws_production.Rdata"))
 
@@
   filter(comprehension!=0, !is.na(sex)) %>% # can't fit children not producing words, or with NA sex in group model
-  arrange(data_id)
+  arrange(data_id) %>%
+  mutate(eth_group = ifelse(ethnicity=="White", "White", "Nonwhite"),
+         ses_group = 
+           ifelse(is.element(mom_ed, c("None", "Primary", "Some Secondary", "Secondary")), "low", "high"))
 
@@
 d_mat <- d_mat %>% filter(is.element(data_id, d_demo$data_id)) %>% arrange(data_id)
+
+# only 4090 kids with demographic variables
 
@@
 
+sex_group = as.character(d_demo$sex)
+ses_group = as.character(d_demo$ses_group)
+eth_group = as.character(d_demo$eth_group)
+
 table(d_demo$mom_ed)
-# high SES = mom_ed >= Some College (739 high, 324 low-SES)
-d_demo <- d_demo %>% mutate(ses_group = ifelse(as.numeric(mom_ed)>4, "high", "low"))
 
-table(en_wg$category) %>% kable(format = "html", table.attr = "style='width:30%;'")
+table(en_ws$category) %>% kable(format = "html", table.attr = "style='width:30%;'")
 ```
 
-We will fit comprehension data from `r nrow(d_demo)` children for the WG form.
+We will fit comprehension data from `r nrow(d_demo)` children for the WS form.
 
@@
 
-save(file = here("data/eng_wg_comp_mod_2pl.Rds"), "m1", "coefs_2pl", "fscores_2pl") # ,
+save(file = here("data/eng_ws_prod_mod_2pl.Rds"), "m1", "coefs_2pl", "fscores_2pl") 
 ```
@@
 ## GLIMMER Plots
-
-```{r fit-rasch-group-models, echo=F, eval=F}
-# first fit Rasch model to each group, and then compare resulting parameters
-
-m1pl_m <- mirt(d_mat[which(d_demo$sex=="Male"),], 1, "Rasch", SE=T, SE.type='Oakes', verbose=T)
-m1pl_f <- mirt(d_mat[which(d_demo$sex=="Female"),], 1, "Rasch", SE=T, SE.type='Oakes', verbose=T)
-
-m1pl_h <- mirt(d_mat[which(d_demo$ses_group=="high"),], 1, "Rasch", SE.type='Oakes', verbose=T)
-m1pl_l <- mirt(d_mat[which(d_demo$ses_group=="low"),], 1, "Rasch", SE.type='Oakes', verbose=T)
-
-save(file = here("data/eng_wg_rasch_group_models.Rds"),
-     "m1pl_m", "m1pl_f", "m1pl_h", "m1pl_l")
-
-res_m = residuals(m1pl_m)
-#m1pl_m_ci = boot.mirt(m1pl_m)
-#itemplot(m1pl_m, 1, CE=T)
-```
-
-```{r, echo=F, include=F}
-load(here("data/eng_wg_rasch_group_models.Rds"))
-
-# coef(m1pl_m) shows CI_2.5 and CI_97.5 for d...iterate over and add to dataframes?
-#ci_lo = coef(m1pl_m)$`baa baa`['CI_2.5','d']
-#ci_hi = coef(m1pl_m)$`baa baa`['CI_97.5','d']
-
-par_m = as_tibble(coef(m1pl_m, simplify = TRUE)$items) %>%
-      mutate(definition = rownames(coef(m1pl_m, simplify = TRUE)$items))
-par_f = as_tibble(coef(m1pl_f, simplify = TRUE)$items) %>%
-      mutate(definition = rownames(coef(m1pl_f, simplify = TRUE)$items))
-
-par_h = as_tibble(coef(m1pl_h, simplify = TRUE)$items) %>%
-      mutate(definition = rownames(coef(m1pl_h, simplify = TRUE)$items))
-par_l = as_tibble(coef(m1pl_l, simplify = TRUE)$items) %>%
-      mutate(definition = rownames(coef(m1pl_l, simplify = TRUE)$items))
-
-rasch_pars = tibble(definition = par_m$definition,
-                    d_Male = par_m$d,
-                    d_Female = par_f$d,
-                    d_LowSES = par_l$d,
-                    d_HighSES = par_h$d) %>%
-  mutate(d_sex_diff = abs(d_Female - d_Male),
-         d_ses_diff = abs(d_HighSES - d_LowSES))
-
-sex_thresh = median(rasch_pars$d_sex_diff) + sd(rasch_pars$d_sex_diff)
-ses_thresh = median(rasch_pars$d_ses_diff) + sd(rasch_pars$d_ses_diff)
-
-p1 <- ggplot(rasch_pars, 
-       aes(x = d_Male, y = d_Female)) + 
-  geom_point(alpha=.7) + 
-  ggrepel::geom_label_repel(data = filter(rasch_pars, d_sex_diff > sex_thresh), 
-                            aes(label = definition), max.overlaps = 20) + theme_bw() +
-  xlab("Item Ease for Males") + 
-  ylab("Item Ease for Females") + 
-  geom_abline(intercept=0, slope=1, linetype='dashed')
-
-p2 <- ggplot(rasch_pars, 
-       aes(x = d_LowSES, y = d_HighSES)) + 
-  geom_point(alpha=.7) + 
-  ggrepel::geom_label_repel(data = filter(rasch_pars, d_ses_diff > ses_thresh), 
-                            aes(label = definition), max.overlaps = 20) + theme_bw() +
-  xlab("Item Ease for Low SES") + 
-  ylab("Item Ease for High SES") + 
-  geom_abline(intercept=0, slope=1, linetype='dashed')
-```
-
-```{r, echo=F, fig.width=10, fig.height=5, include=F}
-ggarrange(p1, p2)
-```
 
@@
 # https://rdrr.io/cran/mirt/man/multipleGroup.html
-
-sex_group = as.character(d_demo$sex)
 
@@
 
-save(file = here("data/eng_wg_comp_sex_sexDIF.Rds"), 
+save(file = here("data/eng_ws_prod_sex_sexDIF.Rds"), 
      mg3_dif, mg1, mg2, mg3, mg4, m_sc2, mod_sc1, mod_fullconstrain) 
@@
 ```{r}
-load(here("data/eng_wg_comp_sex_sexDIF.Rds"))
-load(here("data/eng_wg_comp_mod_2pl.Rds"))
+load(here("data/eng_ws_prod_sex_sexDIF.Rds"))
+load(here("data/eng_ws_prod_mod_2pl.Rds"))
 
@@
 
-residuals(mod_sc1)
+#residuals(mod_sc1)
 
 mg2_dif = DIF(mg2, c('d')) # No hyper-parameters were estimated in the DIF model. For effective DIF testing, freeing the focal group hyper-parameters is recommended.
-mg3_dif = DIF(mg3, c('d')) # (though varying group mean not justified according to AIC/BIC)
+#mg3_dif = DIF(mg3, c('d')) # (though varying group mean not justified according to AIC/BIC)
 
@@
      mg1, mg2, mg3, m_sc2, mod_sc1, mod_fullconstrain,
-     mg2_dif, mg3_dif)
+     mg2_dif) # mg3_dif
 ```

### 2021-08-25 ws_prod_report.Rmd
Code for all three axes and updated the helper functions and made some folders

@@
 output: html_document
+fig_width: 3
+fig_height: 3
+fig_align: "center"
 ---
@@
 require(kableExtra)
+
+library(wordbankr)
+library(ggrepel)
+library(gridExtra)
+library(printr)
 
@@
 #d_demo <- d_demo %>% mutate(relevel() # remove empty "Other" level?
+d_demo <- d_demo %>% 
+  filter(comprehension!=0, !is.na(sex)) %>% # can't fit children not producing words, or with NA sex in group model
+  arrange(data_id) %>%
+  mutate(eth_group = ifelse(ethnicity=="White", "White", "Nonwhite"),
+         ses_group = 
+           ifelse(is.element(mom_ed, c("None", "Primary", "Some Secondary", "Secondary")), "low", "high"))
 
@@
 
+```{r load data}
+load("data/glimmer_prodWS_models.Rds")
+load(here("data/glimmer_prodWS_model_noExt.Rds")) # add this in once I've saved all the new models
+```
+
 ## Sex DIF
@@
 
-```{r show-sex-glimmer, fig.width=5, fig.height=5.5, caption="GLIMMER plot for sample of items showing most DIF in both directions, as well as a random sample from the middle."}
-load("data/glimmer_prodWS_models.Rds")
+```{r show-sex-glimmer,echo= FALSE, warning=FALSE,message=FALSE,error=FALSE, fig.width=3, fig.height=3, fig.cap="GLIMMER plot for sample of items showing most DIF in both directions, as well as a random sample from the middle."}
 
@@
 ### Removing Biased Items
-
-```{r}
+Histogram of Sex Items from GLIMMER
+```{r, fig.width=3, fig.height=3, echo=FALSE,warning=FALSE,message=FALSE,error=FALSE}
 mm_sex <- extract_group_df(mod_intuitive_sex, groups=c("Male","Female"))
@@
 
-orig_sex <- get_extreme_item_difficulty_differences(mm_sex) # min: -.53 max: 1.06 (mean + 2SD)
+orig_sex <- get_extreme_item_difficulty_differences(mm_sex, 2.25) # min: -.53 max: 1.06 (mean + 2SD) mean: .265
 sex_hist
@@
 
-```{r sex-diff-table}
-items_to_remove <- orig_sex %>% filter(extreme==T)
+```{r sex-diff-table, fig.width=3, fig.height=3, echo=FALSE,warning=FALSE,message=FALSE,error=FALSE}
+sex_items_to_remove <- orig_sex %>% filter(extreme==T)
 
-items_to_remove %>% filter(extreme==T) %>%
+sex_items_to_remove %>% filter(extreme==T) %>%
   arrange(d_diff) %>% 
-  select(-d_diff_abs, -extreme, -a1) %>%
-  kable(format = "html", digits = 2, table.attr = "style='width:30%;'")
+  select(-d_diff_abs, -extreme, -a1, -group1, -group2, -d_g1, -d_g2) %>%
+  mutate(sex_bias = case_when(d_diff < 0 ~ "Male", d_diff > 0 ~ "Female")) %>%
+  kable(format = "html", digits = 2, table.attr = "style='width:40%;'")
 ```
-
-ToDo: filter items_to_remove out of d_mat, re-run 1PL IRT model
+List of Sex Items that show DIF outside of the 2SD range.(shows `r nrow(sex_items_to_remove)` items that have DIF)
 
@@
 ```{r, eval=F}
-# chunk to run once:
-mod_intuitive_sex <- fit_mod_intuitive(d_mat, sex_group)
+d_demo <- d_demo %>% 
+  filter(comprehension!=0, !is.na(sex)) %>% # can't fit children not producing words, or with NA sex in group model
+  arrange(data_id) %>%
+  mutate(eth_group = ifelse(ethnicity=="White", "White", "Nonwhite"),
+         ses_group = 
+           ifelse(is.element(mom_ed, c("None", "Primary", "Some Secondary", "Secondary")), "low", "high"))
 
-save()
+sex_group = as.character(d_demo$sex)
+
+sex_words <- c(sex_items_to_remove$definition) 
+#remove words
+d_mat_sex <- d_mat[, ! names(d_mat) %in% sex_words, drop = F]
+# there are more IDs in d_mat_sex than in d_demo; remove those
+d_mat_sex <- d_mat_sex %>% filter(is.element(data_id, d_demo$data_id)) %>% arrange(data_id)
+#copied from George's work may not be nessesary here
+sids = d_mat_sex$data_id
+d_mat_sex$data_id = NULL
+d_mat_sex = data.matrix(d_mat_sex)
+row.names(d_mat_sex) = sids
+
+#refit model now with main DIF items removed
+mod_intuitive_sexn <- fit_mod_intuitive(d_mat_sex, sex_group)
+
+plot_glimmer(mod_intuitive_sexn, colnames(d_mat_sex)[2:ncol(d_mat_sex)])
+
 ```
+```{r show-new-sex-glimmer,echo= FALSE, warning=FALSE,message=FALSE,error=FALSE, fig.width=3, fig.height=3, fig.cap="GLIMMER plot for sample of items showing most DIF in both directions, as well as a random sample from the middle."}
+sex_words <- c(sex_items_to_remove$definition) 
+#remove words
+d_mat_sex <- d_mat[, ! names(d_mat) %in% sex_words, drop = F]
+# there are more IDs in d_mat_sex than in d_demo; remove those
+d_mat_sex <- d_mat_sex %>% filter(is.element(data_id, d_demo$data_id)) %>% arrange(data_id)
+#copied from George's work may not be nessesary here
+sids = d_mat_sex$data_id
+d_mat_sex$data_id = NULL
+d_mat_sex = data.matrix(d_mat_sex)
+row.names(d_mat_sex) = sids
 
-```{r}
-load()
-```
-
-## SES DIF
-
-```{r ses-table}
+plot_glimmer(mod_intuitive_sexn, colnames(d_mat_sex)[2:ncol(d_mat_sex)])
 
@@
 
-```{r}
-plot_glimmer(mod_intuitive_ses, colnames(d_mat), plotName="GLIMMER_ses_prodWS")
+Sex Histogram with the DIF Items removed
+```{r, fig.width=3, fig.height=3, echo=FALSE,warning=FALSE,message=FALSE,error=FALSE}
+mm_sexn <- extract_group_df(mod_intuitive_sexn, groups=c("Male","Female"))
+sex_histn <- item_difficulty_difference_histogram(mm_sexn)
+
+orig_sexn <- get_extreme_item_difficulty_differences(mm_sexn, 2.25) # min: -.20 max: 0.74 (mean + 2SD) mean: 0.268
+sex_histn
+```
+
+```{r sex-diff-table-new, fig.width=3, fig.height=3, echo=FALSE,warning=FALSE,message=FALSE,error=FALSE}
+new_sex_items_to_remove <- orig_sexn %>% filter(extreme==T)
+
+new_sex_items_to_remove %>% filter(extreme==T) %>%
+  arrange(d_diff) %>% 
+  select(-d_diff_abs, -extreme, -a1, -group1, -group2, -d_g1, -d_g2) %>%
+  mutate(sex_bias = case_when(d_diff < 0 ~ "Male", d_diff > 0 ~ "Female")) %>%
+  kable(format = "html", digits = 2, table.attr = "style='width:40%;'")
+```
+New Sex Items that show DIF (shows `r nrow(new_sex_items_to_remove)` new items that have DIF)
+
+```{r, eval=F}
+
+new_sex_words <- c(new_sex_items_to_remove$definition) 
+all_sex_words <- append(sex_words, new_sex_words)
+#remove words
+d_mat_sex_new <- d_mat[, ! names(d_mat) %in% all_sex_words, drop = F]
+# there are more IDs in d_mat_sex than in d_demo; remove those
+d_mat_sex_new <- d_mat_sex_new %>% filter(is.element(data_id, d_demo$data_id)) %>% arrange(data_id)
+#copied from George's work may not be nessesary here
+sids = d_mat_sex_new$data_id
+d_mat_sex_new$data_id = NULL
+d_mat_sex_new = data.matrix(d_mat_sex_new)
+row.names(d_mat_sex_new) = sids
+
+#refit model now with main DIF items removed
+mod_intuitive_sexn_n <- fit_mod_intuitive(d_mat_sex_new, sex_group)
+
+plot_glimmer(mod_intuitive_sexn_n, colnames(d_mat_sex_new)[2:ncol(d_mat_sex_new)])
 
@@
 
+```{r show-nnew-sex-glimmer,echo= FALSE, warning=FALSE,message=FALSE,error=FALSE, fig.width=3, fig.height=3, fig.cap="GLIMMER plot for sample of items showing most DIF in both directions, as well as a random sample from the middle."}
+new_sex_words <- c(new_sex_items_to_remove$definition) 
+all_sex_words <- append(sex_words, new_sex_words)
+#remove words
+d_mat_sex_new <- d_mat[, ! names(d_mat) %in% all_sex_words, drop = F]
+# there are more IDs in d_mat_sex than in d_demo; remove those
+d_mat_sex_new <- d_mat_sex_new %>% filter(is.element(data_id, d_demo$data_id)) %>% arrange(data_id)
+#copied from George's work may not be nessesary here
+sids = d_mat_sex_new$data_id
+d_mat_sex_new$data_id = NULL
+d_mat_sex_new = data.matrix(d_mat_sex_new)
+row.names(d_mat_sex_new) = sids
+
+plot_glimmer(mod_intuitive_sexn_n, colnames(d_mat_sex_new)[2:ncol(d_mat_sex_new)])
+
+```
+
+Sex Histogram with the DIF Items removed
+```{r 3rd sex hist, fig.width=3, fig.height=3, echo=FALSE,warning=FALSE,message=FALSE,error=FALSE}
+mm_sexn_n <- extract_group_df(mod_intuitive_sexn_n, groups=c("Male","Female"))
+sex_histn_n <- item_difficulty_difference_histogram(mm_sexn_n)
+
+orig_sexn_n <- get_extreme_item_difficulty_differences(mm_sexn_n, 2.25) # min: -.20 max: 0.74 (mean + 2SD) mean: 0.268
+sex_histn_n
+```
+From seeing this my take away is that there are more female DIF Items total but that the extreme ones are mostly male.
+```{r sex-words 3rd, fig.width=3, fig.height=3, echo=FALSE,warning=FALSE,message=FALSE,error=FALSE}
+nnew_sex_items_to_remove <- orig_sexn_n %>% filter(extreme==T)
+
+nnew_sex_items_to_remove %>% filter(extreme==T) %>%
+  arrange(d_diff) %>% 
+  select(-d_diff_abs, -extreme, -a1, -group1, -group2, -d_g1, -d_g2) %>%
+  mutate(sex_bias = case_when(d_diff < 0 ~ "Male", d_diff > 0 ~ "Female")) %>%
+  kable(format = "html", digits = 2, table.attr = "style='width:40%;'")
+```
+`r nrow(nnew_sex_items_to_remove)` item that now show DIF
+
+## SES DIF
+```{r, ses-table}
+#d_demo <- d_demo %>% mutate(relevel() # remove empty "Other" level?
+
+table(d_demo$ses_group) %>% kable(format = "html", table.attr = "style='width:30%;'")
+```
+
+```{r show-ses-glimmer, fig.width=3, fig.height=3, fig.cap="GLIMMER plot for sample of items showing most DIF in both directions, as well as a random sample from the middle."}
+plot_glimmer(mod_intuitive_ses, colnames(d_mat)[2:ncol(d_mat)])
+```
+
+Histogram from SES GLIMMER
+```{r, fig.width=3, fig.height=3, echo=FALSE,warning=FALSE,message=FALSE,error=FALSE}
+#used this to find out the that the two groups are 0 1 not high low, not sure why
+#groups <- c("0","1")
+#group_model <- mod_intuitive_ses
+#Mit = as_tibble(coef(group_model, simplify=T)[[groups[1]]]$items) %>%
+#    mutate(definition = rownames(coef(group_model, simplify=T)[[groups[1]]]$items),
+#           group1 = groups[1],
+#          group2 = groups[2])
+
+mm_ses <- extract_group_df(mod_intuitive_ses, groups=c("0","1"))
+ses_hist <- item_difficulty_difference_histogram(mm_ses)
+
+orig_ses <- get_extreme_item_difficulty_differences(mm_ses, 2.25) # min:-0.97  max:1.08  (mean + 2SD) mean:0.051
+ses_hist
+```
+
+```{r ses-diff-table, fig.width=3, fig.height=3, echo=FALSE,warning=FALSE,message=FALSE,error=FALSE}
+ses_items_to_remove <- orig_ses %>% filter(extreme==T)
+
+ses_items_to_remove %>% filter(extreme==T) %>%
+  arrange(d_diff) %>% 
+  select(-d_diff_abs, -extreme, -a1, -group1, -group2, -d_g1, -d_g2) %>%
+  mutate(ses_bias = case_when(d_diff < 0 ~ "Low", d_diff > 0 ~ "High")) %>%
+  kable(format = "html", digits = 2, table.attr = "style='width:40%;'")
+```
+List of SES Items that show DIF outside of the 2SD range. (`r nrow(ses_items_to_remove)` items that show DIF)
+
+```{r, eval=F}
+#do we need to have removed NA sex kids for the ses model
+d_demo <- d_demo %>% 
+  filter(comprehension!=0, !is.na(sex)) %>% # can't fit children not producing words, or with NA sex in group model
+  arrange(data_id) %>%
+  mutate(eth_group = ifelse(ethnicity=="White", "White", "Nonwhite"),
+         ses_group = 
+           ifelse(is.element(mom_ed, c("None", "Primary", "Some Secondary", "Secondary")), "low", "high"))
+
+ses_group = as.character(d_demo$ses_group)
+
+ses_words <- c(ses_items_to_remove$definition) 
+#remove words
+d_mat_ses <- d_mat[, ! names(d_mat) %in% ses_words, drop = F]
+# there are more IDs in d_mat_ses than in d_demo; remove those
+d_mat_ses <- d_mat_ses %>% filter(is.element(data_id, d_demo$data_id)) %>% arrange(data_id)
+#copied from George's work may not be nessesary here
+sids = d_mat_ses$data_id
+d_mat_ses$data_id = NULL
+d_mat_ses = data.matrix(d_mat_ses)
+row.names(d_mat_ses) = sids
+
+#refit model now with main DIF items removed
+mod_intuitive_sesn <- fit_mod_intuitive(d_mat_ses, ses_group)
+
+plot_glimmer(mod_intuitive_sesn, colnames(d_mat_ses)[2:ncol(d_mat_ses)])
+
+```
+```{r show-new-ses-glimmer, fig.width=3, fig.height=3, fig.cap="GLIMMER plot for sample of items showing most DIF in both directions, as well as a random sample from the middle."}
+ses_words <- c(ses_items_to_remove$definition) 
+#remove words
+d_mat_ses <- d_mat[, ! names(d_mat) %in% ses_words, drop = F]
+# there are more IDs in d_mat_ses than in d_demo; remove those
+d_mat_ses <- d_mat_ses %>% filter(is.element(data_id, d_demo$data_id)) %>% arrange(data_id)
+#copied from George's work may not be nessesary here
+sids = d_mat_ses$data_id
+d_mat_ses$data_id = NULL
+d_mat_ses = data.matrix(d_mat_ses)
+row.names(d_mat_ses) = sids
+
+plot_glimmer(mod_intuitive_sesn, colnames(d_mat_ses)[2:ncol(d_mat_ses)])
+```
+
+New Histogram with the DIF Items removed 
+```{r, fig.width=3, fig.height=3, echo=FALSE,warning=FALSE,message=FALSE,error=FALSE}
+#not sure why the groups are now high and low and not 0 and 1
+mm_sesn <- extract_group_df(mod_intuitive_sesn, groups=c("low","high"))
+ses_histn <- item_difficulty_difference_histogram(mm_sesn)
+
+orig_sesn <- get_extreme_item_difficulty_differences(mm_sesn, 2.25) # min:-0.97  max:1.08  (mean + 2SD) mean:0.051
+ses_histn
+```
+
+```{r ses-diff-table-new, fig.width=3, fig.height=3, echo=FALSE,warning=FALSE,message=FALSE,error=FALSE}
+new_ses_items_to_remove <- orig_sesn %>% filter(extreme==T)
+
+new_ses_items_to_remove %>% filter(extreme==T) %>%
+  arrange(d_diff) %>% 
+  select(-d_diff_abs, -extreme, -a1, -group1, -group2, -d_g1, -d_g2) %>%
+  mutate(ses_bias = case_when(d_diff < 0 ~ "Low", d_diff > 0 ~ "High")) %>%
+  kable(format = "html", digits = 2, table.attr = "style='width:40%;'")
+```
+List of new SES Items that show DIF outside of the 2SD range. (shows `r nrow(new_ses_items_to_remove)` new items that have DIF)
+
 ## Race/Ethnicity DIF
+Race is not a stand alone factor; people of each race fall into every combination of the other axes we have analyzed.
+Perhaps whiteness and SES are highly correlated in our data.
+Any binerized set of data will have members of each group that are different across another axis.
+It will perhap be worth while to specify even further such as the effect of white vs.
+non-white but only amongst high SES kids.
+
+```{r, eth-table}
+#d_demo <- d_demo %>% mutate(relevel() # remove empty "Other" level?
+
+table(d_demo$eth_group) %>% kable(format = "html", table.attr = "style='width:30%;'")
+```
+
+```{r show-eth-glimmer, fig.width=3, fig.height=3, fig.cap="GLIMMER plot for sample of items showing most DIF in both directions, as well as a random sample from the middle."}
+plot_glimmer(mod_intuitive_eth, colnames(d_mat)[2:ncol(d_mat)])
+```
+
+Histogram from ETH GLIMMER
+```{r, fig.width=3, fig.height=3, echo=FALSE,warning=FALSE,message=FALSE,error=FALSE}
+#used this to find out the that the two groups are 0 1 not high low, not sure why
+#groups <- c("0","1")
+#group_model <- mod_intuitive_eth
+#Mit = as_tibble(coef(group_model, simplify=T)) %>%
+#    mutate(definition = rownames(coef(group_model, simplify=T)[[groups[1]]]$items),
+#           group1 = groups[1],
+#          group2 = groups[2])
+
+mm_eth <- extract_group_df(mod_intuitive_eth, groups=c("Nonwhite","White"))
+eth_hist <- item_difficulty_difference_histogram(mm_eth)
+
+orig_eth <- get_extreme_item_difficulty_differences(mm_eth, 2.25) # min:-0.97  max:1.08  (mean + 2SD) mean:0.051
+eth_hist
+```
+
+```{r eth-diff-table, fig.width=3, fig.height=3, echo=FALSE,warning=FALSE,message=FALSE,error=FALSE}
+eth_items_to_remove <- orig_eth %>% filter(extreme==T)
+
+eth_items_to_remove %>% filter(extreme==T) %>%
+  arrange(d_diff) %>% 
+  select(-d_diff_abs, -extreme, -a1, -group1, -group2, -d_g1, -d_g2) %>%
+  mutate(ses_bias = case_when(d_diff < 0 ~ "Non-White", d_diff > 0 ~ "White")) %>%
+  kable(format = "html", digits = 2, table.attr = "style='width:40%;'")
+```
+List of Eth Items that show DIF outside of the 2SD range. (`r nrow(eth_items_to_remove)` that show DIF)
+
+```{r, eval=F}
+#do we need to have removed NA sex kids for the ses model
+d_demo <- d_demo %>% 
+  filter(comprehension!=0, !is.na(sex)) %>% # can't fit children not producing words, or with NA sex in group model
+  arrange(data_id) %>%
+  mutate(eth_group = ifelse(ethnicity=="White", "White", "Nonwhite"),
+         ses_group = 
+           ifelse(is.element(mom_ed, c("None", "Primary", "Some Secondary", "Secondary")), "low", "high"))
+#might need this
+d_demo <- d_demo %>% 
+  filter(!is.na(eth_group)
+
+eth_group = as.character(d_demo$eth_group)
+
+eth_words <- c(eth_items_to_remove$definition) 
+#remove words
+d_mat_eth <- d_mat[, ! names(d_mat) %in% eth_words, drop = F]
+# there are more IDs in d_mat_eth than in d_demo; remove those
+d_mat_eth <- d_mat_eth %>% filter(is.element(data_id, d_demo$data_id)) %>% arrange(data_id)
+#copied from George's work may not be nessesary here
+sids = d_mat_eth$data_id
+d_mat_eth$data_id = NULL
+d_mat_eth = data.matrix(d_mat_eth)
+row.names(d_mat_eth) = sids
+
+#refit model now with main DIF items removed
+mod_intuitive_ethn <- fit_mod_intuitive(d_mat_eth, eth_group)
+#recieved this note: NA values in group removed, along with associated rows in data
+#may need to run the "might need this" line to remove those NA kids ahead of time
+plot_glimmer(mod_intuitive_ethn, colnames(d_mat_eth)[2:ncol(d_mat_eth)])
+
+```
+```{r show-new-eth-glimmer, fig.width=3, fig.height=3, fig.cap="GLIMMER plot for sample of items showing most DIF in both directions, as well as a random sample from the middle."}
+eth_words <- c(eth_items_to_remove$definition) 
+#remove words
+d_mat_eth <- d_mat[, ! names(d_mat) %in% eth_words, drop = F]
+# there are more IDs in d_mat_eth than in d_demo; remove those
+d_mat_eth <- d_mat_eth %>% filter(is.element(data_id, d_demo$data_id)) %>% arrange(data_id)
+#copied from George's work may not be nessesary here
+sids = d_mat_eth$data_id
+d_mat_eth$data_id = NULL
+d_mat_eth = data.matrix(d_mat_eth)
+row.names(d_mat_eth) = sids
+
+plot_glimmer(mod_intuitive_ethn, colnames(d_mat_eth)[2:ncol(d_mat_eth)])
+```
+
+New Histogram with the DIF Items removed
+```{r, fig.width=3, fig.height=3, echo=FALSE,warning=FALSE,message=FALSE,error=FALSE}
+mm_ethn <- extract_group_df(mod_intuitive_ethn, groups=c("Nonwhite","White"))
+eth_histn <- item_difficulty_difference_histogram(mm_ethn)
+
+orig_ethn <- get_extreme_item_difficulty_differences(mm_ethn, 2.25) # min:-0.97  max:1.08  (mean + 2SD) mean:0.051
+eth_histn
+```
+
+```{r eth-diff-table-new, fig.width=3, fig.height=3, echo=FALSE,warning=FALSE,message=FALSE,error=FALSE}
+new_eth_items_to_remove <- orig_ethn %>% filter(extreme==T)
+
+new_eth_items_to_remove %>% filter(extreme==T) %>%
+  arrange(d_diff) %>% 
+  select(-d_diff_abs, -extreme, -a1, -group1, -group2, -d_g1, -d_g2) %>%
+  mutate(eth_bias = case_when(d_diff < 0 ~ "Non-White", d_diff > 0 ~ "White")) %>%
+  kable(format = "html", digits = 2, table.attr = "style='width:60%;'")
+```
+List of new Eth Items that show DIF outside of the 2SD range. (`r nrow(new_eth_items_to_remove)` that show DIF)
 
@@
 # breakdown white/non-white
+```
+
+```{r, eval=FALSE}
+#save all the new models
+save(mod_intuitive_sexn, mod_intuitive_ethn, mod_intuitive_sesn, mod_intuitive_sexn_n,
+       file="data/glimmer_prodWS_model_noExt.Rds")
+
 ```

### 2021-08-25 report_helper_functions.R

@@ -1,3 +1,8 @@
+# MIRT FUNCTION HELPERS ---------------------------------------------------
+fit_mod_intuitive <- function(data, groups){
+  multipleGroup(data, 1, itemtype = "Rasch", groups, invariance = "free_var", SE = TRUE, verbose = T)
+}
+
 extract_group_df <- function(group_model, groups=c("Male","Female")) {
   Mit = as_tibble(coef(group_model, simplify=T)[[groups[1]]]$items) %>%
     mutate(definition = rownames(coef(group_model, simplify=T)[[groups[1]]]$items),
@@ -16,11 +21,11 @@ extract_group_df <- function(group_model, groups=c("Male","Female")) {
   return(combo)
 }
 
-get_extreme_item_difficulty_differences <- function(mm) {
+get_extreme_item_difficulty_differences <- function(mm, SD=2) {
   #yfit <- dnorm(mm$d_diff, mean = mean(mm$d_diff), sd = sd(mm$d_diff)) 
   mm <- mm %>% mutate(d_diff = d_g2 - d_g1)
-  max_dif = mean(mm$d_diff) + 2*sd(mm$d_diff)
-  min_dif = mean(mm$d_diff) - 2*sd(mm$d_diff)
+  max_dif = mean(mm$d_diff) + SD*sd(mm$d_diff)
+  min_dif = mean(mm$d_diff) - SD*sd(mm$d_diff)
   mm <- mm %>% mutate(extreme = ifelse((d_diff > max_dif) | (d_diff < min_dif), T, F))
   print(paste("mininum difference:",min_dif, "maximum difference:",max_dif))
   return(mm)
@@ -63,7 +68,7 @@ mod_intuitive_to_draws_df <- function(mod){
 plot_glimmer <- function(mod_intuitive, item_names) {
   draws_df <- mod_intuitive %>% mod_intuitive_to_draws_df()
   names(draws_df) = c("run", item_names)
-  
+
   p_sample <- draws_df %>%
     clean_names() %>%
     gather(var, val, -run) %>%
@@ -74,9 +79,10 @@ plot_glimmer <- function(mod_intuitive, item_names) {
     arrange(mean_val)
   
   middle_index = floor(nrow(word_avg) / 2) # actual middle?
-  
+
+  mid <- sample(6:(nrow(word_avg)-5), 5, replace=F)
   desired_words <- bind_rows(word_avg %>% head(5),
-                             word_avg %>% sample_n(5), # might sample head/tail
[…]

### 2021-08-25 ws_prod_report.Rmd
added some notes

@@
 #d_demo <- d_demo %>% mutate(relevel() # remove empty "Other" level?
+#need this so that d_demo extists in the right form with you knit.
 d_demo <- d_demo %>% 
@@
 sex_hist <- item_difficulty_difference_histogram(mm_sex)
-
+#added the option to change SD threshold
 orig_sex <- get_extreme_item_difficulty_differences(mm_sex, 2.25) # min: -.53 max: 1.06 (mean + 2SD) mean: .265
@@
 ```{r, eval=F}
+#may no longer be nesesary now that its up top and I have already fit the models but I left it for now. 
 d_demo <- d_demo %>% 
@@
 #Mit = as_tibble(coef(group_model, simplify=T)[[groups[1]]]$items) %>%
-#    mutate(definition = rownames(coef(group_model, simplify=T)[[groups[1]]]$items),
-#           group1 = groups[1],
-#          group2 = groups[2])
 
@@
 ```{r, eval=F}
-#do we need to have removed NA sex kids for the ses model
+#do we need to have removed NA sex kids for the ses model?
 d_demo <- d_demo %>% 

### 2021-08-31 word_freq.Rmd
code to find the number of times a word is spoken to one group. plus updated helper functions

---
title: "Word_freq"
author: "Nathan, George, Mike"
date: "8/27/2021"
output: html_document
---

```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(cache = TRUE, warning = FALSE, message = FALSE)
library(dplyr)
library(dbplyr)
library(wordbankr)
library(childesr)
source("report_helper_functions.R")
```

```{r get-data, include=FALSE}
d_eng_na <- get_transcripts(collection = "Eng-NA")
head(d_eng_na)
sex_range <- get_participants(sex = "male", role_exclude = "Target_Child")
```

```{r wordbank}
data_mode <- "local"
languages <- "English"
admins <- get_administration_data(mode = data_mode) %>%
  select(language, form, sex, age, data_id) %>%
  filter(language %in% languages)
words <- get_item_data(mode = data_mode) %>%
  filter(type == "word", language %in% languages) %>%
  select(language, form, lexical_class, category, uni_lemma, definition,
         item_id) %>%
  mutate(num_item_id = as.numeric(substr(item_id, 6, nchar(item_id))),
         definition = tolower(definition))
```

```{r counts, include=FALSE}
exc_target_types <- get_types(#target_child = "Adam",
                          role_exclude = "target_child")
                          #type = c("dog", "ball"))
sex_group <- as.character(exc_target_types$target_child_sex)

items <- get_item_data(language = "English (American)", form = "WS")
CDI_words <- c(items$definition)
#potential issue in that the words they count include "accidentally" and "called" rather than the general form so we may need to lemmatize all the words in both or something. plus church* and mommy*
#plus id 120 is the name nathan for the child nathan which is child's_name in CDI but isn't reported as such in the wordbank and "drink (action)" is not recorded as such. 
target_types <- exc_target_types[exc_target_types$gloss %in% CDI_words, ] 
#sum the counts for spoken to boys removing the na sex values. 
#note truck had a huge difference in favor of boys and purse had a very small differnece for girls so I wonder if this is perfect
word_total <- sum(target_types[target_types$gloss == "nurse",]$count, na.rm=TRUE)
boy_tmw <- sum(target_types[(target_types$target_child_sex == "male")&(target_types$gloss == "nurse"),]$count, na.rm=TRUE)
girl_tmw <- sum(target_types[(target_types$target_child_sex == "female")&(target_types$gloss == "nurse"),]$count, na.rm=TRUE)
#freq_dif returns 
num <- freq_dif(target_types, "truck")
```

```{r find words, include=FALSE}
best_words = c()
for(word in CDI_words) {
  if((word %in% target_types$gloss)& (abs(freq_dif(target_types, word)) < 10) ){
    #print(abs(freq_dif(target_types, word)))
    best_words <- append(best_words, word)
  }
}
#finds no words that have a difference of 0 exactly 
```

### 2021-08-31 report_helper_functions.R

@@ -1,4 +1,11 @@
 # MIRT FUNCTION HELPERS ---------------------------------------------------
+#returns the difference between the number of times a words is spoke to boys vs. girls
+freq_dif <- function(data, target_word){
+  boy <- sum(data[(data$target_child_sex == "male")&(data$gloss == target_word),]$count, na.rm=TRUE)
+  girl <- sum(data[(data$target_child_sex == "female")&(data$gloss == target_word),]$count, na.rm=TRUE)
+  return(boy-girl)
+}
+
 fit_mod_intuitive <- function(data, groups){
   multipleGroup(data, 1, itemtype = "Rasch", groups, invariance = "free_var", SE = TRUE, verbose = T)
 }

### 2021-09-07 word_freq.Rmd
special cases counts and smoothing and log(freq)

@@
 knitr::opts_chunk$set(cache = TRUE, warning = FALSE, message = FALSE)
-library(dplyr)
+library(tidyverse)
 library(dbplyr)
@@
 library(childesr)
+library(glue)
 source("report_helper_functions.R")
@@
 ```{r wordbank}
-data_mode <- "local"
-languages <- "English"
+data_mode <- "remote"
+languages <- "English (American)"
 admins <- get_administration_data(mode = data_mode) %>%
@@
 ```{r counts, include=FALSE}
+#only english
 exc_target_types <- get_types(#target_child = "Adam",
-                          role_exclude = "target_child")
-                          #type = c("dog", "ball"))
+                          role_exclude = "target_child",
+                          language = "eng")
+                          # type = c("dog", "ball"))
+
 sex_group <- as.character(exc_target_types$target_child_sex)
 
+#have to nest because code expects nested DF
+#uni_lemma and definition are basically the same in english 
 items <- get_item_data(language = "English (American)", form = "WS")
+items_nested <- items %>%
+  filter(type == "word", !str_detect(definition, "name")) %>%
+  mutate(uni_lemma = definition) %>%
+  nest(items = -c(language, uni_lemma))
+uni_lemma_map <- build_uni_lemma_map(items_nested)
+
+# target_types <- exc_target_types[exc_target_types$gloss %in% CDI_words, ]
+
+#groups by sex and words(lower case) so now theres indevisual words with total count per gender. 
+target_type_counts <- exc_target_types %>%
+  filter(!is.na(target_child_sex)) %>%
+  mutate(gloss = str_to_lower(gloss)) %>%
+  group_by(language, gloss, target_child_sex) %>%
+  summarise(count = sum(count))
+
+#stem the gloss(the words in Childes)
+type_counts_stemmed <- target_type_counts %>%
+  mutate(gloss_stem = stem(gloss, convert_lang_stemmer("English (American)")))
+
+#joins the uni_lemmas from CDI with both gloss and stemmed gloss in Childes to get only the overlapping words
+type_counts_mapped <- type_counts_stemmed %>%
+  ungroup() %>%
+  select(-language) %>%
+  inner_join(uni_lemma_map %>% mutate(option = tolower(option)) %>% rename(gloss = option)) %>%
+  inner_join(uni_lemma_map %>% mutate(option = tolower(option)) %>% rename(gloss_stem = option)) %>%
+  select(-language, -gloss_stem)
+
+# TODO: normalize counts (over CDI words or over all of CHILDES? per sex or joined?)
+#re-summarizes by counts 
+type_counts <- type_counts_mapped %>%
+  group_by(target_child_sex, uni_lemma) %>%
+  summarise(count = sum(count))
+
+#finds missing items and makes their counts = 1
+missing_items <- items_nested %>%
+  filter(!(uni_lemma %in% type_counts$uni_lemma)) %>%
+  select(uni_lemma) %>%
+  mutate(female = 1, male = 1)
+
+#adds back in the unfound words with count 1 and adds ones to all other words
+type_counts_smoothed <- type_counts %>%
+  mutate(count = count + 1) %>%
+  pivot_wider(names_from = target_child_sex, values_from = count, values_fill = 1) %>%
+  bind_rows(missing_items) %>%
+  pivot_longer(cols = c(female, male), names_to = "sex", values_to = "count")
+
+# TODO: normalize counts (over CDI words or over all of CHILDES? per sex or joined?)
+#smoothing
+type_freqs <- type_counts_smoothed %>%
+  mutate(freq = log(count / sum(count)))
+
+# saveRDS(type_counts, "data/type_counts.rds")
+# saveRDS(type_freqs, "data/type_freqs.rds")
+
+# TODO: this is log ratio of probabilities, if doing log odds ratio transform to odds before taking difference
+type_diffs <- type_freqs %>%
+  select(-count) %>%
+  pivot_wider(names_from = sex, values_from = freq) %>%
+  mutate(freq_ratio = female - male)
+
+qplot(type_diffs$freq_ratio)
+
 CDI_words <- c(items$definition)
@@
 #plus id 120 is the name nathan for the child nathan which is child's_name in CDI but isn't reported as such in the wordbank and "drink (action)" is not recorded as such. 
-target_types <- exc_target_types[exc_target_types$gloss %in% CDI_words, ] 
+#target_types <- exc_target_types[exc_target_types$gloss %in% CDI_words, ] 
 #sum the counts for spoken to boys removing the na sex values. 
-#note truck had a huge difference in favor of boys and purse had a very small differnece for girls so I wonder if this is perfect
+#note truck had a huge difference in favor of boys and purse had a very small difference for girls so I wonder if this is perfect
 word_total <- sum(target_types[target_types$gloss == "nurse",]$count, na.rm=TRUE)

### 2021-09-07 report_helper_functions.R

@@ -100,3 +100,130 @@ plot_glimmer <- function(mod_intuitive, item_names) {
   
   return(p)
 }
+
+## childes_new
+lang_map <- read_csv("resources/language_map.csv")
+
+normalize_language <- function(language) {
+  language %>% str_replace(" ", "_") %>% str_to_lower()
+}
+
+convert_lang_stemmer <- function(lang, method = "snowball") {
+  lang_map %>% filter(wordbank == lang) %>% pull(get(method))
+}
+
+transforms <- list(
+  function(s) str_replace_all(s, "(.*) \\(.*\\)", "\\1"), # foo (bar) -> foo
+  function(s) str_replace_all(s, " ", "_"), # foo bar -> foo_bar
+  function(s) str_replace_all(s, " ", "+"), # foo bar -> foo+bar
+  function(s) str_replace_all(s, "(.+) \\1", "\\1") # (foo) bar -> bar
+  #   function(x) paste0(x, "e+moi"),
+  #   function(x) paste0(x, "e-moi"),
+  #   function(x) paste0(x, "+moi"),
+  #   function(x) paste0(x, "-moi"),
+  #   function(x) paste0(x, "ent"),
+  #   function(x) paste0(x, "e-l"),
+  #   function(x) paste0(x, "e-toi"),
+  #   function(x) paste0(x, "-l"),
+  #   function(x) paste0(x, "-toi"),
+  #   function(x) paste0(x, "es-tu"),
+  #   function(x) paste0(x, "s+tu"),
+  #   function(x) paste0(x, "s-moi")
+)
+
+build_special_case_map <- function(lang) {
+  norm_lang <- normalize_language(lang)
+  special_case_file <- glue("resources/{norm_lang}.csv")
+  if (file.exists(special_case_file)) {
+    a<-read_csv(special_case_file, col_names = FALSE) %>%
+      rename(uni_lemma = X1, definition = X2) %>%
+      pivot_longer(-c(uni_lemma, definition),
+                   names_to = "x", values_to = "option") %>%
+      filter(!is.na(option)) %>%
+      select(-x) %>%
+      mutate(language = lang)
+  }
+  else{
+    a<-data.frame(matrix(ncol=4,nrow=0, dimnames=list(NULL, c("uni_lemma", "definition", "option", "language")))) %>%
+      mutate(language = lang, uni_lemma = as.character(uni_lemma), definition = as.character(definition), option = as.character(option) )
+  }
+  return(a)
+}
+
[…]

### 2021-09-09 word_freq.Rmd
smoothed counts and Childes words

@@
 
-#groups by sex and words(lower case) so now theres indevisual words with total count per gender. 
+#groups by sex and words(lower case) so now theres individual words with total count per gender. 
 target_type_counts <- exc_target_types %>%
@@
   summarise(count = sum(count))
+#stem childes words and get counts to use as recomended replacements
 
@@
   mutate(gloss_stem = stem(gloss, convert_lang_stemmer("English (American)")))
+
+#group stemmed words and counts
+childes_replace <- type_counts_stemmed %>%
+  group_by(gloss_stem, target_child_sex) %>%
+  summarise(count = sum(count))
+
+normalized_counts <- childes_replace %>%
+  group_by(target_child_sex) %>%
+  mutate(prob = count/sum(count), 
+         count_per_mil =  prob * 1000000)
+
+#totals of counts
+total_counts <- childes_replace %>%
+   group_by(target_child_sex) %>%
+   summarize(count = sum(count))
 
@@
   summarise(count = sum(count))
+  #1 is female 2 is male
+  #mutate(count_per_mil = 1000000 * (count/ifelse(target_child_sex == "male", total_counts$count[2], total_counts$count[1])))
 
 #finds missing items and makes their counts = 1
+#missing mostly stared words like "penis*" and "church*"
 missing_items <- items_nested %>%
@@
   bind_rows(missing_items) %>%
-  pivot_longer(cols = c(female, male), names_to = "sex", values_to = "count")
+  pivot_longer(cols = c(female, male), names_to = "sex", values_to = "count") %>%
+  #1 is female 2 is male
+  mutate(count_per_mil = 1000000 * (count/ifelse(sex == "male", total_counts$count[2], total_counts$count[1])))
 
@@
 # saveRDS(type_freqs, "data/type_freqs.rds")
+#saveRDS(normalized_counts, "data/normalized_childes_counts.rds")
+#saveRDS(type_counts_smoothed, "data/type_counts_smoothed.rds")
 

### 2021-09-21 ws_prod_report.Rmd
add comparison of by-sex CHILDES word freqs to DIF

@@
 ```
+
 ```{r show-new-sex-glimmer,echo= FALSE, warning=FALSE,message=FALSE,error=FALSE, fig.width=3, fig.height=3, fig.cap="GLIMMER plot for sample of items showing most DIF in both directions, as well as a random sample from the middle."}
@@
 ```
+
 From seeing this my take away is that there are more female DIF Items total but that the extreme ones are mostly male.
+
 ```{r sex-words 3rd, fig.width=3, fig.height=3, echo=FALSE,warning=FALSE,message=FALSE,error=FALSE}
@@
 `r nrow(nnew_sex_items_to_remove)` item that now show DIF
+
+### Suggesting replacement items
+
+```{r}
+type_freqs = readRDS("data/type_freqs.rds")
+type_counts = readRDS("data/type_counts.rds")
+type_counts_smoothed = readRDS("data/type_counts_smoothed.rds")
+
+type_counts_smoothed <- type_counts_smoothed %>%
+  pivot_wider(id_cols=uni_lemma, names_from=sex, values_from=count_per_mil) %>%
+  mutate(propFemale = female / (female + male),
+         FtoM = female / male)
+
+# how biased are the items we suggest removing in the corpus counts?
+sex_items_to_remove$definition
+
+tmp <- left_join(orig_sex, type_counts_smoothed, by=c("definition"="uni_lemma"))
+
+tmp %>% #filter(extreme==T) %>%
+  ggplot(aes(x=propFemale, y=d_diff, color=extreme)) + 
+  geom_point(alpha=.5) + theme_bw() +
+  ylab("Female - Male Word Difficulty") +
+  xlab("Proportion of Utterances to Females in CHILDES")
+
+subset(tmp, extreme==T & propFemale > .47 & propFemale < .53) # hose & dump
+
+cor.test(tmp$propFemale, tmp$d_diff) # .19, p<.001
+
+# for replacements, want equal-frequency by sex (and high-freq overall) words from CHILDES
+normalized_childes_counts = readRDS("data/normalized_childes_counts.rds")
+
+childes_rel_freq <- normalized_childes_counts %>% filter(count_per_mil > 10) %>%
+  pivot_wider(id_cols=gloss_stem, names_from=target_child_sex, values_from=count_per_mil) %>%
+  mutate(propFemale = female / (female + male))
+# 8717 words to choose from
+hist(childes_rel_freq$propFemale)
+mean(childes_rel_freq$propFemale, na.rm=T) # .497
+
+replacements <- childes_rel_freq %>% 
+  filter(propFemale < .51, propFemale > .49) %>%
+  filter(!is.element(gloss_stem, en_ws$definition)) # 182
+```
+
+1. Of the 22 items with extreme sex DIF, only 2 of those fall very close to being equally often spoken to boys and girls ('hose' and 'dump')
+2. There are many items on the CDI with small difficulty difference (DIF) by sex
 

### 2021-10-12 paper/cogsci.Rmd
starting on cogsci paper

---
title: "Investigating Item Bias "
bibliography: library.bib
csl: apa6.csl
document-params: "10pt, letterpaper"

author-information: > 
    \author{{\large \bf Morton Ann Gernsbacher (MAG@Macc.Wisc.Edu)} \\ Department of Psychology, 1202 W. Johnson Street \\ Madison, WI 53706 USA
    \AND {\large \bf Sharon J.~Derry (SDJ@Macc.Wisc.Edu)} \\ Department of Educational Psychology, 1025 W. Johnson Street \\ Madison, WI 53706 USA}

abstract: >
    Include no author information in the initial submission, to facilitate
    blind review.  The abstract should be one paragraph, indented 1/8 inch on both sides,
    in 9~point font with single spacing. The heading 'Abstract'
    should be 10~point, bold, centered, with one line of space below
    it. This one-paragraph abstract section is required only for standard
    six page proceedings papers. Following the abstract should be a blank
    line, followed by the header 'Keywords' and a list of
    descriptive keywords separated by semicolons, all in 9~point font, as
    shown below.

keywords: >
    language development;

output: cogsci2016::cogsci_paper
#final-submission: \cogscifinalcopy
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=3, fig.height=3, fig.crop = F, 
                      fig.pos = "tb", fig.path='figs/',
                      echo=F, warning=F, cache=F, 
                      message=F, sanitize = T)
```

```{r, libraries}
library(png)
library(grid)
library(ggplot2)
library(xtable)
library(tidyverse)

```

# Introduction

- introduce importance of measuring early language development (e.g., links to later educational outcomes)
- introduce CDI
- what is a good measure? 
- IRT
- what does between-group bias look like?

- what is our goal for measuring vocabulary? - identifying language delays, predicting later reading, or talking...
- how you select items influences how well you achieve these goals (and what bias you find)

# Methods

## Dataset

We analyze American English CDI: Words & Sentences administrations from `r nrow(d_demo)` children aged `r min(d_demo$age)` to `r max(d_demo$age)` from Wordbank [@Frank2016].

```{r import-data}
load(here("data/en_ws_production.Rdata")) # d_demo, d_mat, en_wg (items)

```

## Second-Level Headings

Second level headings should be 11 point , initial caps, bold, and
flush left. Leave one line space above the heading and 1/4~ line
space below the heading.

### Third-Level Headings

Third-level headings should be 10 point , initial caps, bold, and flush
left. Leave one line space above the heading, but no space after the
heading.

# Formalities, Footnotes, and Floats

Use standard APA citation format. Citations within the text should
include the author's last name and year. If the authors' names are
included in the sentence, place only the year in parentheses, as in
[-@NewellSimon1972a], but otherwise place the entire reference in
parentheses with the authors and year separated by a comma
[@NewellSimon1972a]. List multiple references alphabetically and
separate them by semicolons [@ChalnickBillman1988a; @NewellSimon1972a]. 
Use the et. al. construction only after listing all the authors to a
publication in an earlier reference and for citations with four or
more authors.

For more information on citations in RMarkdown, see **[here](http://rmarkdown.rstudio.com/authoring_bibliographies_and_citations.html#citations).**

## Footnotes

Indicate footnotes with a number\footnote{Sample of the first
footnote.} in the text. Place the footnotes in 9 point type at the
bottom of the page on which they appear. Precede the footnote with a
horizontal rule.\footnote{Sample of the second footnote.} You can also use 
markdown formatting to include footnotes using this syntax [^1].

[^1]: Sample of a markdown footnote.

## Figures

All artwork must be very dark for purposes of reproduction and should
not be hand drawn. Number figures sequentially, placing the figure
number and caption, in 10 point, after the figure with one line space
above the caption and one line space below it. If necessary, leave extra white space at
the bottom of the page to avoid splitting the figure and figure
caption. You may float figures to the top or bottom of a column, or
set wide figures across both columns.

## Two-column images

You can read local images using png package for example and plot 
it like a regular plot using grid.raster from the grid package. 
With this method you have full control of the size of your image. **Note: Image must be in .png file format for the readPNG function to work.**

You might want to display a wide figure across both columns. To do this, you change the `fig.env` chunk option to `figure*`. To align the image in the center of the page, set `fig.align` option to `center`. To format the width of your caption text, you set the `num.cols.cap` option to `2`.

```{r 2-col-image, fig.env = "figure*", fig.pos = "h", fig.width=4, fig.height=2, fig.align = "center", set.cap.width=T, num.cols.cap=2, fig.cap = "This image spans both columns. And the caption text is limited to 0.8 of the width of the document."}
img <- png::readPNG("figs/walrus.png")
grid::grid.raster(img)
```

## One-column images

Single column is the default option, but if you want set it explicitly, set `fig.env` to `figure`. Notice that the `num.cols` option for the caption width is set to `1`.

```{r image, fig.env = "figure", fig.pos = "H", fig.align='center', fig.width=2, fig.height=2, set.cap.width=T, num.cols.cap=1, fig.cap = "One column image."}
img <- png::readPNG("figs/lab_logo_stanford.png")
grid::grid.raster(img)
```

## R Plots

You can use R chunks directly to plot graphs. And you can use latex floats in the
fig.pos chunk option to have more control over the location of your plot on the page. For more information on latex placement specifiers see **[here](https://en.wikibooks.org/wiki/LaTeX/Floats,_Figures_and_Captions)**

```{r plot, fig.env="figure", fig.pos = "H", fig.align = "center", fig.width=2, fig.height=2, fig.cap = "R plot" }
x <- 0:100
y <- 2 * (x + rnorm(length(x), sd = 3) + 3)

ggplot2::ggplot(data = data.frame(x, y), 
                aes(x = x, y = y)) + 
  geom_point() + 
  geom_smooth(method = "lm")
```

## Tables

Number tables consecutively; place the table number and title (in
10 point) above the table with one line space above the caption and
one line space below it, as in Table 1. You may float
tables to the top or bottom of a column, set wide tables across both
columns.

You can use the xtable function in the xtable package.

```{r xtable, results="asis"}
n <- 100
x <- rnorm(n)
y <- 2*x + rnorm(n)
out <- lm(y ~ x)

tab1 <- xtable::xtable(summary(out)$coef, digits=c(0, 2, 2, 1, 2), 
                       caption = "This table prints across one column.")

print(tab1, type="latex", comment = F, table.placement = "H")
```

# Acknowledgements

Place acknowledgments (including funding information) in a section at
the end of the paper.

# References

```{r}
# References will be generated automatically by Pandoc and included here.
# The following code is some latex to format the bibliography. Do not remove it.
```

\setlength{\parindent}{-0.1in} 
\setlength{\leftskip}{0.125in}
\noindent

### 2021-10-20 paper/cogsci.Rmd
outline and methods updates

@@
 - introduce CDI
-- what is a good measure? 
-- IRT
+- predictive validity: Fenson et al. 1994, Bornstein et al. 2012, Duff et al. 2015
+
+- discuss potential for bias
 - what does between-group bias look like?
+
+- IRT - what is a good measure?
+
+## Fundamental DIF Problem
 
@@
 - how you select items influences how well you achieve these goals (and what bias you find)
+We take the approach put forward by @Stenhaug2021: GLIMMER plots based on the 1-parameter logistic are sufficient to identify bias, without potentially obscuring DIF in a more complex model's additional parameters.
 
@@
 
-We analyze American English CDI: Words & Sentences administrations from `r nrow(d_demo)` children aged `r min(d_demo$age)` to `r max(d_demo$age)` from Wordbank [@Frank2016].
+We analyze American English CDI: Words & Sentences administrations from `r nrow(d_demo)` children aged `r min(d_demo$age)` to `r max(d_demo$age)` from Wordbank [@Frank2016; @Frank2021].
 
@@
 
-## Second-Level Headings
+### Participants
 
-Second level headings should be 11 point , initial caps, bold, and
-flush left. Leave one line space above the heading and 1/4~ line
-space below the heading.
+```{r}
+sex_tab <- table(d_demo$sex)
 
-### Third-Level Headings
+eth_tab <- table(d_demo$ethnicity)
+```
 
-Third-level headings should be 10 point , initial caps, bold, and flush
-left. Leave one line space above the heading, but no space after the
-heading.
+We use Wordbank data from `r sex_tab["Female"]` female and `r sex_tab["Male"]` male participants, and was unreported for an additional `r sum(is.na(d_demo$sex))` participants.
+<!-- same for eth, SES -->
 
-# Formalities, Footnotes, and Floats
-
-Use standard APA citation format. Citations within the text should
-include the author's last name and year. If the authors' names are
-included in the sentence, place only the year in parentheses, as in
-[-@NewellSimon1972a], but otherwise place the entire reference in
-parentheses with the authors and year separated by a comma
-[@NewellSimon1972a]. List multiple references alphabetically and
-separate them by semicolons [@ChalnickBillman1988a; @NewellSimon1972a]. 
-Use the et. al. construction only after listing all the authors to a
-publication in an earlier reference and for citations with four or
-more authors.
-
-For more information on citations in RMarkdown, see **[here](http://rmarkdown.rstudio.com/authoring_bibliographies_and_citations.html#citations).**
-
-## Footnotes
-
-Indicate footnotes with a number\footnote{Sample of the first
-footnote.} in the text. Place the footnotes in 9 point type at the
-bottom of the page on which they appear. Precede the footnote with a
-horizontal rule.\footnote{Sample of the second footnote.} You can also use 
-markdown formatting to include footnotes using this syntax [^1].
-
-[^1]: Sample of a markdown footnote.
-
-## Figures
-
-All artwork must be very dark for purposes of reproduction and should
-not be hand drawn. Number figures sequentially, placing the figure
-number and caption, in 10 point, after the figure with one line space
-above the caption and one line space below it. If necessary, leave extra white space at
-the bottom of the page to avoid splitting the figure and figure
-caption. You may float figures to the top or bottom of a column, or
-set wide figures across both columns.
-
-## Two-column images
-
-You can read local images using png package for example and plot 
-it like a regular plot using grid.raster from the grid package. 
-With this method you have full control of the size of your image. **Note: Image must be in .png file format for the readPNG function to work.**
-
-You might want to display a wide figure across both columns.
-To do this, you change the `fig.env` chunk option to `figure*`.
-To align the image in the center of the page, set `fig.align` option to `center`.
-To format the width of your caption text, you set the `num.cols.cap` option to `2`.
-
+<!-- goal: make custom glimmer plot in R with high / low bias words -->
 ```{r 2-col-image, fig.env = "figure*", fig.pos = "h", fig.width=4, fig.height=2, fig.align = "center", set.cap.width=T, num.cols.cap=2, fig.cap = "This image spans both columns.

### 2021-10-27 paper/cogsci.Rmd
The participants section

@@
 library(tidyverse)
+library(here)
 
@@
 
-We analyze American English CDI: Words & Sentences administrations from `r nrow(d_demo)` children aged `r min(d_demo$age)` to `r max(d_demo$age)` from Wordbank [@Frank2016; @Frank2021].
-
 ```{r import-data}
@@
 
+#Add this exclusion to the Analysis 
+removed <- d_demo %>%
+  filter(comprehension!=0)
+#this is the actually d_demo we used so we must use this in our report
+d_demo <- d_demo %>% 
+  #filter(comprehension!=0, !is.na(sex)) %>% # can't fit children not producing words, or with NA sex in group model
+  arrange(data_id) %>%
+  mutate(eth_group = ifelse(ethnicity=="White", "White", "Nonwhite"),
+         ses_group = 
+           ifelse(is.element(mom_ed, c("None", "Primary", "Some Secondary", "Secondary")), "low", "high"))
 ```
+We analyze American English CDI: Words & Sentences administrations from `r nrow(d_demo)` children aged `r min(d_demo$age)` to `r max(d_demo$age)` from Wordbank [@Frank2016; @Frank2021].
 
@@
 sex_tab <- table(d_demo$sex)
-
+ses_tab <- table(d_demo$ses_group)
 eth_tab <- table(d_demo$ethnicity)
+eth2_tab <- table(d_demo$eth_group)
 ```
 
-We use Wordbank data from `r sex_tab["Female"]` female and `r sex_tab["Male"]` male participants, and was unreported for an additional `r sum(is.na(d_demo$sex))` participants.
+Research was conducted over  `r nrow(d_demo)` participants, whose data was parent reported and collected by the Wordbank Project.
+Unlike other projects Wordback relies on the kindness of others to contribute their data which means that often meta-data for some sets is missing.
+Even so, the project now has data over 29 different languages on children from ages 8-30 months old, though our study focused entirely on American English and Children 16-30 months.
+When they received complete information, the Wordbank data set collected demographic data consisting of Birth Order, Race/Ethnicity, Sex and Mothers Education.
+For our study we focused on the comparing data between demographic groups on the axis of Sex, Ethnicity and Mothers Education, which was a proxy for social economic status.
+Out data included `r sex_tab["Female"]` female and `r sex_tab["Male"]` male participants, and an additional `r sum(is.na(d_demo$sex))` participants whose sex remained unreported.
+In terms of Ethnicity data was recorded from `r eth_tab["White"]` white, `r eth_tab["Asian"]` Asian, `r eth_tab["Black"]` Black, `r eth_tab["Hispanic"]` Hispanic, and `r eth_tab["Other"]` “Other” participants as well as `r sum(is.na(d_demo$ethnicity))` unreported.
+Given our distribution of ethnicities, for our analysis we split participants into White (`r eth2_tab["White"]`) and Non-White (`r eth2_tab["Nonwhite"]`).
+Finally, we split participants into high and low SES, of which we had `r ses_tab["high"]` high SES participants, corresponding to mothers' education of some college or higher and `r ses_tab["low"]` Low SES participants.
 <!-- same for eth, SES -->

### 2021-10-27 paper/cogsci.tex
The participants section

% Template for Cogsci submission with R Markdown

% Stuff changed from original Markdown PLOS Template
\documentclass[10pt, letterpaper]{article}

\usepackage{cogsci}
\usepackage{pslatex}
\usepackage{float}
\usepackage{caption}

% amsmath package, useful for mathematical formulas
\usepackage{amsmath}

% amssymb package, useful for mathematical symbols
\usepackage{amssymb}

% hyperref package, useful for hyperlinks
\usepackage{hyperref}

% graphicx package, useful for including eps and pdf graphics
% include graphics with the command \includegraphics
\usepackage{graphicx}

% Sweave(-like)
\usepackage{fancyvrb}
\DefineVerbatimEnvironment{Sinput}{Verbatim}{fontshape=sl}
\DefineVerbatimEnvironment{Soutput}{Verbatim}{}
\DefineVerbatimEnvironment{Scode}{Verbatim}{fontshape=sl}
\newenvironment{Schunk}{}{}
\DefineVerbatimEnvironment{Code}{Verbatim}{}
\DefineVerbatimEnvironment{CodeInput}{Verbatim}{fontshape=sl}
\DefineVerbatimEnvironment{CodeOutput}{Verbatim}{}
\newenvironment{CodeChunk}{}{}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{apacite}

% KM added 1/4/18 to allow control of blind submission

\usepackage{color}

% Use doublespacing - comment out for single spacing
%\usepackage{setspace}
%\doublespacing

% % Text layout
% \topmargin 0.0cm
% \oddsidemargin 0.5cm
% \evensidemargin 0.5cm
% \textwidth 16cm
% \textheight 21cm

\title{Investigating Item Bias}

\author{{\large \bf Morton Ann Gernsbacher (MAG@Macc.Wisc.Edu)} \\ Department of Psychology, 1202 W. Johnson Street \\ Madison, WI 53706 USA \AND {\large \bf Sharon J.~Derry (SDJ@Macc.Wisc.Edu)} \\ Department of Educational Psychology, 1025 W. Johnson Street \\ Madison, WI 53706 USA}

\begin{document}

\maketitle

\begin{abstract}
Include no author information in the initial submission, to facilitate
blind review. The abstract should be one paragraph, indented 1/8 inch on
both sides, in 9\textasciitilde point font with single spacing. The
heading `Abstract' should be 10\textasciitilde point, bold, centered,
with one line of space below it. This one-paragraph abstract section is
required only for standard six page proceedings papers. Following the
abstract should be a blank line, followed by the header `Keywords' and a
list of descriptive keywords separated by semicolons, all in
9\textasciitilde point font, as shown below.

\textbf{Keywords:}
language development;
\end{abstract}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

\begin{itemize}
\item
  introduce importance of measuring early language development (e.g.,
  links to later educational outcomes)
\item
  introduce CDI
\item
  predictive validity: Fenson et al.~1994, Bornstein et al.~2012, Duff
  et al.~2015
\item
  discuss potential for bias
\item
  what does between-group bias look like?
\item
  IRT - what is a good measure?
\end{itemize}

\hypertarget{fundamental-dif-problem}{%
\subsection{Fundamental DIF Problem}\label{fundamental-dif-problem}}

\begin{itemize}
\tightlist
\item
  what is our goal for measuring vocabulary? - identifying language
  delays, predicting later reading, or talking\ldots{}
\item
  how you select items influences how well you achieve these goals (and
  what bias you find) We take the approach put forward by
  (\textbf{Stenhaug2021?}): GLIMMER plots based on the 1-parameter
  logistic are sufficient to identify bias, without potentially
  obscuring DIF in a more complex model's additional parameters.
\end{itemize}

\hypertarget{methods}{%
\section{Methods}\label{methods}}

\hypertarget{dataset}{%
\subsection{Dataset}\label{dataset}}

We analyze American English CDI: Words \& Sentences administrations from
5520 children aged 16 to 30 from Wordbank (\textbf{Frank2016?};
\textbf{Frank2021?}).

\hypertarget{participants}{%
\subsubsection{Participants}\label{participants}}

Research was conducted over 5520 participants, whose data was parent
reported and collected by the Wordbank Project. Unlike other projects
Wordback relies on the kindness of others to contribute their data which
means that often meta-data for some sets is missing. Even so, the
project now has data over 29 different languages on children from ages
8-30 months old, though our study focused entirely on American English
and Children 16-30 months. When they received complete information, the
Wordbank data set collected demographic data consisting of Birth Order,
Race/Ethnicity, Sex and Mothers Education. For our study we focused on
the comparing data between demographic groups on the axis of Sex,
Ethnicity and Mothers Education, which was a proxy for social economic
status. Out data included 1989 female and 2105 male participants, and an
additional 1426 participants whose sex remained unreported. In terms of
Ethnicity data was recorded from 2202 white, 67 Asian, 222 Black, 131
Hispanic, and 93 ``Other'' participants as well as 2805 unreported.
Given our distribution of ethnicities, for our analysis we split
participants into White (2202) and Non-White (513). Finally, we split
participants into high and low SES, of which we had NA high SES
participants, corresponding to mothers' education of some college or
higher and NA Low SES participants.

\bibliographystyle{apacite}

\end{document}

### 2021-10-27 paper/cogsci.Rmd
adding refs, some edits and fixes

@@
 ---
-title: "Investigating Item Bias "
+title: "Words aren't created equal: Investigating bias on the CDI"
 bibliography: library.bib
@@
 - how you select items influences how well you achieve these goals (and what bias you find)
-We take the approach put forward by @Stenhaug2021: GLIMMER plots based on the 1-parameter logistic are sufficient to identify bias, without potentially obscuring DIF in a more complex model's additional parameters.
+We take the approach put forward by @stenhaug2021treading: GLIMMER plots based on the 1-parameter logistic are sufficient to identify bias, without potentially obscuring DIF in a more complex model's additional parameters.
+
+Differential item functioning (DIF) is a technique in IRT used to identify items that show bias against demographic groups. 
+DIF is a statistical characteristic of an item that shows the extent to which the item might operate differently or measure varying abilities for subgroups and members of separate demographic groups. 
+DIF is, however, a challenging metric for identifying bias for a couple reasons. 
+Firstly, DIF is extremely hard to accurately root out. 
+The process of finding DIF involves using the test you used to do IRT analysis in order to look for bias within the very same test. 
+Secondly, the presence of DIF does not necessarily indicate the presence of bias; it can indicate that one demographic has a higher average ability level than the other. 
+These two issues are the basis upon which we get the Fundamental DIF Identification problem.
+
+This problem can be be understood through a concrete example that is well-known from research on early language learning: consider the fact that females show a larger vocabulary than males across early development [see @Frank2021 Ch.
+6?; OTHER REFS].
+The question however is whether or not girls actually have a higher ability level or if the tests have words that are bias toward girls ie.
+are easier for girls, or perhaps the test is biased toward boys but the natural ability of girls is enough to overcome that bias.
+To answer this question we need to know the ability levels of girls and boys in order to confirm that if a word is learned earlier by girls it is simply because of an ability difference rather than a bias inherent to the word.
+The problem arises from the fact that we measure ability level using the very same test that we are trying to check for biased words.
+If the boys and girls were of the same ability level this wouldn’t be an issue but given the evidence showing that girls have a higher ability level, we need to know the difference in ability so that when we find DIF for a specific word can be sure it is outside of the expecting DIF that results naturally from their difference in ability.
 
@@
 ```
-We analyze American English CDI: Words & Sentences administrations from `r nrow(d_demo)` children aged `r min(d_demo$age)` to `r max(d_demo$age)` from Wordbank [@Frank2016; @Frank2021].
+We analyze American English CDI: Words & Sentences administrations from `r nrow(d_demo)` children aged `r min(d_demo$age)` to `r max(d_demo$age)` from Wordbank [@Frank2017; @Frank2021].
 
@@
 
-```{r}
+```{r, echo=F}
 sex_tab <- table(d_demo$sex)
@@
 Research was conducted over  `r nrow(d_demo)` participants, whose data was parent reported and collected by the Wordbank Project. 
-Unlike other projects Wordback relies on the kindness of others to contribute their data which means that often meta-data for some sets is missing.
+Unlike other projects Wordbank relies on the kindness of others to contribute their data which means that often meta-data for some sets is missing. 
 Even so, the project now has data over 29 different languages on children from ages 8-30 months old, though our study focused entirely on American English and Children 16-30 months. 
 When they received complete information, the Wordbank data set collected demographic data consisting of Birth Order, Race/Ethnicity, Sex and Mothers Education. 
-For our study we focused on the comparing data between demographic groups on the axis of Sex, Ethnicity and Mothers Education, which was a proxy for social economic status.
-Out data included `r sex_tab["Female"]` female and `r sex_tab["Male"]` male participants, and an additional `r sum(is.na(d_demo$sex))` participants whose sex remained unreported.
+We focused on comparing data between demographic groups on the axes of Sex, Ethnicity and Mother's Education, a proxy for socioeconomic status (henceforth, SES). 
+Our data included `r sex_tab["Female"]` female and `r sex_tab["Male"]` male participants, and an additional `r sum(is.na(d_demo$sex))` participants whose sex remained unreported. 
 In terms of Ethnicity data was recorded from `r eth_tab["White"]` white, `r eth_tab["Asian"]` Asian, `r eth_tab["Black"]` Black, `r eth_tab["Hispanic"]` Hispanic, and `r eth_tab["Other"]` “Other” participants as well as `r sum(is.na(d_demo$ethnicity))` unreported.
@@
 And the caption text is limited to 0.8 of the width of the document."}
-img <- png::readPNG("figs/walrus.png")
-grid::grid.raster(img)
+#img <- png::readPNG("figs/walrus.png")
+#grid::grid.raster(img)
 ```
@@
 
-Single column is the default option, but if you want set it explicitly, set `fig.env` to `figure`. Notice that the `num.cols` option for the caption width is set to `1`.
-
 ```{r image, fig.env = "figure", fig.pos = "H", fig.align='center', fig.width=2, fig.height=2, set.cap.width=T, num.cols.cap=1, fig.cap = "One column image."}
-img <- png::readPNG("figs/lab_logo_stanford.png")
-grid::grid.raster(img)
-```
-
-## R Plots
-
-You can use R chunks directly to plot graphs. And you can use latex floats in the
-fig.pos chunk option to have more control over the location of your plot on the page.
-For more information on latex placement specifiers see **[here](https://en.wikibooks.org/wiki/LaTeX/Floats,_Figures_and_Captions)**
-
-```{r plot, fig.env="figure", fig.pos = "H", fig.align = "center", fig.width=2, fig.height=2, fig.cap = "R plot" }
-x <- 0:100
-y <- 2 * (x + rnorm(length(x), sd = 3) + 3)
-
-ggplot2::ggplot(data = data.frame(x, y), 
-                aes(x = x, y = y)) + 
-  geom_point() + 
-  geom_smooth(method = "lm")
-```
-
-## Tables
-
-Number tables consecutively; place the table number and title (in
-10 point) above the table with one line space above the caption and
-one line space below it, as in Table 1. You may float
-tables to the top or bottom of a column, set wide tables across both
-columns.
-
-You can use the xtable function in the xtable package.
-
-```{r xtable, results="asis"}
-n <- 100
-x <- rnorm(n)
-y <- 2*x + rnorm(n)
-out <- lm(y ~ x)
-
-tab1 <- xtable::xtable(summary(out)$coef, digits=c(0, 2, 2, 1, 2), 
-                       caption = "This table prints across one column.")
-
-print(tab1, type="latex", comment = F, table.placement = "H")
+#img <- png::readPNG("figs/lab_logo_stanford.png")
+#grid::grid.raster(img)
 ```

### 2021-10-29 paper/cogsci.Rmd
wrote an abstract

@@
 abstract: >
-    Include no author information in the initial submission, to facilitate
-    blind review.  The abstract should be one paragraph, indented 1/8 inch on both sides,
-    in 9~point font with single spacing. The heading 'Abstract'
-    should be 10~point, bold, centered, with one line of space below
-    it. This one-paragraph abstract section is required only for standard
-    six page proceedings papers. Following the abstract should be a blank
-    line, followed by the header 'Keywords' and a list of
-    descriptive keywords separated by semicolons, all in 9~point font, as
-    shown below.
+    Children's early language skill has been linked to later educational outcomes, making
+    it important to accurately measure early language. Parent-reported instruments such as the 
+    Communicative Development Inventories (CDIs) have been shown to provide valid, consistent 
+    measures of children's aggregate early language skill. However, CDIs are predominantly 
+    comprised of hundreds of vocabulary items, some of which may not be heard (and thus learned) 
+    equally often by children of varying backgrounds. Here, we use a database of American English 
+    CDIs to identify words that show strong bias for particular groups of children, along the axes
+    of sex (male vs. female), race/ethnicity (white vs. non-white), and socioeconomic status 
+    (high vs. low). For each axis, we identify dozens of strongly biased items, and show that 
+    eliminating these items reduces the expected ability difference between groups. For sex, we
+    consider how to propose replacement words that may show less bias, on the basis of their 
+    relatively equal frequency in adult speech directed to male and female children.
 
 keywords: >
-    language development;
+    language development; early vocabulary; measuring instrument bias;
 
@@
 
-- introduce importance of measuring early language development (e.g., links to later educational outcomes)
-- introduce CDI
-- predictive validity: Fenson et al. 1994, Bornstein et al. 2012, Duff et al. 2015
+introduce importance of measuring early language development (e.g., links to later educational outcomes)
 
-- discuss potential for bias
-- what does between-group bias look like?
+introduce CDI
 
-- IRT - what is a good measure?
+predictive validity: Fenson et al. 1994, Bornstein et al. 2012, Duff et al. 2015
+
+discuss potential for bias
+
+what does between-group bias look like?
+
+IRT - what is a good measure?
 
@@
 
-- what is our goal for measuring vocabulary? - identifying language delays, predicting later reading, or talking...
-- how you select items influences how well you achieve these goals (and what bias you find)
+what is our goal for measuring vocabulary? - identifying language delays, predicting later reading, or talking...
+
+how you select items influences how well you achieve these goals (and what bias you find)
+
 We take the approach put forward by @stenhaug2021treading: GLIMMER plots based on the 1-parameter logistic are sufficient to identify bias, without potentially obscuring DIF in a more complex model's additional parameters.

### 2021-10-29 paper/cogsci.tex
wrote an abstract

@@
 
-\title{Investigating Item Bias}
+\title{Words aren't created equal: Investigating bias on the CDI}
 
@@
 \begin{abstract}
-Include no author information in the initial submission, to facilitate
-blind review. The abstract should be one paragraph, indented 1/8 inch on
-both sides, in 9\textasciitilde point font with single spacing. The
-heading `Abstract' should be 10\textasciitilde point, bold, centered,
-with one line of space below it. This one-paragraph abstract section is
-required only for standard six page proceedings papers. Following the
-abstract should be a blank line, followed by the header `Keywords' and a
-list of descriptive keywords separated by semicolons, all in
-9\textasciitilde point font, as shown below.
+Children's early language skill has been linked to later educational
+outcomes, making it important to accurately measure early language.
+Parent-reported instruments such as the Communicative Development
+Inventories (CDIs) have been shown to provide valid, consistent measures
+of children's aggregate early language skill. However, CDIs are
+predominantly comprised of hundreds of vocabulary items, some of which
+may not be heard (and thus learned) equally often by children of varying
+backgrounds. Here, we use a database of American English CDIs to
+identify words that show strong bias for particular groups of children,
+along the axes of sex (male vs.~female), race/ethnicity (white
+vs.~non-white), and socioeconomic status (high vs.~low). For each axis,
+we identify dozens of strongly biased items, and show that eliminating
+these items reduces the expected ability difference between groups. For
+sex, we consider how to propose replacement words that may show less
+bias, on the basis of their relatively equal frequency in adult speech
+directed to male and female children.
 
 \textbf{Keywords:}
-language development;
+language development; early vocabulary; measuring instrument bias;
 \end{abstract}
@@
 
-\begin{itemize}
-\item
 introduce importance of measuring early language development (e.g.,
 links to later educational outcomes)
-\item
+
 introduce CDI
-\item
-  predictive validity: Fenson et al.~1994, Bornstein et al.~2012, Duff
-  et al.~2015
-\item
+
+predictive validity: Fenson et al.~1994, Bornstein et al.~2012, Duff et
+al.~2015
+
 discuss potential for bias
-\item
+
 what does between-group bias look like?
-\item
+
 IRT - what is a good measure?
-\end{itemize}
 
@@
 
-\begin{itemize}
-\tightlist
-\item
 what is our goal for measuring vocabulary? - identifying language
 delays, predicting later reading, or talking\ldots{}
-\item
+
 how you select items influences how well you achieve these goals (and
-  what bias you find) We take the approach put forward by
-  (\textbf{Stenhaug2021?}): GLIMMER plots based on the 1-parameter
-  logistic are sufficient to identify bias, without potentially
-  obscuring DIF in a more complex model's additional parameters.
-\end{itemize}
+what bias you find)
+
+We take the approach put forward by Stenhaug, Frank, \& Domingue (2021):
+GLIMMER plots based on the 1-parameter logistic are sufficient to
+identify bias, without potentially obscuring DIF in a more complex
+model's additional parameters.
+
+Differential item functioning (DIF) is a technique in IRT used to
+identify items that show bias against demographic groups. DIF is a
+statistical characteristic of an item that shows the extent to which the
+item might operate differently or measure varying abilities for
+subgroups and members of separate demographic groups. DIF is, however, a
+challenging metric for identifying bias for a couple reasons. Firstly,
+DIF is extremely hard to accurately root out. The process of finding DIF
+involves using the test you used to do IRT analysis in order to look for
+bias within the very same test. Secondly, the presence of DIF does not
+necessarily indicate the presence of bias; it can indicate that one
+demographic has a higher average ability level than the other. These two
+issues are the basis upon which we get the Fundamental DIF
+Identification problem.
+
+This problem can be be understood through a concrete example that is
+well-known from research on early language learning: consider the fact
+that females show a larger vocabulary than males across early
+development {[}see Frank, Braginsky, Yurovsky, \& Marchman (2021) Ch.
+6?; OTHER REFS{]}. The question however is whether or not girls actually
+have a higher ability level or if the tests have words that are bias
+toward girls ie. are easier for girls, or perhaps the test is biased
+toward boys but the natural ability of girls is enough to overcome that
+bias. To answer this question we need to know the ability levels of
+girls and boys in order to confirm that if a word is learned earlier by
+girls it is simply because of an ability difference rather than a bias
+inherent to the word. The problem arises from the fact that we measure
+ability level using the very same test that we are trying to check for
+biased words. If the boys and girls were of the same ability level this
+wouldn't be an issue but given the evidence showing that girls have a
+higher ability level, we need to know the difference in ability so that
+when we find DIF for a specific word can be sure it is outside of the
+expecting DIF that results naturally from their difference in ability.
 
@@
 We analyze American English CDI: Words \& Sentences administrations from
-5520 children aged 16 to 30 from Wordbank (\textbf{Frank2016?};
-\textbf{Frank2021?}).
+5520 children aged 16 to 30 from Wordbank (Frank, Braginsky, Yurovsky,
+\& Marchman, 2017; Frank et al., 2021).
 
@@
 reported and collected by the Wordbank Project. Unlike other projects
-Wordback relies on the kindness of others to contribute their data which
+Wordbank relies on the kindness of others to contribute their data which
 means that often meta-data for some sets is missing. Even so, the
@@
 Wordbank data set collected demographic data consisting of Birth Order,
-Race/Ethnicity, Sex and Mothers Education. For our study we focused on
-the comparing data between demographic groups on the axis of Sex,
-Ethnicity and Mothers Education, which was a proxy for social economic
-status. Out data included 1989 female and 2105 male participants, and an
-additional 1426 participants whose sex remained unreported. In terms of
-Ethnicity data was recorded from 2202 white, 67 Asian, 222 Black, 131
-Hispanic, and 93 ``Other'' participants as well as 2805 unreported.
-Given our distribution of ethnicities, for our analysis we split
-participants into White (2202) and Non-White (513). Finally, we split
-participants into high and low SES, of which we had NA high SES
-participants, corresponding to mothers' education of some college or
-higher and NA Low SES participants.
+Race/Ethnicity, Sex and Mothers Education. We focused on comparing data
+between demographic groups on the axes of Sex, Ethnicity and Mother's
+Education, a proxy for socioeconomic status (henceforth, SES). Our data
+included 1989 female and 2105 male participants, and an additional 1426
+participants whose sex remained unreported. In terms of Ethnicity data
+was recorded from 2202 white, 67 Asian, 222 Black, 131 Hispanic, and 93
+``Other'' participants as well as 2805 unreported. Given our
+distribution of ethnicities, for our analysis we split participants into
+White (2202) and Non-White (513). Finally, we split participants into
+high and low SES, of which we had 4973 high SES participants,
+corresponding to mothers' education of some college or higher and 547
+Low SES participants.
+
+\hypertarget{one-column-images}{%
+\subsection{One-column images}\label{one-column-images}}
+
+\hypertarget{acknowledgements}{%
+\section{Acknowledgements}\label{acknowledgements}}
+
+Place acknowledgments (including funding information) in a section at
+the end of the paper.
+
+\hypertarget{references}{%
+\section{References}\label{references}}
+
+\setlength{\parindent}{-0.1in} 
+\setlength{\leftskip}{0.125in}
+
+\noindent
+
+\hypertarget{refs}{}
+\leavevmode\hypertarget{ref-Frank2017}{}%
+Frank, M. C., Braginsky, M., Yurovsky, D., \& Marchman, V. A. (2017).
+Wordbank: An open repository for developmental vocabulary data.
+\emph{Journal of Child Language}, \emph{44}(3), 677--694.
+
+\leavevmode\hypertarget{ref-Frank2021}{}%
+Frank, M. C., Braginsky, M., Yurovsky, D., \& Marchman, V. A. (2021).
+\emph{Variability and consistency in early language learning: The
+wordbank project}. MIT Press.
+
+\leavevmode\hypertarget{ref-stenhaug2021treading}{}%
+Stenhaug, B., Frank, M. C., \& Domingue, B. (2021). Treading carefully:
+Agnostic identification as the first step of detecting differential item
+functioning.
 

### 2022-01-05 GLIMMER_production.R
small glimmer plots for paper

@@ -88,21 +88,22 @@ draws_df_to_logit_plot <- function(draws_df){
 }
 
 # CREATE GLIMMER PLOT ----------------------------------------------------------
-plot_glimmer <- function(mod_intuitive, item_names, plotName='') {
+plot_glimmer <- function(mod_intuitive, item_names, items_to_plot, plotName='', width=8, height=80) {
   draws_df <- mod_intuitive %>% mod_intuitive_to_draws_df()
   names(draws_df) = c("run", item_names)
   
   
-  p <- draws_df %>%
+  p <- draws_df %>% 
     clean_names() %>%
     gather(var, val, -run) %>%
+    filter(is.element(var, items_to_plot)) %>%
     mutate(var = fct_reorder(var, val)) %>%
     ggplot(aes(x = val, y = var)) +
     ggridges::geom_density_ridges() +
     geom_vline(aes(xintercept=0), linetype='dashed', alpha=.5) +
     labs(x = "", y = "") + theme_classic()
   if(plotName!='') ggsave(paste0(plotName,'.pdf'), 
-                          width=8, height=80, # .2 * length(item_names)
+                          width=width, height=height, # .2 * length(item_names)
                           limitsize = F) 
   return(p)
 }
@@ -121,10 +122,24 @@ run_once <- function() {
 load("data/glimmer_prodWS_models.Rds")
 
 # GLIMMER TIME
-plot_glimmer(mod_intuitive_sex, colnames(d_mat), plotName="GLIMMER_sex_prodWS")
-plot_glimmer(mod_intuitive_ses, colnames(d_mat), plotName="GLIMMER_ses_prodWS")
-plot_glimmer(mod_intuitive_eth, colnames(d_mat), plotName="GLIMMER_eth_prodWS")
+plot_glimmer(mod_intuitive_sex, colnames(d_mat), colnames(d_mat), plotName="GLIMMER_sex_prodWS")
+plot_glimmer(mod_intuitive_ses, colnames(d_mat), colnames(d_mat), plotName="GLIMMER_ses_prodWS")
+plot_glimmer(mod_intuitive_eth, colnames(d_mat), colnames(d_mat), plotName="GLIMMER_eth_prodWS")
 
+# small glimmer plots for paper
+plot_glimmer(mod_intuitive_sex, colnames(d_mat), 
+             items_to_plot = c("vagina","tights","dress_object","doll","necklace","pretty","underpants","purse","baby",
[…]

### 2022-01-05 GLIMMER_production.R
small SES GLIMMER plot

@@ -134,7 +134,9 @@ plot_glimmer(mod_intuitive_sex, colnames(d_mat),
              plotName="smGLIMMER_sex_prodWS", height=4.5, width=4)
 
 plot_glimmer(mod_intuitive_ses, colnames(d_mat), 
-             items_to_plot = c("gum","walker","so"), 
+             items_to_plot = c("gum","walker","so","hate","soda_pop","each","candy","why","can_object",
+                               "game","egg","pizza","touch","crayon","rain","dirty",
+                               "daddy","cockadoodledoo","moo","duck","woof_woof","uh_oh","vroom","quack_quack","grrr"), 
              plotName="smGLIMMER_ses_prodWS", height=4.5, width=4)
 
 plot_glimmer(mod_intuitive_eth, colnames(d_mat), 

### 2022-01-05 GLIMMER_production.R
small ethnicity GLIMMER plot

@@ -129,18 +129,20 @@ plot_glimmer(mod_intuitive_eth, colnames(d_mat), colnames(d_mat), plotName="GLIM
 # small glimmer plots for paper
 plot_glimmer(mod_intuitive_sex, colnames(d_mat), 
              items_to_plot = c("vagina","tights","dress_object","doll","necklace","pretty","underpants","purse","baby",
-                 "bear","hug","some","over","cookie","tiny","why",
-                 "truck","police","dump","firetruck","bat","hammer","tractor","vroom","penis"), 
+                 "bear","hug","some","over","cookie","tiny","why", # middling bias
+                 "truck","police","dump","firetruck","bat","hammer","tractor","vroom","penis"), # male bias
              plotName="smGLIMMER_sex_prodWS", height=4.5, width=4)
 
 plot_glimmer(mod_intuitive_ses, colnames(d_mat), 
-             items_to_plot = c("gum","walker","so","hate","soda_pop","each","candy","why","can_object",
-                               "game","egg","pizza","touch","crayon","rain","dirty",
+             items_to_plot = c("gum","walker","so","hate","soda_pop","each","candy","why","can_object", # low-SES bias
+                               "game","egg","pizza","touch","crayon","rain","dirty", # middling bias
                                "daddy","cockadoodledoo","moo","duck","woof_woof","uh_oh","vroom","quack_quack","grrr"), 
              plotName="smGLIMMER_ses_prodWS", height=4.5, width=4)
 
 plot_glimmer(mod_intuitive_eth, colnames(d_mat), 
-             items_to_plot = c(), 
+             items_to_plot = c("sofa","so","trash","gum","walker","wish","give_me_five","finish","if", # non-white bias
+                               "hello","quiet","church","leg","dish","story","bicycle",
+                               "moo","bye","uh_oh","all_gone","quack_quack","pets_name","grrr","mommy","daddy"), 
              plotName="smGLIMMER_eth_prodWS", height=4.5, width=4)
 
 # histograms of item difficulty differences

### 2022-01-05 paper/cogsci.Rmd
adding GLIMMER plots, editing

@@
 6?; OTHER REFS].
-The question however is whether or not girls actually have a higher ability level or if the tests have words that are bias toward girls ie.
-are easier for girls, or perhaps the test is biased toward boys but the natural ability of girls is enough to overcome that bias.
-To answer this question we need to know the ability levels of girls and boys in order to confirm that if a word is learned earlier by girls it is simply because of an ability difference rather than a bias inherent to the word.
+The question is whether or not females actually have a higher ability level, or if the tests predominantly have words that are easier for females (i.e., biased).
+(It could even be that the test is biased toward males but the language ability of females is large enough to overcome that bias.)
+To address this question, we need to know the ability levels of girls and boys in order to confirm that if a word is learned earlier by girls it is simply because of an ability difference rather than a bias inherent to the word.
 The problem arises from the fact that we measure ability level using the very same test that we are trying to check for biased words.
-If the boys and girls were of the same ability level this wouldn’t be an issue but given the evidence showing that girls have a higher ability level, we need to know the difference in ability so that when we find DIF for a specific word can be sure it is outside of the expecting DIF that results naturally from their difference in ability.
+If the boys and girls were of the same average language ability, this would not be an issue but given the evidence showing that girls have a higher ability level, we need to know the difference in ability so that when we find DIF for a specific word can be sure it is outside of the expecting DIF that results naturally from their difference in ability.
 
@@
 
-## One-column images
+# Results
 
-```{r image, fig.env = "figure", fig.pos = "H", fig.align='center', fig.width=2, fig.height=2, set.cap.width=T, num.cols.cap=1, fig.cap = "One column image."}
+```{r sex-glimmer, out.width="\\linewidth", fig.env = "figure", fig.pos = "H", fig.align='center', fig.width=3.2, fig.height=3.5, set.cap.width=T, num.cols.cap=1, fig.cap = "GLIMMER plot of a sample of CDI:WS words from the sex bias model.
+Words at the top are easier to learn for females, while those at the bottom are easier for males."}
 #img <- png::readPNG("figs/lab_logo_stanford.png")
 #grid::grid.raster(img)
+knitr::include_graphics("figs/smGLIMMER_sex_prodWS.pdf")
 ```
+
+```{r ses-glimmer, out.width="\\linewidth", fig.env = "figure", fig.pos = "H", fig.align='center', fig.width=3, fig.height=3.5, set.cap.width=T, num.cols.cap=1, fig.cap = "GLIMMER plot of a sample of CDI:WS words from the SES bias model.
+Words at the top are easier to learn for children from low-SES families, while those at the bottom are easier for those from high-SES families."}
+knitr::include_graphics("figs/smGLIMMER_ses_prodWS.pdf")
+```
+
+```{r eth-glimmer, out.width="\\linewidth", fig.env = "figure", fig.pos = "H", fig.align='center', fig.width=3, fig.height=3.5, set.cap.width=T, num.cols.cap=1, fig.cap = "GLIMMER plot of a sample of CDI:WS words from the ethnicity bias model.
+Words at the top are easier to learn for children from low-SES families, while those at the bottom are easier for those from high-SES families."}
+knitr::include_graphics("figs/smGLIMMER_eth_prodWS.pdf")
+```
+
+# Discussion
 

### 2022-01-05 paper/cogsci.Rmd
some writing

@@
 keywords: >
-    language development; early vocabulary; measuring instrument bias;
+    language acquisition; word learning; measuring instrument bias; development;
 
@@
 
+Researchers, clinicians, and parents have long been fascinated with measuring and theorizing about the growth of young children's vocabulary.
+Children's early vocabulary growth is assumed to reflect not only their exposure to child-directed speech, but also the varying difficulty of different types of words, and the aptitude of the child -- including potential language deficits.
+For example, @bornstein2016stability found stability in language skills
+
+Children's demographics have also been found to be predictive of their language skills.
+
 introduce importance of measuring early language development (e.g., links to later educational outcomes)
 
-introduce CDI
-
-predictive validity: Fenson et al. 1994, Bornstein et al. 2012, Duff et al. 2015
+The MacArthur-Bates Communicative Development Inventories [CDIs; @Fenson2007] are a set of parent-reported measures of children's productive and receptive language skills, which produce a low-cost and reliable way to estimate children's early language skills [@fenson1994].
+CDIs have shown good predictive validity [e.g., @fenson1994; Bornstein et al. 2012, Duff et al. 2015].
 
@@
 
-This problem can be be understood through a concrete example that is well-known from research on early language learning: consider the fact that females show a larger vocabulary than males across early development [see @Frank2021 Ch.
+This problem can be be understood through a concrete example that is well-known from research on early language learning: consider the fact that females show a larger vocabulary than males across early development [see @frank2021 Ch.
 6?; OTHER REFS].
@@
 
+The outline of this paper is as follows.
+First, we will introduce the Wordbank data and the Rasch model we use to analyze the data.
+
 # Methods
 
-## Dataset
+## Vocabulary Data
 
@@
 ```
-We analyze American English CDI: Words & Sentences administrations from `r nrow(d_demo)` children aged `r min(d_demo$age)` to `r max(d_demo$age)` from Wordbank [@Frank2017; @Frank2021].
 
@@
 
-Research was conducted over `r nrow(d_demo)` participants, whose data was parent reported and collected by the Wordbank Project. 
+We analyze parent-reported Wordbank data from `r nrow(d_demo)` American English CDI: Words & Sentences administrations for children `r min(d_demo$age)` to `r max(d_demo$age)` months of age [@frank2017; @frank2021].
 Unlike other projects Wordbank relies on the kindness of others to contribute their data which means that often meta-data for some sets is missing. 
-Even so, the project now has data over 29 different languages on children from ages 8-30 months old, though our study focused entirely on American English and Children 16-30 months. 
 When they received complete information, the Wordbank data set collected demographic data consisting of Birth Order, Race/Ethnicity, Sex and Mothers Education. 
@@
 
-<!-- goal: make custom glimmer plot in R with high / low bias words -->
+## Rasch Model
+
+The Rasch model, also known as the 1-parameter logistic (1PL) model, is the simplest Item Response Theory model, and is thus the easiest to use to investigate potential differences in item function across different groups of participants.
+The goal of the Rasch model is to jointly estimate for each child $j$ a latent ability $\theta_j$, and for each item $i$ a difficulty parameter $b_i$.
+In the model, the probability of child $j$ knowing (i.e., producing or understanding) a given item $i$ is
+
+$$P_{i}(x_i = 1 | b_{i},\theta_j ) = \frac{1}{1 + e^{-D(\theta_j - b_i )}}$$
+
+where $D$ is a constant scaling parameter ($D=1.702$) which makes the logistic closely match the ogive function in traditional factor analysis [@R-mirt; @reckase2009].
+Child ability ($\theta$) and item difficulty ($b$) distributions are standardized (i.e., mean of 0), and expected to be normally-distributed. 
+Children with high latent ability ($\theta$) will be more likely to produce any given item than children with lower latent ability, and more difficult items will be produced by fewer children (at any given $\theta$) than easier items.
+
+For each demographic dimension (sex, socioeconomic status, and ethnicity) we fit a separate Rasch model for each demographic group.
+
+# Results
+
+<!-- put the GLIMMERs side-by-side?
 ```{r 2-col-image, fig.env = "figure*", fig.pos = "h", fig.width=4, fig.height=2, fig.align = "center", set.cap.width=T, num.cols.cap=2, fig.cap = "This image spans both columns.
 And the caption text is limited to 0.8 of the width of the document."}
-#img <- png::readPNG("figs/walrus.png")
-#grid::grid.raster(img)
+
 ```
-
-# Results
+-->
 

### 2022-01-10 paper/cogsci.Rmd
intro writing

@@
 
-Researchers, clinicians, and parents have long been fascinated with measuring and theorizing about the growth of young children's vocabulary.
-Children's early vocabulary growth is assumed to reflect not only their exposure to child-directed speech, but also the varying difficulty of different types of words, and the aptitude of the child -- including potential language deficits.
-For example, @bornstein2016stability found stability in language skills
-
-Children's demographics have also been found to be predictive of their language skills.
-
-introduce importance of measuring early language development (e.g., links to later educational outcomes)
+Researchers, clinicians, and parents have long been fascinated with the surprising speed and variability in the growth of young children's vocabulary.
+Children's early vocabulary growth is assumed to reflect not only their exposure to child-directed speech, but also the varying difficulty of different types of words, and individual differences in the aptitude of the child -- including potential language deficits.
+Research has uncovered both within-child consistency in development, as well as significant influence from external factors.
+For example, @bornstein2016stability found stability in core language skills across 10 years of children's development, despite changes in maternal income and education over the study period.
+Yet socioeconomic status (SES) has also been found to be predictive of children's early language skill, and of later educational outcomes [for a review, see @schwab2016].
+Biological factors are also predictive of language skill: first-born children tend to outpace their siblings, and female children tend to have better language skills than their age-matched male counterparts [@fenson1994].
+However, it is also difficult to measure language skill: in any short recording children are unlikely to use all of the words and constructions that they know, and any comprehensive battery of language measures will likely exhaust children's attention span.
 
@@
 CDIs have shown good predictive validity [e.g., @fenson1994; Bornstein et al. 2012, Duff et al. 2015].
+Due to these good properties, CDIs have been adapted to dozens of languages, and a central repository of CDIs with datasets contributed from all over the world has been created [Wordbank; @frank2017; @frank2021].
+Our focus will be on the vocabulary checklist portion of the CDI Words & Sentences (CDI:WS) form, comprised of 680 items across 22 categories (e.g., animals, vehicles, action words, pronouns) designed to assess the productive vocabulary of children 16 to 30 months of age.
+For each item on the CDI:WS, caregivers are asked to respond whether the target child has been heard to say (i.e. produce) the given item.
+Children's total vocabulary score on the CDI:WS is tightly correlated with other facets of early language (e.g., grammatical competence and gesture), suggesting that the language system is "tightly woven" [@frank2021].
 
-discuss potential for bias
+Together, these desirable properties of CDI:WS vocabulary items have led researchers to
 

### 2022-01-11 paper/cogsci.Rmd
intro writing

@@
 Yet socioeconomic status (SES) has also been found to be predictive of children's early language skill, and of later educational outcomes [for a review, see @schwab2016].
-Biological factors are also predictive of language skill: first-born children tend to outpace their siblings, and female children tend to have better language skills than their age-matched male counterparts [@fenson1994].
+Biological factors are also predictive of language skill: first-born children tend to outpace their siblings, and female children tend to have better language skills than their age-matched male counterparts [@fenson1994] -- an sex-based verbal advantage that continues through high school [see @petersen2018gender for a review].
 However, it is also difficult to measure language skill: in any short recording children are unlikely to use all of the words and constructions that they know, and any comprehensive battery of language measures will likely exhaust children's attention span.
@@
 CDIs have shown good predictive validity [e.g., @fenson1994; Bornstein et al. 2012, Duff et al. 2015].
-Due to these good properties, CDIs have been adapted to dozens of languages, and a central repository of CDIs with datasets contributed from all over the world has been created [Wordbank; @frank2017; @frank2021].
-Our focus will be on the vocabulary checklist portion of the CDI Words & Sentences (CDI:WS) form, comprised of 680 items across 22 categories (e.g., animals, vehicles, action words, pronouns) designed to assess the productive vocabulary of children 16 to 30 months of age.
-For each item on the CDI:WS, caregivers are asked to respond whether the target child has been heard to say (i.e. produce) the given item.
+Our focus will be on the vocabulary checklist portion of the CDI Words & Sentences (CDI:WS) form, comprised of 680 early-learned words across 22 categories (e.g., animals, vehicles, action words, pronouns) selected to assess the productive vocabulary of children 16 to 30 months of age.
+For each item on the CDI:WS, caregivers are asked to respond whether the target child has been heard to say (i.e. produce) the given item.[^1]
 Children's total vocabulary score on the CDI:WS is tightly correlated with other facets of early language (e.g., grammatical competence and gesture), suggesting that the language system is "tightly woven" [@frank2021].
+Due to these desirable properties, CDIs have been adapted to dozens of languages, and a central repository of CDI data contributed from all over the world has been created [Wordbank; @frank2017; @frank2021].
 
-Together, these desirable properties of CDI:WS vocabulary items have led researchers to
+[^1]: The CDI: Words and Gestures form includes a subset of the CDI:WS items and targets children 12 to 18 months of age, measuring both comprehension and production.
 
-what does between-group bias look like?
+Inspired by the utility and widespread use of the CDI, researchers have recently been using psychometric models on CDI data to construct short, adaptive tests to reliably assess language ability using only a small subset of the CDI items [e.g., @mayor2019, @kachergis2021cat].
+These psychometric models typically come from the Item-Response Theory (IRT) framework [@Baker2001], which assumes that not only test-takers (here, children) have normally-distributed ability, but that items (words on the CDI) have normally-distributed difficulty.
+The very efficacy of these IRT-based models depends on words varying in difficulty, and hence being more/less informative of the ability level of different individuals.
+For example, asking whether a 22-month-old produces the word "ball" is far less informative of that child's language ability than asking whether they produce "table", as 96% of 22-month-olds can produce the former, while only 47% produce the latter.
 
-IRT - what is a good measure?
+While it is quite reasonable to expect that some CDI words are easier than others, and even to use these varying difficulties to predict variation in children's language ability, the use of psychometric models highlights the possibility that some CDI items may function differently (i.e., be more/less difficult) for different groups of children.
+The idea that some items on a test may show bias, favoring one group over another, is known as Differential Item Function (DIF; @holland1993differential]. 
+A variety of
 

### 2022-01-11 paper/cogsci.Rmd
writing

@@
 The idea that some items on a test may show bias, favoring one group over another, is known as Differential Item Function (DIF; @holland1993differential]. 
-A variety of
-
-## Fundamental DIF Problem
-
-what is our goal for measuring vocabulary? - identifying language delays, predicting later reading, or talking...
-
-how you select items influences how well you achieve these goals (and what bias you find)
+A variety of statistical methods for detecting DIF have been proposed, and investigations have in several instances identified DIF for many items on tests favoring particular one group over another (e.g., rural vs.
+urban).
+Our goal here is to test the items on the American CDI:WS for DIF along three main axes: sex (male vs. female), socioeconomic status (low- vs. high-SES), and ethnicity (white vs. non-white).
 
@@
 
-Differential item functioning (DIF) is a technique in IRT used to identify items that show bias against demographic groups. 
-DIF is a statistical characteristic of an item that shows the extent to which the item might operate differently or measure varying abilities for subgroups and members of separate demographic groups. 
 DIF is, however, a challenging metric for identifying bias for a couple reasons. 
@@
 The outline of this paper is as follows.
-First, we will introduce the Wordbank data and the Rasch model we use to analyze the data.
+First, we will introduce the Wordbank data and the IRT model we use to analyze the data.
 
@@
 We analyze parent-reported Wordbank data from `r nrow(d_demo)` American English CDI: Words & Sentences administrations for children `r min(d_demo$age)` to `r max(d_demo$age)` months of age [@frank2017; @frank2021].
-Unlike other projects Wordbank relies on the kindness of others to contribute their data which means that often meta-data for some sets is missing. 
-When they received complete information, the Wordbank data set collected demographic data consisting of Birth Order, Race/Ethnicity, Sex and Mothers Education. 
+Full demographic data are not reported in some datasets contributed to Wordbank: sex is available for `r nrow(d_demo)-sum(is.na(d_demo$sex))` children, `r nrow(d_demo)-sum(is.na(d_demo$sex))`.
+
 We focused on comparing data between demographic groups on the axes of Sex, Ethnicity and Mother's Education, a proxy for socioeconomic status (henceforth, SES). 
-Our data included `r sex_tab["Female"]` female and `r sex_tab["Male"]` male participants, and an additional `r sum(is.na(d_demo$sex))` participants whose sex remained unreported. 
+Our data included `r sex_tab["Female"]` female and `r sex_tab["Male"]` male participants.
 In terms of Ethnicity data was recorded from `r eth_tab["White"]` white, `r eth_tab["Asian"]` Asian, `r eth_tab["Black"]` Black, `r eth_tab["Hispanic"]` Hispanic, and `r eth_tab["Other"]` “Other” participants as well as `r sum(is.na(d_demo$ethnicity))` unreported.

### 2022-01-11 paper/cogsci.Rmd
writing

@@
 
-For each demographic dimension (sex, socioeconomic status, and ethnicity) we fit a separate Rasch model for each demographic group.
+In the multigroup Rasch model, an item's difficulty is allowed to vary depending on which group an individual child belongs to. 
+For example, in the sex-based multigroup model, item $i$'s difficulty is $b_{i}^{female}$ for females, and $b_{i}^{male}$ for males.
+We fitted a multigroup Rasch model for each demographic dimension (sex, socioeconomic status, and ethnicity).
 

### 2022-01-12 paper/cogsci.tex
title abstract

@@
 
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newenvironment{CSLReferences}%
+  {}%
+  {\par}
+
 \begin{document}
@@
 \textbf{Keywords:}
-language development; early vocabulary; measuring instrument bias;
+language acquisition; word learning; measuring instrument bias;
+development;
 \end{abstract}
@@
 
-introduce importance of measuring early language development (e.g.,
-links to later educational outcomes)
+Researchers, clinicians, and parents have long been fascinated with the
+surprising speed and variability in the growth of young children's
+vocabulary. Children's early vocabulary growth is assumed to reflect not
+only their exposure to child-directed speech, but also the varying
+difficulty of different types of words, and individual differences in
+the aptitude of the child -- including potential language deficits.
+Research has uncovered both within-child consistency in development, as
+well as significant influence from external factors. For example,
+Bornstein, Hahn, \& Putnick (2016) found stability in core language
+skills across 10 years of children's development, despite changes in
+maternal income and education over the study period. Yet socioeconomic
+status (SES) has also been found to be predictive of children's early
+language skill, and of later educational outcomes (for a review, see
+Schwab \& Lew-Williams, 2016). Biological factors are also predictive of
+language skill: first-born children tend to outpace their siblings, and
+female children tend to have better language skills than their
+age-matched male counterparts (Larry Fenson et al., 1994) -- an
+sex-based verbal advantage that continues through high school (see
+Petersen, 2018 for a review). However, it is also difficult to measure
+language skill: in any short recording children are unlikely to use all
+of the words and constructions that they know, and any comprehensive
+battery of language measures will likely exhaust children's attention
+span.
 
-introduce CDI
+The MacArthur-Bates Communicative Development Inventories {[}CDIs; L.
+Fenson et al. (2007){]} are a set of parent-reported measures of
+children's productive and receptive language skills, which produce a
+low-cost and reliable way to estimate children's early language skills
+(Larry Fenson et al., 1994). CDIs have shown good predictive validity
+{[}e.g., Larry Fenson et al. (1994); Bornstein et al.~2012, Duff et
+al.~2015{]}. Our focus will be on the vocabulary checklist portion of
+the CDI Words \& Sentences (CDI:WS) form, comprised of 680 early-learned
+words across 22 categories (e.g., animals, vehicles, action words,
+pronouns) selected to assess the productive vocabulary of children 16 to
+30 months of age. For each item on the CDI:WS, caregivers are asked to
+respond whether the target child has been heard to say (i.e.~produce)
+the given item.\footnote{The CDI: Words and Gestures form includes a
+  subset of the CDI:WS items and targets children 12 to 18 months of
+  age, measuring both comprehension and production.} Children's total
+vocabulary score on the CDI:WS is tightly correlated with other facets
+of early language (e.g., grammatical competence and gesture), suggesting
+that the language system is ``tightly woven'' (Frank, Braginsky,
+Yurovsky, \& Marchman, 2021). Due to these desirable properties, CDIs
+have been adapted to dozens of languages, and a central repository of
+CDI data contributed from all over the world has been created
+{[}Wordbank; Frank, Braginsky, Yurovsky, \& Marchman (2017); Frank,
+Braginsky, Yurovsky, \& Marchman (2021){]}.
 
-predictive validity: Fenson et al.~1994, Bornstein et al.~2012, Duff et
-al.~2015
+Inspired by the utility and widespread use of the CDI, researchers have
+recently been using psychometric models on CDI data to construct short,
+adaptive tests to reliably assess language ability using only a small
+subset of the CDI items Kachergis, Marchman, Dale, Mankewitz, \& Frank
+(2021). These psychometric models typically come from the Item-Response
+Theory (IRT) framework (Baker, 2001), which assumes that not only
+test-takers (here, children) have normally-distributed ability, but that
+items (words on the CDI) have normally-distributed difficulty. The very
+efficacy of these IRT-based models depends on words varying in
+difficulty, and hence being more/less informative of the ability level
+of different individuals. For example, asking whether a 22-month-old
+produces the word ``ball'' is far less informative of that child's
+language ability than asking whether they produce ``table,'' as 96\% of
+22-month-olds can produce the former, while only 47\% produce the
+latter.
 
-discuss potential for bias
-
-what does between-group bias look like?
-
-IRT - what is a good measure?
-
-\hypertarget{fundamental-dif-problem}{%
-\subsection{Fundamental DIF Problem}\label{fundamental-dif-problem}}
-
-what is our goal for measuring vocabulary? - identifying language
-delays, predicting later reading, or talking\ldots{}
-
-how you select items influences how well you achieve these goals (and
-what bias you find)
+While it is quite reasonable to expect that some CDI words are easier
+than others, and even to use these varying difficulties to predict
+variation in children's language ability, the use of psychometric models
+highlights the possibility that some CDI items may function differently
+(i.e., be more/less difficult) for different groups of children. The
+idea that some items on a test may show bias, favoring one group over
+another, is known as Differential Item Function (DIF; Holland \& Wainer
+(1993){]}. A variety of statistical methods for detecting DIF have been
+proposed, and investigations have in several instances identified DIF
+for many items on tests favoring particular one group over another
+(e.g., rural vs.~urban). Our goal here is to test the items on the
+American CDI:WS for DIF along three main axes: sex (male vs.~female),
+socioeconomic status (low- vs.~high-SES), and ethnicity (white
+vs.~non-white).
 
@@
 
-Differential item functioning (DIF) is a technique in IRT used to
-identify items that show bias against demographic groups. DIF is a
-statistical characteristic of an item that shows the extent to which the
-item might operate differently or measure varying abilities for
-subgroups and members of separate demographic groups. DIF is, however, a
-challenging metric for identifying bias for a couple reasons. Firstly,
-DIF is extremely hard to accurately root out. The process of finding DIF
-involves using the test you used to do IRT analysis in order to look for
-bias within the very same test. Secondly, the presence of DIF does not
-necessarily indicate the presence of bias; it can indicate that one
-demographic has a higher average ability level than the other. These two
-issues are the basis upon which we get the Fundamental DIF
-Identification problem.
+DIF is, however, a challenging metric for identifying bias for a couple
+reasons. Firstly, DIF is extremely hard to accurately root out. The
+process of finding DIF involves using the test you used to do IRT
+analysis in order to look for bias within the very same test. Secondly,
+the presence of DIF does not necessarily indicate the presence of bias;
+it can indicate that one demographic has a higher average ability level
+than the other. These two issues are the basis upon which we get the
+Fundamental DIF Identification problem.
 
@@
 development {[}see Frank, Braginsky, Yurovsky, \& Marchman (2021) Ch.
-6?; OTHER REFS{]}. The question however is whether or not girls actually
-have a higher ability level or if the tests have words that are bias
-toward girls ie. are easier for girls, or perhaps the test is biased
-toward boys but the natural ability of girls is enough to overcome that
-bias. To answer this question we need to know the ability levels of
-girls and boys in order to confirm that if a word is learned earlier by
-girls it is simply because of an ability difference rather than a bias
-inherent to the word. The problem arises from the fact that we measure
-ability level using the very same test that we are trying to check for
-biased words. If the boys and girls were of the same ability level this
-wouldn't be an issue but given the evidence showing that girls have a
-higher ability level, we need to know the difference in ability so that
-when we find DIF for a specific word can be sure it is outside of the
-expecting DIF that results naturally from their difference in ability.
+6?; OTHER REFS{]}. The question is whether or not females actually have
+a higher ability level, or if the tests predominantly have words that
+are easier for females (i.e., biased). (It could even be that the test
+is biased toward males but the language ability of females is large
+enough to overcome that bias.) To address this question, we need to know
+the ability levels of girls and boys in order to confirm that if a word
+is learned earlier by girls it is simply because of an ability
+difference rather than a bias inherent to the word. The problem arises
+from the fact that we measure ability level using the very same test
+that we are trying to check for biased words. If the boys and girls were
+of the same average language ability, this would not be an issue but
+given the evidence showing that girls have a higher ability level, we
+need to know the difference in ability so that when we find DIF for a
+specific word can be sure it is outside of the expecting DIF that
+results naturally from their difference in ability.
+
+The outline of this paper is as follows. First, we will introduce the
+Wordbank data and the IRT model we use to analyze the data.
 
@@
 
-\hypertarget{dataset}{%
-\subsection{Dataset}\label{dataset}}
-
-We analyze American English CDI: Words \& Sentences administrations from
-5520 children aged 16 to 30 from Wordbank (Frank, Braginsky, Yurovsky,
-\& Marchman, 2017; Frank et al., 2021).
+\hypertarget{vocabulary-data}{%
+\subsection{Vocabulary Data}\label{vocabulary-data}}
 
@@
 
-Research was conducted over 5520 participants, whose data was parent
-reported and collected by the Wordbank Project. Unlike other projects
-Wordbank relies on the kindness of others to contribute their data which
-means that often meta-data for some sets is missing. Even so, the
-project now has data over 29 different languages on children from ages
-8-30 months old, though our study focused entirely on American English
-and Children 16-30 months. When they received complete information, the
-Wordbank data set collected demographic data consisting of Birth Order,
-Race/Ethnicity, Sex and Mothers Education. We focused on comparing data
-between demographic groups on the axes of Sex, Ethnicity and Mother's
-Education, a proxy for socioeconomic status (henceforth, SES). Our data
-included 1989 female and 2105 male participants, and an additional 1426
-participants whose sex remained unreported. In terms of Ethnicity data
-was recorded from 2202 white, 67 Asian, 222 Black, 131 Hispanic, and 93
-``Other'' participants as well as 2805 unreported. Given our
-distribution of ethnicities, for our analysis we split participants into
-White (2202) and Non-White (513). Finally, we split participants into
-high and low SES, of which we had 4973 high SES participants,
-corresponding to mothers' education of some college or higher and 547
-Low SES participants.
+We analyze parent-reported Wordbank data from 5520 American English CDI:
+Words \& Sentences administrations for children 16 to 30 months of age
+(Frank, Braginsky, Yurovsky, \& Marchman, 2017, 2021). Full demographic
+data are not reported in some datasets contributed to Wordbank: sex is
+available for 4094 children, 4094.
 
-\hypertarget{one-column-images}{%
-\subsection{One-column images}\label{one-column-images}}
+We focused on comparing data between demographic groups on the axes of
+Sex, Ethnicity and Mother's Education, a proxy for socioeconomic status
+(henceforth, SES). Our data included 1989 female and 2105 male
+participants. In terms of Ethnicity data was recorded from 2202 white,
+67 Asian, 222 Black, 131 Hispanic, and 93 ``Other'' participants as well
+as 2805 unreported. Given our distribution of ethnicities, for our
+analysis we split participants into White (2202) and Non-White (513).
+Finally, we split participants into high and low SES, of which we had
+4973 high SES participants, corresponding to mothers' education of some
+college or higher and 547 Low SES participants.
+
+\hypertarget{rasch-model}{%
+\subsection{Rasch Model}\label{rasch-model}}
+
+The Rasch model, also known as the 1-parameter logistic (1PL) model, is
+the simplest Item Response Theory model, and is thus the easiest to use
+to investigate potential differences in item function across different
+groups of participants. The goal of the Rasch model is to jointly
+estimate for each child \(j\) a latent ability \(\theta_j\), and for
+each item \(i\) a difficulty parameter \(b_i\). In the model, the
+probability of child \(j\) knowing (i.e., producing or understanding) a
+given item \(i\) is
+
+\[P_{i}(x_i = 1 | b_{i},\theta_j ) = \frac{1}{1 + e^{-D(\theta_j - b_i )}}\]
+
+where \(D\) is a constant scaling parameter (\(D=1.702\)) which makes
+the logistic closely match the ogive function in traditional factor
+analysis (Chalmers, 2012; Reckase, 2009). Child ability (\(\theta\)) and
+item difficulty (\(b\)) distributions are standardized (i.e., mean of
+0), and expected to be normally-distributed. Children with high latent
+ability (\(\theta\)) will be more likely to produce any given item than
+children with lower latent ability, and more difficult items will be
+produced by fewer children (at any given \(\theta\)) than easier items.
+
+In the multigroup Rasch model, an item's difficulty is allowed to vary
+depending on which group an individual child belongs to. For example, in
+the sex-based multigroup model, item \(i\)'s difficulty is
+\(b_{i}^{female}\) for females, and \(b_{i}^{male}\) for males. We
+fitted a multigroup Rasch model for each demographic dimension (sex,
+socioeconomic status, and ethnicity).
+
+\hypertarget{results}{%
+\section{Results}\label{results}}
+
+\begin{CodeChunk}
+\begin{figure}[H]
+
+{\centering \includegraphics[width=\linewidth]{figs/smGLIMMER_sex_prodWS}
+
+}
+
+\caption[GLIMMER plot of a sample of CDI:WS words from the sex bias model]{GLIMMER plot of a sample of CDI:WS words from the sex bias model.
+Words at the top are easier to learn for females, while those at the bottom are easier for males.}\label{fig:sex-glimmer}
+\end{figure}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{figure}[H]
+
+{\centering \includegraphics[width=\linewidth]{figs/smGLIMMER_ses_prodWS}
+
+}
+
+\caption[GLIMMER plot of a sample of CDI:WS words from the SES bias model]{GLIMMER plot of a sample of CDI:WS words from the SES bias model.
+Words at the top are easier to learn for children from low-SES families, while those at the bottom are easier for those from high-SES families.}\label{fig:ses-glimmer}
+\end{figure}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{figure}[H]
+
+{\centering \includegraphics[width=\linewidth]{figs/smGLIMMER_eth_prodWS}
+
+}
+
+\caption[GLIMMER plot of a sample of CDI:WS words from the ethnicity bias model]{GLIMMER plot of a sample of CDI:WS words from the ethnicity bias model.
+Words at the top are easier to learn for children from low-SES families, while those at the bottom are easier for those from high-SES families.}\label{fig:eth-glimmer}
+\end{figure}
+\end{CodeChunk}
+
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
 
@@
 \hypertarget{refs}{}
-\leavevmode\hypertarget{ref-Frank2017}{}%
+\begin{CSLReferences}{1}{0}
+\leavevmode\hypertarget{ref-Baker2001}{}%
+Baker, F. B. (2001). \emph{The basics of item response theory}. ERIC.
+
+\leavevmode\hypertarget{ref-bornstein2016stability}{}%
+Bornstein, M. H., Hahn, C.-S., \& Putnick, D. L. (2016). Stability of
+core language skill across the first decade of life in children at
+biological and social risk. \emph{Journal of Child Psychology and
+Psychiatry}, \emph{57}(12), 1434--1443.
+
+\leavevmode\hypertarget{ref-R-mirt}{}%
+Chalmers, R. P. (2012). {mirt}: A multidimensional item response theory
+package for the {R} environment. \emph{Journal of Statistical Software},
+\emph{48}(6), 1--29.
+http://doi.org/\href{https://doi.org/10.18637/jss.v048.i06}{10.18637/jss.v048.i06}
+
+\leavevmode\hypertarget{ref-fenson1994}{}%
+Fenson, Larry, Dale, P. S., Reznick, J. S., Bates, E., Thal, D. J.,
+Pethick, S. J., \ldots{} Stiles, J. (1994). Variability in early
+communicative development. \emph{Monographs of the Society for Research
+in Child Development}, i--185.
+
+\leavevmode\hypertarget{ref-Fenson2007}{}%
+Fenson, L., Marchman, V. A., Thal, D. J., Dale, P. S., Reznick, J. S.,
+\& Bates, E. (2007). \emph{{M}ac{A}rthur-{B}ates {C}ommunicative
+{D}evelopment {I}nventories: User's guide and technical manual (2nd
+ed.)}. Baltimore, MD: Brookes.
+
+\leavevmode\hypertarget{ref-frank2017}{}%
 Frank, M. C., Braginsky, M., Yurovsky, D., \& Marchman, V. A. (2017).
 Wordbank: An open repository for developmental vocabulary data.
-\emph{Journal of Child Language}, \emph{44}(3), 677--694.
+\emph{Journal of Child Language}, \emph{44}(3), 677.
 
-\leavevmode\hypertarget{ref-Frank2021}{}%
+\leavevmode\hypertarget{ref-frank2021}{}%
 Frank, M. C., Braginsky, M., Yurovsky, D., \& Marchman, V. A. (2021).
@@
 wordbank project}. MIT Press.
+
+\leavevmode\hypertarget{ref-holland1993differential}{}%
+Holland, P. W., \& Wainer, H. (1993). \emph{Differential item
+functioning}. Routledge.
+
+\leavevmode\hypertarget{ref-kachergis2021cat}{}%
+Kachergis, G., Marchman, V. A., Dale, P., Mankewitz, J., \& Frank, M. C.
+(2021). Online computerized adaptive tests (CAT) of children's
+vocabulary development in english and mexican spanish.
+
+\leavevmode\hypertarget{ref-mayor2019}{}%
+Mayor, J., \& Mani, N. (2019). A short version of the
+{M}ac{A}rthur-{B}ates {C}ommunicative {D}evelopment {I}nventories with
+high validity. \emph{Behavior Research Methods}, \emph{51}(5),
+2248--2255.
+
+\leavevmode\hypertarget{ref-petersen2018gender}{}%
+Petersen, J. (2018). Gender difference in verbal performance: A
+meta-analysis of united states state performance assessments.
+\emph{Educational Psychology Review}. Springer.
+
+\leavevmode\hypertarget{ref-reckase2009}{}%
+Reckase, M. D. (2009). Multidimensional item response theory models. In
+\emph{Multidimensional item response theory} (pp. 79--112). Springer.
+
+\leavevmode\hypertarget{ref-schwab2016}{}%
+Schwab, J. F., \& Lew-Williams, C. (2016). Language learning,
+socioeconomic status, and child-directed speech. \emph{WIREs Cognitive
+Science}, \emph{7}, 264--275.
+http://doi.org/\href{https://doi.org/10.1002/wcs.1393}{10.1002/wcs.1393}
 
@@
 
+\end{CSLReferences}
+
 \bibliographystyle{apacite}

### 2022-01-13 paper/cogsci.Rmd
writing

@@
 ---
-title: "Words aren't created equal: Investigating bias on the CDI"
+title: "Estimating demographic bias on tests of children's early vocabulary"
 bibliography: library.bib
@@
 author-information: > 
-    \author{{\large \bf Morton Ann Gernsbacher (MAG@Macc.Wisc.Edu)} \\ Department of Psychology, 1202 W. Johnson Street \\ Madison, WI 53706 USA
-    \AND {\large \bf Sharon J.~Derry (SDJ@Macc.Wisc.Edu)} \\ Department of Educational Psychology, 1025 W. Johnson Street \\ Madison, WI 53706 USA}
+    \author{{\large \bf George Kachergis (kachergis@stanford.edu)}
+    \AND {\large \bf Nathan Francis (nathan99@stanford.edu)}
+    \AND {\large \bf Michael C. Frank (mcfrank@stanford.edu)} \\ Department of Psychology, Stanford Unviersity \\ Stanford, CA 94305 USA
+    }
 
@@
     equally often by children of varying backgrounds. Here, we use a database of American English 
-    CDIs to identify words that show strong bias for particular groups of children, along the axes
+    CDIs to identify words that show strong bias for particular groups of children, on dimensions
     of sex (male vs. female), race/ethnicity (white vs. non-white), and socioeconomic status 
-    (high vs. low). For each axis, we identify dozens of strongly biased items, and show that 
+    (high vs. low). For each dimension, we identify dozens of strongly biased items, and show that 
     eliminating these items reduces the expected ability difference between groups. For sex, we
@@
 Yet socioeconomic status (SES) has also been found to be predictive of children's early language skill, and of later educational outcomes [for a review, see @schwab2016].
-Biological factors are also predictive of language skill: first-born children tend to outpace their siblings, and female children tend to have better language skills than their age-matched male counterparts [@fenson1994] -- an sex-based verbal advantage that continues through high school [see @petersen2018gender for a review].
+Demographic factors are also predictive of language skill: first-born children tend to outpace their siblings, and female children tend to have better language skills than their age-matched male counterparts [@fenson1994] -- a sex-based verbal advantage that continues through high school [see @petersen2018gender for a review].
 However, it is also difficult to measure language skill: in any short recording children are unlikely to use all of the words and constructions that they know, and any comprehensive battery of language measures will likely exhaust children's attention span.
@@
 The idea that some items on a test may show bias, favoring one group over another, is known as Differential Item Function (DIF; @holland1993differential]. 
+On any given test, it is clearly undesirable to have more items favoring one group (say, children from rural households) over another (urban children), as the test will overestimate the ability of test-takers in the former (rural) group -- and not because of any underlying mean difference in ability between the groups, but simply because the test is unfair.
 A variety of statistical methods for detecting DIF have been proposed, and investigations have in several instances identified DIF for many items on tests favoring particular one group over another (e.g., rural vs.
@@
 
-We take the approach put forward by @stenhaug2021treading: GLIMMER plots based on the 1-parameter logistic are sufficient to identify bias, without potentially obscuring DIF in a more complex model's additional parameters.
+However, DIF is fundamentally difficult to identify for multiple reasons [for an overview, see @stenhaug2021treading].
+First, most techniques to identify DIF rely on defining a set of "anchor" test items that are assumed to be equally difficult (i.e., unbiased) for both groups of interest.
+Identifying anchor items is at best fraught when there may in fact be a difference in ability between groups (e.g., the female language advantage), and is further confounded when the magnitude of this ability difference is unknown.
+A reason that DIF is particularly tricky in measuring children's early language ability is that there is a finite universe of early-learned words to choose from--and we may expect many of them to be biased for various environmental reasons (e.g., children in Florida may not use mittens or skis).
+Hence, the presence of DIF on a wide variety of items may not indicate the presence of bias; it could indicate that one demographic has a higher average ability level than the other.
 
-DIF is, however, a challenging metric for identifying bias for a couple reasons. 
-Firstly, DIF is extremely hard to accurately root out. 
-The process of finding DIF involves using the test you used to do IRT analysis in order to look for bias within the very same test. 
-Secondly, the presence of DIF does not necessarily indicate the presence of bias; it can indicate that one demographic has a higher average ability level than the other. 
-These two issues are the basis upon which we get the Fundamental DIF Identification problem.
-
+<!--
 This problem can be be understood through a concrete example that is well-known from research on early language learning: consider the fact that females show a larger vocabulary than males across early development [see @frank2021 Ch.
@@
 If the boys and girls were of the same average language ability, this would not be an issue but given the evidence showing that girls have a higher ability level, we need to know the difference in ability so that when we find DIF for a specific word can be sure it is outside of the expecting DIF that results naturally from their difference in ability.
+-->
 
@@
 First, we will introduce the Wordbank data and the IRT model we use to analyze the data.
+We then fit the IRT model to each group along each dimension, and examine the parameters for evidence of DIF
 
@@
 We analyze parent-reported Wordbank data from `r nrow(d_demo)` American English CDI: Words & Sentences administrations for children `r min(d_demo$age)` to `r max(d_demo$age)` months of age [@frank2017; @frank2021].
-Full demographic data are not reported in some datasets contributed to Wordbank: sex is available for `r nrow(d_demo)-sum(is.na(d_demo$sex))` children, `r nrow(d_demo)-sum(is.na(d_demo$sex))`.
+Full demographic data are not reported in some datasets contributed to Wordbank: sex was available for `r nrow(d_demo)-sum(is.na(d_demo$sex))` children, race/ethnicity was available for `r nrow(d_demo)-sum(is.na(d_demo$ethnicity))` children, and mother's education (a proxy for socioeconomic status; SES) was available for `r nrow(d_demo)-sum(is.na(d_demo$ses_group))` children.
 
-We focused on comparing data between demographic groups on the axes of Sex, Ethnicity and Mother's Education, a proxy for socioeconomic status (henceforth, SES). 
-Our data included `r sex_tab["Female"]` female and `r sex_tab["Male"]` male participants.
-In terms of Ethnicity data was recorded from `r eth_tab["White"]` white, `r eth_tab["Asian"]` Asian, `r eth_tab["Black"]` Black, `r eth_tab["Hispanic"]` Hispanic, and `r eth_tab["Other"]` “Other” participants as well as `r sum(is.na(d_demo$ethnicity))` unreported.
-Given our distribution of ethnicities, for our analysis we split participants into White (`r eth2_tab["White"]`) and Non-White (`r eth2_tab["Nonwhite"]`).
-Finally, we split participants into high and low SES, of which we had `r ses_tab["high"]` high SES participants, corresponding to mothers' education of some college or higher and `r ses_tab["low"]` Low SES participants.
-<!-- same for eth, SES -->
+The analysis of sex-based differences included CDI administrations from `r sex_tab["Female"]` female and `r sex_tab["Male"]` male children.
+The analysis of race/ethnicity-based differences included data from `r eth_tab["White"]` white, `r eth_tab["Asian"]` Asian, `r eth_tab["Black"]` Black, `r eth_tab["Hispanic"]` Hispanic, and `r eth_tab["Other"]` "Other" children.
+For this analysis we categorized participants' race as White (`r eth2_tab["White"]`) or Non-White (`r eth2_tab["Nonwhite"]`). 
+For the SES-based analysis, we categorized the `r ses_tab["high"]` children whose mothers had at least some college or more as high-SES, and those whose mothers had at most high school (`r ses_tab["low"]` children) as low-SES.
 

### 2022-01-13 paper/cogsci.tex
writing

@@
 
-\title{Words aren't created equal: Investigating bias on the CDI}
+\title{Estimating demographic bias on tests of children's early vocabulary}
 
-\author{{\large \bf Morton Ann Gernsbacher (MAG@Macc.Wisc.Edu)} \\ Department of Psychology, 1202 W.
-Johnson Street \\ Madison, WI 53706 USA \AND {\large \bf Sharon J.~Derry (SDJ@Macc.Wisc.Edu)} \\ Department of Educational Psychology, 1025 W.
-Johnson Street \\ Madison, WI 53706 USA}
-
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newenvironment{CSLReferences}%
-  {}%
-  {\par}
+\author{{\large \bf George Kachergis (kachergis@stanford.edu)} \AND {\large \bf Nathan Francis (nathan99@stanford.edu)} \AND {\large \bf Michael C.
+Frank (mcfrank@stanford.edu)} \\ Department of Psychology, Stanford Unviersity \\ Stanford, CA 94305 USA }
 
@@
 identify words that show strong bias for particular groups of children,
-along the axes of sex (male vs.~female), race/ethnicity (white
-vs.~non-white), and socioeconomic status (high vs.~low). For each axis,
-we identify dozens of strongly biased items, and show that eliminating
-these items reduces the expected ability difference between groups. For
-sex, we consider how to propose replacement words that may show less
-bias, on the basis of their relatively equal frequency in adult speech
-directed to male and female children.
+on dimensions of sex (male vs.~female), race/ethnicity (white
+vs.~non-white), and socioeconomic status (high vs.~low). For each
+dimension, we identify dozens of strongly biased items, and show that
+eliminating these items reduces the expected ability difference between
+groups. For sex, we consider how to propose replacement words that may
+show less bias, on the basis of their relatively equal frequency in
+adult speech directed to male and female children.
 
@@
 language skill, and of later educational outcomes (for a review, see
-Schwab \& Lew-Williams, 2016). Biological factors are also predictive of
-language skill: first-born children tend to outpace their siblings, and
-female children tend to have better language skills than their
-age-matched male counterparts (Larry Fenson et al., 1994) -- an
-sex-based verbal advantage that continues through high school (see
-Petersen, 2018 for a review). However, it is also difficult to measure
-language skill: in any short recording children are unlikely to use all
-of the words and constructions that they know, and any comprehensive
-battery of language measures will likely exhaust children's attention
-span.
+Schwab \& Lew-Williams, 2016). Demographic factors are also predictive
+of language skill: first-born children tend to outpace their siblings,
+and female children tend to have better language skills than their
+age-matched male counterparts (Fenson et al., 1994) -- a sex-based
+verbal advantage that continues through high school (see Petersen, 2018
+for a review). However, it is also difficult to measure language skill:
+in any short recording children are unlikely to use all of the words and
+constructions that they know, and any comprehensive battery of language
+measures will likely exhaust children's attention span.
 
-The MacArthur-Bates Communicative Development Inventories {[}CDIs; L.
-Fenson et al. (2007){]} are a set of parent-reported measures of
-children's productive and receptive language skills, which produce a
-low-cost and reliable way to estimate children's early language skills
-(Larry Fenson et al., 1994). CDIs have shown good predictive validity
-{[}e.g., Larry Fenson et al. (1994); Bornstein et al.~2012, Duff et
-al.~2015{]}. Our focus will be on the vocabulary checklist portion of
-the CDI Words \& Sentences (CDI:WS) form, comprised of 680 early-learned
-words across 22 categories (e.g., animals, vehicles, action words,
-pronouns) selected to assess the productive vocabulary of children 16 to
-30 months of age. For each item on the CDI:WS, caregivers are asked to
-respond whether the target child has been heard to say (i.e.~produce)
-the given item.\footnote{The CDI: Words and Gestures form includes a
-  subset of the CDI:WS items and targets children 12 to 18 months of
-  age, measuring both comprehension and production.} Children's total
-vocabulary score on the CDI:WS is tightly correlated with other facets
-of early language (e.g., grammatical competence and gesture), suggesting
-that the language system is ``tightly woven'' (Frank, Braginsky,
-Yurovsky, \& Marchman, 2021). Due to these desirable properties, CDIs
-have been adapted to dozens of languages, and a central repository of
-CDI data contributed from all over the world has been created
-{[}Wordbank; Frank, Braginsky, Yurovsky, \& Marchman (2017); Frank,
-Braginsky, Yurovsky, \& Marchman (2021){]}.
+The MacArthur-Bates Communicative Development Inventories (CDIs; Fenson
+et al., 2007) are a set of parent-reported measures of children's
+productive and receptive language skills, which produce a low-cost and
+reliable way to estimate children's early language skills (Fenson et
+al., 1994). CDIs have shown good predictive validity {[}e.g., Fenson et
+al. (1994); Bornstein et al.~2012, Duff et al.~2015{]}. Our focus will
+be on the vocabulary checklist portion of the CDI Words \& Sentences
+(CDI:WS) form, comprised of 680 early-learned words across 22 categories
+(e.g., animals, vehicles, action words, pronouns) selected to assess the
+productive vocabulary of children 16 to 30 months of age. For each item
+on the CDI:WS, caregivers are asked to respond whether the target child
+has been heard to say (i.e.~produce) the given item.\footnote{The CDI:
+  Words and Gestures form includes a subset of the CDI:WS items and
+  targets children 12 to 18 months of age, measuring both comprehension
+  and production.} Children's total vocabulary score on the CDI:WS is
+tightly correlated with other facets of early language (e.g.,
+grammatical competence and gesture), suggesting that the language system
+is ``tightly woven'' (Frank, Braginsky, Yurovsky, \& Marchman, 2021).
+Due to these desirable properties, CDIs have been adapted to dozens of
+languages, and a central repository of CDI data contributed from all
+over the world has been created (Wordbank; Frank, Braginsky, Yurovsky,
+\& Marchman, 2017; Frank et al., 2021).
 
@@
 adaptive tests to reliably assess language ability using only a small
-subset of the CDI items Kachergis, Marchman, Dale, Mankewitz, \& Frank
-(2021). These psychometric models typically come from the Item-Response
-Theory (IRT) framework (Baker, 2001), which assumes that not only
-test-takers (here, children) have normally-distributed ability, but that
-items (words on the CDI) have normally-distributed difficulty. The very
-efficacy of these IRT-based models depends on words varying in
+subset of the CDI items (e.g., Mayor \& Mani, 2019, p.
+@kachergis2021cat). These psychometric models typically come from the
+Item-Response Theory (IRT) framework (Baker, 2001), which assumes that
+not only test-takers (here, children) have normally-distributed ability,
+but that items (words on the CDI) have normally-distributed difficulty.
+The very efficacy of these IRT-based models depends on words varying in
 difficulty, and hence being more/less informative of the ability level
@@
 produces the word ``ball'' is far less informative of that child's
-language ability than asking whether they produce ``table,'' as 96\% of
+language ability than asking whether they produce ``table'', as 96\% of
 22-month-olds can produce the former, while only 47\% produce the
@@
 another, is known as Differential Item Function (DIF; Holland \& Wainer
-(1993){]}. A variety of statistical methods for detecting DIF have been
-proposed, and investigations have in several instances identified DIF
-for many items on tests favoring particular one group over another
-(e.g., rural vs.~urban). Our goal here is to test the items on the
-American CDI:WS for DIF along three main axes: sex (male vs.~female),
-socioeconomic status (low- vs.~high-SES), and ethnicity (white
-vs.~non-white).
+(1993){]}. On any given test, it is clearly undesirable to have more
+items favoring one group (say, children from rural households) over
+another (urban children), as the test will overestimate the ability of
+test-takers in the former (rural) group -- and not because of any
+underlying mean difference in ability between the groups, but simply
+because the test is unfair. A variety of statistical methods for
+detecting DIF have been proposed, and investigations have in several
+instances identified DIF for many items on tests favoring particular one
+group over another (e.g., rural vs.~urban). Our goal here is to test the
+items on the American CDI:WS for DIF along three main axes: sex (male
+vs.~female), socioeconomic status (low- vs.~high-SES), and ethnicity
+(white vs.~non-white).
 
-We take the approach put forward by Stenhaug, Frank, \& Domingue (2021):
-GLIMMER plots based on the 1-parameter logistic are sufficient to
-identify bias, without potentially obscuring DIF in a more complex
-model's additional parameters.
-
-DIF is, however, a challenging metric for identifying bias for a couple
-reasons. Firstly, DIF is extremely hard to accurately root out. The
-process of finding DIF involves using the test you used to do IRT
-analysis in order to look for bias within the very same test. Secondly,
-the presence of DIF does not necessarily indicate the presence of bias;
-it can indicate that one demographic has a higher average ability level
-than the other. These two issues are the basis upon which we get the
-Fundamental DIF Identification problem.
-
-This problem can be be understood through a concrete example that is
-well-known from research on early language learning: consider the fact
-that females show a larger vocabulary than males across early
-development {[}see Frank, Braginsky, Yurovsky, \& Marchman (2021) Ch.
-6?; OTHER REFS{]}. The question is whether or not females actually have
-a higher ability level, or if the tests predominantly have words that
-are easier for females (i.e., biased). (It could even be that the test
-is biased toward males but the language ability of females is large
-enough to overcome that bias.) To address this question, we need to know
-the ability levels of girls and boys in order to confirm that if a word
-is learned earlier by girls it is simply because of an ability
-difference rather than a bias inherent to the word. The problem arises
-from the fact that we measure ability level using the very same test
-that we are trying to check for biased words. If the boys and girls were
-of the same average language ability, this would not be an issue but
-given the evidence showing that girls have a higher ability level, we
-need to know the difference in ability so that when we find DIF for a
-specific word can be sure it is outside of the expecting DIF that
-results naturally from their difference in ability.
+However, DIF is fundamentally difficult to identify for multiple reasons
+(for an overview, see Stenhaug, Frank, \& Domingue, 2021). First, most
+techniques to identify DIF rely on defining a set of ``anchor'' test
+items that are assumed to be equally difficult (i.e., unbiased) for both
+groups of interest. Identifying anchor items is at best fraught when
+there may in fact be a difference in ability between groups (e.g., the
+female language advantage), and is further confounded when the magnitude
+of this ability difference is unknown. A reason that DIF is particularly
+tricky in measuring children's early language ability is that there is a
+finite universe of early-learned words to choose from--and we may expect
+many of them to be biased for various environmental reasons (e.g.,
+children in Florida may not use mittens or skis). Hence, the presence of
+DIF on a wide variety of items may not indicate the presence of bias; it
+could indicate that one demographic has a higher average ability level
+than the other.
 
 The outline of this paper is as follows. First, we will introduce the
-Wordbank data and the IRT model we use to analyze the data.
+Wordbank data and the IRT model we use to analyze the data. We then fit
+the IRT model to each group along each dimension, and examine the
+parameters for evidence of DIF
 
@@
 Words \& Sentences administrations for children 16 to 30 months of age
-(Frank, Braginsky, Yurovsky, \& Marchman, 2017, 2021). Full demographic
-data are not reported in some datasets contributed to Wordbank: sex is
-available for 4094 children, 4094.
+(Frank et al., 2017, 2021). Full demographic data are not reported in
+some datasets contributed to Wordbank: sex was available for 4094
+children, race/ethnicity was available for 2715 children, and mother's
+education (a proxy for socioeconomic status; SES) was available for 5520
+children.
 
-We focused on comparing data between demographic groups on the axes of
-Sex, Ethnicity and Mother's Education, a proxy for socioeconomic status
-(henceforth, SES). Our data included 1989 female and 2105 male
-participants. In terms of Ethnicity data was recorded from 2202 white,
-67 Asian, 222 Black, 131 Hispanic, and 93 ``Other'' participants as well
-as 2805 unreported. Given our distribution of ethnicities, for our
-analysis we split participants into White (2202) and Non-White (513).
-Finally, we split participants into high and low SES, of which we had
-4973 high SES participants, corresponding to mothers' education of some
-college or higher and 547 Low SES participants.
+The analysis of sex-based differences included CDI administrations from
+1989 female and 2105 male children. The analysis of race/ethnicity-based
+differences included data from 2202 white, 67 Asian, 222 Black, 131
+Hispanic, and 93 ``Other'' children. For this analysis we categorized
+participants' race as White (2202) or Non-White (513). For the SES-based
+analysis, we categorized the 4973 children whose mothers had at least
+some college or more as high-SES, and those whose mothers had at most
+high school (547 children) as low-SES.
 
@@
 \hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
 \leavevmode\hypertarget{ref-Baker2001}{}%
@@
 \leavevmode\hypertarget{ref-R-mirt}{}%
-Chalmers, R. P. (2012). {mirt}: A multidimensional item response theory
-package for the {R} environment. \emph{Journal of Statistical Software},
-\emph{48}(6), 1--29.
-http://doi.org/\href{https://doi.org/10.18637/jss.v048.i06}{10.18637/jss.v048.i06}
+Chalmers, R. P. (2012). mirt: A multidimensional item response theory
+package for the R environment. \emph{Journal of Statistical Software},
+\emph{48}(6), 1--29. \url{http://doi.org/10.18637/jss.v048.i06}
 
 \leavevmode\hypertarget{ref-fenson1994}{}%
-Fenson, Larry, Dale, P. S., Reznick, J. S., Bates, E., Thal, D. J.,
+Fenson, L., Dale, P. S., Reznick, J. S., Bates, E., Thal, D. J.,
 Pethick, S. J., \ldots{} Stiles, J. (1994). Variability in early
@@
 Fenson, L., Marchman, V. A., Thal, D. J., Dale, P. S., Reznick, J. S.,
-\& Bates, E. (2007). \emph{{M}ac{A}rthur-{B}ates {C}ommunicative
-{D}evelopment {I}nventories: User's guide and technical manual (2nd
-ed.)}. Baltimore, MD: Brookes.
+\& Bates, E. (2007). \emph{MacArthur-Bates Communicative Development
+Inventories: User's guide and technical manual (2nd ed.)}. Baltimore,
+MD: Brookes.
 
@@
 Kachergis, G., Marchman, V. A., Dale, P., Mankewitz, J., \& Frank, M. C.
-(2021). Online computerized adaptive tests (CAT) of children's
+(2021). Online computerized adaptive tests (cat) of children's
 vocabulary development in english and mexican spanish.
@@
 \leavevmode\hypertarget{ref-mayor2019}{}%
-Mayor, J., \& Mani, N. (2019). A short version of the
-{M}ac{A}rthur-{B}ates {C}ommunicative {D}evelopment {I}nventories with
-high validity. \emph{Behavior Research Methods}, \emph{51}(5),
-2248--2255.
+Mayor, J., \& Mani, N. (2019). A short version of the MacArthur-Bates
+Communicative Development Inventories with high validity. \emph{Behavior
+Research Methods}, \emph{51}(5), 2248--2255.
 
@@
 socioeconomic status, and child-directed speech. \emph{WIREs Cognitive
-Science}, \emph{7}, 264--275.
-http://doi.org/\href{https://doi.org/10.1002/wcs.1393}{10.1002/wcs.1393}
+Science}, \emph{7}, 264--275. \url{http://doi.org/10.1002/wcs.1393}
 
@@
 
-\end{CSLReferences}
-
 \bibliographystyle{apacite}

### 2022-01-16 paper/cogsci.Rmd
writing

@@
 First, we will introduce the Wordbank data and the IRT model we use to analyze the data.
-We then fit the IRT model to each group along each dimension, and examine the parameters for evidence of DIF
+We then fit the IRT model to each group along each demographic dimension, and examine the parameters for evidence of DIF, noting in particular how many items are significantly biased in favor of each group.
+We identify a set of suspect items to eliminate (or in the future, replace), and provide updated estimates of the effect size of these demographic variables, were the biased items to be eliminated.
+Next, using a corpus of child-directed speech, we examine the extent to which variation in word frequencies heard by boys vs. girls predicts sex-based DIF.
+Finally, we provide recommendations for next steps to be taken to identify and replace biased items on the CDI.
 
@@
 
+```{r load-data}
+load("../data/glimmer_prodWS_models.Rds")
+
+library(mirt)
+
+mod_intuitive_eth
+mod_intuitive_ses
+mod_intuitive_sex
+```
+
+```{r}
+# baseline Rasch model differences in ability by demographic groups
+```
+
+```{r}
+mod_intuitive_ses <- fit_mod_intuitive(d_mat, ses_group)
+```
+
+```{r demo-ability}
+# show density plots of child ability based on group models
+# sex
+ses <- d_demo %>% filter(!is.na(sex_group), production!=0) %>%
+  mutate(theta_sex = fscores(mod_intuitive_sex)) %>%
+  rename(`Sex Group`=sex_group)
+# cor.test(eth$production, eth$theta_eth) # .98
+
+sex_m <- sex %>% group_by(`Sex Group`) %>%
+  summarise(ability = mean(theta_sex, na.rm=T))
+
+p1 <- sex %>%
+  ggplot(aes(x=theta_sex, color=`Sex Group`, fill=`Sex Group`)) +
+    geom_density(alpha=.2) + theme_classic() + xlab("Ability") + 
+    geom_vline(data=sex_m, aes(xintercept=ability, color=`Sex Group`))
+
+# SES
+ses <- d_demo %>% filter(!is.na(ses_group), production!=0) %>%
+  mutate(theta_ses = fscores(mod_intuitive_ses)) %>%
+  rename(`SES Group`=ses_group)
+# cor.test(eth$production, eth$theta_eth) # .98
+
+ses_m <- ses %>% group_by(`Ethnic Group`) %>%
+  summarise(ability = mean(theta_ses, na.rm=T))
+
+p2 <- ses %>%
+  ggplot(aes(x=theta_ses, color=`SES Group`, fill=`SES Group`)) +
+    geom_density(alpha=.2) + theme_classic() + xlab("Ability") + 
+    geom_vline(data=ses_m, aes(xintercept=ability, color=`SES Group`))
+
+# ethnicity
+eth <- d_demo %>% filter(!is.na(eth_group), production!=0) %>%
+  mutate(theta_eth = fscores(mod_intuitive_eth)) %>%
+  rename(`Ethnic Group`=eth_group)
+# cor.test(eth$production, eth$theta_eth) # .98
+
+eth_m <- eth %>% group_by(`Ethnic Group`) %>%
+  summarise(ability = mean(theta_eth, na.rm=T))
+
+p3 <- eth %>%
+  ggplot(aes(x=theta_eth, color=`Ethnic Group`, fill=`Ethnic Group`)) +
+    geom_density(alpha=.2) + theme_classic() + xlab("Ability") + 
+    geom_vline(data=eth_m, aes(xintercept=ability, color=`Ethnic Group`))
+
+ggarrange(p1, p2, p3)
+```
+
 <!-- put the GLIMMERs side-by-side?

### 2022-01-16 GLIMMER_production.R

@@ -19,7 +19,7 @@ get_production_data <- function() {
   d_demo <- get_administration_data(language = "English (American)", form = "WS")
   
   # subjects not producing any words yet (can't fit model with them)
-  bad_Ss = d_demo[which(d_demo$comprehension==0),]$data_id # 9 bad production Ss
+  bad_Ss = d_demo[which(d_demo$production==0),]$data_id # 9 bad production Ss
   
   d_long_ws <- d_long_ws %>% left_join(en_ws %>% select(num_item_id, definition)) %>%
     filter(!is.na(definition))
@@ -39,7 +39,7 @@ get_production_data <- function() {
 load(here("data/en_ws_production.Rdata")) # d_demo, d_mat, en_wg (items)
 
 d_demo <- d_demo %>% 
-  filter(comprehension!=0, !is.na(sex)) %>% # can't fit children not producing words, or with NA sex in group model
+  filter(production!=0) %>% # can't fit children not producing words, or with NA sex in group model
   arrange(data_id) %>%
   mutate(eth_group = ifelse(ethnicity=="White", "White", "Nonwhite"),
          ses_group = 

### 2022-01-16 03-preliminary-WSprod.Rmd
writing

@@
 We will fit comprehension data from `r nrow(d_demo)` children for the WS form.
+
+## Baseline 1PL model
+
+```{r}
+m0 <- mirt(d_mat %>% filter(rowSums(d_mat)!=0) %>% select(-data_id), 1, "Rasch", 
+           technical=list(NCYCLES=1000))
+
+coefs_1pl <- as_tibble(coef(m0, simplify = TRUE)$items) %>%
+  mutate(definition = rownames(coef(m0, simplify = TRUE)$items))
+fscores_1pl <- tibble(data_id = d_mat %>% filter(rowSums(d_mat)!=0) %>% select(data_id), 
+                             ability = fscores(m0, method = "MAP")[,1])
+
+save(file = here("data/eng_Ws_prod_mod_1pl.Rds"), "m0", "coefs_1pl", "fscores_1pl")
+```
 

### 2022-01-16 paper/cogsci.Rmd

--- 2022-01-13 paper/cogsci.Rmd
@@
 library(here)
-
+library(mirt)
+library(ggpubr)
 ```
@@
 We fitted a multigroup Rasch model for each demographic dimension (sex, socioeconomic status, and ethnicity).
+To contextualize these results, we also first fitted a baseline Rasch model
 
 # Results
+
+```{r load-data}
+load(here("data/glimmer_prodWS_models.Rds"))
+load(here("data/eng_Ws_prod_mod_1pl.Rds")) # m0h - baseline Rasch model, with coefs_1pl and fscores_1pl
+#cor.test(d_demo$production, d_demo$ability) # .972
+d_demo <- d_demo %>% left_join(fscores_1pl)
+
+#d_demo %>% filter(!is.na(sex)) %>%
+#  ggplot(aes(x=sex, y=ability)) +
+#  geom_boxplot() + theme_classic()
+
+sex_m <- d_demo %>% filter(!is.na(sex), !is.na(ability), age<26, age>22) %>% 
+  group_by(sex) %>% summarise(ability=mean(ability),
+                              production=mean(production))
+
+ses_m <- d_demo %>% filter(!is.na(ses_group), !is.na(ability), age<26, age>22) %>% 
+  group_by(ses_group) %>% summarise(ability=mean(ability),
+                                    production=mean(production))
+
+eth_m <- d_demo %>% filter(!is.na(eth_group), age<26, age>22) %>% 
+  group_by(eth_group) %>% summarise(ability=mean(ability),
+                                    production=mean(production))
+
+# sex_m[1,2:3] - sex_m[2,2:3] # average female vs. male advantage at 23-25 months of age: 0.78 theta, 69 words
+# ses_m[1,2:3] - ses_m[2,2:3] # high vs. low advantage: 0.33 theta, 33 words
+# eth_m[2,2:3] - eth_m[1,2:3] # white vs. nonwhite advantage: 0.52 theta, 45
+
+# should we report t-tests? 
+table(subset(d_demo, age>22 & age<26)$sex) # 395 female, 408 male
+table(subset(d_demo, age>22 & age<26)$ses_group) # 1038 high-SES, 140 low-SES
+table(subset(d_demo, age>22 & age<26)$eth_group) # 144 nonwhite, 603 white
+```
+
+Figure 1 shows children's language ability vs. age by demographic group from the baseline Rasch model, which assumes no DIF.
+Although the demographic language ability advantages go in the expected directions (i.e., favoring female, high-SES, and white children), note that magnitude of the differences vary somewhat across the age range, especially for SES and ethnicity.
+Thus, although the IRT models are fitted to the entire age range, our comparisons of average ability will focus on the middle of the CDI:WS' intended age range, where there is great variation in the size and composition of children's vocabulary.
+At ages 23-25 months, the average female vs. male advantage is an ability difference of $\theta_{f-m} = 0.78$, amounting to females knowing 69 more words, on average.
+The average advantage for high-SES children at 23-25 months is $\theta_{high-low} = 0.33$, amounting to 33 more produced CDI items.
+The average advantage for white vs. non-white children in this age range is $\theta_{w-nonw} = 0.52$, an average of 45 more produced CDI items.
+
+```{r baseline-ability-vs-age, out.width="\\linewidth", fig.env = "figure", fig.pos = "H", fig.align='center', fig.width=3.3, fig.height=6, set.cap.width=T, num.cols.cap=1, fig.cap = "Language ability vs.
+age by demographic group, from the Rasch model."}
+p1 <- d_demo %>% filter(!is.na(sex)) %>%
+  ggplot(aes(x=jitter(age), y=ability, color=sex)) + 
+  geom_point(alpha=.1) + geom_smooth() + theme_classic() + 
+  xlab("Age (months)") + ylab("Language Ability")
+
+p2 <- d_demo %>% filter(!is.na(ses_group)) %>%
+  ggplot(aes(x=jitter(age), y=ability, color=ses_group)) + 
+  geom_point(alpha=.1) + geom_smooth() + theme_classic() + 
+  xlab("Age (months)") + ylab("Language Ability") +
+  scale_color_discrete(name = "SES")
+
+p3 <- d_demo %>% filter(!is.na(eth_group)) %>%
+  ggplot(aes(x=jitter(age), y=ability, color=eth_group)) + 
+  geom_point(alpha=.1) + geom_smooth() + theme_classic() + 
+  xlab("Age (months)") + ylab("Language Ability") + 
+  scale_color_discrete(name = "ethnicity")
+
+ggarrange(p1, p2, p3, ncol=1)
+```
+
+```{r demo-ability}
+# show density plots of child ability based on group models
+# sex
+sex <- d_demo %>% filter(!is.na(sex), production!=0) %>%
+  mutate(theta_sex = fscores(mod_intuitive_sex)) 
+# cor.test(eth$production, eth$theta_eth) # .98
+
+sex_m <- sex %>% group_by(sex) %>%
+  summarise(ability = mean(theta_sex, na.rm=T))
+
+p1 <- sex %>%
+  ggplot(aes(x=theta_sex, color=sex, fill=sex)) +
+    geom_density(alpha=.2) + theme_classic() + xlab("Ability") + 
+    geom_vline(data=sex_m, aes(xintercept=ability, color=sex))
+
+# SES
+ses <- d_demo %>% filter(!is.na(ses_group), production!=0) %>%
+  mutate(theta_ses = fscores(mod_intuitive_ses)) %>%
+  rename(SES=ses_group)
+# cor.test(eth$production, eth$theta_eth) # .98
+
+ses_m <- ses %>% group_by(SES) %>%
+  summarise(ability = mean(theta_ses, na.rm=T))
+
+p2 <- ses %>%
+  ggplot(aes(x=theta_ses, color=SES, fill=SES)) +
+    geom_density(alpha=.2) + theme_classic() + xlab("Ability") + 
+    geom_vline(data=ses_m, aes(xintercept=ability, color=SES))
+
+# ethnicity
+eth <- d_demo %>% filter(!is.na(eth_group), production!=0) %>%
+  mutate(theta_eth = fscores(mod_intuitive_eth)) %>%
+  rename(`Ethnic Group`=eth_group)
+# cor.test(eth$production, eth$theta_eth) # .98
+
+eth_m <- eth %>% group_by(`Ethnic Group`) %>%
+  summarise(ability = mean(theta_eth, na.rm=T))
+
+p3 <- eth %>%
+  ggplot(aes(x=theta_eth, color=`Ethnic Group`, fill=`Ethnic Group`)) +
+    geom_density(alpha=.2) + theme_classic() + xlab("Ability") + 
+    geom_vline(data=eth_m, aes(xintercept=ability, color=`Ethnic Group`))
+
+ggarrange(p1, p2, p3)
+```
 

### 2022-01-16 paper/cogsci.Rmd
color-blind-friendly colors

@@
 
+```{r colors}
+# color-blind palette
+cbp1 <- c("#999999", "#E69F00", "#56B4E9", "#009E73",
+          "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
+sex_colors = cbp1[2:3]
+eth_colors = cbp1[4:5]
+ses_colors = cbp1[6:7]
+```
+
 ```{r load-data}
@@
 # should we report t-tests? 
-table(subset(d_demo, age>22 & age<26)$sex) # 395 female, 408 male
-table(subset(d_demo, age>22 & age<26)$ses_group) # 1038 high-SES, 140 low-SES
-table(subset(d_demo, age>22 & age<26)$eth_group) # 144 nonwhite, 603 white
+#table(subset(d_demo, age>22 & age<26)$sex) # 395 female, 408 male
+#table(subset(d_demo, age>22 & age<26)$ses_group) # 1038 high-SES, 140 low-SES
+#table(subset(d_demo, age>22 & age<26)$eth_group) # 144 nonwhite, 603 white
 ```
@@
   ggplot(aes(x=jitter(age), y=ability, color=sex)) + 
-  geom_point(alpha=.1) + geom_smooth() + theme_classic() + 
-  xlab("Age (months)") + ylab("Language Ability")
+  geom_point(alpha=.1) + geom_smooth() + theme_classic(base_size=9) + 
+  xlab("Age (months)") + ylab("Language Ability") + 
+  scale_color_discrete(type=sex_colors)
+  #theme(legend.position = "top")
 
@@
   ggplot(aes(x=jitter(age), y=ability, color=ses_group)) + 
-  geom_point(alpha=.1) + geom_smooth() + theme_classic() + 
+  geom_point(alpha=.1) + geom_smooth() + theme_classic(base_size=9) + 
   xlab("Age (months)") + ylab("Language Ability") +
-  scale_color_discrete(name = "SES")
+  scale_color_discrete(name = "SES", type=ses_colors) #+ 
+  #theme(legend.position = "top")
 
@@
   ggplot(aes(x=jitter(age), y=ability, color=eth_group)) + 
-  geom_point(alpha=.1) + geom_smooth() + theme_classic() + 
+  geom_point(alpha=.1) + geom_smooth() + theme_classic(base_size=9) + 
   xlab("Age (months)") + ylab("Language Ability") + 
-  scale_color_discrete(name = "ethnicity")
+  scale_color_discrete(name = "ethnicity", type=eth_colors) #+ 
+  #theme(legend.position = "top")
 
@@
 
-```{r demo-ability}
+```{r demo-ability, out.width="\\linewidth", fig.env = "figure", fig.pos = "H", fig.align='center', fig.width=3.3, fig.height=5.5, set.cap.width=T, num.cols.cap=1, fig.cap = "Density plots of language ability for children aged 23-25 months, based on the group models."}
 # show density plots of child ability based on group models
@@
 sex_m <- sex %>% group_by(sex) %>%
+  filter(age<26, age>22) %>%
   summarise(ability = mean(theta_sex, na.rm=T))
 
-p1 <- sex %>%
+p1 <- sex %>% filter(age<26, age>22) %>%
   ggplot(aes(x=theta_sex, color=sex, fill=sex)) +
     geom_density(alpha=.2) + theme_classic() + xlab("Ability") + 
-    geom_vline(data=sex_m, aes(xintercept=ability, color=sex))
+    geom_vline(data=sex_m, aes(xintercept=ability, color=sex)) +
+    scale_color_discrete(type=sex_colors) + 
+    scale_fill_discrete(type=sex_colors)
 
@@
   mutate(theta_ses = fscores(mod_intuitive_ses)) %>%
+  filter(age<26, age>22) %>%
   rename(SES=ses_group)
@@
 ses_m <- ses %>% group_by(SES) %>%
+  filter(age<26, age>22) %>%
   summarise(ability = mean(theta_ses, na.rm=T))
@@
     geom_density(alpha=.2) + theme_classic() + xlab("Ability") + 
-    geom_vline(data=ses_m, aes(xintercept=ability, color=SES))
+    geom_vline(data=ses_m, aes(xintercept=ability, color=SES)) + 
+    scale_color_discrete(type=ses_colors) +
+    scale_fill_discrete(type=ses_colors)
 
@@
   mutate(theta_eth = fscores(mod_intuitive_eth)) %>%
+  filter(age<26, age>22) %>%
   rename(`Ethnic Group`=eth_group)
@@
 eth_m <- eth %>% group_by(`Ethnic Group`) %>%
+  filter(age<26, age>22) %>%
   summarise(ability = mean(theta_eth, na.rm=T))
@@
     geom_density(alpha=.2) + theme_classic() + xlab("Ability") + 
-    geom_vline(data=eth_m, aes(xintercept=ability, color=`Ethnic Group`))
+    geom_vline(data=eth_m, aes(xintercept=ability, color=`Ethnic Group`)) + 
+    scale_color_discrete(type=eth_colors) + 
+    scale_fill_discrete(type=eth_colors)
 
-ggarrange(p1, p2, p3)
+ggarrange(p1, p2, p3, ncol=1)
 ```

### 2022-01-16 paper/cogsci.Rmd
writing results

@@
     CDIs to identify words that show strong bias for particular groups of children, on dimensions
-    of sex (male vs. female), race/ethnicity (white vs. non-white), and socioeconomic status 
+    of sex (male vs. female), race (white vs. non-white), and socioeconomic status 
     (high vs. low). For each dimension, we identify dozens of strongly biased items, and show that 
@@
 library(ggpubr)
+
+theme_set(theme_classic(base_size=10))
 ```
@@
 urban).
-Our goal here is to test the items on the American CDI:WS for DIF along three main axes: sex (male vs. female), socioeconomic status (low- vs. high-SES), and ethnicity (white vs. non-white).
+Our goal here is to test the items on the American CDI:WS for DIF along three main axes: sex (male vs. female), socioeconomic status (low- vs. high-SES), and race (white vs. non-white).
 
@@
 The analysis of sex-based differences included CDI administrations from `r sex_tab["Female"]` female and `r sex_tab["Male"]` male children.
-The analysis of race/ethnicity-based differences included data from `r eth_tab["White"]` white, `r eth_tab["Asian"]` Asian, `r eth_tab["Black"]` Black, `r eth_tab["Hispanic"]` Hispanic, and `r eth_tab["Other"]` "Other" children.
-For this analysis we categorized participants' race as White (`r eth2_tab["White"]`) or Non-White (`r eth2_tab["Nonwhite"]`). 
+The analysis of race-based differences included data from `r eth_tab["White"]` white, `r eth_tab["Asian"]` Asian, `r eth_tab["Black"]` Black, `r eth_tab["Hispanic"]` Hispanic, and `r eth_tab["Other"]` "Other" children.
+For this analysis we categorized participants' race/ethnicity as White (`r eth2_tab["White"]`) or Non-white (`r eth2_tab["Nonwhite"]`). 
 For the SES-based analysis, we categorized the `r ses_tab["high"]` children whose mothers had at least some college or more as high-SES, and those whose mothers had at most high school (`r ses_tab["low"]` children) as low-SES.
@@
   group_by(sex) %>% summarise(ability=mean(ability),
-                              production=mean(production))
+                              production=mean(production),
+                              n=n()) # 394 female, 408 male
 
@@
   group_by(ses_group) %>% summarise(ability=mean(ability),
-                                    production=mean(production))
+                                    production=mean(production),
+                                    n=n()) # 1014 high-SES, 140 low-SES
 
@@
   group_by(eth_group) %>% summarise(ability=mean(ability),
-                                    production=mean(production))
+                                    production=mean(production),
+                                    n=n()) # 144 nonwhite, 603 white
 
@@
 # should we report t-tests? 
-#table(subset(d_demo, age>22 & age<26)$sex) # 395 female, 408 male
-#table(subset(d_demo, age>22 & age<26)$ses_group) # 1038 high-SES, 140 low-SES
-#table(subset(d_demo, age>22 & age<26)$eth_group) # 144 nonwhite, 603 white
+sex_diff_test = t.test(subset(d_demo, age>22 & age<26 & sex=="Male")$ability,
+       subset(d_demo, age>22 & age<26 & sex=="Female")$ability) # p<.001
+ses_diff_test = t.test(subset(d_demo, age>22 & age<26 & ses_group=="high")$ability,
+       subset(d_demo, age>22 & age<26 & ses_group=="low")$ability) # p=.1 (larger difference at >ages)
+eth_diff_test = t.test(subset(d_demo, age>22 & age<26 & eth_group=="White")$ability,
+       subset(d_demo, age>22 & age<26 & eth_group=="Nonwhite")$ability) # p=.01
 ```
+
+## Demographic effects in baseline Rasch model
 
@@
 
+## Demographic effects in group models
+
+maybe only do this after we've eliminated the most biased items?
+
 ```{r demo-ability, out.width="\\linewidth", fig.env = "figure", fig.pos = "H", fig.align='center', fig.width=3.3, fig.height=5.5, set.cap.width=T, num.cols.cap=1, fig.cap = "Density plots of language ability for children aged 23-25 months, based on the group models."}
 # show density plots of child ability based on group models
+
+# ToDo: maybe make these violin plots and squeeze all 6 groups side-by-side?
+
 # sex
@@
 
-sex_m <- sex %>% group_by(sex) %>%
+sex_m2 <- sex %>% group_by(sex) %>%
   filter(age<26, age>22) %>%
   summarise(ability = mean(theta_sex, na.rm=T)) 
-
+# now Female ability = 0.68, male=0.14
 p1 <- sex %>% filter(age<26, age>22) %>%
   ggplot(aes(x=theta_sex, color=sex, fill=sex)) +
-    geom_density(alpha=.2) + theme_classic() + xlab("Ability") + 
-    geom_vline(data=sex_m, aes(xintercept=ability, color=sex)) +
+    geom_density(alpha=.2) + xlab("Ability") + 
+    geom_vline(data=sex_m2, aes(xintercept=ability, color=sex)) +
     scale_color_discrete(type=sex_colors) + 
@@
 
-ses_m <- ses %>% group_by(SES) %>%
+ses_m2 <- ses %>% group_by(SES) %>%
   filter(age<26, age>22) %>%
   summarise(ability = mean(theta_ses, na.rm=T)) 
-
+# now high ability=.41, low=.08
 p2 <- ses %>%
   ggplot(aes(x=theta_ses, color=SES, fill=SES)) +
-    geom_density(alpha=.2) + theme_classic() + xlab("Ability") + 
-    geom_vline(data=ses_m, aes(xintercept=ability, color=SES)) + 
+    geom_density(alpha=.2) + xlab("Ability") + 
+    geom_vline(data=ses_m2, aes(xintercept=ability, color=SES)) + 
     scale_color_discrete(type=ses_colors) +
@@
 
-eth_m <- eth %>% group_by(`Ethnic Group`) %>%
+eth_m2 <- eth %>% group_by(`Ethnic Group`) %>%
   filter(age<26, age>22) %>%
   summarise(ability = mean(theta_eth, na.rm=T))
-
+# now white=.52, nonwhite=.47 (was .59, .06)
 p3 <- eth %>%
   ggplot(aes(x=theta_eth, color=`Ethnic Group`, fill=`Ethnic Group`)) +
-    geom_density(alpha=.2) + theme_classic() + xlab("Ability") + 
+    geom_density(alpha=.2) + xlab("Ability") + 
     geom_vline(data=eth_m, aes(xintercept=ability, color=`Ethnic Group`)) + 
@@
     scale_fill_discrete(type=eth_colors)
+
+# Note: can extract group parameters:
+# coef(mod_intuitive_eth)$White$GroupPars - but means both 0, just SD varies?
 
@@
 -->
+
+## Identifying biased CDI items
+
+define GLIMMER
+
+summarize findings
+
+```{r get-dif-dfs}
+extract_group_df <- function(group_model, groups=c("Male","Female")) {
+  Mit = as_tibble(coef(group_model, simplify=T)[[groups[1]]]$items) %>%
+    mutate(definition = rownames(coef(group_model, simplify=T)[[groups[1]]]$items),
+           group1 = groups[1],
+           group2 = groups[2]) %>%
+    select(-g, -u) %>% 
+    rename(d_g1 = d)
+  Fit = as_tibble(coef(group_model, simplify=T)[[groups[2]]]$items) %>%
+    mutate(definition = rownames(coef(group_model, simplify=T)[[groups[2]]]$items)) %>%
+    select(-g, -u) %>%
+    rename(d_g2 = d)
+
+  combo <- Mit %>% left_join(Fit) %>%
+    mutate(d_diff = d_g2 - d_g1,
+           d_diff_abs = abs(d_diff)) 
+  return(combo)
+}
+
+get_extreme_item_difficulty_differences <- function(mm, sd_thresh=2) {
+  #yfit <- dnorm(mm$d_diff, mean = mean(mm$d_diff), sd = sd(mm$d_diff)) 
+  mm <- mm %>% mutate(d_diff = d_g2 - d_g1)
+  max_dif = mean(mm$d_diff) + sd_thresh*sd(mm$d_diff)
+  min_dif = mean(mm$d_diff) - sd_thresh*sd(mm$d_diff)
+  mm <- mm %>% mutate(extreme = ifelse((d_diff > max_dif) | (d_diff < min_dif), T, F))
+  print(paste("mininum difference:",round(min_dif,3), "maximum difference:",round(max_dif,3)))
+  return(mm)
+}
+
+mm_sex <- extract_group_df(mod_intuitive_sex, groups=c("Male","Female"))
+mm_ses <- extract_group_df(mod_intuitive_ses, groups=c("high","low"))
+mm_eth <- extract_group_df(mod_intuitive_eth, groups=c("Nonwhite","White"))
+
+mm_sex <- get_extreme_item_difficulty_differences(mm_sex) # min: -.53 max: 1.06
+mm_ses <- get_extreme_item_difficulty_differences(mm_ses) # min: -1.13 max: 1.07
+mm_eth <- get_extreme_item_difficulty_differences(mm_eth) # min: -.64 max: 1.56
+
+View(subset(mm_sex, extreme)) # 25 extreme (outside 2sd) - 7 favoring girls, 18 favoring boys
+# possibly underestimating the female advantage?
+View(subset(mm_ses, extreme)) # 39 extreme - 13 favoring low-SES, 26 favoring high-SES
+View(subset(mm_eth, extreme)) # 31 extreme - 11 favoring nonwhite, 20 favoring white
+# overestimating both high-SES and white advantages?
+
+# median sex difference: 
+# median(mm_sex$d_diff) # 0.28
+
+# median SES difference: 
+# median(mm_ses$d_diff) # 0.009 - very small!
+
+# median race difference:
+# median(mm_eth$d_diff) # 0.40 - large!
+
+# report # of items above and below 0 for each demographic?
+
+all_extreme <- rbind(subset(mm_sex, extreme),
+                     subset(mm_ses, extreme),
+                     subset(mm_eth, extreme)) 
+# length(unique(all_extreme$definition)) # 76 - so 19 items extreme on >1 dimension
+# sort(table(all_extreme$definition)) # candy, choo choo, daddy, duck, ... tractor, vroom
+```
 
@@
 
+We investigated a popular test of children's early vocabulary for potential demographic bias, examining the distribution of words' estimated difficulties for high- vs.
+low-SES children, females vs.
+males, and for white vs.
+non-white children.
+
 # Acknowledgements

### 2022-01-16 paper/cogsci.Rmd
writing

@@
 The idea that some items on a test may show bias, favoring one group over another, is known as Differential Item Function (DIF; @holland1993differential]. 
-On any given test, it is clearly undesirable to have more items favoring one group (say, children from rural households) over another (urban children), as the test will overestimate the ability of test-takers in the former (rural) group -- and not because of any underlying mean difference in ability between the groups, but simply because the test is unfair.
+On any given test, it is clearly undesirable to have more items favoring one group (say, children from rural households) over another (urban children), as the test will overestimate the ability of test-takers in the former (rural) group -- and not because of any underlying mean difference in ability between the groups, but simply because the test is unfair [@camilli2013ongoing].
 A variety of statistical methods for detecting DIF have been proposed, and investigations have in several instances identified DIF for many items on tests favoring particular one group over another (e.g., rural vs.
@@
   filter(age<26, age>22) %>%
-  rename(`Ethnic Group`=eth_group)
+  mutate(Race=eth_group)
 # cor.test(eth$production, eth$theta_eth) # .98
 
-eth_m2 <- eth %>% group_by(`Ethnic Group`) %>%
+eth_m2 <- eth %>% group_by(Race) %>%
   filter(age<26, age>22) %>%
@@
 p3 <- eth %>%
-  ggplot(aes(x=theta_eth, color=`Ethnic Group`, fill=`Ethnic Group`)) +
+  ggplot(aes(x=theta_eth, color=Race, fill=Race)) +
     geom_density(alpha=.2) + xlab("Ability") + 
-    geom_vline(data=eth_m, aes(xintercept=ability, color=`Ethnic Group`)) + 
+    geom_vline(data=eth_m2, aes(xintercept=ability, color=Race)) + 
     scale_color_discrete(type=eth_colors) + 
@@
 
-define GLIMMER
+The first step we take to identify CDI items with DIF is to create GLIMMERs (Graphs of Logits Imputed Multiply with Means Equal; @stenhaug2021treading), which visualize between-item variation in group performance differences.
+
+To give analysts a sense of the uncertainty about the existence of DIF, GLIMMER plots show distributions of parameter differences rather than point estimates. 
+These distributions are generated by drawing 10,000 imputations from the item parameter covariance matrix.
+If there is a cluster of items in the GLIMMER plot, the analyst may conclude that these items 
+If there is no cluster, but rather a continuum of smoothly varying differences, the identification of DIF is fraught:
 

### 2022-01-17 paper/cogsci.Rmd
writing

@@
 
-Inspired by the utility and widespread use of the CDI, researchers have recently been using psychometric models on CDI data to construct short, adaptive tests to reliably assess language ability using only a small subset of the CDI items [e.g., @mayor2019, @kachergis2021cat].
+Inspired by the utility and widespread use of the CDI, researchers have recently been using psychometric models on CDI data to construct short, adaptive tests to reliably assess language ability using only a small subset of the CDI items [e.g., @mayor2019; @kachergis2021cat].
 These psychometric models typically come from the Item-Response Theory (IRT) framework [@Baker2001], which assumes that not only test-takers (here, children) have normally-distributed ability, but that items (words on the CDI) have normally-distributed difficulty.
@@
 Hence, the presence of DIF on a wide variety of items may not indicate the presence of bias; it could indicate that one demographic has a higher average ability level than the other.
+For example, let male language ability be drawn from a standard normal $\mu_{male} ~ N(0,1)$, with female language ability slightly higher, on average ($\mu_{female} = \mu_{male} + 0.1$).
+Then we would expect items to be an average of $0.1$ easier for females than for males, and we might identify items that are instead easier for males than females as showing undesirable DIF.
+And yet, without knowing the actual ability difference between two demographic groups--for which we also rely upon our tests, it is difficult to adjudicate which items show DIF, and which do not.
 
@@
 The outline of this paper is as follows.
-First, we will introduce the Wordbank data and the IRT model we use to analyze the data.
-We then fit the IRT model to each group along each demographic dimension, and examine the parameters for evidence of DIF, noting in particular how many items are significantly biased in favor of each group.
+First, we introduce the Wordbank data and the IRT model, and use it examine the overall size of demographic differences in early word learning.
+We then fit the IRT model to each group along each demographic dimension, and examine the item parameters for evidence of DIF, noting in particular how many items are significantly biased in favor of each group.
 We identify a set of suspect items to eliminate (or in the future, replace), and provide updated estimates of the effect size of these demographic variables, were the biased items to be eliminated.
-Next, using a corpus of child-directed speech, we examine the extent to which variation in word frequencies heard by boys vs. girls predicts sex-based DIF.
+<!--Next, using a corpus of child-directed speech, we examine the extent to which variation in word frequencies heard by boys vs. girls predicts sex-based DIF.-->
 Finally, we provide recommendations for next steps to be taken to identify and replace biased items on the CDI.
@@
 
-get_extreme_item_difficulty_differences <- function(mm, sd_thresh=2) {
+get_extreme_item_difficulty_differences <- function(mm, sd_thresh=2, verbose=F) {
   #yfit <- dnorm(mm$d_diff, mean = mean(mm$d_diff), sd = sd(mm$d_diff)) 
@@
   mm <- mm %>% mutate(extreme = ifelse((d_diff > max_dif) | (d_diff < min_dif), T, F))
-  print(paste("mininum difference:",round(min_dif,3), "maximum difference:",round(max_dif,3)))
+  if(verbose) print(paste("mininum difference:",round(min_dif,3), "maximum difference:",round(max_dif,3)))
   return(mm)
@@
 mm_eth <- get_extreme_item_difficulty_differences(mm_eth) # min: -.64 max: 1.56
-
-View(subset(mm_sex, extreme)) # 25 extreme (outside 2sd) - 7 favoring girls, 18 favoring boys
-# possibly underestimating the female advantage?
-View(subset(mm_ses, extreme)) # 39 extreme - 13 favoring low-SES, 26 favoring high-SES
-View(subset(mm_eth, extreme)) # 31 extreme - 11 favoring nonwhite, 20 favoring white
-# overestimating both high-SES and white advantages?
 
@@
 # sort(table(all_extreme$definition)) # candy, choo choo, daddy, duck, ... tractor, vroom
+```
+
+```{r, eval=F, include=F}
+View(subset(mm_sex, extreme)) # 25 extreme (outside 2sd) - 7 favoring girls, 18 favoring boys
+# possibly underestimating the female advantage?
+View(subset(mm_ses, extreme)) # 39 extreme - 13 favoring low-SES, 26 favoring high-SES
+View(subset(mm_eth, extreme)) # 31 extreme - 11 favoring nonwhite, 20 favoring white
+# overestimating both high-SES and white advantages?
 ```

### 2022-01-17 paper/cogsci.Rmd
writing

@@
 
-# should we report t-tests? 
-sex_diff_test = t.test(subset(d_demo, age>22 & age<26 & sex=="Male")$ability,
-       subset(d_demo, age>22 & age<26 & sex=="Female")$ability) # p<.001
-ses_diff_test = t.test(subset(d_demo, age>22 & age<26 & ses_group=="high")$ability,
-       subset(d_demo, age>22 & age<26 & ses_group=="low")$ability) # p=.1 (larger difference at >ages)
-eth_diff_test = t.test(subset(d_demo, age>22 & age<26 & eth_group=="White")$ability,
-       subset(d_demo, age>22 & age<26 & eth_group=="Nonwhite")$ability) # p=.01
+# regression - report coefficients again after we cut some items?
+require(lmerTest)
+d_demo$eth_group = factor(d_demo$eth_group, levels=c("Nonwhite","White"))
+d_demo$sex = factor(d_demo$sex, levels=c("Male","Female"))
+d_demo$ses_group = factor(d_demo$ses_group, levels=c("low","high"))
+d_demo <- d_demo %>% mutate(age_sc = scale(age, scale=F))
+eth_reg <- summary(lm(ability ~ age_sc * eth_group, data=d_demo))$coefficients # R^2=.48 white-, age*white+
+ses_reg <- summary(lm(ability ~ age_sc * ses_group, data=d_demo))$coefficients # R^2=.56
+sex_reg <- summary(lm(ability ~ age_sc * sex, data=d_demo))$coefficients # R^2=.59
+
+#ab_reg <- summary(lm(ability ~ age_sc * (eth_group + ses_group + sex), data=d_demo)) # R^2=.49
+#prod_reg <- summary(lm(production ~ age_sc * (eth_group + ses_group + sex), data=d_demo)) # R^2=.51
 ```
@@
 
-Figure 1 shows children's language ability vs. age by demographic group from the baseline Rasch model, which assumes no DIF.
-Although the demographic language ability advantages go in the expected directions (i.e., favoring female, high-SES, and white children), note that magnitude of the differences vary somewhat across the age range, especially for SES and ethnicity.
+Figure 1 shows children's language ability vs. age by demographic group from the baseline Rasch model, which assumes no DIF (i.e., equal item parameters for all groups).
+A linear regression for each demographic group, with age (centered) and its interaction, showed significant effects in the expected directions, as can also be seen in the figure. 
+Female children had higher language ability than male children ($\beta=0.56$, $p<.001$), with no significant interaction with age ($\beta=0.02$, $p=.10$.
+High-SES children had higher language ability than low-SES children ($\beta=0.23$, $p=.02$), an advantage that grew with age ($\beta=0.11$, $p<.001$).
+White children had higher language ability than non-white children ($\beta=0.50$, $p<.001$), an advantage that grew with age ($\beta=0.08$, $p<.001$).
+We will re-examine these demographic regressions after trimming items showing extreme DIF, which is what we turn to now.
+
+<!-- Although the demographic language ability advantages go in the expected directions (i.e., favoring female, high-SES, and white children), note that magnitude of the differences vary somewhat across the age range, especially for SES and ethnicity.
 Thus, although the IRT models are fitted to the entire age range, our comparisons of average ability will focus on the middle of the CDI:WS' intended age range, where there is great variation in the size and composition of children's vocabulary.
@@
 The average advantage for white vs. non-white children in this age range is $\theta_{w-nonw} = 0.52$, an average of 45 more produced CDI items.
+-->
 
@@
 
-## Demographic effects in group models
+<!-- put the GLIMMERs side-by-side?
+```{r 2-col-image, fig.env = "figure*", fig.pos = "h", fig.width=4, fig.height=2, fig.align = "center", set.cap.width=T, num.cols.cap=2, fig.cap = "This image spans both columns.
+And the caption text is limited to 0.8 of the width of the document."}
 
-maybe only do this after we've eliminated the most biased items?
+```
+-->
+
+## Identifying biased CDI items
+
+The first step we take to identify CDI items with DIF is to create GLIMMERs (Graphs of Logits Imputed Multiply with Means Equal; @stenhaug2021treading), which visualize between-item variation in group performance differences.
+These parameters are drawn from a fitted multigroup Rasch model for each demographic variable (sex, SES, and race), with the assumption that the mean language ability in each group is the same (e.g.
+for sex, $\mu_{male}=\mu_{female}=0$), thus pushing all between-group variation into the item difficulty parameters.
+For the case of sex, where we believe $\mu_{female}>\mu_{male}$, this means we may expect to find many items with difficulty $b_j^{female} < b_j^{male}$, but we can still examine the distribution of differences in item difficulty ($d_j = b_j^{male} - b_j^{female}$) for outliers.
+To give analysts a sense of the uncertainty about the existence of DIF, GLIMMER plots show distributions of parameter differences rather than point estimates. 
+These distributions are generated by drawing 10,000 imputations from the item parameter covariance matrix.
+
+Figures 2-4 show GLIMMERs for a selection of CDI items for sex, SES, and race, respectively.
+The full GLIMMERs, with all 680 CDI:WS items, are available on OSF, but were too large to include here.
+Nonetheless, it is important to inspect the full plots, for if there is a cluster of items in a GLIMMER, the analyst may conclude that these items are strong candidates for DIF on that dimension.
+For example, there is some clustering at the top and bottom of Figure 2: at the top, "vagina", "tights", "dress (object)", and "doll" form a cluster of items that are much easier for females, while at the bottom, "penis" stands out as much easier for males.
+For the SES and race GLIMMERs (Figures 3 and 4)--and in the rest of the full sex GLIMMER, there are not clusters, but rather a continuum of smoothly varying differences with overlap. 
+This makes identifying items with DIF quite difficult, as different methods are likely to yield inconsistent results [@stenhaug2021treading].
+
+```{r get-dif-dfs}
+extract_group_df <- function(group_model, groups=c("Male","Female")) {
+  Mit = as_tibble(coef(group_model, simplify=T)[[groups[1]]]$items) %>%
+    mutate(definition = rownames(coef(group_model, simplify=T)[[groups[1]]]$items),
+           group1 = groups[1],
+           group2 = groups[2]) %>%
+    select(-g, -u) %>% 
+    rename(d_g1 = d)
+  Fit = as_tibble(coef(group_model, simplify=T)[[groups[2]]]$items) %>%
+    mutate(definition = rownames(coef(group_model, simplify=T)[[groups[2]]]$items)) %>%
+    select(-g, -u) %>%
+    rename(d_g2 = d)
+
+  combo <- Mit %>% left_join(Fit) %>%
+    mutate(d_diff = d_g2 - d_g1,
+           d_diff_abs = abs(d_diff)) 
+  return(combo)
+}
+
+get_extreme_item_difficulty_differences <- function(mm, sd_thresh=2, verbose=F) {
+  #yfit <- dnorm(mm$d_diff, mean = mean(mm$d_diff), sd = sd(mm$d_diff)) 
+  mm <- mm %>% mutate(d_diff = d_g2 - d_g1)
+  max_dif = mean(mm$d_diff) + sd_thresh*sd(mm$d_diff)
+  min_dif = mean(mm$d_diff) - sd_thresh*sd(mm$d_diff)
+  mm <- mm %>% mutate(extreme = ifelse((d_diff > max_dif) | (d_diff < min_dif), T, F))
+  if(verbose) print(paste("mininum difference:",round(min_dif,3), "maximum difference:",round(max_dif,3)))
+  return(mm)
+}
+
+mm_sex <- extract_group_df(mod_intuitive_sex, groups=c("Male","Female"))
+mm_ses <- extract_group_df(mod_intuitive_ses, groups=c("high","low"))
+mm_eth <- extract_group_df(mod_intuitive_eth, groups=c("Nonwhite","White"))
+
+mm_sex <- get_extreme_item_difficulty_differences(mm_sex) # min: -.53 max: 1.06
+mm_ses <- get_extreme_item_difficulty_differences(mm_ses) # min: -1.13 max: 1.07
+mm_eth <- get_extreme_item_difficulty_differences(mm_eth) # min: -.64 max: 1.56
+
+# median sex difference: 
+# median(mm_sex$d_diff) # 0.28
+
+# median SES difference: 
+# median(mm_ses$d_diff) # 0.009 - very small!
+
+# median race difference:
+# median(mm_eth$d_diff) # 0.40 - large!
+
+# report # of items above and below 0 for each demographic?
+
+all_extreme <- rbind(subset(mm_sex, extreme),
+                     subset(mm_ses, extreme),
+                     subset(mm_eth, extreme)) 
+# length(unique(all_extreme$definition)) # 76 - so 19 items extreme on >1 dimension
+# sort(table(all_extreme$definition)) # candy, choo choo, daddy, duck, ... tractor, vroom
+```
+
+```{r sex-glimmer, out.width="\\linewidth", fig.env = "figure", fig.pos = "H", fig.align='center', fig.width=3.2, fig.height=3.5, set.cap.width=T, num.cols.cap=1, fig.cap = "GLIMMER plot of a sample of CDI:WS words from the sex bias model.
+Words at the top are easier to learn for females, while those at the bottom are easier for males."}
+#img <- png::readPNG("figs/lab_logo_stanford.png")
+#grid::grid.raster(img)
+knitr::include_graphics("figs/smGLIMMER_sex_prodWS.pdf")
+```
+
+```{r ses-glimmer, out.width="\\linewidth", fig.env = "figure", fig.pos = "H", fig.align='center', fig.width=3, fig.height=3.5, set.cap.width=T, num.cols.cap=1, fig.cap = "GLIMMER plot of a sample of CDI:WS words from the SES bias model.
+Words at the top are easier to learn for children from low-SES families, while those at the bottom are easier for those from high-SES families."}
+knitr::include_graphics("figs/smGLIMMER_ses_prodWS.pdf")
+```
+
+```{r eth-glimmer, out.width="\\linewidth", fig.env = "figure", fig.pos = "H", fig.align='center', fig.width=3, fig.height=3.5, set.cap.width=T, num.cols.cap=1, fig.cap = "GLIMMER plot of a sample of CDI:WS words from the ethnicity bias model.
+Words at the top are easier to learn for children from low-SES families, while those at the bottom are easier for those from high-SES families."}
+knitr::include_graphics("figs/smGLIMMER_eth_prodWS.pdf")
+```
+
+```{r, eval=F, include=F}
+View(subset(mm_sex, extreme)) # 25 extreme (outside 2sd) - 7 favoring girls, 18 favoring boys
+# possibly underestimating the female advantage?
+View(subset(mm_ses, extreme)) # 39 extreme - 13 favoring low-SES, 26 favoring high-SES
+View(subset(mm_eth, extreme)) # 31 extreme - 11 favoring nonwhite, 20 favoring white
+# overestimating both high-SES and white advantages?
+```
+
+## Demographic effects after trimming
+
+Having identified 76 CDI items that function differently across demographic groups -- including 19 that
 
@@
 
-<!-- put the GLIMMERs side-by-side?
-```{r 2-col-image, fig.env = "figure*", fig.pos = "h", fig.width=4, fig.height=2, fig.align = "center", set.cap.width=T, num.cols.cap=2, fig.cap = "This image spans both columns.
-And the caption text is limited to 0.8 of the width of the document."}
-
-```
--->
-
-## Identifying biased CDI items
-
-The first step we take to identify CDI items with DIF is to create GLIMMERs (Graphs of Logits Imputed Multiply with Means Equal; @stenhaug2021treading), which visualize between-item variation in group performance differences.
-
-To give analysts a sense of the uncertainty about the existence of DIF, GLIMMER plots show distributions of parameter differences rather than point estimates. 
-These distributions are generated by drawing 10,000 imputations from the item parameter covariance matrix.
-If there is a cluster of items in the GLIMMER plot, the analyst may conclude that these items 
-If there is no cluster, but rather a continuum of smoothly varying differences, the identification of DIF is fraught:
-
-summarize findings
-
-```{r get-dif-dfs}
-extract_group_df <- function(group_model, groups=c("Male","Female")) {
-  Mit = as_tibble(coef(group_model, simplify=T)[[groups[1]]]$items) %>%
-    mutate(definition = rownames(coef(group_model, simplify=T)[[groups[1]]]$items),
-           group1 = groups[1],
-           group2 = groups[2]) %>%
-    select(-g, -u) %>% 
-    rename(d_g1 = d)
-  Fit = as_tibble(coef(group_model, simplify=T)[[groups[2]]]$items) %>%
-    mutate(definition = rownames(coef(group_model, simplify=T)[[groups[2]]]$items)) %>%
-    select(-g, -u) %>%
-    rename(d_g2 = d)
-
-  combo <- Mit %>% left_join(Fit) %>%
-    mutate(d_diff = d_g2 - d_g1,
-           d_diff_abs = abs(d_diff)) 
-  return(combo)
-}
-
-get_extreme_item_difficulty_differences <- function(mm, sd_thresh=2, verbose=F) {
-  #yfit <- dnorm(mm$d_diff, mean = mean(mm$d_diff), sd = sd(mm$d_diff)) 
-  mm <- mm %>% mutate(d_diff = d_g2 - d_g1)
-  max_dif = mean(mm$d_diff) + sd_thresh*sd(mm$d_diff)
-  min_dif = mean(mm$d_diff) - sd_thresh*sd(mm$d_diff)
-  mm <- mm %>% mutate(extreme = ifelse((d_diff > max_dif) | (d_diff < min_dif), T, F))
-  if(verbose) print(paste("mininum difference:",round(min_dif,3), "maximum difference:",round(max_dif,3)))
-  return(mm)
-}
-
-mm_sex <- extract_group_df(mod_intuitive_sex, groups=c("Male","Female"))
-mm_ses <- extract_group_df(mod_intuitive_ses, groups=c("high","low"))
-mm_eth <- extract_group_df(mod_intuitive_eth, groups=c("Nonwhite","White"))
-
-mm_sex <- get_extreme_item_difficulty_differences(mm_sex) # min: -.53 max: 1.06
-mm_ses <- get_extreme_item_difficulty_differences(mm_ses) # min: -1.13 max: 1.07
-mm_eth <- get_extreme_item_difficulty_differences(mm_eth) # min: -.64 max: 1.56
-
-# median sex difference: 
-# median(mm_sex$d_diff) # 0.28
-
-# median SES difference: 
-# median(mm_ses$d_diff) # 0.009 - very small!
-
-# median race difference:
-# median(mm_eth$d_diff) # 0.40 - large!
-
-# report # of items above and below 0 for each demographic?
-
-all_extreme <- rbind(subset(mm_sex, extreme),
-                     subset(mm_ses, extreme),
-                     subset(mm_eth, extreme)) 
-# length(unique(all_extreme$definition)) # 76 - so 19 items extreme on >1 dimension
-# sort(table(all_extreme$definition)) # candy, choo choo, daddy, duck, ... tractor, vroom
-```
-
-```{r, eval=F, include=F}
-View(subset(mm_sex, extreme)) # 25 extreme (outside 2sd) - 7 favoring girls, 18 favoring boys
-# possibly underestimating the female advantage?
-View(subset(mm_ses, extreme)) # 39 extreme - 13 favoring low-SES, 26 favoring high-SES
-View(subset(mm_eth, extreme)) # 31 extreme - 11 favoring nonwhite, 20 favoring white
-# overestimating both high-SES and white advantages?
-```
-
-```{r sex-glimmer, out.width="\\linewidth", fig.env = "figure", fig.pos = "H", fig.align='center', fig.width=3.2, fig.height=3.5, set.cap.width=T, num.cols.cap=1, fig.cap = "GLIMMER plot of a sample of CDI:WS words from the sex bias model.
-Words at the top are easier to learn for females, while those at the bottom are easier for males."}
-#img <- png::readPNG("figs/lab_logo_stanford.png")
-#grid::grid.raster(img)
-knitr::include_graphics("figs/smGLIMMER_sex_prodWS.pdf")
-```
-
-```{r ses-glimmer, out.width="\\linewidth", fig.env = "figure", fig.pos = "H", fig.align='center', fig.width=3, fig.height=3.5, set.cap.width=T, num.cols.cap=1, fig.cap = "GLIMMER plot of a sample of CDI:WS words from the SES bias model.
-Words at the top are easier to learn for children from low-SES families, while those at the bottom are easier for those from high-SES families."}
-knitr::include_graphics("figs/smGLIMMER_ses_prodWS.pdf")
-```
-
-```{r eth-glimmer, out.width="\\linewidth", fig.env = "figure", fig.pos = "H", fig.align='center', fig.width=3, fig.height=3.5, set.cap.width=T, num.cols.cap=1, fig.cap = "GLIMMER plot of a sample of CDI:WS words from the ethnicity bias model.
-Words at the top are easier to learn for children from low-SES families, while those at the bottom are easier for those from high-SES families."}
-knitr::include_graphics("figs/smGLIMMER_eth_prodWS.pdf")
-```
-
 # Discussion

### 2022-01-17 paper/cogsci.Rmd
writing

@@
 For example, in the sex-based multigroup model, item $i$'s difficulty is $b_{i}^{female}$ for females, and $b_{i}^{male}$ for males.
-We fitted a multigroup Rasch model for each demographic dimension (sex, socioeconomic status, and ethnicity).
-To contextualize these results, we also first fitted a baseline Rasch model
+In the quest to identify DIF, a multigroup Rasch model will be fitted for each demographic dimension of interest (sex, socioeconomic status, and ethnicity).
+To contextualize these results, we first fitted a baseline Rasch model to the entire dataset, without demographic information.
 
@@
 A linear regression for each demographic group, with age (centered) and its interaction, showed significant effects in the expected directions, as can also be seen in the figure. 
-Female children had higher language ability than male children ($\beta=0.56$, $p<.001$), with no significant interaction with age ($\beta=0.02$, $p=.10$.
+Female children had higher language ability than male children ($\beta=0.56$, $p<.001$), with no significant interaction with age ($\beta=0.02$, $p=.10$).
 High-SES children had higher language ability than low-SES children ($\beta=0.23$, $p=.02$), an advantage that grew with age ($\beta=0.11$, $p<.001$).
@@
 This makes identifying items with DIF quite difficult, as different methods are likely to yield inconsistent results [@stenhaug2021treading].
+Nonetheless, we describe the distributions of item-level group difficulty differences, and examine overlap of the extrema.
 
@@
 mm_sex <- extract_group_df(mod_intuitive_sex, groups=c("Male","Female"))
-mm_ses <- extract_group_df(mod_intuitive_ses, groups=c("high","low"))
+mm_ses <- extract_group_df(mod_intuitive_ses, groups=c("low","high"))
 mm_eth <- extract_group_df(mod_intuitive_eth, groups=c("Nonwhite","White"))
@@
 mm_sex <- get_extreme_item_difficulty_differences(mm_sex) # min: -.53 max: 1.06
-mm_ses <- get_extreme_item_difficulty_differences(mm_ses) # min: -1.13 max: 1.07
+mm_ses <- get_extreme_item_difficulty_differences(mm_ses) # min: -1.07 max: 1.13
 mm_eth <- get_extreme_item_difficulty_differences(mm_eth) # min: -.64 max: 1.56
-
-# median sex difference: 
-# median(mm_sex$d_diff) # 0.28
-
-# median SES difference: 
-# median(mm_ses$d_diff) # 0.009 - very small!
-
-# median race difference:
-# median(mm_eth$d_diff) # 0.40 - large!
-
-# report # of items above and below 0 for each demographic?
 
@@
                      subset(mm_eth, extreme)) 
-# length(unique(all_extreme$definition)) # 76 - so 19 items extreme on >1 dimension
+# length(unique(all_extreme$definition)) 
 # sort(table(all_extreme$definition)) # candy, choo choo, daddy, duck, ... tractor, vroom
+
+# overlap between extreme SES and race items:
+#intersect(subset(mm_ses, extreme)$definition, subset(mm_eth, extreme)$definition)
+```
+
+For sex, the median difficulty difference (male-female) was `r round(median(mm_sex$d_diff),2)` (M=`r round(mean(mm_sex$d_diff),2)`, sd=`r round(sd(mm_sex$d_diff),2)`), with `r length(which(mm_sex$d_diff>0))`/680 items being easier for females than males.
+The fact that the bulk of this distribution favors females seems to confirm that the female language advantage is pervasive across the CDI:WS, and is thus likely to be a real advantage.
+For SES, the median difficulty difference (low-high) was `r round(median(mm_ses$d_diff),2)` (M=`r round(mean(mm_ses$d_diff),2)`, sd=`r round(sd(mm_ses$d_diff),2)`), with `r length(which(mm_ses$d_diff>0))`/680 items being easier for high-SES than low-SES children.
+With the mean and median difficulty differences close to 0, and roughly half of the words favoring each SES group, it is tempting to conclude that the CDI:WS items are somewhat balanced with respect to SES.
+For race, the median difficulty difference (nonwhite-white) was `r round(median(mm_eth$d_diff),2)` (M=`r round(mean(mm_eth$d_diff),2)`, sd=`r round(sd(mm_eth$d_diff),2)`), with `r length(which(mm_eth$d_diff>0))`/680 items being easier for white than non-white children, revealing a fairly pervasive advantage for white children on CDI items.
+
+Finally, for each distribution we looked at extreme values, flagging items with difficulty differences that were outside 2 standard deviations.
+This identified 25 items with extreme sex-based difficulty differences, only 7 of which were easier for females, with the other 18 items favoring males.
+<!-- thus, removing these extrema would actually lower male children's vocabulary scores..] -->
+For SES, 39 extrema were identified, only 13 of which were easier for low-SES children.
+For race, 31 extrema were identified, only 11 of which were easier for non-white children.
+Thus, while if anything the bulk of the extrema in the sex-based model advantaged the lower language ability group (males), the extrema in the other two models mostly favored the advantaged groups: 26/39 extreme items in the SES model favored high-SES children, and 20/31 items in the race model favored white children.
+17 CDI items were identified as extrema in more than one model: "tractor" and "vroom" were extreme in all three models, "choo choo" was extreme in both SES and sex models, "give me five!" was extreme in both race and sex models, and 13 other items were extreme in both SES and race models ("grrr", "moo", "quack quack", "uh oh", "duck", "owl", "candy", "gum", "walker", "daddy", "pet's name", "up", and "so").
+In our final analysis, we explore the effect of pruning the 15 items that were extreme in both SES and race models.
+
+```{r items-to-remove}
+prune_defs <- c("tractor", "vroom", "grrr", "moo", "quack quack", "uh oh", "duck", "owl", "candy", "gum", "walker", "daddy*", "pet's name", "up", "so") # also "choo choo" and "give me five!" ?
+prune_cols <- which(!is.element(names(d_mat), prune_defs))
+#names(d_mat)[which(is.element(names(d_mat), prune_defs))]
+d_mat_pruned <- d_mat[,prune_cols]
+```
+
+```{r fit-pruned-model, eval=F}
+m0 <- mirt(d_mat_pruned %>% filter(rowSums(d_mat_pruned)!=0) %>% select(-data_id), 1, "Rasch", 
+           technical=list(NCYCLES=1000))
+
+coefs_1pl <- as_tibble(coef(m0, simplify = TRUE)$items) %>%
+  mutate(definition = rownames(coef(m0, simplify = TRUE)$items))
+fscores_1pl <- tibble(data_id = d_mat_pruned %>% filter(rowSums(d_mat_pruned)!=0) %>% select(data_id), 
+                             ability = fscores(m0, method = "MAP")[,1])
+
+save(file = here("data/eng_Ws_prod_mod_1pl_pruned.Rds"), "m0", "coefs_1pl", "fscores_1pl")
 ```
@@
 
-## Demographic effects after trimming
+## Demographic effects after pruning
 
-Having identified 76 CDI items that function differently across demographic groups -- including 19 that
+```{r load-pruned}
+load(here("data/eng_Ws_prod_mod_1pl_pruned.Rds"))
+
+# join with d_demo, do the demo regressions again..
+```
+
+We identified 76 CDI items that function quite differently across demographic groups, and chose to prune the 15 items with large difficulty differences for both SES and race.
 

### 2022-01-17 paper/cogsci.Rmd
wrapping up results

@@
 # regression - report coefficients again after we cut some items?
-require(lmerTest)
 d_demo$eth_group = factor(d_demo$eth_group, levels=c("Nonwhite","White"))
@@
 fscores_1pl <- tibble(data_id = d_mat_pruned %>% filter(rowSums(d_mat_pruned)!=0) %>% select(data_id), 
-                             ability = fscores(m0, method = "MAP")[,1])
-
+                             ability_pr = fscores(m0, method = "MAP")[,1])
+fscores_1pl$data_id = fscores_1pl$data_id$data_id
 save(file = here("data/eng_Ws_prod_mod_1pl_pruned.Rds"), "m0", "coefs_1pl", "fscores_1pl")
@@
 
+d_demo <- d_demo %>% left_join(fscores_1pl)
 # join with d_demo, do the demo regressions again..
+
+eth_reg_pr <- summary(lm(ability_pr ~ age_sc * eth_group, data=d_demo))$coefficients # R^2=.48 
+ses_reg_pr <- summary(lm(ability_pr ~ age_sc * ses_group, data=d_demo))$coefficients # R^2=.56
+sex_reg_pr <- summary(lm(ability_pr ~ age_sc * sex, data=d_demo))$coefficients # R^2=.59
+
+#white=.48
+#highSES=0.20
+#female=0.57
 ```
@@
 We identified 76 CDI items that function quite differently across demographic groups, and chose to prune the 15 items with large difficulty differences for both SES and race. 
+After pruning these 2.2% of CDI:WS items, regressions with each demographic variable (and age, centered) showed smaller effects of SES ($\beta=0.20$ vs.
+$\beta=0.23$) and race ($\beta=0.48$ vs.
+$\beta=0.50$), while the effect of sex remained the same.
 
-```{r demo-ability, out.width="\\linewidth", fig.env = "figure", fig.pos = "H", fig.align='center', fig.width=3.3, fig.height=5.5, set.cap.width=T, num.cols.cap=1, fig.cap = "Density plots of language ability for children aged 23-25 months, based on the group models."}
-# show density plots of child ability based on group models
+```{r demo-ability, out.width="\\linewidth", fig.env = "figure", fig.pos = "H", fig.align='center', fig.width=3.3, fig.height=5.5, set.cap.width=T, num.cols.cap=1, fig.cap = "Density plots of language ability for children aged 23-25 months using the pruned model."}
 
@@
 sex <- d_demo %>% filter(!is.na(sex), production!=0) %>%
-  mutate(theta_sex = fscores(mod_intuitive_sex)) 
+  mutate(theta_sex = ability_pr)
+  #mutate(theta_sex = fscores(mod_intuitive_sex)) 
 # cor.test(eth$production, eth$theta_eth) # .98
@@
 ses <- d_demo %>% filter(!is.na(ses_group), production!=0) %>%
-  mutate(theta_ses = fscores(mod_intuitive_ses)) %>%
+  #mutate(theta_ses = fscores(mod_intuitive_ses)) %>%
+  mutate(theta_ses = ability_pr) %>%
   filter(age<26, age>22) %>%
@@
 eth <- d_demo %>% filter(!is.na(eth_group), production!=0) %>%
-  mutate(theta_eth = fscores(mod_intuitive_eth)) %>%
+  #mutate(theta_eth = fscores(mod_intuitive_eth)) %>%
+  mutate(theta_eth = ability_pr) %>%
   filter(age<26, age>22) %>%

### 2022-01-17 paper/cogsci.Rmd
editing intro

@@
 For example, @bornstein2016stability found stability in core language skills across 10 years of children's development, despite changes in maternal income and education over the study period.
-Yet socioeconomic status (SES) has also been found to be predictive of children's early language skill, and of later educational outcomes [for a review, see @schwab2016].
+Yet socioeconomic status (SES) has also been found to be associated with children's language processing and vocabulary by 18 months [@fernald2013ses], and to be predictive of later educational outcomes [@marchman2008speed; see @schwab2016 for a review].
 Demographic factors are also predictive of language skill: first-born children tend to outpace their siblings, and female children tend to have better language skills than their age-matched male counterparts [@fenson1994] -- a sex-based verbal advantage that continues through high school [see @petersen2018gender for a review].
-However, it is also difficult to measure language skill: in any short recording children are unlikely to use all of the words and constructions that they know, and any comprehensive battery of language measures will likely exhaust children's attention span.
+However, it is also difficult to get a complete measure of young children's language skill: long recordings are prohibitively difficult to collect and transcribe [see @roy2015predicting, the exception that proves the rule], and yet any short recording (e.g., a 1-hour play session) will elicit only a small proportion of the words and constructions that children know.
+Thus, researchers of early word learning have constructed tests with hundreds of words, with an intentional bias to select words that are more likely to be known by children.
 
-The MacArthur-Bates Communicative Development Inventories [CDIs; @Fenson2007] are a set of parent-reported measures of children's productive and receptive language skills, which produce a low-cost and reliable way to estimate children's early language skills [@fenson1994].
+We focus on one set of popular instruments, the MacArthur-Bates Communicative Development Inventories [CDIs; @Fenson2007] are a set of parent-reported measures of children's productive and receptive language skills, which produce a low-cost and reliable way to estimate children's early language skills [@fenson1994].
 CDIs have shown good predictive validity [e.g., @fenson1994; Bornstein et al. 2012, Duff et al. 2015].
@@
 White children had higher language ability than non-white children ($\beta=0.50$, $p<.001$), an advantage that grew with age ($\beta=0.08$, $p<.001$).
-We will re-examine these demographic regressions after trimming items showing extreme DIF, which is what we turn to now.
+We will re-examine these demographic regressions after trimming items showing extreme DIF.
 
@@
 Figures 2-4 show GLIMMERs for a selection of CDI items for sex, SES, and race, respectively.
-The full GLIMMERs, with all 680 CDI:WS items, are available on OSF, but were too large to include here.
+The full GLIMMERs, with all 680 CDI:WS items, are available on [OSF](https://osf.io/57rsw/?view_only=2b6ecb61fe08458293af7421d276932a), but were too large to include here.
 Nonetheless, it is important to inspect the full plots, for if there is a cluster of items in a GLIMMER, the analyst may conclude that these items are strong candidates for DIF on that dimension.

### 2022-01-18 paper/cogsci.Rmd
draft discussion

@@
 
-We identified 76 CDI items that function quite differently across demographic groups, and chose to prune the 15 items with large difficulty differences for both SES and race. 
-After pruning these 2.2% of CDI:WS items, regressions with each demographic variable (and age, centered) showed smaller effects of SES ($\beta=0.20$ vs.
+We identified 76 CDI items that function quite differently across demographic groups, and chose to prune just the 15 items with large difficulty differences for both SES and race. 
+After pruning just these 15 items from the CDI:WS and re-fitting a Rasch IRT model, regressions predicting ability with each demographic variable (and age, centered) showed reduced main effects of SES ($\beta=0.20$ vs.
 $\beta=0.23$) and race ($\beta=0.48$ vs.
-$\beta=0.50$), while the effect of sex remained the same.
+$\beta=0.50$).
+Although these are small reductions in the SES and race effect sizes, we find them striking as we pruned only 2.2% of the CDI:WS items (and only 20% of the 76 extrema we identified), and did not specifically prune items that were biased against the disadvantaged groups (non-white, low-SES).
 
-```{r demo-ability, out.width="\\linewidth", fig.env = "figure", fig.pos = "H", fig.align='center', fig.width=3.3, fig.height=5.5, set.cap.width=T, num.cols.cap=1, fig.cap = "Density plots of language ability for children aged 23-25 months using the pruned model."}
-
+```{r demo-ability, include=F, out.width="\\linewidth", fig.env = "figure", fig.pos = "H", fig.align='center', fig.width=3.3, fig.height=5.5, set.cap.width=T, num.cols.cap=1, fig.cap = "Density plots of language ability for children aged 23-25 months using the pruned model."}
 # ToDo: maybe make these violin plots and squeeze all 6 groups side-by-side?
@@
 
-# Note: can extract group parameters:
-# coef(mod_intuitive_eth)$White$GroupPars - but means both 0, just SD varies?
-
 ggarrange(p1, p2, p3, ncol=1)
@@
 
-We investigated a popular test of children's early vocabulary for potential demographic bias, examining the distribution of words' estimated difficulties for high- vs.
+We investigated the CDI:WS, a popular parent-report measure of children's early vocabulary, for potential demographic bias, examining the distribution of words' estimated difficulties for high- vs.
 low-SES children, females vs.
@@
 non-white children.
+The IRT-based analysis revealed differential item functioning (DIF) for many items along each demographic dimension, but only in the case of sex were clear clusters of items that were more well-known to females (including feminine clothing and genitalia), and a clear item that was more well-known to males (male genitalia).
+For the rest of the items, and for SES- and race-based analysis, there was a smooth continuum of DIF, making the boundary of true DIF subjective, as this would rely on knowing the true difference in language ability between groups--which we reciprocally estimate from instruments like the CDI.
+To move forward, we identified candidate DIF items by looking at the extremes of each distribution, and found that for SES and race, the majority of these extrema (66% in both cases) were easier for the majority demographic group (i.e., high-SES, white), meaning that if these items were removed, SES and race effects on language ability as measured by the CDI:WS would decrease.
+In contrast, the majority of the extrema in the sex-based DIF analysis were easier for the disadvantaged group (males): if these extrema are removed, the female language advantage on the CDI:WS will likely increase.
+To confirm these implications, we pruned a very small number of CDI items (15 of 680) that were DIF extrema in both SES- and race-based analyses, and found that the effect of these dimensions on language ability was reduced.
+
+This investigation is only a first step in measuring demographic bias in the items on the CDI:WS.
+In aggregate, it seems to suggest that we may be slightly overestimating the magnitude of differences in language ability along the dimensions of SES and race, and perhaps underestimating the female advantage.
+However, the best way to estimate endogenous differences in ability (rather than exogenous differences, e.g.
+in the language environment) would be to conduct controlled, in-lab experiments measuring children's ability to learn novel words.
+Future research may also look into the specific items that are easier for one group than another, and determine whether these differences can be accounted for by different environmental contexts. 
+For example, it is striking that many of the extrema favoring high-SES children are animals and animal sounds (e.g., "grrr", "quack quack", "woof woof", "baa baa", "duck", "sheep", "giraffe", "zebra"): do high-SES households visit the zoo more, or do they often engage in other activities related to naming animals and noises they make?
+If certain high-SES activities are driving early word learning for these children, what are the activities (and associated vocabulary) that low-SES households are instead engaging in?
+A fair test of children's early vocabulary would contain a representative sample of words from all activities that children engage in, across demographic groups.
 
@@
 
-Place acknowledgments (including funding information) in a section at
-the end of the paper.
+[Redacted for anonymous review.]
+<!--We thank members of the Language and Cognition lab for their feedback.-->
 

### 2022-01-19 GLIMMER_production.R
AOAA-OAT

@@ -217,3 +217,56 @@ eth_histn <- item_difficulty_difference_histogram(mm_eth, withNormal = T)
 
 ggarrange(sex_histn, ses_histn, eth_histn, nrow=1)
 ggsave(file="item_DIF_hists_withNormals.pdf", width=8, height=3.5)
+
+
+
+
+# constrained multiGroup model, allowing only variance between groups to vary
+constrained_sex <- multipleGroup(d_mat, 1, itemtype = "Rasch", 
+                                 sex_group, invariance = c("intercepts","free_var"), SE = TRUE, verbose = T)
+
+constrained_ses <- multipleGroup(d_mat, 1, itemtype = "Rasch", 
+                                 ses_group, invariance = c("intercepts","free_var"), SE = TRUE, verbose = T)
+
+constrained_eth <- multipleGroup(d_mat, 1, itemtype = "Rasch", 
+                                 eth_group, invariance = c("intercepts","free_var"), SE = TRUE, verbose = T)
+
+
+# AOAA-OAT: all others as anchors - one at a time
+# remove the item showing most DIF (based on chisq LR test) from the anchor set, 
+# re-fit, iterate until no new items display DIF
+AOAA_OAT <- function(d_mat, group, fname=c()) {
+  mirtCluster()
+  constrained <- multipleGroup(d_mat, 1, itemtype = "Rasch", 
+                               group, invariance = c("intercepts","free_var"), SE = TRUE, verbose = T)
+  items_removed <- tibble()
+  difPresent = TRUE
+  while(difPresent) {
+    # figure out which item shows the most DIF (by LRT--chisq)
+    difm <- DIF(constrained, which.par=c('d'), scheme='drop') # ~20 minutes per item..
+    most_dif_indx = which(difm$X2==max(difm$X2))
+    if(difm[most_dif_indx,]$p<.01) {
+      items_removed = bind_rows(items_removed, 
+                                bind_cols(word=row.names(difm)[most_dif_indx], difm[most_dif_indx,]))
+      col_to_remove = which(colnames(d_mat)==row.names(difm)[most_dif_indx])
+      print(paste("removing",row.names(difm)[most_dif_indx]))
+      # remove that item, fit the constrained model again
+      constrained <- multipleGroup(d_mat[,-col_to_remove], 1, itemtype = "Rasch", 
[…]

### 2022-01-19 GLIMMER_production.R
AOAA-OAT saving changes, possible paper figs

@@ -253,7 +253,7 @@ AOAA_OAT <- function(d_mat, group, fname=c()) {
       # remove that item, fit the constrained model again
       constrained <- multipleGroup(d_mat[,-col_to_remove], 1, itemtype = "Rasch", 
                                        group, invariance = c("intercepts","free_var"), SE = TRUE, verbose = T)
-      if(length(fname)>0) save(constrained, items_removed, file=paste0("AOAA-OAT_",fname,".Rdata"))
+      if(length(fname)>0) save(constrained, items_removed, difm, file=paste0("AOAA-OAT_",fname,".Rdata"))
     } else {
       difPresent = FALSE
     }
@@ -261,12 +261,12 @@ AOAA_OAT <- function(d_mat, group, fname=c()) {
   return(list(model=constrained, items_removed=items_removed, dif_stats=difm))
 }
 
-AOAA_OAT(d_mat, ses_group, "ses")
-AOAA_OAT(d_mat, sex_group, "sex")
-AOAA_OAT(d_mat, eth_group, "race")
+ses_aoaa <- AOAA_OAT(d_mat, ses_group, "ses")
+sex_aoaa <- AOAA_OAT(d_mat, sex_group, "sex")
+race_aoaa <- AOAA_OAT(d_mat, eth_group, "race")
 
 # drop uses a constrained model (e.g. same group means) and then tests whether dropping
-dif_itemN = DIF(mod_intuitive_sex, which.par=c('d'), scheme='drop', items2test=1:2)
+#dif_itemN = DIF(mod_intuitive_sex, which.par=c('d'), scheme='drop', items2test=1:2)
 
-dif_itemN = DIF(mod_intuitive_sex, 'd', scheme='add', items2test = 1)
+#dif_itemN = DIF(mod_intuitive_sex, 'd', scheme='add', items2test = 1)
 # Error in constrain[[i]] :  attempt to select less than one element in integerOneIndex

### 2022-01-19 paper/cogsci.Rmd
minor edits from Rondeline

@@
 
-We focus on one set of popular instruments, the MacArthur-Bates Communicative Development Inventories [CDIs; @Fenson2007] are a set of parent-reported measures of children's productive and receptive language skills, which produce a low-cost and reliable way to estimate children's early language skills [@fenson1994].
+We focus on one set of popular instruments, the MacArthur-Bates Communicative Development Inventories [CDIs; @Fenson2007]. 
+The CDIs are a set of parent-reported measures of children's productive and receptive language skills, which produce a low-cost and reliable way to estimate children's early language skills [@fenson1994].
 CDIs have shown good predictive validity [e.g., @fenson1994; Bornstein et al. 2012, Duff et al. 2015].
@@
 On any given test, it is clearly undesirable to have more items favoring one group (say, children from rural households) over another (urban children), as the test will overestimate the ability of test-takers in the former (rural) group -- and not because of any underlying mean difference in ability between the groups, but simply because the test is unfair [@camilli2013ongoing].
-A variety of statistical methods for detecting DIF have been proposed, and investigations have in several instances identified DIF for many items on tests favoring particular one group over another (e.g., rural vs.
+A variety of statistical methods for detecting DIF have been proposed, and investigations have in several instances identified DIF for many items on tests favoring one group over another (e.g., rural vs.
 urban).

### 2022-01-19 paper/cogsci.Rmd
adding CHILDES m vs. f section

@@
 
+## Relating child-directed speech to demographic bias
+
+Demographic differences in language ability are likely to be at least partially explained by differences in linguistic input received by children in different groups.
+Indeed, input quantity (total daily tokens) and some measures of quality (e.g., lexical diversity: ratio of word types vs.
+tokens) have been predictive of language learning outcomes in some demographic studies [@rowe2009differences; @huttenlocher1991early].
+Here, we investigated the extent to which word frequency in child-directed speech to male vs. female children was predictive of the amount of DIF shown by CDI items. 
+Similar to the approach taken by Braginsky, Meylan, and Frank (2016), we used the CHILDES corpus of transcripts from dyadic play sessions [@macwhinney2000childes], which are labeled with the sex of the target child, but not other demographic variables.
+(Number of males, number of females, total word counts, a couple examples...)
+
+Overall, the correlation between the proportion of times a word was spoken to a female child and the size of the female (vs.
+male) advantage for that CDI word was modest, but significant ($r=0.19$, $p<.001$).
+
+```{r}
+type_freqs = readRDS("../data/type_freqs.rds")
+type_counts = readRDS("../data/type_counts.rds")
+type_counts_smoothed = readRDS("../data/type_counts_smoothed.rds")
+
+type_counts_smoothed <- type_counts_smoothed %>%
+  pivot_wider(id_cols=uni_lemma, names_from=sex, values_from=count_per_mil) %>%
+  mutate(propFemale = female / (female + male),
+         FtoM = female / male)
+
+# how biased are the extreme items we suggest removing in the corpus counts?
+#sex_items_to_remove$definition
+
+tmp <- left_join(mm_sex, type_counts_smoothed, by=c("definition"="uni_lemma"))
+
+tmp %>% #filter(extreme==T) %>%
+  ggplot(aes(x=propFemale, y=d_diff, color=extreme)) + 
+  geom_point(alpha=.5) + theme_bw() +
+  ylab("Female - Male Word Difficulty") +
+  xlab("Proportion of Utterances to Females in CHILDES")
+
+#subset(tmp, extreme==T & propFemale > .47 & propFemale < .53) # hose & dump
+
+#cor.test(tmp$propFemale, tmp$d_diff) # .19, p<.001
+
+```
+
 # Discussion

### 2022-01-19 paper/cogsci.Rmd
new refs

@@
 Here, we investigated the extent to which word frequency in child-directed speech to male vs. female children was predictive of the amount of DIF shown by CDI items. 
-Similar to the approach taken by Braginsky, Meylan, and Frank (2016), we used the CHILDES corpus of transcripts from dyadic play sessions [@macwhinney2000childes], which are labeled with the sex of the target child, but not other demographic variables.
+Similar to the approach taken by @braginsky2016gender, we used the CHILDES corpus of transcripts from dyadic play sessions [@macwhinney2000childes], which are labeled with the sex of the target child, but not other demographic variables.
 (Number of males, number of females, total word counts, a couple examples...)

### 2022-01-21 paper/cogsci.Rmd
demographic betas x pruning threshold

@@
     CDIs to identify words that show strong bias for particular groups of children, on dimensions
-    of sex (male vs. female), race (white vs. non-white), and socioeconomic status 
+    of sex (male vs. female), race (white vs. non-white), and maternal education 
     (high vs. low). For each dimension, we identify dozens of strongly biased items, and show that 
@@
 Children's early vocabulary growth is assumed to reflect not only their exposure to child-directed speech, but also the varying difficulty of different types of words, and individual differences in the aptitude of the child -- including potential language deficits.
-Research has uncovered both within-child consistency in development, as well as significant influence from external factors.
+Children show both consistency in some skills across development, as well as significant influence from external factors.
 For example, @bornstein2016stability found stability in core language skills across 10 years of children's development, despite changes in maternal income and education over the study period.
-Yet socioeconomic status (SES) has also been found to be associated with children's language processing and vocabulary by 18 months [@fernald2013ses], and to be predictive of later educational outcomes [@marchman2008speed; see @schwab2016 for a review].
+Yet maternal education, often used as a proxy for socioeconomic status (SES), has also been found to be associated with children's language processing and vocabulary by 18 months [@fernald2013ses], and to be predictive of later educational outcomes [@marchman2008speed; see @schwab2016 for a review].
 Demographic factors are also predictive of language skill: first-born children tend to outpace their siblings, and female children tend to have better language skills than their age-matched male counterparts [@fenson1994] -- a sex-based verbal advantage that continues through high school [see @petersen2018gender for a review].
+
 However, it is also difficult to get a complete measure of young children's language skill: long recordings are prohibitively difficult to collect and transcribe [see @roy2015predicting, the exception that proves the rule], and yet any short recording (e.g., a 1-hour play session) will elicit only a small proportion of the words and constructions that children know.
@@
 
-While it is quite reasonable to expect that some CDI words are easier than others, and even to use these varying difficulties to predict variation in children's language ability, the use of psychometric models highlights the possibility that some CDI items may function differently (i.e., be more/less difficult) for different groups of children.
+While it is quite reasonable to expect that some CDI words are "easier" (i.e., more likely to be known by children) than others, and even to use these varying difficulties to predict variation in children's language ability, the use of psychometric models highlights the possibility that some CDI items may function differently (i.e., be more/less widely known) for different groups of children.
 The idea that some items on a test may show bias, favoring one group over another, is known as Differential Item Function (DIF; @holland1993differential]. 
@@
 urban).
-Our goal here is to test the items on the American CDI:WS for DIF along three main axes: sex (male vs. female), socioeconomic status (low- vs. high-SES), and race (white vs. non-white).
-
-However, DIF is fundamentally difficult to identify for multiple reasons [for an overview, see @stenhaug2021treading].
-First, most techniques to identify DIF rely on defining a set of "anchor" test items that are assumed to be equally difficult (i.e., unbiased) for both groups of interest.
-Identifying anchor items is at best fraught when there may in fact be a difference in ability between groups (e.g., the female language advantage), and is further confounded when the magnitude of this ability difference is unknown.
-A reason that DIF is particularly tricky in measuring children's early language ability is that there is a finite universe of early-learned words to choose from--and we may expect many of them to be biased for various environmental reasons (e.g., children in Florida may not use mittens or skis).
-Hence, the presence of DIF on a wide variety of items may not indicate the presence of bias; it could indicate that one demographic has a higher average ability level than the other.
-For example, let male language ability be drawn from a standard normal $\mu_{male} ~ N(0,1)$, with female language ability slightly higher, on average ($\mu_{female} = \mu_{male} + 0.1$).
-Then we would expect items to be an average of $0.1$ easier for females than for males, and we might identify items that are instead easier for males than females as showing undesirable DIF.
-And yet, without knowing the actual ability difference between two demographic groups--for which we also rely upon our tests, it is difficult to adjudicate which items show DIF, and which do not.
-
-<!--
-This problem can be be understood through a concrete example that is well-known from research on early language learning: consider the fact that females show a larger vocabulary than males across early development [see @frank2021 Ch.
-6?; OTHER REFS].
-The question is whether or not females actually have a higher ability level, or if the tests predominantly have words that are easier for females (i.e., biased).
-(It could even be that the test is biased toward males but the language ability of females is large enough to overcome that bias.)
-To address this question, we need to know the ability levels of girls and boys in order to confirm that if a word is learned earlier by girls it is simply because of an ability difference rather than a bias inherent to the word.
-The problem arises from the fact that we measure ability level using the very same test that we are trying to check for biased words.
-If the boys and girls were of the same average language ability, this would not be an issue but given the evidence showing that girls have a higher ability level, we need to know the difference in ability so that when we find DIF for a specific word can be sure it is outside of the expecting DIF that results naturally from their difference in ability.
--->
+Our goal here is to test the items on the American CDI:WS for DIF along three main axes: sex (male vs.
+female), maternal education (no more than secondary vs.
+at least some college), and race (white vs.
+non-white).
 
@@
 eth2_tab <- table(d_demo$eth_group)
+momed_tab <- table(d_demo$mom_ed)
 ```
@@
 We analyze parent-reported Wordbank data from `r nrow(d_demo)` American English CDI: Words & Sentences administrations for children `r min(d_demo$age)` to `r max(d_demo$age)` months of age [@frank2017; @frank2021].
-Full demographic data are not reported in some datasets contributed to Wordbank: sex was available for `r nrow(d_demo)-sum(is.na(d_demo$sex))` children, race/ethnicity was available for `r nrow(d_demo)-sum(is.na(d_demo$ethnicity))` children, and mother's education (a proxy for socioeconomic status; SES) was available for `r nrow(d_demo)-sum(is.na(d_demo$ses_group))` children.
+Full demographic data are not reported in some datasets contributed to Wordbank: sex was available for `r nrow(d_demo)-sum(is.na(d_demo$sex))` children, race/ethnicity was available for `r nrow(d_demo)-sum(is.na(d_demo$ethnicity))` children, and maternal education (a proxy for socioeconomic status; SES) was available for `r nrow(d_demo)-sum(is.na(d_demo$ses_group))` children.
 
@@
 The analysis of race-based differences included data from `r eth_tab["White"]` white, `r eth_tab["Asian"]` Asian, `r eth_tab["Black"]` Black, `r eth_tab["Hispanic"]` Hispanic, and `r eth_tab["Other"]` "Other" children.
-For this analysis we categorized participants' race/ethnicity as White (`r eth2_tab["White"]`) or Non-white (`r eth2_tab["Nonwhite"]`). 
-For the SES-based analysis, we categorized the `r ses_tab["high"]` children whose mothers had at least some college or more as high-SES, and those whose mothers had at most high school (`r ses_tab["low"]` children) as low-SES.
+Due to sparse data for many categories, we binarized participants' race/ethnicity as White (`r eth2_tab["White"]`) or Non-white (`r eth2_tab["Nonwhite"]`), recognizing that there may be important variation between groups that this will fail to capture.
+Data for the maternal education analysis included CDI data from children whose mother's had the following levels of education: `r momed_tab["Primary"]` with no more than primary school education, `r momed_tab["Some Secondary"]` with some secondary school, `r momed_tab["Secondary"]` with no more than secondary school, `r momed_tab["Some College"]` with some college, `r momed_tab["College"]` with no more than a college degree, `r momed_tab["Some Graduate"]` with no more than some graduate school, and `r momed_tab["Graduate"]` with a graduate degree.
+Again due to data sparsity, we binarized the `r ses_tab["high"]` children whose mothers had at least some college or more as high maternal education (high-ME), and those whose mothers had at most high school (`r ses_tab["low"]` children) as low maternal education (low-ME).
 
@@
 For example, in the sex-based multigroup model, item $i$'s difficulty is $b_{i}^{female}$ for females, and $b_{i}^{male}$ for males.
-In the quest to identify DIF, a multigroup Rasch model will be fitted for each demographic dimension of interest (sex, socioeconomic status, and ethnicity).
-To contextualize these results, we first fitted a baseline Rasch model to the entire dataset, without demographic information.
+In the quest to identify DIF, a multigroup Rasch model will be fitted for each demographic dimension of interest (sex, maternal education, and ethnicity), and we will examine the between-group difficulty difference for each item (e.g., $d_i = b_{i}^{female} - b_{i}^{male}$).
+If there is no DIF for a given item, then $d_i \approx 0$ as the two groups find the item equally difficult.
 
 # Results
+
+We will first examine the size of demographic effects on language ability in a baseline Rasch model fitted without regard to demographic group.
+We then fit a multigroup Rasch model for each examine evidence for DIF, assuming that 
+Finally, we prune the CDI of items showing varying degrees of DIF
 
@@
 
+To contextualize the later DIF results, we first fitted a baseline Rasch model to the entire dataset, without demographic information.
 Figure 1 shows children's language ability vs. age by demographic group from the baseline Rasch model, which assumes no DIF (i.e., equal item parameters for all groups).
@@
 for sex, $\mu_{male}=\mu_{female}=0$), thus pushing all between-group variation into the item difficulty parameters.
-For the case of sex, where we believe $\mu_{female}>\mu_{male}$, this means we may expect to find many items with difficulty $b_j^{female} < b_j^{male}$, but we can still examine the distribution of differences in item difficulty ($d_j = b_j^{male} - b_j^{female}$) for outliers.
+For the case of sex, where we believe $\mu_{female}>\mu_{male}$, this means we may expect to find many items with difficulty $b_i^{female} < b_i^{male}$, but we can still examine the distribution of differences in item difficulty ($d_j = b_i^{male} - b_i^{female}$) for outliers.
 To give analysts a sense of the uncertainty about the existence of DIF, GLIMMER plots show distributions of parameter differences rather than point estimates. 
@@
 For example, there is some clustering at the top and bottom of Figure 2: at the top, "vagina", "tights", "dress (object)", and "doll" form a cluster of items that are much easier for females, while at the bottom, "penis" stands out as much easier for males.
-For the SES and race GLIMMERs (Figures 3 and 4)--and in the rest of the full sex GLIMMER, there are not clusters, but rather a continuum of smoothly varying differences with overlap. 
+For the maternal education and race GLIMMERs (Figures 3 and 4)--and in the rest of the full sex GLIMMER, there are not clusters, but rather a continuum of smoothly varying differences with overlap. 
 This makes identifying items with DIF quite difficult, as different methods are likely to yield inconsistent results [@stenhaug2021treading].
-Nonetheless, we describe the distributions of item-level group difficulty differences, and examine overlap of the extrema.
+Nonetheless, we describe the distributions of item-level group difficulty differences, and identify extreme values, flagging items with difficulty differences that were outside 2 standard deviations of the mean of that distribution.
 
@@
 
-mm_sex <- get_extreme_item_difficulty_differences(mm_sex) # min: -.53 max: 1.06
-mm_ses <- get_extreme_item_difficulty_differences(mm_ses) # min: -1.07 max: 1.13
-mm_eth <- get_extreme_item_difficulty_differences(mm_eth) # min: -.64 max: 1.56
+# varying thresholds
+prune_dat <- tibble()
+for(thresh in seq(.25, 2.5, .25)) {
+  tmp_sex <- get_extreme_item_difficulty_differences(mm_sex, sd_thresh=thresh) # min: -.53 max: 1.06
+  tmp_ses <- get_extreme_item_difficulty_differences(mm_ses, sd_thresh=thresh) # min: -1.07 max: 1.13
+  tmp_eth <- get_extreme_item_difficulty_differences(mm_eth, sd_thresh=thresh) # min: -.64 max: 1.56
+  prune_defs <- intersect(intersect(subset(tmp_ses, extreme)$definition, 
+                       subset(tmp_eth, extreme)$definition),
+                       subset(tmp_sex, extreme)$definition)
 
-all_extreme <- rbind(subset(mm_sex, extreme),
-                     subset(mm_ses, extreme),
-                     subset(mm_eth, extreme)) 
+  # prune the columns
+  keep_cols <- which(!is.element(names(d_mat), prune_defs))
+  #names(d_mat)[which(is.element(names(d_mat), prune_defs))]
+  d_mat_pruned <- d_mat[,keep_cols]
+
+  m0 <- mirt(d_mat_pruned %>% filter(rowSums(d_mat_pruned)!=0) %>% select(-data_id), 1, "Rasch", 
+           technical=list(NCYCLES=1000))
+
+  #coefs_1pl <- as_tibble(coef(m0, simplify = TRUE)$items) %>%
+  #  mutate(definition = rownames(coef(m0, simplify = TRUE)$items))
+  fscores_1pl <- tibble(data_id = subset(d_mat_pruned, rowSums(d_mat_pruned)!=0)$data_id, 
+                               ability_pr = fscores(m0, method = "MAP")[,1])
+
+  d_demo_tmp <- d_demo %>% left_join(fscores_1pl) %>%
+    mutate(age_sc = scale(age, center=T, scale=F))
+  # join with d_demo, do the demo regressions again..
+
+  eth_reg_pr <- summary(lm(ability_pr ~ age_sc * eth_group, data=d_demo_tmp))$coefficients # R^2=.48 
+  ses_reg_pr <- summary(lm(ability_pr ~ age_sc * ses_group, data=d_demo_tmp))$coefficients # R^2=.56
+  sex_reg_pr <- summary(lm(ability_pr ~ age_sc * sex, data=d_demo_tmp))$coefficients # R^2=.59
+
+  prune_dat = bind_rows(prune_dat, 
+                        tibble(SDthreshold=thresh, 
+                               extreme_sex=nrow(subset(tmp_sex, extreme)),
+                               extreme_ses=nrow(subset(tmp_ses, extreme)),
+                               extreme_eth=nrow(subset(tmp_eth, extreme)),
+                               extreme_total = length(prune_defs),
+                               sex_beta = sex_reg_pr[3,1],
+                               ses_beta = ses_reg_pr[3,1],
+                               eth_beta = eth_reg_pr[3,1]))
+}
+
+# 2SD threshold
+mm_sex2 <- get_extreme_item_difficulty_differences(mm_sex) # min: -.53 max: 1.06
+mm_ses2 <- get_extreme_item_difficulty_differences(mm_ses) # min: -1.07 max: 1.13
+mm_eth2 <- get_extreme_item_difficulty_differences(mm_eth) # min: -.64 max: 1.56
+
+all_extreme <- rbind(subset(mm_sex2, extreme),
+                     subset(mm_ses2, extreme),
+                     subset(mm_eth2, extreme)) 
 # length(unique(all_extreme$definition)) 
@@
 
+### Sex
 For sex, the median difficulty difference (male-female) was `r round(median(mm_sex$d_diff),2)` (M=`r round(mean(mm_sex$d_diff),2)`, sd=`r round(sd(mm_sex$d_diff),2)`), with `r length(which(mm_sex$d_diff>0))`/680 items being easier for females than males.
 The fact that the bulk of this distribution favors females seems to confirm that the female language advantage is pervasive across the CDI:WS, and is thus likely to be a real advantage.
-For SES, the median difficulty difference (low-high) was `r round(median(mm_ses$d_diff),2)` (M=`r round(mean(mm_ses$d_diff),2)`, sd=`r round(sd(mm_ses$d_diff),2)`), with `r length(which(mm_ses$d_diff>0))`/680 items being easier for high-SES than low-SES children.
-With the mean and median difficulty differences close to 0, and roughly half of the words favoring each SES group, it is tempting to conclude that the CDI:WS items are somewhat balanced with respect to SES.
+
+### Maternal Education
+For maternal education (ME), the median difficulty difference (low-high) was `r round(median(mm_ses$d_diff),2)` (M=`r round(mean(mm_ses$d_diff),2)`, sd=`r round(sd(mm_ses$d_diff),2)`), with `r length(which(mm_ses$d_diff>0))`/680 items being easier for high-ME than low-ME children.
+With the mean and median difficulty differences close to 0, and roughly half of the words favoring each ME group, it is tempting to conclude that the CDI:WS items are somewhat balanced with respect to ME.
+
+### Race
 For race, the median difficulty difference (nonwhite-white) was `r round(median(mm_eth$d_diff),2)` (M=`r round(mean(mm_eth$d_diff),2)`, sd=`r round(sd(mm_eth$d_diff),2)`), with `r length(which(mm_eth$d_diff>0))`/680 items being easier for white than non-white children, revealing a fairly pervasive advantage for white children on CDI items.
 
-Finally, for each distribution we looked at extreme values, flagging items with difficulty differences that were outside 2 standard deviations.
 This identified 25 items with extreme sex-based difficulty differences, only 7 of which were easier for females, with the other 18 items favoring males.
@@
 
+## The difficulty of DIF
+
+However, DIF is fundamentally difficult to identify for multiple reasons [for an overview, see @stenhaug2021treading].
+First, most techniques to identify DIF rely on defining a set of "anchor" test items that are assumed to be equally difficult (i.e., unbiased) for both groups of interest.
+Identifying anchor items is at best fraught when there may in fact be a difference in ability between groups (e.g., the female language advantage), and is further confounded when the magnitude of this ability difference is unknown.
+A reason that DIF is particularly tricky in measuring children's early language ability is that there is a finite universe of early-learned words to choose from--and we may expect many of them to be biased for various environmental reasons (e.g., children in Florida may not use mittens or skis).
+Hence, the presence of DIF on a wide variety of items may not indicate the presence of bias; it could indicate that one demographic has a higher average ability level than the other.
+For example, let male language ability be drawn from a standard normal $\mu_{male} ~ N(0,1)$, with female language ability slightly higher, on average ($\mu_{female} = \mu_{male} + 0.1$).
+Then we would expect items to be an average of $0.1$ easier for females than for males, and we might identify items that are instead easier for males than females as showing undesirable DIF.
+And yet, without knowing the actual ability difference between two demographic groups--for which we also rely upon our tests, it is difficult to adjudicate which items show DIF, and which do not.
+
 This investigation is only a first step in measuring demographic bias in the items on the CDI:WS.

### 2022-01-21 paper/cogsci.Rmd
added pruning vs. demo effect size plot

@@
 mm_eth <- extract_group_df(mod_intuitive_eth, groups=c("Nonwhite","White"))
+```
 
+```{r train-pruned-models, eval=F}
 # varying thresholds
 prune_dat <- tibble()
-for(thresh in seq(.25, 2.5, .25)) {
+for(thresh in seq(.25, 3.0, .25)) {
   tmp_sex <- get_extreme_item_difficulty_differences(mm_sex, sd_thresh=thresh) # min: -.53 max: 1.06
@@
   tmp_eth <- get_extreme_item_difficulty_differences(mm_eth, sd_thresh=thresh) # min: -.64 max: 1.56
-  prune_defs <- intersect(intersect(subset(tmp_ses, extreme)$definition, 
-                       subset(tmp_eth, extreme)$definition),
-                       subset(tmp_sex, extreme)$definition)
+  prune_defs <- unique(c(subset(tmp_ses, extreme)$definition, 
+                       subset(tmp_eth, extreme)$definition,
+                       subset(tmp_sex, extreme)$definition))
 
@@
 
+save(prune_dat, file=here("data/pruned_demo_coefs.Rdata"))
+```
+
+```{r}
 # 2SD threshold
-mm_sex2 <- get_extreme_item_difficulty_differences(mm_sex) # min: -.53 max: 1.06
-mm_ses2 <- get_extreme_item_difficulty_differences(mm_ses) # min: -1.07 max: 1.13
-mm_eth2 <- get_extreme_item_difficulty_differences(mm_eth) # min: -.64 max: 1.56
+mm_sex <- get_extreme_item_difficulty_differences(mm_sex) # min: -.53 max: 1.06
+mm_ses <- get_extreme_item_difficulty_differences(mm_ses) # min: -1.07 max: 1.13
+mm_eth <- get_extreme_item_difficulty_differences(mm_eth) # min: -.64 max: 1.56
 
-all_extreme <- rbind(subset(mm_sex2, extreme),
-                     subset(mm_ses2, extreme),
-                     subset(mm_eth2, extreme)) 
+all_extreme <- rbind(subset(mm_sex, extreme),
+                     subset(mm_ses, extreme),
+                     subset(mm_eth, extreme)) 
 # length(unique(all_extreme$definition)) 
@@
 #intersect(subset(mm_ses, extreme)$definition, subset(mm_eth, extreme)$definition)
+```
+
+```{r pruned-coefs, out.width="\\linewidth", fig.env = "figure", fig.pos = "H", fig.align='center', fig.width=3.5, fig.height=3.5, set.cap.width=T, num.cols.cap=1, fig.cap = "Size of demographic effects (regression coefficients) vs.
+the threshold of pruned items."}
+load(here("data/pruned_demo_coefs.Rdata"))
+
+prune_dat %>% rename(c(beta_sex=sex_beta, beta_ses=ses_beta, beta_eth=eth_beta)) %>%
+  pivot_longer(c("extreme_sex","extreme_ses","extreme_eth","beta_sex","beta_ses","beta_eth"), 
+                           names_to=c("type", "demo"), names_sep="_") %>%
+  pivot_wider(names_from=type, values_from=value) %>%
+  mutate(Dimension = ifelse(demo=="ses", "maternal ed", ifelse(demo=="eth", "race", demo))) %>%
+  ggplot(aes(x=SDthreshold, y=beta, color=extreme)) + facet_wrap(. ~ Dimension) + 
+  geom_point() + theme(legend.position = "bottom") +
+  labs(color="Excluded items") + 
+  xlab("Exclusion Threshold (SDs)") + ylab("Regression coefficient (beta)")
 ```

### 2022-01-22 paper/cogsci.Rmd
working on plots

@@
 library(ggpubr)
+library(ggrepel)
 
@@
 
-<!-- put the GLIMMERs side-by-side?
-```{r 2-col-image, fig.env = "figure*", fig.pos = "h", fig.width=4, fig.height=2, fig.align = "center", set.cap.width=T, num.cols.cap=2, fig.cap = "This image spans both columns.
-And the caption text is limited to 0.8 of the width of the document."}
-
-```
--->
-
 ## Identifying biased CDI items
@@
 
-```{r pruned-coefs, out.width="\\linewidth", fig.env = "figure", fig.pos = "H", fig.align='center', fig.width=3.5, fig.height=3.5, set.cap.width=T, num.cols.cap=1, fig.cap = "Size of demographic effects (regression coefficients) vs.
+```{r pruned-coefs, out.width="\\linewidth", fig.env = "figure", fig.pos = "H", fig.align='center', fig.width=3.4, fig.height=4.75, set.cap.width=T, num.cols.cap=1, fig.cap = "Size of demographic effects (regression coefficients) vs.
 the threshold of pruned items."}
@@
 
-prune_dat %>% rename(c(beta_sex=sex_beta, beta_ses=ses_beta, beta_eth=eth_beta)) %>%
+# prune_dat %>% rename(c(beta_sex=sex_beta, beta_ses=ses_beta, beta_eth=eth_beta)) %>%
+#   pivot_longer(c("extreme_sex","extreme_ses","extreme_eth","beta_sex","beta_ses","beta_eth"), 
+#                            names_to=c("type", "demo"), names_sep="_") %>%
+#   pivot_wider(names_from=type, values_from=value) %>%
+#   mutate(Dimension = ifelse(demo=="ses", "maternal ed", ifelse(demo=="eth", "race", demo))) %>%
+#   ggplot(aes(x=SDthreshold, y=beta, color=extreme)) + facet_wrap(. ~ Dimension, scales="free") + 
+#   geom_point() + theme(legend.position = "bottom") + # geom_line() +
+#   labs(color="Excluded items") + 
+#   xlab("Exclusion Threshold (SDs)") + ylab("Regression coefficient (beta)")
+
+pdat <- prune_dat %>% rename(c(beta_sex=sex_beta, beta_ses=ses_beta, beta_eth=eth_beta)) %>%
   pivot_longer(c("extreme_sex","extreme_ses","extreme_eth","beta_sex","beta_ses","beta_eth"), 
@@
   pivot_wider(names_from=type, values_from=value) %>%
-  mutate(Dimension = ifelse(demo=="ses", "maternal ed", ifelse(demo=="eth", "race", demo))) %>%
-  ggplot(aes(x=SDthreshold, y=beta, color=extreme)) + facet_wrap(. ~ Dimension) + 
+  mutate(Dimension = ifelse(demo=="ses", "maternal ed", ifelse(demo=="eth", "race", demo)))
+
+pr1 <- pdat %>% filter(Dimension=="sex") %>%
+  ggplot(aes(x=SDthreshold, y=beta, color=extreme)) + 
   geom_point() + theme(legend.position = "bottom") + 
-  labs(color="Excluded items") + 
-  xlab("Exclusion Threshold (SDs)") + ylab("Regression coefficient (beta)")
+  labs(color="Excluded items") + ylim(-.7, -.4) + ggtitle("Sex") +
+  xlab("Exclusion Threshold (SDs)") + ylab("Coefficient (beta)")
+
+pr2 <- pdat %>% filter(Dimension=="maternal ed") %>%
+  ggplot(aes(x=SDthreshold, y=beta, color=extreme)) + 
+  geom_point() + theme(legend.position = "bottom") + 
+  labs(color="Excluded items") + ylim(-.3, -.1) + ggtitle("Maternal Education") +
+  xlab("Exclusion Threshold (SDs)") + ylab("Coefficient (beta)")
+
+pr3 <- pdat %>% filter(Dimension=="race") %>%
+  ggplot(aes(x=SDthreshold, y=beta, color=extreme)) + 
+  geom_point() + theme(legend.position = "bottom") + 
+  labs(color="Excluded items") + ylim(.35, .55) + ggtitle("Race") +
+  xlab("Exclusion Threshold (SDs)") + ylab("Coefficient (beta)")
+
+ggarrange(pr1, pr2, pr3, ncol=1, common.legend = T)
 ```
@@
 ```{r load-pruned}
+# (2sd threshold)
 load(here("data/eng_Ws_prod_mod_1pl_pruned.Rds"))
@@
 #female=0.57
+
+# table proportion of DIF items by lexical class -
+# if due to environmental 
 ```
@@
 
-```{r demo-ability, include=F, out.width="\\linewidth", fig.env = "figure", fig.pos = "H", fig.align='center', fig.width=3.3, fig.height=5.5, set.cap.width=T, num.cols.cap=1, fig.cap = "Density plots of language ability for children aged 23-25 months using the pruned model."}
-# ToDo: maybe make these violin plots and squeeze all 6 groups side-by-side?
+```{r demo-ability, include=T, out.width="\\linewidth", fig.env = "figure", fig.pos = "H", fig.align='center', fig.width=3.5, fig.height=7.5, set.cap.width=T, num.cols.cap=1, fig.cap = "Scatterplots showing item difficulty per demographic group along each dimension."}
 
-# sex
-sex <- d_demo %>% filter(!is.na(sex), production!=0) %>%
-  mutate(theta_sex = ability_pr)
-  #mutate(theta_sex = fscores(mod_intuitive_sex)) 
-# cor.test(eth$production, eth$theta_eth) # .98
+p1 <- mm_sex %>%
+  ggplot(aes(x=-d_g1, y=-d_g2)) + geom_point(alpha=.2) + # color=extreme
+  xlab("Male Difficulty") + ylab("Female Difficulty") +
+  geom_text_repel(aes(label=definition)) +
+  geom_abline(aes(slope=1, intercept=0), linetype="dashed")
 
-sex_m2 <- sex %>% group_by(sex) %>%
-  filter(age<26, age>22) %>%
-  summarise(ability = mean(theta_sex, na.rm=T)) 
-# now Female ability = 0.68, male=0.14
-p1 <- sex %>% filter(age<26, age>22) %>%
-  ggplot(aes(x=theta_sex, color=sex, fill=sex)) +
-    geom_density(alpha=.2) + xlab("Ability") + 
-    geom_vline(data=sex_m2, aes(xintercept=ability, color=sex)) +
-    scale_color_discrete(type=sex_colors) + 
-    scale_fill_discrete(type=sex_colors)
+p2 <- mm_ses %>%
+  ggplot(aes(x=-d_g1, y=-d_g2)) + geom_point(alpha=.2) +
+  xlab("Low-ME Difficulty") + ylab("High-ME Difficulty") +
+  geom_text_repel(aes(label=definition)) +
+  geom_abline(aes(slope=1, intercept=0), linetype="dashed")
 
-# SES
-ses <- d_demo %>% filter(!is.na(ses_group), production!=0) %>%
-  #mutate(theta_ses = fscores(mod_intuitive_ses)) %>%
-  mutate(theta_ses = ability_pr) %>%
-  filter(age<26, age>22) %>%
-  rename(SES=ses_group)
-# cor.test(eth$production, eth$theta_eth) # .98
+p3 <- mm_eth %>%
+  ggplot(aes(x=-d_g1, y=-d_g2)) + geom_point(alpha=.2) +
+  xlab("Non-white Difficulty") + ylab("White Difficulty") +
+  geom_text_repel(aes(label=definition)) +
+  geom_abline(aes(slope=1, intercept=0), linetype="dashed")
 
-ses_m2 <- ses %>% group_by(SES) %>%
-  filter(age<26, age>22) %>%
-  summarise(ability = mean(theta_ses, na.rm=T)) 
-# now high ability=.41, low=.08
-p2 <- ses %>%
-  ggplot(aes(x=theta_ses, color=SES, fill=SES)) +
-    geom_density(alpha=.2) + xlab("Ability") + 
-    geom_vline(data=ses_m2, aes(xintercept=ability, color=SES)) + 
-    scale_color_discrete(type=ses_colors) +
-    scale_fill_discrete(type=ses_colors)
-
-# ethnicity
-eth <- d_demo %>% filter(!is.na(eth_group), production!=0) %>%
-  #mutate(theta_eth = fscores(mod_intuitive_eth)) %>%
-  mutate(theta_eth = ability_pr) %>%
-  filter(age<26, age>22) %>%
-  mutate(Race=eth_group)
-# cor.test(eth$production, eth$theta_eth) # .98
-
-eth_m2 <- eth %>% group_by(Race) %>%
-  filter(age<26, age>22) %>%
-  summarise(ability = mean(theta_eth, na.rm=T))
-# now white=.52, nonwhite=.47 (was .59, .06)
-p3 <- eth %>%
-  ggplot(aes(x=theta_eth, color=Race, fill=Race)) +
-    geom_density(alpha=.2) + xlab("Ability") + 
-    geom_vline(data=eth_m2, aes(xintercept=ability, color=Race)) + 
-    scale_color_discrete(type=eth_colors) + 
-    scale_fill_discrete(type=eth_colors)
+# p1 <- mm_sex %>% 
+#   ggplot(aes(x=d_diff)) + # , color=extreme, fill=extreme
+#     geom_density(alpha=.2) + xlab("Ability") 
+# 
+# p2 <- mm_ses %>%
+#   ggplot(aes(x=d_diff)) + 
+#     geom_density(alpha=.2) + xlab("Ability") 
+# 
+# p3 <- mm_eth %>%
+#   ggplot(aes(x=d_diff)) + 
+#     geom_density(alpha=.2) + xlab("Ability")
 
@@
 
-```{r}
+```{r, include=T, out.width="\\linewidth", fig.env = "figure", fig.pos = "H", fig.align='center', fig.width=3.5, fig.height=3.5, set.cap.width=T, num.cols.cap=1, fig.cap = "CHILDES frequency of input to male vs.
+female children."}
 type_freqs = readRDS("../data/type_freqs.rds")
@@
   ylab("Female - Male Word Difficulty") +
-  xlab("Proportion of Utterances to Females in CHILDES")
+  xlab("Proportion of Utterances to Females in CHILDES") +
+  theme(legend.position = "bottom")
 

### 2022-01-24 paper/cogsci.Rmd
edits from Nathan, new refs

@@
     Communicative Development Inventories (CDIs) have been shown to provide valid, consistent 
-    measures of children's aggregate early language skill. However, CDIs are predominantly 
-    comprised of hundreds of vocabulary items, some of which may not be heard (and thus learned) 
-    equally often by children of varying backgrounds. Here, we use a database of American English 
-    CDIs to identify words that show strong bias for particular groups of children, on dimensions
-    of sex (male vs. female), race (white vs. non-white), and maternal education 
-    (high vs. low). For each dimension, we identify dozens of strongly biased items, and show that 
-    eliminating these items reduces the expected ability difference between groups. For sex, we
-    consider how to propose replacement words that may show less bias, on the basis of their 
-    relatively equal frequency in adult speech directed to male and female children.
+    measures of children's aggregate early language skill. However, CDIs contain
+    hundreds of vocabulary items, some of which may not be heard (and thus learned) equally
+    often by children of varying backgrounds. This study used a database of American English 
+    CDIs to identify words that showed strong bias for particular demographic groups of children, on 
+    dimensions of sex (male vs. female), race (white vs. non-white), and maternal education 
+    (high vs. low). For each dimension, we identified dozens of strongly biased items, and showed that 
+    eliminating these items reduced the expected ability difference between groups. Additionally,
+    we investigated how well the relative frequency of words spoken to young girls vs. boys predicted 
+    sex-based word learning bias, and discuss possible sources of demographic bias in early word learning.
 
 keywords: >
-    language acquisition; word learning; measuring instrument bias; development;
+    language acquisition; word learning; measuring instrument bias; demographics;
 
@@
 However, it is also difficult to get a complete measure of young children's language skill: long recordings are prohibitively difficult to collect and transcribe [see @roy2015predicting, the exception that proves the rule], and yet any short recording (e.g., a 1-hour play session) will elicit only a small proportion of the words and constructions that children know.
-Thus, researchers of early word learning have constructed tests with hundreds of words, with an intentional bias to select words that are more likely to be known by children.
+Thus, researchers of early word learning have constructed tests with hundreds of words, intentionally oversampling words that are more likely to be known by young children.
 
-We focus on one set of popular instruments, the MacArthur-Bates Communicative Development Inventories [CDIs; @Fenson2007]. 
-The CDIs are a set of parent-reported measures of children's productive and receptive language skills, which produce a low-cost and reliable way to estimate children's early language skills [@fenson1994].
-CDIs have shown good predictive validity [e.g., @fenson1994; Bornstein et al. 2012, Duff et al. 2015].
-Our focus will be on the vocabulary checklist portion of the CDI Words & Sentences (CDI:WS) form, comprised of 680 early-learned words across 22 categories (e.g., animals, vehicles, action words, pronouns) selected to assess the productive vocabulary of children 16 to 30 months of age.
+We focus on the MacArthur-Bates Communicative Development Inventories [CDIs; @Fenson2007], a set of parent-reported measures of children's productive and receptive language skills, which offer a low-cost and reliable way to estimate children's early language skills [@fenson1994].
+CDIs have shown good predictive validity [e.g., @fenson1994; @bornstein2012stability; @duff2015infant].
+Our primary focus is the vocabulary checklist portion of the CDI Words & Sentences (CDI:WS) form, comprised of 680 early-learned words across 22 categories (e.g., animals, vehicles, action words, pronouns) selected to assess the productive vocabulary of children 16 to 30 months of age.
 For each item on the CDI:WS, caregivers are asked to respond whether the target child has been heard to say (i.e. produce) the given item.[^1]
@@
 If certain high-SES activities are driving early word learning for these children, what are the activities (and associated vocabulary) that low-SES households are instead engaging in?
-A fair test of children's early vocabulary would contain a representative sample of words from all activities that children engage in, across demographic groups.
+A truly fair test of children's early vocabulary would contain a representative sample of words from all activities that children engage in, across demographic groups.
 

### 2022-01-24 paper/cogsci.Rmd
edits

@@
 Our primary focus is the vocabulary checklist portion of the CDI Words & Sentences (CDI:WS) form, comprised of 680 early-learned words across 22 categories (e.g., animals, vehicles, action words, pronouns) selected to assess the productive vocabulary of children 16 to 30 months of age.
-For each item on the CDI:WS, caregivers are asked to respond whether the target child has been heard to say (i.e. produce) the given item.[^1]
+For each item on the CDI:WS, caregivers are asked to respond whether the target child has been heard to say (i.e. produce) the given item.
 Children's total vocabulary score on the CDI:WS is tightly correlated with other facets of early language (e.g., grammatical competence and gesture), suggesting that the language system is "tightly woven" [@frank2021].
@@
 
-[^1]: The CDI: Words and Gestures form includes a subset of the CDI:WS items and targets children 12 to 18 months of age, measuring both comprehension and production.
+<!-- [^1]: The CDI: Words and Gestures form includes a subset of the CDI:WS items and targets children 12 to 18 months of age, measuring both comprehension and production. -->
 
 Inspired by the utility and widespread use of the CDI, researchers have recently been using psychometric models on CDI data to construct short, adaptive tests to reliably assess language ability using only a small subset of the CDI items [e.g., @mayor2019; @kachergis2021cat].
-These psychometric models typically come from the Item-Response Theory (IRT) framework [@Baker2001], which assumes that not only test-takers (here, children) have normally-distributed ability, but that items (words on the CDI) have normally-distributed difficulty.
-The very efficacy of these IRT-based models depends on words varying in difficulty, and hence being more/less informative of the ability level of different individuals.
+These psychometric models typically come from the Item-Response Theory (IRT) framework [@Baker2001], which assumes that not only test-takers (here, children being reported on) have normally-distributed ability, but that items (words on the CDI) have normally-distributed difficulty.
+The efficacy of these IRT-based models depends on words varying in difficulty, in order for the test to be more informative of the ability level of each individual.
 For example, asking whether a 22-month-old produces the word "ball" is far less informative of that child's language ability than asking whether they produce "table", as 96% of 22-month-olds can produce the former, while only 47% produce the latter.
@@
 The outline of this paper is as follows.
-First, we introduce the Wordbank data and the IRT model, and use it examine the overall size of demographic differences in early word learning.
+First, we introduce the Wordbank data and the IRT model, and use it to examine the overall size of demographic differences in early word learning.
 We then fit the IRT model to each group along each demographic dimension, and examine the item parameters for evidence of DIF, noting in particular how many items are significantly biased in favor of each group.
@@
 The Rasch model, also known as the 1-parameter logistic (1PL) model, is the simplest Item Response Theory model, and is thus the easiest to use to investigate potential differences in item function across different groups of participants.
-The goal of the Rasch model is to jointly estimate for each child $j$ a latent ability $\theta_j$, and for each item $i$ a difficulty parameter $b_i$.
+The Rasch model jointly estimates for each child $j$ a latent ability $\theta_j$, and for each item $i$ a difficulty parameter $b_i$.
 In the model, the probability of child $j$ knowing (i.e., producing or understanding) a given item $i$ is
@@
 
-In the multigroup Rasch model, an item's difficulty is allowed to vary depending on which group an individual child belongs to. 
+In the multigroup Rasch model, an item's difficulty is allowed to vary by group. 
 For example, in the sex-based multigroup model, item $i$'s difficulty is $b_{i}^{female}$ for females, and $b_{i}^{male}$ for males.
@@
 Figure 1 shows children's language ability vs. age by demographic group from the baseline Rasch model, which assumes no DIF (i.e., equal item parameters for all groups).
-A linear regression for each demographic group, with age (centered) and its interaction, showed significant effects in the expected directions, as can also be seen in the figure. 
+A linear regression for each demographic group, with age (centered) and its interaction, showed significant effects. 
 Female children had higher language ability than male children ($\beta=0.56$, $p<.001$), with no significant interaction with age ($\beta=0.02$, $p=.10$).

### 2022-01-24 paper/cogsci.Rmd
intro done

@@
 Yet maternal education, often used as a proxy for socioeconomic status (SES), has also been found to be associated with children's language processing and vocabulary by 18 months [@fernald2013ses], and to be predictive of later educational outcomes [@marchman2008speed; see @schwab2016 for a review].
-Demographic factors are also predictive of language skill: first-born children tend to outpace their siblings, and female children tend to have better language skills than their age-matched male counterparts [@fenson1994] -- a sex-based verbal advantage that continues through high school [see @petersen2018gender for a review].
+Other demographic factors are also predictive of language skill: first-born children tend to outpace their siblings, and female children tend to have better language skills than their age-matched male counterparts [@frank2021; @eriksson2012differences] -- a sex-based verbal advantage that continues through high school [see @petersen2018gender for a review].
 
@@
 
-We focus on the MacArthur-Bates Communicative Development Inventories [CDIs; @Fenson2007], a set of parent-reported measures of children's productive and receptive language skills, which offer a low-cost and reliable way to estimate children's early language skills [@fenson1994].
+We focus on the MacArthur-Bates Communicative Development Inventories [CDIs; @fenson1994; @Fenson2007], a set of parent-reported measures of children's productive and receptive language skills, which offer a low-cost and reliable way to estimate children's early language skills [@fenson1994].
 CDIs have shown good predictive validity [e.g., @fenson1994; @bornstein2012stability; @duff2015infant].
@@
 
+<!-- nix the CAT transition, raise the issue of potential test bias somehow-->
+
 Inspired by the utility and widespread use of the CDI, researchers have recently been using psychometric models on CDI data to construct short, adaptive tests to reliably assess language ability using only a small subset of the CDI items [e.g., @mayor2019; @kachergis2021cat].
@@
 The idea that some items on a test may show bias, favoring one group over another, is known as Differential Item Function (DIF; @holland1993differential]. 
-On any given test, it is clearly undesirable to have more items favoring one group (say, children from rural households) over another (urban children), as the test will overestimate the ability of test-takers in the former (rural) group -- and not because of any underlying mean difference in ability between the groups, but simply because the test is unfair [@camilli2013ongoing].
-A variety of statistical methods for detecting DIF have been proposed, and investigations have in several instances identified DIF for many items on tests favoring one group over another (e.g., rural vs.
-urban).
-Our goal here is to test the items on the American CDI:WS for DIF along three main axes: sex (male vs.
+The validity of any test with many items favoring one group over another (say, children from rural vs.
+urban households) is questionable, as the test may be overestimating the ability of test-takers in the former (rural) group, and/or underestimating the ability of the latter group -- and not because of any underlying mean difference in ability between the groups, but simply because the test is unfair [@camilli2013ongoing].
+For example, if a vocabulary test is composed almost entirely of the names of farm equipment, the test will underestimate the knowledge of urban children, who have experience of other contexts, despite being less familiar with farm equipment than their rural peers.
+Of course, there may in fact be some meaningful ability difference between demographic groups, as is likely the case for the female language advntage, which has been documented in many languages [@eriksson2012differences; @frank2021].
+However, these demographic differences in early language learning have not been investigated as potentially due to DIF.
+That is the contribution of the present study.
+Specifically, our goal is to test the 680 vocabulary items on the American CDI:WS for DIF along three dichotomized demographic dimensions: sex (male vs.
 female), maternal education (no more than secondary vs.
@@
 
+<!-- A variety of statistical methods for detecting DIF have been proposed, and investigations have in several instances identified DIF for many items on tests favoring one group over another. -->
+
 The outline of this paper is as follows.
-First, we introduce the Wordbank data and the IRT model, and use it to examine the overall size of demographic differences in early word learning.
-We then fit the IRT model to each group along each demographic dimension, and examine the item parameters for evidence of DIF, noting in particular how many items are significantly biased in favor of each group.
-We identify a set of suspect items to eliminate (or in the future, replace), and provide updated estimates of the effect size of these demographic variables, were the biased items to be eliminated.
-<!--Next, using a corpus of child-directed speech, we examine the extent to which variation in word frequencies heard by boys vs. girls predicts sex-based DIF.-->
-Finally, we provide recommendations for next steps to be taken to identify and replace biased items on the CDI.
+First, we introduce the Wordbank data and the IRT model, and examine the overall size of demographic differences in early word learning based on the full CDI:WS.
+We then fit the IRT model to both groups for each demographic factor, and look at the item parameters for evidence of DIF, noting in particular how many items are significantly biased in favor of each group, and describing them qualitatively.
+Next, we use these item parameters to systematically prune biased items from the CDI:WS, and examine how the estimated size of demographic differences change as we prune more items.
+<!-- We identify a set of suspect items to eliminate (or in the future, replace), and provide updated estimates of the effect size of these demographic variables, were the biased items to be eliminated.-->
+Finally, using a corpus of child-directed speech, we examine the extent to which variation in word frequencies heard by boys vs. girls predicts sex-based DIF.
+Based on these analyses, we provide recommendations for next steps to be taken to identify and potentially replace items on the CDI showing demographic bias.
 
@@
 For example, in the sex-based multigroup model, item $i$'s difficulty is $b_{i}^{female}$ for females, and $b_{i}^{male}$ for males.
-In the quest to identify DIF, a multigroup Rasch model will be fitted for each demographic dimension of interest (sex, maternal education, and ethnicity), and we will examine the between-group difficulty difference for each item (e.g., $d_i = b_{i}^{female} - b_{i}^{male}$).
+To identify DIF, a multigroup Rasch model will be fitted for each demographic dimension of interest (sex, maternal education, and ethnicity), and we will examine the between-group difficulty difference for each item (e.g., $d_i = b_{i}^{female} - b_{i}^{male}$).
 If there is no DIF for a given item, then $d_i \approx 0$ as the two groups find the item equally difficult.
@@
 The full GLIMMERs, with all 680 CDI:WS items, are available on [OSF](https://osf.io/57rsw/?view_only=2b6ecb61fe08458293af7421d276932a), but were too large to include here.
-Nonetheless, it is important to inspect the full plots, for if there is a cluster of items in a GLIMMER, the analyst may conclude that these items are strong candidates for DIF on that dimension.
+It is important, however, to inspect the full plots, for if there is a cluster of items in a GLIMMER, the analyst may conclude that these items are strong candidates for DIF on that dimension.
 For example, there is some clustering at the top and bottom of Figure 2: at the top, "vagina", "tights", "dress (object)", and "doll" form a cluster of items that are much easier for females, while at the bottom, "penis" stands out as much easier for males.

### 2022-01-24 paper/cogsci.Rmd
more intro writing

@@
 
-However, it is also difficult to get a complete measure of young children's language skill: long recordings are prohibitively difficult to collect and transcribe [see @roy2015predicting, the exception that proves the rule], and yet any short recording (e.g., a 1-hour play session) will elicit only a small proportion of the words and constructions that children know.
+However, it is also difficult to get a complete measure of young children's language skill: long recordings are prohibitively difficult to collect and transcribe [see the singular exception: @roy2015predicting], and yet any short recording (e.g., a 1-hour play session) will elicit only a small proportion of the words and constructions that children know.
 Thus, researchers of early word learning have constructed tests with hundreds of words, intentionally oversampling words that are more likely to be known by young children.
@@
 
-<!-- nix the CAT transition, raise the issue of potential test bias somehow-->
-
-Inspired by the utility and widespread use of the CDI, researchers have recently been using psychometric models on CDI data to construct short, adaptive tests to reliably assess language ability using only a small subset of the CDI items [e.g., @mayor2019; @kachergis2021cat].
+<!--Inspired by the utility and widespread use of the CDI, researchers have recently been using psychometric models on CDI data to construct short, adaptive tests to reliably assess language ability using only a small subset of the CDI items [e.g., @mayor2019; @kachergis2021cat].
 These psychometric models typically come from the Item-Response Theory (IRT) framework [@Baker2001], which assumes that not only test-takers (here, children being reported on) have normally-distributed ability, but that items (words on the CDI) have normally-distributed difficulty.
@@
 For example, asking whether a 22-month-old produces the word "ball" is far less informative of that child's language ability than asking whether they produce "table", as 96% of 22-month-olds can produce the former, while only 47% produce the latter.
+-->
 
-While it is quite reasonable to expect that some CDI words are "easier" (i.e., more likely to be known by children) than others, and even to use these varying difficulties to predict variation in children's language ability, the use of psychometric models highlights the possibility that some CDI items may function differently (i.e., be more/less widely known) for different groups of children.
-The idea that some items on a test may show bias, favoring one group over another, is known as Differential Item Function (DIF; @holland1993differential]. 
+<!--While it is quite reasonable to expect that some CDI words are "easier" (i.e., more likely to be known by children) than others, and even to use these varying difficulties to predict variation in children's language ability, the use of psychometric models highlights the possibility that some CDI items may function differently (i.e., be more/less widely known) for different groups of children.
+-->
+
+While many languages show similar demographic effects on early word learning [e.g., the female advantage: @eriksson2012differences; and the first-born advantage: @frank2021], the size of these demographic effects vary across languages [see @frank2021, Ch.
+6].
+Why might we see such variation?
+While it is possible that there is an underlying difference in language ability between demographic groups, could it also be possible that the tests are biased to some degree--and perhaps to a different degree in different languages?
+The idea that some items on a test may show bias, favoring one group over another, is known in psychometrics as Differential Item Function (DIF; @holland1993differential]. 
 The validity of any test with many items favoring one group over another (say, children from rural vs.

### 2022-01-24 paper/cogsci.Rmd
writing pruning section

@@
 theme_set(theme_classic(base_size=10))
+
+# color-blind palette
+cbp1 <- c("#999999", "#E69F00", "#56B4E9", "#009E73",
+          "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
+sex_colors = cbp1[2:3]
+eth_colors = cbp1[4:5]
+ses_colors = cbp1[6:7]
 ```
@@
 Why might we see such variation?
-While it is possible that there is an underlying difference in language ability between demographic groups, could it also be possible that the tests are biased to some degree--and perhaps to a different degree in different languages?
+While it is possible that there is an underlying difference in mean language ability between demographic groups, and that cross-linguistic variation is driven by differences in the language environments of different cultures, could it also be possible that the tests are biased to some degree--and perhaps to a different degree in different languages?
 The idea that some items on a test may show bias, favoring one group over another, is known in psychometrics as Differential Item Function (DIF; @holland1993differential]. 
@@
 
-We will first examine the size of demographic effects on language ability in a baseline Rasch model fitted without regard to demographic group.
-We then fit a multigroup Rasch model for each examine evidence for DIF, assuming that 
-Finally, we prune the CDI of items showing varying degrees of DIF
-
-```{r colors}
-# color-blind palette
-cbp1 <- c("#999999", "#E69F00", "#56B4E9", "#009E73",
-          "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
-sex_colors = cbp1[2:3]
-eth_colors = cbp1[4:5]
-ses_colors = cbp1[6:7]
-```
+First, we will examine the size of demographic effects on language ability in a baseline Rasch model fitted without regard to demographic group.
+Then we will fit a multigroup Rasch model for each demographic factor, and characterize the between-group differences in item difficulties.
+Next, we will re-examine the size of demographic effects after pruning biased items from the CDI:WS using a varying threshold.
+Finally, we measure the strength of association between sex-related differences in language input and the degree of sex-related DIF.
 
@@
 ## Demographic effects in baseline Rasch model
-
-To contextualize the later DIF results, we first fitted a baseline Rasch model to the entire dataset, without demographic information.
+A baseline Rasch model was fitted to the entire dataset, without demographic information.
 Figure 1 shows children's language ability vs. age by demographic group from the baseline Rasch model, which assumes no DIF (i.e., equal item parameters for all groups).
@@
 
-```{r baseline-ability-vs-age, out.width="\\linewidth", fig.env = "figure", fig.pos = "H", fig.align='center', fig.width=3.3, fig.height=6, set.cap.width=T, num.cols.cap=1, fig.cap = "Language ability vs.
+```{r baseline-ability-vs-age, out.width="\\linewidth", fig.env = "figure", fig.pos = "h", fig.align='center', fig.width=3.3, fig.height=6, set.cap.width=T, num.cols.cap=1, fig.cap = "Language ability vs.
 age by demographic group, from the Rasch model."}
@@
 
-The first step we take to identify CDI items with DIF is to create GLIMMERs (Graphs of Logits Imputed Multiply with Means Equal; @stenhaug2021treading), which visualize between-item variation in group performance differences.
+To aid in identifying CDI items with DIF we created GLIMMERs (Graphs of Logits Imputed Multiply with Means Equal; @stenhaug2021treading), which visualize between-item variation in group performance differences.
 These parameters are drawn from a fitted multigroup Rasch model for each demographic variable (sex, SES, and race), with the assumption that the mean language ability in each group is the same (e.g.
@@
 For the case of sex, where we believe $\mu_{female}>\mu_{male}$, this means we may expect to find many items with difficulty $b_i^{female} < b_i^{male}$, but we can still examine the distribution of differences in item difficulty ($d_j = b_i^{male} - b_i^{female}$) for outliers.
-To give analysts a sense of the uncertainty about the existence of DIF, GLIMMER plots show distributions of parameter differences rather than point estimates. 
+GLIMMER plots show distributions of parameter differences rather than point estimates to convey the uncertainty about the existence of DIF.
 These distributions are generated by drawing 10,000 imputations from the item parameter covariance matrix.
@@
 This makes identifying items with DIF quite difficult, as different methods are likely to yield inconsistent results [@stenhaug2021treading].
-Nonetheless, we describe the distributions of item-level group difficulty differences, and identify extreme values, flagging items with difficulty differences that were outside 2 standard deviations of the mean of that distribution.
+Hence, we next characterized the distributions of the item-level group difficulty differences, and measured the influence of pruning a varying number of items on the demographic effect sizes.
+
+```{r sex-glimmer, out.width="\\linewidth", fig.env = "figure", fig.pos = "h", fig.align='center', fig.width=3.2, fig.height=3.5, set.cap.width=T, num.cols.cap=1, fig.cap = "GLIMMER plot of a sample of CDI:WS words from the sex bias model.
+Words at the top are easier to learn for females, while those at the bottom are easier for males."}
+#img <- png::readPNG("figs/lab_logo_stanford.png")
+#grid::grid.raster(img)
+knitr::include_graphics("figs/smGLIMMER_sex_prodWS.pdf")
+```
+
+```{r ses-glimmer, out.width="\\linewidth", fig.env = "figure", fig.pos = "h", fig.align='center', fig.width=3, fig.height=3.5, set.cap.width=T, num.cols.cap=1, fig.cap = "GLIMMER plot of a sample of CDI:WS words from the SES bias model.
+Words at the top are easier to learn for children from low-SES families, while those at the bottom are easier for those from high-SES families."}
+knitr::include_graphics("figs/smGLIMMER_ses_prodWS.pdf")
+```
+
+```{r eth-glimmer, out.width="\\linewidth", fig.env = "figure", fig.pos = "h", fig.align='center', fig.width=3, fig.height=3.5, set.cap.width=T, num.cols.cap=1, fig.cap = "GLIMMER plot of a sample of CDI:WS words from the ethnicity bias model.
+Words at the top are easier to learn for children from low-SES families, while those at the bottom are easier for those from high-SES families."}
+knitr::include_graphics("figs/smGLIMMER_eth_prodWS.pdf")
+```
 
@@
 ```{r train-pruned-models, eval=F}
-# varying thresholds
+# varying SD thresholds (SLOW TO RUN)
 prune_dat <- tibble()
@@
 
-```{r pruned-coefs, out.width="\\linewidth", fig.env = "figure", fig.pos = "H", fig.align='center', fig.width=3.4, fig.height=4.75, set.cap.width=T, num.cols.cap=1, fig.cap = "Size of demographic effects (regression coefficients) vs.
-the threshold of pruned items."}
+```{r pruned-coefs, out.width="\\linewidth", fig.env = "figure", fig.pos = "h", fig.align='center', fig.width=3.4, fig.height=4.75, set.cap.width=T, num.cols.cap=1, fig.cap = "Size of demographic effects (regression coefficients) with different pruning thresholds."}
 load(here("data/pruned_demo_coefs.Rdata"))
-
-# prune_dat %>% rename(c(beta_sex=sex_beta, beta_ses=ses_beta, beta_eth=eth_beta)) %>%
-#   pivot_longer(c("extreme_sex","extreme_ses","extreme_eth","beta_sex","beta_ses","beta_eth"), 
-#                            names_to=c("type", "demo"), names_sep="_") %>%
-#   pivot_wider(names_from=type, values_from=value) %>%
-#   mutate(Dimension = ifelse(demo=="ses", "maternal ed", ifelse(demo=="eth", "race", demo))) %>%
-#   ggplot(aes(x=SDthreshold, y=beta, color=extreme)) + facet_wrap(. ~ Dimension, scales="free") + 
-#   geom_point() + theme(legend.position = "bottom") + # geom_line() +
-#   labs(color="Excluded items") + 
-#   xlab("Exclusion Threshold (SDs)") + ylab("Regression coefficient (beta)")
 
@@
   ggplot(aes(x=SDthreshold, y=beta, color=extreme)) + 
-  geom_point() + theme(legend.position = "bottom") + 
+  geom_point() + #theme(legend.position = "bottom") + 
   labs(color="Excluded items") + ylim(-.7, -.4) + ggtitle("Sex") +
@@
   ggplot(aes(x=SDthreshold, y=beta, color=extreme)) + 
-  geom_point() + theme(legend.position = "bottom") + 
+  geom_point() + 
   labs(color="Excluded items") + ylim(-.3, -.1) + ggtitle("Maternal Education") +
@@
   ggplot(aes(x=SDthreshold, y=beta, color=extreme)) + 
-  geom_point() + theme(legend.position = "bottom") + 
+  geom_point() + 
   labs(color="Excluded items") + ylim(.35, .55) + ggtitle("Race") +
@@
 
-This identified 25 items with extreme sex-based difficulty differences, only 7 of which were easier for females, with the other 18 items favoring males.
-<!-- thus, removing these extrema would actually lower male children's vocabulary scores..] -->
+<!--
+## Recommended Pruning Threshold: 2SD
+Considering just those items whose difficulty was >2SD from the mean difficulty difference identified 25 items with extreme sex-based difficulty differences, only 7 of which were easier for females, with the other 18 items favoring males.
 For SES, 39 extrema were identified, only 13 of which were easier for low-SES children.
@@
 In our final analysis, we explore the effect of pruning the 15 items that were extreme in both SES and race models.
+-->
 
@@
 
-```{r sex-glimmer, out.width="\\linewidth", fig.env = "figure", fig.pos = "H", fig.align='center', fig.width=3.2, fig.height=3.5, set.cap.width=T, num.cols.cap=1, fig.cap = "GLIMMER plot of a sample of CDI:WS words from the sex bias model.
-Words at the top are easier to learn for females, while those at the bottom are easier for males."}
-#img <- png::readPNG("figs/lab_logo_stanford.png")
-#grid::grid.raster(img)
-knitr::include_graphics("figs/smGLIMMER_sex_prodWS.pdf")
-```
-
-```{r ses-glimmer, out.width="\\linewidth", fig.env = "figure", fig.pos = "H", fig.align='center', fig.width=3, fig.height=3.5, set.cap.width=T, num.cols.cap=1, fig.cap = "GLIMMER plot of a sample of CDI:WS words from the SES bias model.
-Words at the top are easier to learn for children from low-SES families, while those at the bottom are easier for those from high-SES families."}
-knitr::include_graphics("figs/smGLIMMER_ses_prodWS.pdf")
-```
-
-```{r eth-glimmer, out.width="\\linewidth", fig.env = "figure", fig.pos = "H", fig.align='center', fig.width=3, fig.height=3.5, set.cap.width=T, num.cols.cap=1, fig.cap = "GLIMMER plot of a sample of CDI:WS words from the ethnicity bias model.
-Words at the top are easier to learn for children from low-SES families, while those at the bottom are easier for those from high-SES families."}
-knitr::include_graphics("figs/smGLIMMER_eth_prodWS.pdf")
-```
-
 ```{r, eval=F, include=F}
@@
 
-```{r load-pruned}
+Given the smoothly-varying distributions of DIF shown by each demographic factor, we chose to evaluate multiple thresholds for pruning the extreme-valued items from each distribution.
+For each demographic factor, we pruned items with difficulty difference from >.25 SD to >3 SD.
+Shown in Figure 5, the regression coefficient for sex was closest to 0 when almost all items were trimmed (>.25SD; 661/680 items trimmed)
+
+```{r old-load-pruned, include=F, eval=F}
 # (2sd threshold)
@@
 
+<!-- OLD
 We identified 76 CDI items that function quite differently across demographic groups, and chose to prune just the 15 items with large difficulty differences for both SES and race. 
@@
 Although these are small reductions in the SES and race effect sizes, we find them striking as we pruned only 2.2% of the CDI:WS items (and only 20% of the 76 extrema we identified), and did not specifically prune items that were biased against the disadvantaged groups (non-white, low-SES).
+-->
 
-```{r demo-ability, include=T, out.width="\\linewidth", fig.env = "figure", fig.pos = "H", fig.align='center', fig.width=3.5, fig.height=7.5, set.cap.width=T, num.cols.cap=1, fig.cap = "Scatterplots showing item difficulty per demographic group along each dimension."}
+```{r demo-ability, include=T, out.width="\\linewidth", fig.env = "figure", fig.pos = "h", fig.align='center', fig.width=3.5, fig.height=7.5, set.cap.width=T, num.cols.cap=1, fig.cap = "Scatterplots showing item difficulty per demographic group along each dimension."}
 
@@
 #     geom_density(alpha=.2) + xlab("Ability")
-# 
-# p2 <- mm_ses %>%
-#   ggplot(aes(x=d_diff)) + 
-#     geom_density(alpha=.2) + xlab("Ability") 
-# 
-# p3 <- mm_eth %>%
-#   ggplot(aes(x=d_diff)) + 
-#     geom_density(alpha=.2) + xlab("Ability")
 
@@
 
-```{r, include=T, out.width="\\linewidth", fig.env = "figure", fig.pos = "H", fig.align='center', fig.width=3.5, fig.height=3.5, set.cap.width=T, num.cols.cap=1, fig.cap = "CHILDES frequency of input to male vs.
+```{r, include=F, out.width="\\linewidth", fig.env = "figure", fig.pos = "h", fig.align='center', fig.width=3.5, fig.height=3.5, set.cap.width=T, num.cols.cap=1, fig.cap = "CHILDES frequency of input to male vs.
 female children."}

### 2022-01-28 paper/cogsci.Rmd
writing pruning section

@@
 It is important, however, to inspect the full plots, for if there is a cluster of items in a GLIMMER, the analyst may conclude that these items are strong candidates for DIF on that dimension.
-For example, there is some clustering at the top and bottom of Figure 2: at the top, "vagina", "tights", "dress (object)", and "doll" form a cluster of items that are much easier for females, while at the bottom, "penis" stands out as much easier for males.
+For example, there is some clustering at the top and bottom of Figure 2: at the top, "vagina", "tights", "dress (object)", and "doll" form a cluster of items that are much more well-known for females, while at the bottom, "penis" stands out as much more well-known by males.
 For the maternal education and race GLIMMERs (Figures 3 and 4)--and in the rest of the full sex GLIMMER, there are not clusters, but rather a continuum of smoothly varying differences with overlap. 
@@
 ```{r sex-glimmer, out.width="\\linewidth", fig.env = "figure", fig.pos = "h", fig.align='center', fig.width=3.2, fig.height=3.5, set.cap.width=T, num.cols.cap=1, fig.cap = "GLIMMER plot of a sample of CDI:WS words from the sex bias model.
-Words at the top are easier to learn for females, while those at the bottom are easier for males."}
+Words at the top are more well-known by females, while those at the bottom are more known for males."}
 #img <- png::readPNG("figs/lab_logo_stanford.png")
@@
 ```{r ses-glimmer, out.width="\\linewidth", fig.env = "figure", fig.pos = "h", fig.align='center', fig.width=3, fig.height=3.5, set.cap.width=T, num.cols.cap=1, fig.cap = "GLIMMER plot of a sample of CDI:WS words from the SES bias model.
-Words at the top are easier to learn for children from low-SES families, while those at the bottom are easier for those from high-SES families."}
+Words at the top are more known by children in low-SES families, while those at the bottom are more known by those in high-SES families."}
 knitr::include_graphics("figs/smGLIMMER_ses_prodWS.pdf")
@@
 
-```{r eth-glimmer, out.width="\\linewidth", fig.env = "figure", fig.pos = "h", fig.align='center', fig.width=3, fig.height=3.5, set.cap.width=T, num.cols.cap=1, fig.cap = "GLIMMER plot of a sample of CDI:WS words from the ethnicity bias model.
-Words at the top are easier to learn for children from low-SES families, while those at the bottom are easier for those from high-SES families."}
+```{r eth-glimmer, out.width="\\linewidth", fig.env = "figure", fig.pos = "h", fig.align='center', fig.width=3, fig.height=3.5, set.cap.width=T, num.cols.cap=1, fig.cap = "GLIMMER plot of a sample of CDI:WS words from the race bias model.
+Words at the top are more known by non-white children, while those at the bottom are more known by white children."}
 knitr::include_graphics("figs/smGLIMMER_eth_prodWS.pdf")
@@
 For sex, the median difficulty difference (male-female) was `r round(median(mm_sex$d_diff),2)` (M=`r round(mean(mm_sex$d_diff),2)`, sd=`r round(sd(mm_sex$d_diff),2)`), with `r length(which(mm_sex$d_diff>0))`/680 items being easier for females than males.
-The fact that the bulk of this distribution favors females seems to confirm that the female language advantage is pervasive across the CDI:WS, and is thus likely to be a real advantage.
+The fact that the bulk of this distribution favors females shows that the female language advantage is pervasive across the CDI:WS, and suggests that it is likely to be a real ability advantage, unless the bulk of these items are actually spoken more often to female children than to male children.
 
@@
 For maternal education (ME), the median difficulty difference (low-high) was `r round(median(mm_ses$d_diff),2)` (M=`r round(mean(mm_ses$d_diff),2)`, sd=`r round(sd(mm_ses$d_diff),2)`), with `r length(which(mm_ses$d_diff>0))`/680 items being easier for high-ME than low-ME children.
-With the mean and median difficulty differences close to 0, and roughly half of the words favoring each ME group, it is tempting to conclude that the CDI:WS items are somewhat balanced with respect to ME.
+With the mean and median difficulty differences close to 0, and roughly half of the words favoring each ME group, it is tempting to conclude that the CDI:WS items are somewhat balanced with respect to ME (and thus SES, by proxy).
 
@@
 Given the smoothly-varying distributions of DIF shown by each demographic factor, we chose to evaluate multiple thresholds for pruning the extreme-valued items from each distribution.
-For each demographic factor, we pruned items with difficulty difference from >.25 SD to >3 SD.
-Shown in Figure 5, the regression coefficient for sex was closest to 0 when almost all items were trimmed (>.25SD; 661/680 items trimmed)
+For each demographic factor, we pruned items with difficulty difference from the mean at varying thresholds, from >.25 SD to >3 SD, in increments of .25 SD. 
+At each SD threshold, a potentially different subset of items are excluded for each model, but with the goal of creating a single CDI that is less biased on all dimensions, we pruned the union of the subsets excluded from each model.
+For example, pruning >2SD from the mean of each model excluded 25 sex-biased items, 39 ME-biased items, and 31 race-biased items, with their union being 76 unique items.
+Figure 5 shows the size of each demographic effect at different exclusion thresholds.
+
+Ideally, we would find a single threshold that minimized the magnitude of coefficients (and thus bias) for all three demographics simultaneously.
+However, 
+The regression coefficient for sex was closest to 0 when almost all items were trimmed (>.25SD; 661/680 items trimmed).
 

### 2022-01-29 paper/cogsci.Rmd
discussified

@@
 # 2SD threshold
-mm_sex <- get_extreme_item_difficulty_differences(mm_sex) # min: -.53 max: 1.06
-mm_ses <- get_extreme_item_difficulty_differences(mm_ses) # min: -1.07 max: 1.13
-mm_eth <- get_extreme_item_difficulty_differences(mm_eth) # min: -.64 max: 1.56
+mm_sex <- get_extreme_item_difficulty_differences(mm_sex, sd_thresh=2.25) # min: -.53 max: 1.06
+mm_ses <- get_extreme_item_difficulty_differences(mm_ses, sd_thresh=2.25) # min: -1.07 max: 1.13
+mm_eth <- get_extreme_item_difficulty_differences(mm_eth, sd_thresh=2.25) # min: -.64 max: 1.56
 
@@
 
+```{r, include=F}
+# find minimal effect sizes on each demographich dimension
+opt_sex = pdat[which(pdat$beta==max(subset(pdat, Dimension=="sex")$beta)),]$beta
+opt_ME = pdat[which(pdat$beta==max(subset(pdat, Dimension=="maternal ed")$beta)),]$beta
+opt_race = pdat[which(pdat$beta==min(subset(pdat, Dimension=="race")$beta)),]$beta
+
+prune_dat <- prune_dat %>%
+  mutate(SSE = (sex_beta-opt_sex)^2 + (ses_beta-opt_ME)^2 + (eth_beta-opt_race)^2,
+         pct_overlap = extreme_total / (extreme_ses + extreme_eth + extreme_sex)) %>%
+  arrange(SSE)
+```
+
 Given the smoothly-varying distributions of DIF shown by each demographic factor, we chose to evaluate multiple thresholds for pruning the extreme-valued items from each distribution.
@@
 Ideally, we would find a single threshold that minimized the magnitude of coefficients (and thus bias) for all three demographics simultaneously.
-However, 
-The regression coefficient for sex was closest to 0 when almost all items were trimmed (>.25SD; 661/680 items trimmed).
+Unfortunately, the effect size for each demographic variable was smallest (closest to 0) at different exclusion thresholds. 
+The effect size of sex was smallest when almost all items were trimmed (>0.25 SD; $\beta_{sex} = -.47$; 661/680 total items trimmed; 433 due to sex extremity).
+The effect size of maternal education (SES) was smallest when items >1.25SD in difficulty difference were trimmed ($\beta_{ME} = -.17$; 234 total items trimmed; 126 due to ME extremity). 
+The effect size of race was minimized when items more extreme than 0.5 SD were trimmed ($\beta_{race} = .42$; 577 total items trimmed; 425 due to race extremity).
+
+Where is the exclusion threshold that jointly minimizes the effect sizes of these demographic variables? 
+The strict exclusion of all items with >.25 SD difficulty difference -- 661/680 (97%) of the CDI:WS -- best optimized this, but seems too extreme a culling.
+The next best thresholds are >.5 SD -- again, too extreme -- or 3 SD, a remarkably lax criterion that only excludes 19 items in total (11 due to sex bias).
+Only slightly worse than these is >2.25 SD, which excludes 59 items (9% of the CDI:WS), and shows a modest effect of both race ($\beta=0.48$) and maternal education ($\beta=.19$), and a near-median effect size for sex ($\beta=-.59$).
+We characterize these 59 items below, and consider whether we might recommend pruning them from the CDI:WS.
+
+## Characterizing biased items
+
+Considering just those items whose difficulty was >2.25 SD from the mean difficulty difference on each dimension identified 22 items with extreme sex-based difficulty differences, only 7x of which were easier for females, with the other 18x items favoring males.
+For maternal education, 27 extrema were identified, only 13x of which were easier for low-SES children.
+For race, 21 extrema were identified, only 11x of which were easier for non-white children.
+Thus, while if anything the bulk of the extrema in the sex-based model advantaged the lower language ability group (males), the extrema in the other two models mostly favored the advantaged groups: 26/39x extreme items in the SES model favored high-SES children, and 20/31x items in the race model favored white children.
+17 CDI items were identified as extrema in more than one model: "tractor" and "vroom" were extreme in all three models, "choo choo" was extreme in both SES and sex models, "give me five!" was extreme in both race and sex models, and 13 other items were extreme in both SES and race models ("grrr", "moo", "quack quack", "uh oh", "duck", "owl", "candy", "gum", "walker", "daddy", "pet's name", "up", and "so").
+[Count by lexical class?]
 

### 2022-01-30 paper/cogsci.Rmd
making figures horizontal

@@
 
-```{r baseline-ability-vs-age, out.width="\\linewidth", fig.env = "figure", fig.pos = "h", fig.align='center', fig.width=3.3, fig.height=6, set.cap.width=T, num.cols.cap=1, fig.cap = "Language ability vs.
-age by demographic group, from the Rasch model."}
+```{r baseline-ability-vs-age-hor, fig.env = "figure*", out.width="\\linewidth", fig.pos = "h", fig.align='center', fig.width=6.5, fig.height=2.8, set.cap.width=T, num.cols.cap=2, fig.cap = "Language ability vs.
+age by demographic group, from the baseline Rasch model."}
 p1 <- d_demo %>% filter(!is.na(sex)) %>%
@@
   xlab("Age (months)") + ylab("Language Ability") + 
-  scale_color_discrete(type=sex_colors)
-  #theme(legend.position = "top")
+  scale_color_discrete(type=sex_colors) +
+  theme(legend.position = "top")
 
@@
   xlab("Age (months)") + ylab("Language Ability") +
-  scale_color_discrete(name = "SES", type=ses_colors) #+ 
-  #theme(legend.position = "top")
+  scale_color_discrete(name = "SES", type=ses_colors) + 
+  theme(legend.position = "top")
 
@@
   xlab("Age (months)") + ylab("Language Ability") + 
-  scale_color_discrete(name = "ethnicity", type=eth_colors) #+ 
-  #theme(legend.position = "top")
+  scale_color_discrete(name = "ethnicity", type=eth_colors) + 
+  theme(legend.position = "top")
 
-ggarrange(p1, p2, p3, ncol=1)
+ggarrange(p1, p2, p3, ncol=3, nrow=1)
 ```
@@
 
-Figures 2-4 show GLIMMERs for a selection of CDI items for sex, SES, and race, respectively.
+Figure 2 shows GLIMMERs for a selection of CDI items for sex (left), maternal education (middle), and race (right).
 The full GLIMMERs, with all 680 CDI:WS items, are available on [OSF](https://osf.io/57rsw/?view_only=2b6ecb61fe08458293af7421d276932a), but were too large to include here.
@@
 
-```{r sex-glimmer, out.width="\\linewidth", fig.env = "figure", fig.pos = "h", fig.align='center', fig.width=3.2, fig.height=3.5, set.cap.width=T, num.cols.cap=1, fig.cap = "GLIMMER plot of a sample of CDI:WS words from the sex bias model.
+```{r glimmer-combo, fig.env = "figure*", out.width="\\linewidth", fig.pos = "h", fig.width=6.5, fig.align = "center", set.cap.width=T, num.cols.cap=2, fig.cap = "GLIMMER plot of a sample of CDI:WS words from the sex bias model.
 Words at the top are more well-known by females, while those at the bottom are more known for males."}
-#img <- png::readPNG("figs/lab_logo_stanford.png")
-#grid::grid.raster(img)
+knitr::include_graphics("figs/smGLIMMER_combo.pdf")
+```
+
+```{r sex-glimmer, include=F, out.width="\\linewidth", fig.env = "figure", fig.pos = "H", fig.align='center', fig.width=3.2, fig.height=3.5, set.cap.width=T, num.cols.cap=1, fig.cap = "GLIMMER plot of a sample of CDI:WS words from the sex bias model.
+Words at the top are more well-known by females, while those at the bottom are more known for males."}
 knitr::include_graphics("figs/smGLIMMER_sex_prodWS.pdf")
@@
 
-```{r ses-glimmer, out.width="\\linewidth", fig.env = "figure", fig.pos = "h", fig.align='center', fig.width=3, fig.height=3.5, set.cap.width=T, num.cols.cap=1, fig.cap = "GLIMMER plot of a sample of CDI:WS words from the SES bias model.
+```{r ses-glimmer, include=F, out.width="\\linewidth", fig.env = "figure", fig.pos = "H", fig.align='center', fig.width=3, fig.height=3.5, set.cap.width=T, num.cols.cap=1, fig.cap = "GLIMMER plot of a sample of CDI:WS words from the SES bias model.
 Words at the top are more known by children in low-SES families, while those at the bottom are more known by those in high-SES families."}
@@
 
-```{r eth-glimmer, out.width="\\linewidth", fig.env = "figure", fig.pos = "h", fig.align='center', fig.width=3, fig.height=3.5, set.cap.width=T, num.cols.cap=1, fig.cap = "GLIMMER plot of a sample of CDI:WS words from the race bias model.
+```{r eth-glimmer, include=F, out.width="\\linewidth", fig.env = "figure", fig.pos = "H", fig.align='center', fig.width=3, fig.height=3.5, set.cap.width=T, num.cols.cap=1, fig.cap = "GLIMMER plot of a sample of CDI:WS words from the race bias model.
 Words at the top are more known by non-white children, while those at the bottom are more known by white children."}
@@
 
-```{r pruned-coefs, out.width="\\linewidth", fig.env = "figure", fig.pos = "h", fig.align='center', fig.width=3.4, fig.height=4.75, set.cap.width=T, num.cols.cap=1, fig.cap = "Size of demographic effects (regression coefficients) with different pruning thresholds."}
+```{r pruned-coefs-hor, fig.env = "figure*", out.width="\\linewidth", fig.pos = "h", fig.align='center', fig.width=6.5, fig.height=2.8, set.cap.width=T, num.cols.cap=2, fig.cap = "Size of demographic effects (regression coefficients) with different pruning thresholds."}
 load(here("data/pruned_demo_coefs.Rdata"))
-
+# horizontal version
 pdat <- prune_dat %>% rename(c(beta_sex=sex_beta, beta_ses=ses_beta, beta_eth=eth_beta)) %>%
@@
 
-ggarrange(pr1, pr2, pr3, ncol=1, common.legend = T)
+ggarrange(pr1, pr2, pr3, nrow=1, ncol=3, common.legend = T)
 ```
@@
 For example, pruning >2SD from the mean of each model excluded 25 sex-biased items, 39 ME-biased items, and 31 race-biased items, with their union being 76 unique items.
-Figure 5 shows the size of each demographic effect at different exclusion thresholds.
+Figure 3 shows the demographic effects ($\beta$s) at different exclusion thresholds.
 
@@
 
-```{r demo-ability, include=T, out.width="\\linewidth", fig.env = "figure", fig.pos = "h", fig.align='center', fig.width=3.5, fig.height=7.5, set.cap.width=T, num.cols.cap=1, fig.cap = "Scatterplots showing item difficulty per demographic group along each dimension."}
+```{r demo-ability, include=F, out.width="\\linewidth", fig.env = "figure", fig.pos = "h", fig.align='center', fig.width=3.5, fig.height=7.5, set.cap.width=T, num.cols.cap=1, fig.cap = "Scatterplots showing item difficulty per demographic group along each dimension."}
 

### 2022-01-31 paper/cogsci.Rmd
discussion rewrite, pruned items Venn diagram

@@
                      subset(mm_eth, extreme)) 
-# length(unique(all_extreme$definition)) 
+# length(unique(all_extreme$definition)) # 59
 # sort(table(all_extreme$definition)) # candy, choo choo, daddy, duck, ... tractor, vroom
 
-# overlap between extreme SES and race items:
-#intersect(subset(mm_ses, extreme)$definition, subset(mm_eth, extreme)$definition)
+# overlaps extreme SES and race items:
+# intersect(subset(mm_ses, extreme)$definition, subset(mm_eth, extreme)$definition)
+# intersect(subset(mm_sex, extreme)$definition, subset(mm_eth, extreme)$definition)
+# intersect(subset(mm_sex, extreme)$definition, subset(mm_ses, extreme)$definition)
+#subset(all_extreme, group1=="Nonwhite")
 ```
@@
 Where is the exclusion threshold that jointly minimizes the effect sizes of these demographic variables? 
-The strict exclusion of all items with >.25 SD difficulty difference -- 661/680 (97%) of the CDI:WS -- best optimized this, but seems too extreme a culling.
+The strict exclusion of all items with >.25 SD difficulty difference -- 661/680 (97%) of the CDI:WS -- best optimized this, but seems too extreme of a culling.
 The next best thresholds are >.5 SD -- again, too extreme -- or 3 SD, a remarkably lax criterion that only excludes 19 items in total (11 due to sex bias).
 Only slightly worse than these is >2.25 SD, which excludes 59 items (9% of the CDI:WS), and shows a modest effect of both race ($\beta=0.48$) and maternal education ($\beta=.19$), and a near-median effect size for sex ($\beta=-.59$).
-We characterize these 59 items below, and consider whether we might recommend pruning them from the CDI:WS.
+We characterize these 59 potentially biased items below, and consider whether we might recommend pruning them from the CDI:WS.
 
@@
 
-Considering just those items whose difficulty was >2.25 SD from the mean difficulty difference on each dimension identified 22 items with extreme sex-based difficulty differences, only 7x of which were easier for females, with the other 18x items favoring males.
-For maternal education, 27 extrema were identified, only 13x of which were easier for low-SES children.
-For race, 21 extrema were identified, only 11x of which were easier for non-white children.
-Thus, while if anything the bulk of the extrema in the sex-based model advantaged the lower language ability group (males), the extrema in the other two models mostly favored the advantaged groups: 26/39x extreme items in the SES model favored high-SES children, and 20/31x items in the race model favored white children.
-17 CDI items were identified as extrema in more than one model: "tractor" and "vroom" were extreme in all three models, "choo choo" was extreme in both SES and sex models, "give me five!" was extreme in both race and sex models, and 13 other items were extreme in both SES and race models ("grrr", "moo", "quack quack", "uh oh", "duck", "owl", "candy", "gum", "walker", "daddy", "pet's name", "up", and "so").
-[Count by lexical class?]
+Figure 4 shows the 59 items showing extreme bias (i.e., whose difficulty was >2.25 SD from the mean difficulty difference on one or more demographic dimension).
+There were 22 items with extreme sex-based difficulty differences, only 7 of which were more known by females, with the other 15 items favoring males.
+All but one of the sex-biased items are nouns, many of which are stereotypically associated with one gender more than the other, including stereotypically-male professions (e.g., "fireman").
+For maternal education, 27 extrema were identified, only 10 of which were biased for low-SES children.
+As for sex, most of the ME-biased items are nouns (15/27). 
+Notably, many of the items more known by high-ME children were animals ("zebra", "owl") and animal sounds ("baa baa", "moo").
+Items more known by low-ME children include a few sweet treats: "candy", "soda/pop", and "gum".
+For race, 21 extrema were identified, only 8 of which were easier for non-white children.
+The lexical class of the race-biased items was more varied: 10 were nouns, but many others were early-learned words and phrases (e.g., "up", "all gone", "uh-oh", "bye", "mommy").
+Only 10 of the 59 items were extreme in more than one model, with most of the overlap being between race and ME (8 items).
+Only 1 item ("vroom") was extreme for all three demographics.
 
@@
 ```
-
-<!-- OLD
-We identified 76 CDI items that function quite differently across demographic groups, and chose to prune just the 15 items with large difficulty differences for both SES and race. 
-After pruning just these 15 items from the CDI:WS and re-fitting a Rasch IRT model, regressions predicting ability with each demographic variable (and age, centered) showed reduced main effects of SES ($\beta=0.20$ vs.
-$\beta=0.23$) and race ($\beta=0.48$ vs.
-$\beta=0.50$).
-Although these are small reductions in the SES and race effect sizes, we find them striking as we pruned only 2.2% of the CDI:WS items (and only 20% of the 76 extrema we identified), and did not specifically prune items that were biased against the disadvantaged groups (non-white, low-SES).
--->
 
@@
 
+```{r biased-words, include=T, out.width="\\linewidth", fig.env = "figure", fig.pos = "H", fig.align='center', fig.width=3.5, fig.height=3.2, set.cap.width=T, num.cols.cap=1, fig.cap = "The 59 items showing extreme bias (difficulty difference $>2.25$ SD) for one or more demographic."}
+knitr::include_graphics("figs/extrema2p25sd.pdf")
+```
+
 ## Relating child-directed speech to demographic bias
@@
 Indeed, input quantity (total daily tokens) and some measures of quality (e.g., lexical diversity: ratio of word types vs.
-tokens) have been predictive of language learning outcomes in some demographic studies [@rowe2009differences; @huttenlocher1991early].
-Here, we investigated the extent to which word frequency in child-directed speech to male vs. female children was predictive of the amount of DIF shown by CDI items. 
+tokens) have often been predictive of language learning outcomes in  demographic studies [@rowe2009differences; @huttenlocher1991early].
+We used a corpus-based analysis to investigate the extent to which word frequency in child-directed speech to male vs.
+female children was predictive of the amount of sex-based bias shown by CDI items.
 Similar to the approach taken by @braginsky2016gender, we used the CHILDES corpus of transcripts from dyadic play sessions [@macwhinney2000childes], which are labeled with the sex of the target child, but not other demographic variables.
-(Number of males, number of females, total word counts, a couple examples...)
-
-Overall, the correlation between the proportion of times a word was spoken to a female child and the size of the female (vs.
-male) advantage for that CDI word was modest, but significant ($r=0.19$, $p<.001$).
+We found a total of 5213750 female-directed tokens and 6091950 male-directed tokens that matched 662 of the 680 CDI:WS items, normalized the word frequencies to tokens per million for each target sex, and calculated the log-odds of each word being spoken to a female (i.e., $log(p_{f} / p_{m}$), meaning unbiased words will have log-odds of 0, while those spoken more often to females will have log-odds $>0$, and those spoken more often males will have log-odds $<0$.
+For example "doll" was spoken 156 times to females, and 138 times to males, thus $p_{f} = 156/(156+138) = .53$, $p_{m} = 1-p_{f} = 0.47$, and $log(p_{f}/p_{m})$ = 0.12), while "police" was spoken 138 times to females, and 225 times to males, and thus has log-odds = -0.49).
+Overall, the correlation between the log-odds of a word being spoken to a female vs.
+a male child and the size of the female (vs.
+male) advantage for that CDI word was modest, but significant ($r=0.18$, $p<.001$), suggesting that some of the sex bias seen in CDI items is explained by differences in language input received by girls vs.
+boys.
 
@@
 type_counts_smoothed = readRDS("../data/type_counts_smoothed.rds")
+
+# total tokens:
+type_counts %>% group_by(target_child_sex) %>% 
+  summarise(total = sum(count))
 
@@
 
-#cor.test(tmp$propFemale, tmp$d_diff) # .19, p<.001
-
+# subset(type_counts_smoothed, uni_lemma=="police")
+#cor.test(log(tmp$FtoM), tmp$d_diff) # .18
+#subset(type_counts_smoothed, uni_lemma=="police")
 ```
@@
 
-We investigated the CDI:WS, a popular parent-report measure of children's early vocabulary, for potential demographic bias, examining the distribution of words' estimated difficulties for high- vs.
-low-SES children, females vs.
+We investigated the CDI:WS, a popular parent-report measure of children's early vocabulary, for potential demographic bias, examining the distribution of words' estimated difficulties for children of mothers with high- vs.
+low-education children (a proxy for family SES), females vs.
 males, and for white vs.
@@
 For the rest of the items, and for SES- and race-based analysis, there was a smooth continuum of DIF, making the boundary of true DIF subjective, as this would rely on knowing the true difference in language ability between groups--which we reciprocally estimate from instruments like the CDI.
-To move forward, we identified candidate DIF items by looking at the extremes of each distribution, and found that for SES and race, the majority of these extrema (66% in both cases) were easier for the majority demographic group (i.e., high-SES, white), meaning that if these items were removed, SES and race effects on language ability as measured by the CDI:WS would decrease.
-In contrast, the majority of the extrema in the sex-based DIF analysis were easier for the disadvantaged group (males): if these extrema are removed, the female language advantage on the CDI:WS will likely increase.
-To confirm these implications, we pruned a very small number of CDI items (15 of 680) that were DIF extrema in both SES- and race-based analyses, and found that the effect of these dimensions on language ability was reduced.
+To move forward, we identified candidate DIF items systematically pruning the extremes of each distribution, excluding outliers for each demographic across a range of thresholds, and seeking to minimize the size of demographic effects.
+Although many exclusion thresholds decreased the size of maternal education and race effects, the female advantage actually grew under most prunings, with the majority of excluded extrema being stereotypically-male nouns (e.g, "truck", "fireman").
 
-## The difficulty of DIF
+This analysis highlights a fundamental difficulty of identifying DIF: an analyst must either know the expected magnitude of an ability difference between the two demographic groups, or know a set of "anchor" items that are equally difficult (i.e., unbiased) for both groups [for an overview, see @stenhaug2021treading].
+But in the universe of children's early language, there is only a finite set of early-learned words to choose from--and we may expect many of them to be biased for various environmental reasons (e.g., boys and girls know their respective genitalia; children in Florida may not use mittens or skis).
+Hence, the presence of DIF on a set of items could indicate a degree of varying linguistic input, rather than an actual advantage for one demographic over another.
 
-However, DIF is fundamentally difficult to identify for multiple reasons [for an overview, see @stenhaug2021treading].
-First, most techniques to identify DIF rely on defining a set of "anchor" test items that are assumed to be equally difficult (i.e., unbiased) for both groups of interest.
-Identifying anchor items is at best fraught when there may in fact be a difference in ability between groups (e.g., the female language advantage), and is further confounded when the magnitude of this ability difference is unknown.
-A reason that DIF is particularly tricky in measuring children's early language ability is that there is a finite universe of early-learned words to choose from--and we may expect many of them to be biased for various environmental reasons (e.g., children in Florida may not use mittens or skis).
-Hence, the presence of DIF on a wide variety of items may not indicate the presence of bias; it could indicate that one demographic has a higher average ability level than the other.
-For example, let male language ability be drawn from a standard normal $\mu_{male} ~ N(0,1)$, with female language ability slightly higher, on average ($\mu_{female} = \mu_{male} + 0.1$).
+Choosing an exclusion threshold that struck a balance of reducing race- and ME-based bias and not eliminating too many items, we identified 59 suspect items--the majority of which favored white/high-ME groups.
+These extrema were predominantly nouns, including many animals (esp. for maternal education), vehicles and professions (esp. for sex). 
+Notably, no verbs and few syntactically complex items were identified as outliers, which may suggest that these lexical classes are less prone to environmental influence, and perhaps more generally unbiased.
+<!--or something..seems important, as nouns are more referential/context-dependent-->
+Pruning these 59 outliers would decrease the demographic effects of maternal education (unpruned $\beta=.23$; pruned $\beta=.19$) and race (unpruned $\beta=.50$; pruned $\beta=.48$).
+In contrast, the majority of the sex-biased outliers were easier for the disadvantaged group (males), meaning that the removal of these extrema increased the female language advantage on the CDI:WS (unpruned $\beta=0.56$; pruned $\beta=.59$).
+In aggregate, these analyses seem to suggest that the CDI:WS may slightly overestimate the magnitude of differences in language ability along the dimensions of maternal education and race, and perhaps underestimate the female advantage.
+
+This investigation is only a first step in measuring demographic bias in the items on the CDI:WS.
+More research is needed to determine which CDI items that show DIF are more subject to environmental influence, as we found only a modest association between the relative frequency of words in child-directed speech to boys vs.
+girls and the amount of sex-based DIF.
+Examining the relative frequency of CDI words across households with varying maternal education and race may reveal similarly weak associations, or there may be strong associations only among particular categories of words.
+It may be that particular activity contexts--popular with some demographic groups, and not others--may be predictive. 
+For example, many of the extrema favoring high-ME children are animals and animal sounds (e.g., "quack quack", "woof woof", "baa baa", "duck", "sheep", "giraffe", "zebra"): do high-SES households visit the zoo more often, or perhaps engage in other activities related to naming animals and noises they make (e.g., animal noise toys)?
+If certain activities are driving early word learning for high-ME children, what are the activities (and associated vocabulary) that low-SES households are instead engaging in?
+Note, too, that SES, race, and sex are only three of many possible demographic dimensions of interest.
+For example, geographic region is likely predictive of children's early experience with (and thus knowledge of) CDI items related to winter weather (e.g., "snow", "mittens") or outdoors activities (e.g.
+"camping").
+A truly fair test of children's early vocabulary would contain a representative and balanced sample of words from all activities that children engage in, across demographic groups.
+
+<!--However, the best way to estimate endogenous differences in ability (rather than exogenous differences, e.g.
+in the language environment) would be to conduct controlled, in-lab experiments measuring children's ability to learn novel words.-->
+
+<!-- For example, let male language ability be drawn from a standard normal $\mu_{male} ~ N(0,1)$, with female language ability slightly higher, on average ($\mu_{female} = \mu_{male} + 0.1$).
 Then we would expect items to be an average of $0.1$ easier for females than for males, and we might identify items that are instead easier for males than females as showing undesirable DIF.
 And yet, without knowing the actual ability difference between two demographic groups--for which we also rely upon our tests, it is difficult to adjudicate which items show DIF, and which do not.
-
-This investigation is only a first step in measuring demographic bias in the items on the CDI:WS.
-In aggregate, it seems to suggest that we may be slightly overestimating the magnitude of differences in language ability along the dimensions of SES and race, and perhaps underestimating the female advantage.
-However, the best way to estimate endogenous differences in ability (rather than exogenous differences, e.g.
-in the language environment) would be to conduct controlled, in-lab experiments measuring children's ability to learn novel words.
-Future research may also look into the specific items that are easier for one group than another, and determine whether these differences can be accounted for by different environmental contexts. 
-For example, it is striking that many of the extrema favoring high-SES children are animals and animal sounds (e.g., "grrr", "quack quack", "woof woof", "baa baa", "duck", "sheep", "giraffe", "zebra"): do high-SES households visit the zoo more, or do they often engage in other activities related to naming animals and noises they make?
-If certain high-SES activities are driving early word learning for these children, what are the activities (and associated vocabulary) that low-SES households are instead engaging in?
-A truly fair test of children's early vocabulary would contain a representative sample of words from all activities that children engage in, across demographic groups.
+-->
 

### 2022-02-01 paper/cogsci.Rmd
minor edits

@@
     often by children of varying backgrounds. This study used a database of American English 
-    CDIs to identify words that showed strong bias for particular demographic groups of children, on 
+    CDIs to identify words demonstrating strong bias for particular demographic groups of children, on 
     dimensions of sex (male vs. female), race (white vs. non-white), and maternal education 
@@
 keywords: >
-    language acquisition; word learning; measuring instrument bias; demographics;
+    language acquisition; word learning; measuring instrument bias; demographic bias
 
@@
 While it is possible that there is an underlying difference in mean language ability between demographic groups, and that cross-linguistic variation is driven by differences in the language environments of different cultures, could it also be possible that the tests are biased to some degree--and perhaps to a different degree in different languages?
-The idea that some items on a test may show bias, favoring one group over another, is known in psychometrics as Differential Item Function (DIF; @holland1993differential]. 
+The idea that some items on a test may show bias, favoring one group over another, is known in psychometrics as Differential Item Function [DIF; @holland1993differential]. 
 The validity of any test with many items favoring one group over another (say, children from rural vs.
-urban households) is questionable, as the test may be overestimating the ability of test-takers in the former (rural) group, and/or underestimating the ability of the latter group -- and not because of any underlying mean difference in ability between the groups, but simply because the test is unfair [@camilli2013ongoing].
+urban households) is questionable, as the test may be overestimating the ability of test-takers in the former (rural) group, and/or underestimating the ability of the latter group -- not because of any underlying mean difference in ability between the groups, but simply because the test is unfair [@camilli2013ongoing].
 For example, if a vocabulary test is composed almost entirely of the names of farm equipment, the test will underestimate the knowledge of urban children, who have experience of other contexts, despite being less familiar with farm equipment than their rural peers.
@@
 
-We analyze parent-reported Wordbank data from `r nrow(d_demo)` American English CDI: Words & Sentences administrations for children `r min(d_demo$age)` to `r max(d_demo$age)` months of age [@frank2017; @frank2021].
+We analyzed parent-reported Wordbank data from `r nrow(d_demo)` American English CDI: Words & Sentences administrations for children `r min(d_demo$age)` to `r max(d_demo$age)` months of age [@frank2017; @frank2021].
 Full demographic data are not reported in some datasets contributed to Wordbank: sex was available for `r nrow(d_demo)-sum(is.na(d_demo$sex))` children, race/ethnicity was available for `r nrow(d_demo)-sum(is.na(d_demo$ethnicity))` children, and maternal education (a proxy for socioeconomic status; SES) was available for `r nrow(d_demo)-sum(is.na(d_demo$ses_group))` children.
@@
 The analysis of race-based differences included data from `r eth_tab["White"]` white, `r eth_tab["Asian"]` Asian, `r eth_tab["Black"]` Black, `r eth_tab["Hispanic"]` Hispanic, and `r eth_tab["Other"]` "Other" children.
-Due to sparse data for many categories, we binarized participants' race/ethnicity as White (`r eth2_tab["White"]`) or Non-white (`r eth2_tab["Nonwhite"]`), recognizing that there may be important variation between groups that this will fail to capture.
-Data for the maternal education analysis included CDI data from children whose mother's had the following levels of education: `r momed_tab["Primary"]` with no more than primary school education, `r momed_tab["Some Secondary"]` with some secondary school, `r momed_tab["Secondary"]` with no more than secondary school, `r momed_tab["Some College"]` with some college, `r momed_tab["College"]` with no more than a college degree, `r momed_tab["Some Graduate"]` with no more than some graduate school, and `r momed_tab["Graduate"]` with a graduate degree.
-Again due to data sparsity, we binarized the `r ses_tab["high"]` children whose mothers had at least some college or more as high maternal education (high-ME), and those whose mothers had at most high school (`r ses_tab["low"]` children) as low maternal education (low-ME).
+Due to sparse data for many categories, we dichotomized participants' race/ethnicity as White (`r eth2_tab["White"]`) or Non-white (`r eth2_tab["Nonwhite"]`), recognizing that there may be important variation between groups that this will fail to capture.
+Data for the maternal education analysis included CDI data from children whose mothers' had the following levels of education: `r momed_tab["Primary"]` with no more than primary school education, `r momed_tab["Some Secondary"]` with some secondary school, `r momed_tab["Secondary"]` with no more than secondary school, `r momed_tab["Some College"]` with some college, `r momed_tab["College"]` with no more than a college degree, `r momed_tab["Some Graduate"]` with no more than some graduate school, and `r momed_tab["Graduate"]` with a graduate degree.
+Due to data sparsity, we again dichotomized the `r ses_tab["high"]` children whose mothers had at least some college or more as high maternal education (high-ME), and those whose mothers had at most high school (`r ses_tab["low"]` children) as low maternal education (low-ME).
 
@@
 
-First, we will examine the size of demographic effects on language ability in a baseline Rasch model fitted without regard to demographic group.
-Then we will fit a multigroup Rasch model for each demographic factor, and characterize the between-group differences in item difficulties.
-Next, we will re-examine the size of demographic effects after pruning biased items from the CDI:WS using a varying threshold.
+First, we examine the size of demographic effects on language ability in a baseline Rasch model fitted without regard to demographic group.
+Then we fit a multigroup Rasch model for each demographic factor, and characterize the between-group differences in item difficulties.
+Next, we re-examine the size of demographic effects after pruning biased items from the CDI:WS using different criteria, and recommend a threshold.
 Finally, we measure the strength of association between sex-related differences in language input and the degree of sex-related DIF.

### 2022-02-01 paper/cogsci.Rmd
tweaking

--- 2022-01-31 paper/cogsci.Rmd
@@
     Children's early language skill has been linked to later educational outcomes, making
-    it important to accurately measure early language. Parent-reported instruments such as the 
-    Communicative Development Inventories (CDIs) have been shown to provide valid, consistent 
+    it important to measure early language accurately. Parent-reported instruments such as the 
+    Communicative Development Inventories (CDIs) have been shown to provide reliable and valid
     measures of children's aggregate early language skill. However, CDIs contain
@@
     we investigated how well the relative frequency of words spoken to young girls vs. boys predicted 
-    sex-based word learning bias, and discuss possible sources of demographic bias in early word learning.
+    sex-based word learning bias, and discuss possible sources of demographic differences in early word learning.
 
@@
 For example, @bornstein2016stability found stability in core language skills across 10 years of children's development, despite changes in maternal income and education over the study period.
-Yet maternal education, often used as a proxy for socioeconomic status (SES), has also been found to be associated with children's language processing and vocabulary by 18 months [@fernald2013ses], and to be predictive of later educational outcomes [@marchman2008speed; see @schwab2016 for a review].
-Other demographic factors are also predictive of language skill: first-born children tend to outpace their siblings, and female children tend to have better language skills than their age-matched male counterparts [@frank2021; @eriksson2012differences] -- a sex-based verbal advantage that continues through high school [see @petersen2018gender for a review].
 
-However, it is also difficult to get a complete measure of young children's language skill: long recordings are prohibitively difficult to collect and transcribe [see the singular exception: @roy2015predicting], and yet any short recording (e.g., a 1-hour play session) will elicit only a small proportion of the words and constructions that children know.
+Yet there are consistent demographic predictors of early vocabulary.
+Maternal education, often used as a proxy for socioeconomic status (SES), is associated with children's language processing and vocabulary by 18 months [@fernald2013ses], and is predictive of later educational outcomes [@marchman2008speed; see @schwab2016 for review].
+Other factors are also predictive of language skill: first-born children tend to outpace their siblings, and female children tend to have better language skills than their age-matched male counterparts [@frank2021; @eriksson2012differences] -- a sex-based verbal advantage that continues through high school [see @petersen2018gender for a review].
+
+One challenge in comprehensive assessment of these differences is that is difficult to get a complete measure of young children's language skill.
+Long recordings are prohibitively difficult to collect and transcribe, and yet any short recording (e.g., a 1-hour play session) will elicit only a small proportion of the words and constructions that children know.
 Thus, researchers of early word learning have constructed tests with hundreds of words, intentionally oversampling words that are more likely to be known by young children.
 
-We focus on the MacArthur-Bates Communicative Development Inventories [CDIs; @fenson1994; @Fenson2007], a set of parent-reported measures of children's productive and receptive language skills, which offer a low-cost and reliable way to estimate children's early language skills [@fenson1994].
+We focus here on the MacArthur-Bates Communicative Development Inventories [CDIs; @fenson1994; @Fenson2007], a set of parent-reported measures of children's productive and receptive language skills, which offer a low-cost and reliable way to estimate children's early language skills [@fenson1994].
 CDIs have shown good predictive validity [e.g., @fenson1994; @bornstein2012stability; @duff2015infant].
@@
 
-While many languages show similar demographic effects on early word learning [e.g., the female advantage: @eriksson2012differences; and the first-born advantage: @frank2021], the size of these demographic effects vary across languages [see @frank2021, Ch.
-6].
-Why might we see such variation?
-While it is possible that there is an underlying difference in mean language ability between demographic groups, and that cross-linguistic variation is driven by differences in the language environments of different cultures, could it also be possible that the tests are biased to some degree--and perhaps to a different degree in different languages?
-The idea that some items on a test may show bias, favoring one group over another, is known in psychometrics as Differential Item Function (DIF; @holland1993differential]. 
-The validity of any test with many items favoring one group over another (say, children from rural vs.
-urban households) is questionable, as the test may be overestimating the ability of test-takers in the former (rural) group, and/or underestimating the ability of the latter group -- and not because of any underlying mean difference in ability between the groups, but simply because the test is unfair [@camilli2013ongoing].
-For example, if a vocabulary test is composed almost entirely of the names of farm equipment, the test will underestimate the knowledge of urban children, who have experience of other contexts, despite being less familiar with farm equipment than their rural peers.
-Of course, there may in fact be some meaningful ability difference between demographic groups, as is likely the case for the female language advntage, which has been documented in many languages [@eriksson2012differences; @frank2021].
-However, these demographic differences in early language learning have not been investigated as potentially due to DIF.
-That is the contribution of the present study.
-Specifically, our goal is to test the 680 vocabulary items on the American CDI:WS for DIF along three dichotomized demographic dimensions: sex (male vs.
-female), maternal education (no more than secondary vs.
-at least some college), and race (white vs.
-non-white).
-
-<!-- A variety of statistical methods for detecting DIF have been proposed, and investigations have in several instances identified DIF for many items on tests favoring one group over another. -->
-
-The outline of this paper is as follows.
-First, we introduce the Wordbank data and the IRT model, and examine the overall size of demographic differences in early word learning based on the full CDI:WS.
-We then fit the IRT model to both groups for each demographic factor, and look at the item parameters for evidence of DIF, noting in particular how many items are significantly biased in favor of each group, and describing them qualitatively.
-Next, we use these item parameters to systematically prune biased items from the CDI:WS, and examine how the estimated size of demographic differences change as we prune more items.
-<!-- We identify a set of suspect items to eliminate (or in the future, replace), and provide updated estimates of the effect size of these demographic variables, were the biased items to be eliminated.-->
-Finally, using a corpus of child-directed speech, we examine the extent to which variation in word frequencies heard by boys vs. girls predicts sex-based DIF.
-Based on these analyses, we provide recommendations for next steps to be taken to identify and potentially replace items on the CDI showing demographic bias.
-
-# Methods
-
-## Vocabulary Data
-
 ```{r import-data}
@@
 ```
-
-### Participants
-
-```{r, echo=F}
-sex_tab <- table(d_demo$sex)
-ses_tab <- table(d_demo$ses_group)
-eth_tab <- table(d_demo$ethnicity)
-eth2_tab <- table(d_demo$eth_group)
-momed_tab <- table(d_demo$mom_ed)
-```
-
-We analyze parent-reported Wordbank data from `r nrow(d_demo)` American English CDI: Words & Sentences administrations for children `r min(d_demo$age)` to `r max(d_demo$age)` months of age [@frank2017; @frank2021].
-Full demographic data are not reported in some datasets contributed to Wordbank: sex was available for `r nrow(d_demo)-sum(is.na(d_demo$sex))` children, race/ethnicity was available for `r nrow(d_demo)-sum(is.na(d_demo$ethnicity))` children, and maternal education (a proxy for socioeconomic status; SES) was available for `r nrow(d_demo)-sum(is.na(d_demo$ses_group))` children.
-
-The analysis of sex-based differences included CDI administrations from `r sex_tab["Female"]` female and `r sex_tab["Male"]` male children.
-The analysis of race-based differences included data from `r eth_tab["White"]` white, `r eth_tab["Asian"]` Asian, `r eth_tab["Black"]` Black, `r eth_tab["Hispanic"]` Hispanic, and `r eth_tab["Other"]` "Other" children.
-Due to sparse data for many categories, we binarized participants' race/ethnicity as White (`r eth2_tab["White"]`) or Non-white (`r eth2_tab["Nonwhite"]`), recognizing that there may be important variation between groups that this will fail to capture.
-Data for the maternal education analysis included CDI data from children whose mother's had the following levels of education: `r momed_tab["Primary"]` with no more than primary school education, `r momed_tab["Some Secondary"]` with some secondary school, `r momed_tab["Secondary"]` with no more than secondary school, `r momed_tab["Some College"]` with some college, `r momed_tab["College"]` with no more than a college degree, `r momed_tab["Some Graduate"]` with no more than some graduate school, and `r momed_tab["Graduate"]` with a graduate degree.
-Again due to data sparsity, we binarized the `r ses_tab["high"]` children whose mothers had at least some college or more as high maternal education (high-ME), and those whose mothers had at most high school (`r ses_tab["low"]` children) as low maternal education (low-ME).
-
-## Rasch Model
-
-The Rasch model, also known as the 1-parameter logistic (1PL) model, is the simplest Item Response Theory model, and is thus the easiest to use to investigate potential differences in item function across different groups of participants.
-The Rasch model jointly estimates for each child $j$ a latent ability $\theta_j$, and for each item $i$ a difficulty parameter $b_i$.
-In the model, the probability of child $j$ knowing (i.e., producing or understanding) a given item $i$ is
-
-$$P_{i}(x_i = 1 | b_{i},\theta_j ) = \frac{1}{1 + e^{-D(\theta_j - b_i )}}$$
-
-where $D$ is a constant scaling parameter ($D=1.702$) which makes the logistic closely match the ogive function in traditional factor analysis [@R-mirt; @reckase2009].
-Child ability ($\theta$) and item difficulty ($b$) distributions are standardized (i.e., mean of 0), and expected to be normally-distributed. 
-Children with high latent ability ($\theta$) will be more likely to produce any given item than children with lower latent ability, and more difficult items will be produced by fewer children (at any given $\theta$) than easier items.
-
-In the multigroup Rasch model, an item's difficulty is allowed to vary by group. 
-For example, in the sex-based multigroup model, item $i$'s difficulty is $b_{i}^{female}$ for females, and $b_{i}^{male}$ for males.
-To identify DIF, a multigroup Rasch model will be fitted for each demographic dimension of interest (sex, maternal education, and ethnicity), and we will examine the between-group difficulty difference for each item (e.g., $d_i = b_{i}^{female} - b_{i}^{male}$).
-If there is no DIF for a given item, then $d_i \approx 0$ as the two groups find the item equally difficult.
-
-# Results
-
-First, we will examine the size of demographic effects on language ability in a baseline Rasch model fitted without regard to demographic group.
-Then we will fit a multigroup Rasch model for each demographic factor, and characterize the between-group differences in item difficulties.
-Next, we will re-examine the size of demographic effects after pruning biased items from the CDI:WS using a varying threshold.
-Finally, we measure the strength of association between sex-related differences in language input and the degree of sex-related DIF.
 
@@
 
+```{r baseline-ability-vs-age-hor, fig.env = "figure*", out.width="\\linewidth", fig.pos = "t", fig.align='center', fig.width=6.5, fig.height=2.8, set.cap.width=T, num.cols.cap=2, fig.cap = "Language ability vs.
+age by demographic group, from the baseline Rasch model."}
+p1 <- d_demo %>% filter(!is.na(sex)) %>%
+  ggplot(aes(x=jitter(age), y=ability, color=sex)) + 
+  geom_point(alpha=.1) + geom_smooth() + theme_classic(base_size=9) + 
+  xlab("Age (months)") + ylab("Language Ability") + 
+  scale_color_discrete(type=sex_colors) +
+  theme(legend.position = "top")
+
+p2 <- d_demo %>% filter(!is.na(ses_group)) %>%
+  ggplot(aes(x=jitter(age), y=ability, color=ses_group)) + 
+  geom_point(alpha=.1) + geom_smooth() + theme_classic(base_size=9) + 
+  xlab("Age (months)") + ylab("Language Ability") +
+  scale_color_discrete(name = "Maternal Ed.", type=ses_colors) + 
+  theme(legend.position = "top")
+
+p3 <- d_demo %>% filter(!is.na(eth_group)) %>%
+  ggplot(aes(x=jitter(age), y=ability, color=eth_group)) + 
+  geom_point(alpha=.1) + geom_smooth() + theme_classic(base_size=9) + 
+  xlab("Age (months)") + ylab("Language Ability") + 
+  scale_color_discrete(name = "Race", type=eth_colors) + 
+  theme(legend.position = "top")
+
+ggarrange(p1, p2, p3, ncol=3, nrow=1)
+```
+CDIs are a useful tool for measuring demographic effects because of their wide use across languages. 
+<!-- While many languages show similar demographic effects on early word learning [e.g., the female advantage: @eriksson2012differences; and the first-born advantage: @frank2021], the size of these demographic effects vary across languages [see @frank2021, Ch.
+6].
+-->
+<!-- Why might we see such variation? -->
+But their use in this context poses a chicken and egg problem. 
+While it is possible that there is an underlying difference in mean language ability between demographic groups, CDIs could also have bias for or against particular groups based on the words they include.
+<!-- If a hypothetical vocabulary test only consisted of words for cold-weather gear, it would naturally show bias against children who grew up in hot climates (or who were born at certain times of year).
+-->
+<!-- , and that cross-linguistic variation is driven by differences in the language environments of different cultures, could it also be possible that the  -->
+
+The idea that some items on a test may show bias, favoring one group over another, is known in psychometrics as Differential Item Function (DIF; @holland1993differential]. 
+DIF can decrease the validity of a test, as the test may be overestimating the ability of test-takers in one group and underestimating the ability of another group -- not because of any underlying mean difference in ability between the groups, but simply because the test is unfair [@camilli2013ongoing].
+For example, if a vocabulary test is composed almost entirely of the names of farm equipment, the test will underestimate the knowledge of urban children, who have experience of other contexts, despite being less familiar with farm equipment than their rural peers.
+Of course, there may be an ability difference between demographic groups, as is likely the case for the female language advantage, which has been documented in many languages [@eriksson2012differences; @frank2021].
+And both observations could be true -- a test could be biased to inflate a true demographic difference.
+
+Our goal in the current study is to use DIF analyses to assess these possibilities. 
+Specifically, we test the 680 vocabulary items on the American CDI:WS for DIF along three dichotomized demographic dimensions: sex (male vs.
+female), maternal education (no more than secondary vs.
+at least some college), and race (white vs.
+non-white).
+
+<!-- A variety of statistical methods for detecting DIF have been proposed, and investigations have in several instances identified DIF for many items on tests favoring one group over another. -->
+
+The outline of this paper is as follows.
+First, we introduce the Wordbank data and the IRT model, and examine the overall size of demographic differences in early word learning based on the full CDI:WS.
+We then fit the IRT model to both groups for each demographic factor, and look at the item parameters for evidence of DIF, noting in particular how many items are significantly biased in favor of each group, and describing them qualitatively.
+Next, we use these item parameters to systematically prune biased items from the CDI:WS, and examine how the estimated size of demographic differences change as we prune more items.
+<!-- We identify a set of suspect items to eliminate (or in the future, replace), and provide updated estimates of the effect size of these demographic variables, were the biased items to be eliminated.-->
+Finally, using a corpus of child-directed speech, we examine the extent to which variation in word frequencies heard by boys vs. girls predicts sex-based DIF.
+Based on these analyses, we provide recommendations for next steps to be taken to identify and potentially replace items on the CDI showing demographic bias.
+
+# Methods
+
+## Vocabulary Data
+
+### Participants
+
+```{r, echo=F}
+sex_tab <- table(d_demo$sex)
+ses_tab <- table(d_demo$ses_group)
+eth_tab <- table(d_demo$ethnicity)
+eth2_tab <- table(d_demo$eth_group)
+momed_tab <- table(d_demo$mom_ed)
+```
+
+We analyze parent-reported Wordbank data from `r nrow(d_demo)` American English CDI: Words & Sentences administrations for children `r min(d_demo$age)` to `r max(d_demo$age)` months of age [@frank2017; @frank2021].
+Full demographic data are not reported in some datasets contributed to Wordbank: sex was available for `r nrow(d_demo)-sum(is.na(d_demo$sex))` children, race/ethnicity was available for `r nrow(d_demo)-sum(is.na(d_demo$ethnicity))` children, and maternal education (a proxy for socioeconomic status; SES) was available for `r nrow(d_demo)-sum(is.na(d_demo$ses_group))` children.
+
+The analysis of sex-based differences included CDI administrations from `r sex_tab["Female"]` female and `r sex_tab["Male"]` male children.
+The analysis of race-based differences included data from `r eth_tab["White"]` white, `r eth_tab["Asian"]` Asian, `r eth_tab["Black"]` Black, `r eth_tab["Hispanic"]` Hispanic, and `r eth_tab["Other"]` "Other" children.
+Due to sparse data for many categories, we binarized participants' race/ethnicity as White (`r eth2_tab["White"]`) or Non-white (`r eth2_tab["Nonwhite"]`), recognizing that there may be important variation between groups that this will fail to capture.
+Data for the maternal education analysis included CDI data from children whose mother's had the following levels of education: `r momed_tab["Primary"]` with no more than primary school education, `r momed_tab["Some Secondary"]` with some secondary school, `r momed_tab["Secondary"]` with no more than secondary school, `r momed_tab["Some College"]` with some college, `r momed_tab["College"]` with no more than a college degree, `r momed_tab["Some Graduate"]` with no more than some graduate school, and `r momed_tab["Graduate"]` with a graduate degree.
+Again due to data sparsity, we binarized the `r ses_tab["high"]` children whose mothers had at least some college or more as high maternal education (high-ME), and those whose mothers had at most high school (`r ses_tab["low"]` children) as low maternal education (low-ME).
+
+## Rasch Model
+
+The Rasch model, also known as the 1-parameter logistic (1PL) model, is the simplest Item Response Theory model, and is thus the easiest to use to investigate potential differences in item function across different groups of participants.
+The Rasch model jointly estimates for each child $j$ a latent ability $\theta_j$, and for each item $i$ a difficulty parameter $b_i$.
+In the model, the probability of child $j$ knowing (i.e., producing or understanding) a given item $i$ is
+
+$$P_{i}(x_i = 1 | b_{i},\theta_j ) = \frac{1}{1 + e^{-D(\theta_j - b_i )}}$$
+
+\noindent where $D$ is a constant scaling parameter ($D=1.702$) which makes the logistic closely match the ogive function in traditional factor analysis [@R-mirt; @reckase2009].
+Child ability ($\theta$) and item difficulty ($b$) distributions are standardized (i.e., mean of 0), and expected to be normally-distributed. 
+Children with high latent ability ($\theta$) will be more likely to produce any given item than children with lower latent ability, and more difficult items will be produced by fewer children (at any given $\theta$) than easier items.
+
+In the multigroup Rasch model, an item's difficulty is allowed to vary by group. 
+For example, in the sex-based multigroup model, item $i$'s difficulty is $b_{i}^{female}$ for females, and $b_{i}^{male}$ for males.
+To identify DIF, a multigroup Rasch model will be fitted for each demographic dimension of interest (sex, maternal education, and ethnicity), and we will examine the between-group difficulty difference for each item (e.g., $d_i = b_{i}^{female} - b_{i}^{male}$).
+If there is no DIF for a given item, then $d_i \approx 0$ as the two groups find the item equally difficult.
+
+# Results
+
+First, we will examine the size of demographic effects on language ability in a baseline Rasch model fitted without regard to demographic group.
+Then we will fit a multigroup Rasch model for each demographic factor, and characterize the between-group differences in item difficulties.
+Next, we will re-examine the size of demographic effects after pruning biased items from the CDI:WS using a varying threshold.
+Finally, we measure the strength of association between sex-related differences in language input and the degree of sex-related DIF.
+
 ## Demographic effects in baseline Rasch model
-A baseline Rasch model was fitted to the entire dataset, without demographic information.
+We fit a baseline Rasch model to the entire dataset, without demographic information.
 Figure 1 shows children's language ability vs. age by demographic group from the baseline Rasch model, which assumes no DIF (i.e., equal item parameters for all groups).
@@
 
-```{r baseline-ability-vs-age-hor, fig.env = "figure*", out.width="\\linewidth", fig.pos = "h", fig.align='center', fig.width=6.5, fig.height=2.8, set.cap.width=T, num.cols.cap=2, fig.cap = "Language ability vs.
-age by demographic group, from the baseline Rasch model."}
-p1 <- d_demo %>% filter(!is.na(sex)) %>%
-  ggplot(aes(x=jitter(age), y=ability, color=sex)) + 
-  geom_point(alpha=.1) + geom_smooth() + theme_classic(base_size=9) + 
-  xlab("Age (months)") + ylab("Language Ability") + 
-  scale_color_discrete(type=sex_colors) +
-  theme(legend.position = "top")
-
-p2 <- d_demo %>% filter(!is.na(ses_group)) %>%
-  ggplot(aes(x=jitter(age), y=ability, color=ses_group)) + 
-  geom_point(alpha=.1) + geom_smooth() + theme_classic(base_size=9) + 
-  xlab("Age (months)") + ylab("Language Ability") +
-  scale_color_discrete(name = "SES", type=ses_colors) + 
-  theme(legend.position = "top")
-
-p3 <- d_demo %>% filter(!is.na(eth_group)) %>%
-  ggplot(aes(x=jitter(age), y=ability, color=eth_group)) + 
-  geom_point(alpha=.1) + geom_smooth() + theme_classic(base_size=9) + 
-  xlab("Age (months)") + ylab("Language Ability") + 
-  scale_color_discrete(name = "ethnicity", type=eth_colors) + 
-  theme(legend.position = "top")
-
-ggarrange(p1, p2, p3, ncol=3, nrow=1)
-```
-
 ## Identifying biased CDI items
@@
 It is important, however, to inspect the full plots, for if there is a cluster of items in a GLIMMER, the analyst may conclude that these items are strong candidates for DIF on that dimension.
-For example, there is some clustering at the top and bottom of Figure 2: at the top, "vagina", "tights", "dress (object)", and "doll" form a cluster of items that are much more well-known for females, while at the bottom, "penis" stands out as much more well-known by males.
+For example, there is some clustering at the top and bottom of Figure 2: at the top, "vagina"^[Parents using the CDI are instructed to check genital items if the child produces whatever word the family uses.], "tights", "dress (object)", and "doll" form a cluster of items that are much more well-known for females, while at the bottom, "penis" stands out as much more well-known by males.
 For the maternal education and race GLIMMERs (Figures 3 and 4)--and in the rest of the full sex GLIMMER, there are not clusters, but rather a continuum of smoothly varying differences with overlap. 

### 2022-02-01 paper/cogsci.tex
tweaking

--- 2021-10-29 paper/cogsci.tex
@@
 
-\title{Words aren't created equal: Investigating bias on the CDI}
+\title{Estimating demographic bias on tests of children's early
+vocabulary}
 
-\author{{\large \bf Morton Ann Gernsbacher (MAG@Macc.Wisc.Edu)} \\ Department of Psychology, 1202 W.
-Johnson Street \\ Madison, WI 53706 USA \AND {\large \bf Sharon J.~Derry (SDJ@Macc.Wisc.Edu)} \\ Department of Educational Psychology, 1025 W.
-Johnson Street \\ Madison, WI 53706 USA}
+\author{{\large \bf George Kachergis (kachergis@stanford.edu)} \AND {\large \bf Nathan Francis (nathan99@stanford.edu)} \AND {\large \bf Michael C.
+Frank (mcfrank@stanford.edu)} \\ Department of Psychology, Stanford Unviersity \\ Stanford, CA 94305 USA }
+
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newenvironment{CSLReferences}%
+  {}%
+  {\par}
 
@@
 Children's early language skill has been linked to later educational
-outcomes, making it important to accurately measure early language.
+outcomes, making it important to measure early language accurately.
 Parent-reported instruments such as the Communicative Development
-Inventories (CDIs) have been shown to provide valid, consistent measures
-of children's aggregate early language skill. However, CDIs are
-predominantly comprised of hundreds of vocabulary items, some of which
-may not be heard (and thus learned) equally often by children of varying
-backgrounds. Here, we use a database of American English CDIs to
-identify words that show strong bias for particular groups of children,
-along the axes of sex (male vs.~female), race/ethnicity (white
-vs.~non-white), and socioeconomic status (high vs.~low). For each axis,
-we identify dozens of strongly biased items, and show that eliminating
-these items reduces the expected ability difference between groups. For
-sex, we consider how to propose replacement words that may show less
-bias, on the basis of their relatively equal frequency in adult speech
-directed to male and female children.
+Inventories (CDIs) have been shown to provide reliable and valid
+measures of children's aggregate early language skill. However, CDIs
+contain hundreds of vocabulary items, some of which may not be heard
+(and thus learned) equally often by children of varying backgrounds.
+This study used a database of American English CDIs to identify words
+that showed strong bias for particular demographic groups of children,
+on dimensions of sex (male vs.~female), race (white vs.~non-white), and
+maternal education (high vs.~low). For each dimension, we identified
+dozens of strongly biased items, and showed that eliminating these items
+reduced the expected ability difference between groups. Additionally, we
+investigated how well the relative frequency of words spoken to young
+girls vs.~boys predicted sex-based word learning bias, and discuss
+possible sources of demographic differences in early word learning.
 
 \textbf{Keywords:}
-language development; early vocabulary; measuring instrument bias;
+language acquisition; word learning; measuring instrument bias;
+demographics;
 \end{abstract}
@@
 
-introduce importance of measuring early language development (e.g.,
-links to later educational outcomes)
+Researchers, clinicians, and parents have long been fascinated with the
+surprising speed and variability in the growth of young children's
+vocabulary. Children's early vocabulary growth is assumed to reflect not
+only their exposure to child-directed speech, but also the varying
+difficulty of different types of words, and individual differences in
+the aptitude of the child -- including potential language deficits.
+Children show both consistency in some skills across development, as
+well as significant influence from external factors. For example,
+Bornstein, Hahn, \& Putnick (2016) found stability in core language
+skills across 10 years of children's development, despite changes in
+maternal income and education over the study period.
 
-introduce CDI
+Yet there are consistent demographic predictors of early vocabulary.
+Maternal education, often used as a proxy for socioeconomic status
+(SES), is associated with children's language processing and vocabulary
+by 18 months (Fernald, Marchman, \& Weisleder, 2013), and is predictive
+of later educational outcomes (Marchman \& Fernald, 2008; see Schwab \&
+Lew-Williams, 2016 for review). Other factors are also predictive of
+language skill: first-born children tend to outpace their siblings, and
+female children tend to have better language skills than their
+age-matched male counterparts (Eriksson et al., 2012; Frank, Braginsky,
+Yurovsky, \& Marchman, 2021) -- a sex-based verbal advantage that
+continues through high school (see Petersen, 2018 for a review).
 
-predictive validity: Fenson et al.~1994, Bornstein et al.~2012, Duff et
-al.~2015
+One challenge in comprehensive assessment of these differences is that
+is difficult to get a complete measure of young children's language
+skill. Long recordings are prohibitively difficult to collect and
+transcribe, and yet any short recording (e.g., a 1-hour play session)
+will elicit only a small proportion of the words and constructions that
+children know. Thus, researchers of early word learning have constructed
+tests with hundreds of words, intentionally oversampling words that are
+more likely to be known by young children.
 
-discuss potential for bias
+We focus here on the MacArthur-Bates Communicative Development
+Inventories {[}CDIs; Larry Fenson et al. (1994); L. Fenson et al.
+(2007){]}, a set of parent-reported measures of children's productive
+and receptive language skills, which offer a low-cost and reliable way
+to estimate children's early language skills (Larry Fenson et al.,
+1994). CDIs have shown good predictive validity (Bornstein \& Putnick,
+2012; Duff, Reen, Plunkett, \& Nation, 2015; e.g., Larry Fenson et al.,
+1994). Our primary focus is the vocabulary checklist portion of the CDI
+Words \& Sentences (CDI:WS) form, comprised of 680 early-learned words
+across 22 categories (e.g., animals, vehicles, action words, pronouns)
+selected to assess the productive vocabulary of children 16 to 30 months
+of age. For each item on the CDI:WS, caregivers are asked to respond
+whether the target child has been heard to say (i.e.~produce) the given
+item. Children's total vocabulary score on the CDI:WS is tightly
+correlated with other facets of early language (e.g., grammatical
+competence and gesture), suggesting that the language system is
+``tightly woven'' (Frank, Braginsky, Yurovsky, \& Marchman, 2021). Due
+to these desirable properties, CDIs have been adapted to dozens of
+languages, and a central repository of CDI data contributed from all
+over the world has been created {[}Wordbank; Frank, Braginsky, Yurovsky,
+\& Marchman (2017); Frank, Braginsky, Yurovsky, \& Marchman (2021){]}.
 
-what does between-group bias look like?
+\begin{CodeChunk}
+\begin{figure*}[t]
 
-IRT - what is a good measure?
+{\centering \includegraphics[width=\linewidth]{figs/baseline-ability-vs-age-hor-1}
 
-\hypertarget{fundamental-dif-problem}{%
-\subsection{Fundamental DIF Problem}\label{fundamental-dif-problem}}
+}
 
-what is our goal for measuring vocabulary? - identifying language
-delays, predicting later reading, or talking\ldots{}
+\caption[Language ability vs]{Language ability vs. age by demographic group, from the baseline Rasch model.}\label{fig:baseline-ability-vs-age-hor}
+\end{figure*}
+\end{CodeChunk}
 
-how you select items influences how well you achieve these goals (and
-what bias you find)
+CDIs are a useful tool for measuring demographic effects because of
+their wide use across languages. But their use in this context poses a
+chicken and egg problem. While it is possible that there is an
+underlying difference in mean language ability between demographic
+groups, CDIs could also have bias for or against particular groups based
+on the words they include.
 
-We take the approach put forward by Stenhaug, Frank, \& Domingue (2021):
-GLIMMER plots based on the 1-parameter logistic are sufficient to
-identify bias, without potentially obscuring DIF in a more complex
-model's additional parameters.
+The idea that some items on a test may show bias, favoring one group
+over another, is known in psychometrics as Differential Item Function
+(DIF; Holland \& Wainer (1993){]}. DIF can decrease the validity of a
+test, as the test may be overestimating the ability of test-takers in
+one group and underestimating the ability of another group -- not
+because of any underlying mean difference in ability between the groups,
+but simply because the test is unfair (Camilli, 2013). For example, if a
+vocabulary test is composed almost entirely of the names of farm
+equipment, the test will underestimate the knowledge of urban children,
+who have experience of other contexts, despite being less familiar with
+farm equipment than their rural peers. Of course, there may be an
+ability difference between demographic groups, as is likely the case for
+the female language advantage, which has been documented in many
+languages (Eriksson et al., 2012; Frank, Braginsky, Yurovsky, \&
+Marchman, 2021). And both observations could be true -- a test could be
+biased to inflate a true demographic difference.
 
-Differential item functioning (DIF) is a technique in IRT used to
-identify items that show bias against demographic groups. DIF is a
-statistical characteristic of an item that shows the extent to which the
-item might operate differently or measure varying abilities for
-subgroups and members of separate demographic groups. DIF is, however, a
-challenging metric for identifying bias for a couple reasons. Firstly,
-DIF is extremely hard to accurately root out. The process of finding DIF
-involves using the test you used to do IRT analysis in order to look for
-bias within the very same test. Secondly, the presence of DIF does not
-necessarily indicate the presence of bias; it can indicate that one
-demographic has a higher average ability level than the other. These two
-issues are the basis upon which we get the Fundamental DIF
-Identification problem.
+Our goal in the current study is to use DIF analyses to assess these
+possibilities. Specifically, we test the 680 vocabulary items on the
+American CDI:WS for DIF along three dichotomized demographic dimensions:
+sex (male vs.~female), maternal education (no more than secondary vs.~at
+least some college), and race (white vs.~non-white).
 
-This problem can be be understood through a concrete example that is
-well-known from research on early language learning: consider the fact
-that females show a larger vocabulary than males across early
-development {[}see Frank, Braginsky, Yurovsky, \& Marchman (2021) Ch.
-6?; OTHER REFS{]}. The question however is whether or not girls actually
-have a higher ability level or if the tests have words that are bias
-toward girls ie. are easier for girls, or perhaps the test is biased
-toward boys but the natural ability of girls is enough to overcome that
-bias. To answer this question we need to know the ability levels of
-girls and boys in order to confirm that if a word is learned earlier by
-girls it is simply because of an ability difference rather than a bias
-inherent to the word. The problem arises from the fact that we measure
-ability level using the very same test that we are trying to check for
-biased words. If the boys and girls were of the same ability level this
-wouldn't be an issue but given the evidence showing that girls have a
-higher ability level, we need to know the difference in ability so that
-when we find DIF for a specific word can be sure it is outside of the
-expecting DIF that results naturally from their difference in ability.
+The outline of this paper is as follows. First, we introduce the
+Wordbank data and the IRT model, and examine the overall size of
+demographic differences in early word learning based on the full CDI:WS.
+We then fit the IRT model to both groups for each demographic factor,
+and look at the item parameters for evidence of DIF, noting in
+particular how many items are significantly biased in favor of each
+group, and describing them qualitatively. Next, we use these item
+parameters to systematically prune biased items from the CDI:WS, and
+examine how the estimated size of demographic differences change as we
+prune more items. Finally, using a corpus of child-directed speech, we
+examine the extent to which variation in word frequencies heard by boys
+vs.~girls predicts sex-based DIF. Based on these analyses, we provide
+recommendations for next steps to be taken to identify and potentially
+replace items on the CDI showing demographic bias.
 
@@
 
-Research was conducted over 5520 participants, whose data was parent
-reported and collected by the Wordbank Project. Unlike other projects
-Wordbank relies on the kindness of others to contribute their data which
-means that often meta-data for some sets is missing. Even so, the
-project now has data over 29 different languages on children from ages
-8-30 months old, though our study focused entirely on American English
-and Children 16-30 months. When they received complete information, the
-Wordbank data set collected demographic data consisting of Birth Order,
-Race/Ethnicity, Sex and Mothers Education. We focused on comparing data
-between demographic groups on the axes of Sex, Ethnicity and Mother's
-Education, a proxy for socioeconomic status (henceforth, SES). Our data
-included 1989 female and 2105 male participants, and an additional 1426
-participants whose sex remained unreported. In terms of Ethnicity data
-was recorded from 2202 white, 67 Asian, 222 Black, 131 Hispanic, and 93
-``Other'' participants as well as 2805 unreported. Given our
-distribution of ethnicities, for our analysis we split participants into
-White (2202) and Non-White (513). Finally, we split participants into
-high and low SES, of which we had 4973 high SES participants,
-corresponding to mothers' education of some college or higher and 547
-Low SES participants.
+We analyze parent-reported Wordbank data from 5520 American English CDI:
+Words \& Sentences administrations for children 16 to 30 months of age
+(Frank, Braginsky, Yurovsky, \& Marchman, 2017, 2021). Full demographic
+data are not reported in some datasets contributed to Wordbank: sex was
+available for 4094 children, race/ethnicity was available for 2715
+children, and maternal education (a proxy for socioeconomic status; SES)
+was available for 5520 children.
 
-\hypertarget{one-column-images}{%
-\subsection{One-column images}\label{one-column-images}}
+The analysis of sex-based differences included CDI administrations from
+1989 female and 2105 male children. The analysis of race-based
+differences included data from 2202 white, 67 Asian, 222 Black, 131
+Hispanic, and 93 ``Other'' children. Due to sparse data for many
+categories, we binarized participants' race/ethnicity as White (2202) or
+Non-white (513), recognizing that there may be important variation
+between groups that this will fail to capture. Data for the maternal
+education analysis included CDI data from children whose mother's had
+the following levels of education: 8 with no more than primary school
+education, 123 with some secondary school, 416 with no more than
+secondary school, 613 with some college, 870 with no more than a college
+degree, 162 with no more than some graduate school, and 584 with a
+graduate degree. Again due to data sparsity, we binarized the 4973
+children whose mothers had at least some college or more as high
+maternal education (high-ME), and those whose mothers had at most high
+school (547 children) as low maternal education (low-ME).
+
+\hypertarget{rasch-model}{%
+\subsection{Rasch Model}\label{rasch-model}}
+
+The Rasch model, also known as the 1-parameter logistic (1PL) model, is
+the simplest Item Response Theory model, and is thus the easiest to use
+to investigate potential differences in item function across different
+groups of participants. The Rasch model jointly estimates for each child
+\(j\) a latent ability \(\theta_j\), and for each item \(i\) a
+difficulty parameter \(b_i\). In the model, the probability of child
+\(j\) knowing (i.e., producing or understanding) a given item \(i\) is
+
+\[P_{i}(x_i = 1 | b_{i},\theta_j ) = \frac{1}{1 + e^{-D(\theta_j - b_i )}}\]
+
+\noindent where \(D\) is a constant scaling parameter (\(D=1.702\))
+which makes the logistic closely match the ogive function in traditional
+factor analysis (Chalmers, 2012; Reckase, 2009). Child ability
+(\(\theta\)) and item difficulty (\(b\)) distributions are standardized
+(i.e., mean of 0), and expected to be normally-distributed. Children
+with high latent ability (\(\theta\)) will be more likely to produce any
+given item than children with lower latent ability, and more difficult
+items will be produced by fewer children (at any given \(\theta\)) than
+easier items.
+
+In the multigroup Rasch model, an item's difficulty is allowed to vary
+by group. For example, in the sex-based multigroup model, item \(i\)'s
+difficulty is \(b_{i}^{female}\) for females, and \(b_{i}^{male}\) for
+males. To identify DIF, a multigroup Rasch model will be fitted for each
+demographic dimension of interest (sex, maternal education, and
+ethnicity), and we will examine the between-group difficulty difference
+for each item (e.g., \(d_i = b_{i}^{female} - b_{i}^{male}\)). If there
+is no DIF for a given item, then \(d_i \approx 0\) as the two groups
+find the item equally difficult.
+
+\hypertarget{results}{%
+\section{Results}\label{results}}
+
+First, we will examine the size of demographic effects on language
+ability in a baseline Rasch model fitted without regard to demographic
+group. Then we will fit a multigroup Rasch model for each demographic
+factor, and characterize the between-group differences in item
+difficulties. Next, we will re-examine the size of demographic effects
+after pruning biased items from the CDI:WS using a varying threshold.
+Finally, we measure the strength of association between sex-related
+differences in language input and the degree of sex-related DIF.
+
+\hypertarget{demographic-effects-in-baseline-rasch-model}{%
+\subsection{Demographic effects in baseline Rasch
+model}\label{demographic-effects-in-baseline-rasch-model}}
+
+We fit a baseline Rasch model to the entire dataset, without demographic
+information. Figure 1 shows children's language ability vs.~age by
+demographic group from the baseline Rasch model, which assumes no DIF
+(i.e., equal item parameters for all groups). A linear regression for
+each demographic group, with age (centered) and its interaction, showed
+significant effects. Female children had higher language ability than
+male children (\(\beta=0.56\), \(p<.001\)), with no significant
+interaction with age (\(\beta=0.02\), \(p=.10\)). High-SES children had
+higher language ability than low-SES children (\(\beta=0.23\),
+\(p=.02\)), an advantage that grew with age (\(\beta=0.11\),
+\(p<.001\)). White children had higher language ability than non-white
+children (\(\beta=0.50\), \(p<.001\)), an advantage that grew with age
+(\(\beta=0.08\), \(p<.001\)). We will re-examine these demographic
+regressions after trimming items showing extreme DIF.
+
+\hypertarget{identifying-biased-cdi-items}{%
+\subsection{Identifying biased CDI
+items}\label{identifying-biased-cdi-items}}
+
+To aid in identifying CDI items with DIF we created GLIMMERs (Graphs of
+Logits Imputed Multiply with Means Equal; Stenhaug, Frank, \& Domingue
+(2021)), which visualize between-item variation in group performance
+differences. These parameters are drawn from a fitted multigroup Rasch
+model for each demographic variable (sex, SES, and race), with the
+assumption that the mean language ability in each group is the same
+(e.g.~for sex, \(\mu_{male}=\mu_{female}=0\)), thus pushing all
+between-group variation into the item difficulty parameters. For the
+case of sex, where we believe \(\mu_{female}>\mu_{male}\), this means we
+may expect to find many items with difficulty
+\(b_i^{female} < b_i^{male}\), but we can still examine the distribution
+of differences in item difficulty (\(d_j = b_i^{male} - b_i^{female}\))
+for outliers. GLIMMER plots show distributions of parameter differences
+rather than point estimates to convey the uncertainty about the
+existence of DIF. These distributions are generated by drawing 10,000
+imputations from the item parameter covariance matrix.
+
+Figure 2 shows GLIMMERs for a selection of CDI items for sex (left),
+maternal education (middle), and race (right). The full GLIMMERs, with
+all 680 CDI:WS items, are available on
+\href{https://osf.io/57rsw/?view_only=2b6ecb61fe08458293af7421d276932a}{OSF},
+but were too large to include here. It is important, however, to inspect
+the full plots, for if there is a cluster of items in a GLIMMER, the
+analyst may conclude that these items are strong candidates for DIF on
+that dimension. For example, there is some clustering at the top and
+bottom of Figure 2: at the top, ``vagina''\footnote{Parents using the
+  CDI are instructed to check genital items if the child produces
+  whatever word the family uses.}, ``tights,'' ``dress (object),'' and
+``doll'' form a cluster of items that are much more well-known for
+females, while at the bottom, ``penis'' stands out as much more
+well-known by males. For the maternal education and race GLIMMERs
+(Figures 3 and 4)--and in the rest of the full sex GLIMMER, there are
+not clusters, but rather a continuum of smoothly varying differences
+with overlap. This makes identifying items with DIF quite difficult, as
+different methods are likely to yield inconsistent results (Stenhaug,
+Frank, \& Domingue, 2021). Hence, we next characterized the
+distributions of the item-level group difficulty differences, and
+measured the influence of pruning a varying number of items on the
+demographic effect sizes.
+
+\begin{CodeChunk}
+\begin{figure*}[h]
+
+{\centering \includegraphics[width=\linewidth]{figs/smGLIMMER_combo}
+
+}
+
+\caption[GLIMMER plot of a sample of CDI:WS words from the sex bias model]{GLIMMER plot of a sample of CDI:WS words from the sex bias model.
+Words at the top are more well-known by females, while those at the bottom are more known for males.}\label{fig:glimmer-combo}
+\end{figure*}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{figure*}[h]
+
+{\centering \includegraphics[width=\linewidth]{figs/pruned-coefs-hor-1}
+
+}
+
+\caption[Size of demographic effects (regression coefficients) with different pruning thresholds]{Size of demographic effects (regression coefficients) with different pruning thresholds.}\label{fig:pruned-coefs-hor}
+\end{figure*}
+\end{CodeChunk}
+
+\hypertarget{sex}{%
+\subsubsection{Sex}\label{sex}}
+
+For sex, the median difficulty difference (male-female) was 0.28
+(M=0.27, sd=0.4), with 593/680 items being easier for females than
+males. The fact that the bulk of this distribution favors females shows
+that the female language advantage is pervasive across the CDI:WS, and
+suggests that it is likely to be a real ability advantage, unless the
+bulk of these items are actually spoken more often to female children
+than to male children.
+
+\hypertarget{maternal-education}{%
+\subsubsection{Maternal Education}\label{maternal-education}}
+
+For maternal education (ME), the median difficulty difference (low-high)
+was -0.01 (M=0.03, sd=0.55), with 337/680 items being easier for high-ME
+than low-ME children. With the mean and median difficulty differences
+close to 0, and roughly half of the words favoring each ME group, it is
+tempting to conclude that the CDI:WS items are somewhat balanced with
+respect to ME (and thus SES, by proxy).
+
+\hypertarget{race}{%
+\subsubsection{Race}\label{race}}
+
+For race, the median difficulty difference (nonwhite-white) was 0.4
+(M=0.46, sd=0.55), with 549/680 items being easier for white than
+non-white children, revealing a fairly pervasive advantage for white
+children on CDI items.
+
+\hypertarget{demographic-effects-after-pruning}{%
+\subsection{Demographic effects after
+pruning}\label{demographic-effects-after-pruning}}
+
+Given the smoothly-varying distributions of DIF shown by each
+demographic factor, we chose to evaluate multiple thresholds for pruning
+the extreme-valued items from each distribution. For each demographic
+factor, we pruned items with difficulty difference from the mean at
+varying thresholds, from \textgreater.25 SD to \textgreater3 SD, in
+increments of .25 SD. At each SD threshold, a potentially different
+subset of items are excluded for each model, but with the goal of
+creating a single CDI that is less biased on all dimensions, we pruned
+the union of the subsets excluded from each model. For example, pruning
+\textgreater2SD from the mean of each model excluded 25 sex-biased
+items, 39 ME-biased items, and 31 race-biased items, with their union
+being 76 unique items. Figure 3 shows the demographic effects
+(\(\beta\)s) at different exclusion thresholds.
+
+Ideally, we would find a single threshold that minimized the magnitude
+of coefficients (and thus bias) for all three demographics
+simultaneously. Unfortunately, the effect size for each demographic
+variable was smallest (closest to 0) at different exclusion thresholds.
+The effect size of sex was smallest when almost all items were trimmed
+(\textgreater0.25 SD; \(\beta_{sex} = -.47\); 661/680 total items
+trimmed; 433 due to sex extremity). The effect size of maternal
+education (SES) was smallest when items \textgreater1.25SD in difficulty
+difference were trimmed (\(\beta_{ME} = -.17\); 234 total items trimmed;
+126 due to ME extremity). The effect size of race was minimized when
+items more extreme than 0.5 SD were trimmed (\(\beta_{race} = .42\); 577
+total items trimmed; 425 due to race extremity).
+
+Where is the exclusion threshold that jointly minimizes the effect sizes
+of these demographic variables? The strict exclusion of all items with
+\textgreater.25 SD difficulty difference -- 661/680 (97\%) of the CDI:WS
+-- best optimized this, but seems too extreme of a culling. The next
+best thresholds are \textgreater.5 SD -- again, too extreme -- or 3 SD,
+a remarkably lax criterion that only excludes 19 items in total (11 due
+to sex bias). Only slightly worse than these is \textgreater2.25 SD,
+which excludes 59 items (9\% of the CDI:WS), and shows a modest effect
+of both race (\(\beta=0.48\)) and maternal education (\(\beta=.19\)),
+and a near-median effect size for sex (\(\beta=-.59\)). We characterize
+these 59 potentially biased items below, and consider whether we might
+recommend pruning them from the CDI:WS.
+
+\hypertarget{characterizing-biased-items}{%
+\subsection{Characterizing biased
+items}\label{characterizing-biased-items}}
+
+Figure 4 shows the 59 items showing extreme bias (i.e., whose difficulty
+was \textgreater2.25 SD from the mean difficulty difference on one or
+more demographic dimension). There were 22 items with extreme sex-based
+difficulty differences, only 7 of which were more known by females, with
+the other 15 items favoring males. All but one of the sex-biased items
+are nouns, many of which are stereotypically associated with one gender
+more than the other, including stereotypically-male professions (e.g.,
+``fireman''). For maternal education, 27 extrema were identified, only
+10 of which were biased for low-SES children. As for sex, most of the
+ME-biased items are nouns (15/27). Notably, many of the items more known
+by high-ME children were animals (``zebra,'' ``owl'') and animal sounds
+(``baa baa,'' ``moo''). Items more known by low-ME children include a
+few sweet treats: ``candy,'' ``soda/pop,'' and ``gum.'' For race, 21
+extrema were identified, only 8 of which were easier for non-white
+children. The lexical class of the race-biased items was more varied: 10
+were nouns, but many others were early-learned words and phrases (e.g.,
+``up,'' ``all gone,'' ``uh-oh,'' ``bye,'' ``mommy''). Only 10 of the 59
+items were extreme in more than one model, with most of the overlap
+being between race and ME (8 items). Only 1 item (``vroom'') was extreme
+for all three demographics.
+
+\begin{CodeChunk}
+\begin{figure}[H]
+
+{\centering \includegraphics[width=\linewidth]{figs/extrema2p25sd}
+
+}
+
+\caption[The 59 items showing extreme bias (difficulty difference $>2.25$ SD) for one or more demographic]{The 59 items showing extreme bias (difficulty difference $>2.25$ SD) for one or more demographic.}\label{fig:biased-words}
+\end{figure}
+\end{CodeChunk}
+
+\hypertarget{relating-child-directed-speech-to-demographic-bias}{%
+\subsection{Relating child-directed speech to demographic
+bias}\label{relating-child-directed-speech-to-demographic-bias}}
+
+Demographic differences in language ability are likely to be at least
+partially explained by differences in linguistic input received by
+children in different groups. Indeed, input quantity (total daily
+tokens) and some measures of quality (e.g., lexical diversity: ratio of
+word types vs.~tokens) have often been predictive of language learning
+outcomes in demographic studies (Huttenlocher, Haight, Bryk, Seltzer, \&
+Lyons, 1991; Rowe \& Goldin-Meadow, 2009). We used a corpus-based
+analysis to investigate the extent to which word frequency in
+child-directed speech to male vs.~female children was predictive of the
+amount of sex-based bias shown by CDI items. Similar to the approach
+taken by Braginsky, Meylan, \& Frank (2016), we used the CHILDES corpus
+of transcripts from dyadic play sessions (MacWhinney, 2000), which are
+labeled with the sex of the target child, but not other demographic
+variables. We found a total of 5213750 female-directed tokens and
+6091950 male-directed tokens that matched 662 of the 680 CDI:WS items,
+normalized the word frequencies to tokens per million for each target
+sex, and calculated the log-odds of each word being spoken to a female
+(i.e., \(log(p_{f} / p_{m}\)), meaning unbiased words will have log-odds
+of 0, while those spoken more often to females will have log-odds
+\(>0\), and those spoken more often males will have log-odds \(<0\). For
+example ``doll'' was spoken 156 times to females, and 138 times to
+males, thus \(p_{f} = 156/(156+138) = .53\), \(p_{m} = 1-p_{f} = 0.47\),
+and \(log(p_{f}/p_{m})\) = 0.12), while ``police'' was spoken 138 times
+to females, and 225 times to males, and thus has log-odds = -0.49).
+Overall, the correlation between the log-odds of a word being spoken to
+a female vs.~a male child and the size of the female (vs.~male)
+advantage for that CDI word was modest, but significant (\(r=0.18\),
+\(p<.001\)), suggesting that some of the sex bias seen in CDI items is
+explained by differences in language input received by girls vs.~boys.
+
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
+
+We investigated the CDI:WS, a popular parent-report measure of
+children's early vocabulary, for potential demographic bias, examining
+the distribution of words' estimated difficulties for children of
+mothers with high- vs.~low-education children (a proxy for family SES),
+females vs.~males, and for white vs.~non-white children. The IRT-based
+analysis revealed differential item functioning (DIF) for many items
+along each demographic dimension, but only in the case of sex were clear
+clusters of items that were more well-known to females (including
+feminine clothing and genitalia), and a clear item that was more
+well-known to males (male genitalia). For the rest of the items, and for
+SES- and race-based analysis, there was a smooth continuum of DIF,
+making the boundary of true DIF subjective, as this would rely on
+knowing the true difference in language ability between groups--which we
+reciprocally estimate from instruments like the CDI. To move forward, we
+identified candidate DIF items systematically pruning the extremes of
+each distribution, excluding outliers for each demographic across a
+range of thresholds, and seeking to minimize the size of demographic
+effects. Although many exclusion thresholds decreased the size of
+maternal education and race effects, the female advantage actually grew
+under most prunings, with the majority of excluded extrema being
+stereotypically-male nouns (e.g, ``truck,'' ``fireman'').
+
+This analysis highlights a fundamental difficulty of identifying DIF: an
+analyst must either know the expected magnitude of an ability difference
+between the two demographic groups, or know a set of ``anchor'' items
+that are equally difficult (i.e., unbiased) for both groups (for an
+overview, see Stenhaug, Frank, \& Domingue, 2021). But in the universe
+of children's early language, there is only a finite set of
+early-learned words to choose from--and we may expect many of them to be
+biased for various environmental reasons (e.g., boys and girls know
+their respective genitalia; children in Florida may not use mittens or
+skis). Hence, the presence of DIF on a set of items could indicate a
+degree of varying linguistic input, rather than an actual advantage for
+one demographic over another.
+
+Choosing an exclusion threshold that struck a balance of reducing race-
+and ME-based bias and not eliminating too many items, we identified 59
+suspect items--the majority of which favored white/high-ME groups. These
+extrema were predominantly nouns, including many animals (esp.~for
+maternal education), vehicles and professions (esp.~for sex). Notably,
+no verbs and few syntactically complex items were identified as
+outliers, which may suggest that these lexical classes are less prone to
+environmental influence, and perhaps more generally unbiased. Pruning
+these 59 outliers would decrease the demographic effects of maternal
+education (unpruned \(\beta=.23\); pruned \(\beta=.19\)) and race
+(unpruned \(\beta=.50\); pruned \(\beta=.48\)). In contrast, the
+majority of the sex-biased outliers were easier for the disadvantaged
+group (males), meaning that the removal of these extrema increased the
+female language advantage on the CDI:WS (unpruned \(\beta=0.56\); pruned
+\(\beta=.59\)). In aggregate, these analyses seem to suggest that the
+CDI:WS may slightly overestimate the magnitude of differences in
+language ability along the dimensions of maternal education and race,
+and perhaps underestimate the female advantage.
+
+This investigation is only a first step in measuring demographic bias in
+the items on the CDI:WS. More research is needed to determine which CDI
+items that show DIF are more subject to environmental influence, as we
+found only a modest association between the relative frequency of words
+in child-directed speech to boys vs.~girls and the amount of sex-based
+DIF. Examining the relative frequency of CDI words across households
+with varying maternal education and race may reveal similarly weak
+associations, or there may be strong associations only among particular
+categories of words. It may be that particular activity
+contexts--popular with some demographic groups, and not others--may be
+predictive. For example, many of the extrema favoring high-ME children
+are animals and animal sounds (e.g., ``quack quack,'' ``woof woof,''
+``baa baa,'' ``duck,'' ``sheep,'' ``giraffe,'' ``zebra''): do high-SES
+households visit the zoo more often, or perhaps engage in other
+activities related to naming animals and noises they make (e.g., animal
+noise toys)? If certain activities are driving early word learning for
+high-ME children, what are the activities (and associated vocabulary)
+that low-SES households are instead engaging in? Note, too, that SES,
+race, and sex are only three of many possible demographic dimensions of
+interest. For example, geographic region is likely predictive of
+children's early experience with (and thus knowledge of) CDI items
+related to winter weather (e.g., ``snow,'' ``mittens'') or outdoors
+activities (e.g.~``camping''). A truly fair test of children's early
+vocabulary would contain a representative and balanced sample of words
+from all activities that children engage in, across demographic groups.
 
@@
 
-Place acknowledgments (including funding information) in a section at
-the end of the paper.
+{[}Redacted for anonymous review.{]}
 
@@
 \hypertarget{refs}{}
-\leavevmode\hypertarget{ref-Frank2017}{}%
+\begin{CSLReferences}{1}{0}
+\leavevmode\hypertarget{ref-bornstein2016stability}{}%
+Bornstein, M. H., Hahn, C.-S., \& Putnick, D. L. (2016). Stability of
+core language skill across the first decade of life in children at
+biological and social risk. \emph{Journal of Child Psychology and
+Psychiatry}, \emph{57}(12), 1434--1443.
+
+\leavevmode\hypertarget{ref-bornstein2012stability}{}%
+Bornstein, M. H., \& Putnick, D. L. (2012). Stability of language in
+childhood: A multiage, multidomain, multimeasure, and multisource study.
+\emph{Developmental Psychology}, \emph{48}(2), 477.
+
+\leavevmode\hypertarget{ref-braginsky2016gender}{}%
+Braginsky, M., Meylan, S. C., \& Frank, M. C. (2016). Gender differences
+in lexical input and acquisition. Boston.
+
+\leavevmode\hypertarget{ref-camilli2013ongoing}{}%
+Camilli, G. (2013). Ongoing issues in test fairness. \emph{Educational
+Research and Evaluation}, \emph{19}(2-3), 104--120.
+
+\leavevmode\hypertarget{ref-R-mirt}{}%
+Chalmers, R. P. (2012). {mirt}: A multidimensional item response theory
+package for the {R} environment. \emph{Journal of Statistical Software},
+\emph{48}(6), 1--29.
+http://doi.org/\href{https://doi.org/10.18637/jss.v048.i06}{10.18637/jss.v048.i06}
+
+\leavevmode\hypertarget{ref-duff2015infant}{}%
+Duff, F. J., Reen, G., Plunkett, K., \& Nation, K. (2015). Do infant
+vocabulary skills predict school-age language and literacy outcomes?
+\emph{Journal of Child Psychology and Psychiatry}, \emph{56}(8),
+848--856.
+
+\leavevmode\hypertarget{ref-eriksson2012differences}{}%
+Eriksson, M., Marschik, P. B., Tulviste, T., Almgren, M., Pérez Pereira,
+M., Wehberg, S., \ldots{} Gallego, C. (2012). Differences between girls
+and boys in emerging language skills: Evidence from 10 language
+communities. \emph{British Journal of Developmental Psychology},
+\emph{30}(2), 326--343.
+
+\leavevmode\hypertarget{ref-fenson1994}{}%
+Fenson, Larry, Dale, P. S., Reznick, J. S., Bates, E., Thal, D. J.,
+Pethick, S. J., \ldots{} Stiles, J. (1994). Variability in early
+communicative development. \emph{Monographs of the Society for Research
+in Child Development}, i--185.
+
+\leavevmode\hypertarget{ref-Fenson2007}{}%
+Fenson, L., Marchman, V. A., Thal, D. J., Dale, P. S., Reznick, J. S.,
+\& Bates, E. (2007). \emph{{M}ac{A}rthur-{B}ates {C}ommunicative
+{D}evelopment {I}nventories: User's guide and technical manual (2nd
+ed.)}. Baltimore, MD: Brookes.
+
+\leavevmode\hypertarget{ref-fernald2013ses}{}%
+Fernald, A., Marchman, V. A., \& Weisleder, A. (2013). SES differences
+in language processing skill and vocabulary are evident at 18 months.
+\emph{Developmental Science}, \emph{16}(2), 234--248.
+
+\leavevmode\hypertarget{ref-frank2017}{}%
 Frank, M. C., Braginsky, M., Yurovsky, D., \& Marchman, V. A. (2017).
 Wordbank: An open repository for developmental vocabulary data.
-\emph{Journal of Child Language}, \emph{44}(3), 677--694.
+\emph{Journal of Child Language}, \emph{44}(3), 677.
 
-\leavevmode\hypertarget{ref-Frank2021}{}%
+\leavevmode\hypertarget{ref-frank2021}{}%
 Frank, M. C., Braginsky, M., Yurovsky, D., \& Marchman, V. A. (2021).
@@
 wordbank project}. MIT Press.
+
+\leavevmode\hypertarget{ref-holland1993differential}{}%
+Holland, P. W., \& Wainer, H. (1993). \emph{Differential item
+functioning}. Routledge.
+
+\leavevmode\hypertarget{ref-huttenlocher1991early}{}%
+Huttenlocher, J., Haight, W., Bryk, A., Seltzer, M., \& Lyons, T.
+(1991). Early vocabulary growth: Relation to language input and gender.
+\emph{Developmental Psychology}, \emph{27}(2), 236.
+
+\leavevmode\hypertarget{ref-macwhinney2000childes}{}%
+MacWhinney, B. (2000). \emph{The CHILDES project: Tools for analyzing
+talk. Transcription format and programs} (Vol. 1). Psychology Press.
+
+\leavevmode\hypertarget{ref-marchman2008speed}{}%
+Marchman, V. A., \& Fernald, A. (2008). Speed of word recognition and
+vocabulary knowledge in infancy predict cognitive and language outcomes
+in later childhood. \emph{Developmental Science}, \emph{11}(3), F9--F16.
+
+\leavevmode\hypertarget{ref-petersen2018gender}{}%
+Petersen, J. (2018). Gender difference in verbal performance: A
+meta-analysis of united states state performance assessments.
+\emph{Educational Psychology Review}. Springer.
+
+\leavevmode\hypertarget{ref-reckase2009}{}%
+Reckase, M. D. (2009). Multidimensional item response theory models. In
+\emph{Multidimensional item response theory} (pp. 79--112). Springer.
+
+\leavevmode\hypertarget{ref-rowe2009differences}{}%
+Rowe, M. L., \& Goldin-Meadow, S. (2009). Differences in early gesture
+explain SES disparities in child vocabulary size at school entry.
+\emph{Science}, \emph{323}(5916), 951--953.
+
+\leavevmode\hypertarget{ref-schwab2016}{}%
+Schwab, J. F., \& Lew-Williams, C. (2016). Language learning,
+socioeconomic status, and child-directed speech. \emph{WIREs Cognitive
+Science}, \emph{7}, 264--275.
+http://doi.org/\href{https://doi.org/10.1002/wcs.1393}{10.1002/wcs.1393}
 

### 2022-02-01 paper/cogsci.tex

@@
 This study used a database of American English CDIs to identify words
-that showed strong bias for particular demographic groups of children,
+demonstrating strong bias for particular demographic groups of children,
 on dimensions of sex (male vs.~female), race (white vs.~non-white), and
@@
 language acquisition; word learning; measuring instrument bias;
-demographics;
+demographic bias
 \end{abstract}
@@
 
-\begin{CodeChunk}
-\begin{figure*}[t]
-
-{\centering \includegraphics[width=\linewidth]{figs/baseline-ability-vs-age-hor-1}
-
-}
-
-\caption[Language ability vs]{Language ability vs. age by demographic group, from the baseline Rasch model.}\label{fig:baseline-ability-vs-age-hor}
-\end{figure*}
-\end{CodeChunk}
-
 CDIs are a useful tool for measuring demographic effects because of
@@
 
-We analyze parent-reported Wordbank data from 5520 American English CDI:
-Words \& Sentences administrations for children 16 to 30 months of age
-(Frank, Braginsky, Yurovsky, \& Marchman, 2017, 2021). Full demographic
-data are not reported in some datasets contributed to Wordbank: sex was
-available for 4094 children, race/ethnicity was available for 2715
-children, and maternal education (a proxy for socioeconomic status; SES)
-was available for 5520 children.
+We analyzed parent-reported Wordbank data from 5520 American English
+CDI: Words \& Sentences administrations for children 16 to 30 months of
+age (Frank, Braginsky, Yurovsky, \& Marchman, 2017, 2021). Full
+demographic data are not reported in some datasets contributed to
+Wordbank: sex was available for 4094 children, race/ethnicity was
+available for 2715 children, and maternal education (a proxy for
+socioeconomic status; SES) was available for 5520 children.
 
@@
 Hispanic, and 93 ``Other'' children. Due to sparse data for many
-categories, we binarized participants' race/ethnicity as White (2202) or
-Non-white (513), recognizing that there may be important variation
+categories, we binarized participants' race andethnicity as White (2202)
+or Non-white (513), recognizing that there is important variation
 between groups that this will fail to capture. Data for the maternal
-education analysis included CDI data from children whose mother's had
-the following levels of education: 8 with no more than primary school
+education analysis included CDI data from children whose mothers had the
+following levels of education: 8 with no more than primary school
 education, 123 with some secondary school, 416 with no more than
@@
 
-First, we will examine the size of demographic effects on language
-ability in a baseline Rasch model fitted without regard to demographic
-group. Then we will fit a multigroup Rasch model for each demographic
-factor, and characterize the between-group differences in item
-difficulties. Next, we will re-examine the size of demographic effects
-after pruning biased items from the CDI:WS using a varying threshold.
+First, we examine the size of demographic effects on language ability in
+a baseline Rasch model fitted without regard to demographic group. Then
+we fit a multigroup Rasch model for each demographic factor, and
+characterize the between-group differences in item difficulties. Next,
+we re-examine the size of demographic effects after pruning biased items
+from the CDI:WS using different criteria, and recommend a threshold.
 Finally, we measure the strength of association between sex-related
@@
 
-\hypertarget{demographic-effects-in-baseline-rasch-model}{%
-\subsection{Demographic effects in baseline Rasch
-model}\label{demographic-effects-in-baseline-rasch-model}}
+\begin{CodeChunk}
+\begin{figure*}[t]
+
+{\centering \includegraphics[width=\linewidth]{figs/baseline-ability-vs-age-hor-1}
+
+}
+
+\caption[Language ability vs]{Language ability vs. age by demographic group, from the baseline Rasch model.}\label{fig:baseline-ability-vs-age-hor}
+\end{figure*}
+\end{CodeChunk}
+
+\hypertarget{demographic-effects-in-the-baseline-rasch-model}{%
+\subsection{Demographic effects in the baseline Rasch
+model}\label{demographic-effects-in-the-baseline-rasch-model}}
 
@@
 well-known by males. For the maternal education and race GLIMMERs
-(Figures 3 and 4)--and in the rest of the full sex GLIMMER, there are
+(Figures 3 and 4) -- and in the rest of the full sex GLIMMER, there are
 not clusters, but rather a continuum of smoothly varying differences
-with overlap. This makes identifying items with DIF quite difficult, as
-different methods are likely to yield inconsistent results (Stenhaug,
-Frank, \& Domingue, 2021). Hence, we next characterized the
+with overlap. This pattern makes identifying items with DIF quite
+difficult, as different methods are likely to yield inconsistent results
+(Stenhaug, Frank, \& Domingue, 2021). Hence, we next characterized the
 distributions of the item-level group difficulty differences, and
@@
 that the female language advantage is pervasive across the CDI:WS, and
-suggests that it is likely to be a real ability advantage, unless the
-bulk of these items are actually spoken more often to female children
-than to male children.
+suggests that it is likely to reflect more than the effect of a small
+number of biased words (cf. Frank, Braginsky, Yurovsky, \& Marchman,
+2021).
 
@@
 than low-ME children. With the mean and median difficulty differences
-close to 0, and roughly half of the words favoring each ME group, it is
-tempting to conclude that the CDI:WS items are somewhat balanced with
-respect to ME (and thus SES, by proxy).
+close to 0, and roughly half of the words favoring each ME group, it
+seems that the CDI:WS items are somewhat balanced with respect to ME
+(and thus SES, by proxy).
 
@@
 
-Given the smoothly-varying distributions of DIF shown by each
-demographic factor, we chose to evaluate multiple thresholds for pruning
-the extreme-valued items from each distribution. For each demographic
-factor, we pruned items with difficulty difference from the mean at
-varying thresholds, from \textgreater.25 SD to \textgreater3 SD, in
-increments of .25 SD. At each SD threshold, a potentially different
-subset of items are excluded for each model, but with the goal of
-creating a single CDI that is less biased on all dimensions, we pruned
-the union of the subsets excluded from each model. For example, pruning
-\textgreater2SD from the mean of each model excluded 25 sex-biased
-items, 39 ME-biased items, and 31 race-biased items, with their union
-being 76 unique items. Figure 3 shows the demographic effects
-(\(\beta\)s) at different exclusion thresholds.
+If only a small number of items show DIF, pruning decisions are simple;
+this situation was not the case for any of the three characteristics we
+examined. Instead, distributions of DIF varied smoothly by each
+demographic factor. Thus, we chose to evaluate multiple thresholds for
+pruning the extreme-valued items from each distribution. For each
+demographic factor, we pruned items with a between-group difficulty
+difference from the mean difference at varying thresholds, from
+\textgreater.25 SD to \textgreater3 SD, in increments of .25 SD. At each
+SD threshold, a potentially different subset of items are excluded for
+each model, but with the goal of creating a single CDI that is less
+biased on all dimensions, we pruned the union of the subsets excluded
+from each model. For example, pruning \textgreater2SD from the mean of
+each model excluded 25 sex-biased items, 39 ME-biased items, and 31
+race-biased items, with their union being 76 unique items. Figure 3
+shows the demographic effects (\(\beta\)s) at different exclusion
+thresholds.
 
@@
 \textgreater.25 SD difficulty difference -- 661/680 (97\%) of the CDI:WS
--- best optimized this, but seems too extreme of a culling. The next
+-- best optimized this, but seems far too extreme of a culling. The next
 best thresholds are \textgreater.5 SD -- again, too extreme -- or 3 SD,
@@
 to sex bias). Only slightly worse than these is \textgreater2.25 SD,
-which excludes 59 items (9\% of the CDI:WS), and shows a modest effect
-of both race (\(\beta=0.48\)) and maternal education (\(\beta=.19\)),
-and a near-median effect size for sex (\(\beta=-.59\)). We characterize
-these 59 potentially biased items below, and consider whether we might
+which excludes 59 items (9\%), and shows a modest effect of both race
+(\(\beta=0.48\)) and maternal education (\(\beta=.19\)), and a
+near-median effect size for sex (\(\beta=-.59\)). We characterize these
+59 potentially biased items below, and consider whether we might
 recommend pruning them from the CDI:WS.
@@
 advantage for that CDI word was modest, but significant (\(r=0.18\),
-\(p<.001\)), suggesting that some of the sex bias seen in CDI items is
-explained by differences in language input received by girls vs.~boys.
+\(p<.001\)). Some of the sex bias seen in CDI items is related to
+differences in language input received by girls vs.~boys -- though of
+course this differential input could be elicited due to different
+interests.
 
@@
 children's early vocabulary, for potential demographic bias, examining
-the distribution of words' estimated difficulties for children of
-mothers with high- vs.~low-education children (a proxy for family SES),
-females vs.~males, and for white vs.~non-white children. The IRT-based
-analysis revealed differential item functioning (DIF) for many items
-along each demographic dimension, but only in the case of sex were clear
-clusters of items that were more well-known to females (including
-feminine clothing and genitalia), and a clear item that was more
-well-known to males (male genitalia). For the rest of the items, and for
-SES- and race-based analysis, there was a smooth continuum of DIF,
-making the boundary of true DIF subjective, as this would rely on
-knowing the true difference in language ability between groups--which we
-reciprocally estimate from instruments like the CDI. To move forward, we
-identified candidate DIF items systematically pruning the extremes of
-each distribution, excluding outliers for each demographic across a
-range of thresholds, and seeking to minimize the size of demographic
-effects. Although many exclusion thresholds decreased the size of
-maternal education and race effects, the female advantage actually grew
-under most prunings, with the majority of excluded extrema being
+the distribution of word difficulties for children of mothers with high
+vs.~low education (a proxy for family SES), for female vs.~male
+children, and for white vs.~non-white children. Our IRT-based analysis
+revealed differential item functioning (DIF) for many items along each
+demographic dimension, but only in the case of sex were clear clusters
+of items that were more well-known to females (including feminine
+clothing and genitalia), and a clear item that was more well-known to
+males (male genitalia). For the rest of the items, and for SES- and
+race-based analysis, there was a smooth continuum of DIF, making the
+boundary of true DIF subjective. To move forward, we identified
+candidate DIF items by systematically pruning the extremes of each
+distribution, excluding outliers for each demographic across a range of
+thresholds, and seeking to minimize the size of demographic effects.
+Although many exclusion thresholds decreased the size of maternal
+education and race effects, the female advantage actually grew under
+most prunings, with the majority of excluded extrema being
 stereotypically-male nouns (e.g, ``truck,'' ``fireman'').
@@
 of children's early language, there is only a finite set of
-early-learned words to choose from--and we may expect many of them to be
-biased for various environmental reasons (e.g., boys and girls know
-their respective genitalia; children in Florida may not use mittens or
-skis). Hence, the presence of DIF on a set of items could indicate a
-degree of varying linguistic input, rather than an actual advantage for
-one demographic over another.
+early-learned words to choose from -- and we may expect many of them to
+be biased for various environmental reasons (e.g., boys and girls know
+the names for their respective anatomy; children in Florida may not use
+mittens or skis). Hence, the presence of DIF on a set of items could
+indicate varying linguistic and environmental input, rather than an
+actual advantage for one demographic over another.
 
@@
 and ME-based bias and not eliminating too many items, we identified 59
-suspect items--the majority of which favored white/high-ME groups. These
+items -- the majority of which favored white/high-ME groups. These
 extrema were predominantly nouns, including many animals (esp.~for
@@
 outliers, which may suggest that these lexical classes are less prone to
-environmental influence, and perhaps more generally unbiased. Pruning
-these 59 outliers would decrease the demographic effects of maternal
-education (unpruned \(\beta=.23\); pruned \(\beta=.19\)) and race
-(unpruned \(\beta=.50\); pruned \(\beta=.48\)). In contrast, the
-majority of the sex-biased outliers were easier for the disadvantaged
-group (males), meaning that the removal of these extrema increased the
-female language advantage on the CDI:WS (unpruned \(\beta=0.56\); pruned
-\(\beta=.59\)). In aggregate, these analyses seem to suggest that the
-CDI:WS may slightly overestimate the magnitude of differences in
-language ability along the dimensions of maternal education and race,
-and perhaps underestimate the female advantage.
+bias via the internal or external mechanisms leading to the effects we
+observed. Pruning these 59 outliers would slightly decrease the
+demographic effects of maternal education (unpruned \(\beta=.23\);
+pruned \(\beta=.19\)) and race (unpruned \(\beta=.50\); pruned
+\(\beta=.48\)). In contrast, the majority of the sex-biased outliers
+were easier for the disadvantaged group (males), meaning that the
+removal of these extrema would increase the female language advantage on
+the CDI:WS (unpruned \(\beta=0.56\); pruned \(\beta=.59\)). In
+aggregate, these analyses seem to suggest that the CDI:WS may very
+slightly overestimate the magnitude of differences in language ability
+along the dimensions of maternal education and race, and perhaps
+underestimate the female advantage.
 
 This investigation is only a first step in measuring demographic bias in
-the items on the CDI:WS. More research is needed to determine which CDI
-items that show DIF are more subject to environmental influence, as we
-found only a modest association between the relative frequency of words
-in child-directed speech to boys vs.~girls and the amount of sex-based
-DIF. Examining the relative frequency of CDI words across households
-with varying maternal education and race may reveal similarly weak
-associations, or there may be strong associations only among particular
-categories of words. It may be that particular activity
+the items on the CDI:WS. More research is needed to determine whether
+CDI items that show DIF are more related to environmental frequencies,
+as we found only a modest association between the relative frequency of
+words in child-directed speech to boys vs.~girls and the amount of
+sex-based DIF. Examining the relative frequency of CDI words across
+households with varying maternal education and race may reveal similarly
+weak associations, or there may be strong associations only among
+particular categories of words. It may be that particular activity
 contexts--popular with some demographic groups, and not others--may be
@@
 high-ME children, what are the activities (and associated vocabulary)
-that low-SES households are instead engaging in? Note, too, that SES,
-race, and sex are only three of many possible demographic dimensions of
-interest. For example, geographic region is likely predictive of
-children's early experience with (and thus knowledge of) CDI items
-related to winter weather (e.g., ``snow,'' ``mittens'') or outdoors
-activities (e.g.~``camping''). A truly fair test of children's early
-vocabulary would contain a representative and balanced sample of words
-from all activities that children engage in, across demographic groups.
+that low-SES households are instead engaging in? Critically this
+enterprise will require a greater diversity of publicly-available
+corpora of child language; we view the lack of such data as a
+substantial issue.
+
+SES, race, and sex are only three of many possible demographic
+dimensions of interest. For example, geographic region is likely
+predictive of children's early experience with (and thus knowledge of)
+CDI items related to winter weather (e.g., ``snow,'' ``mittens'') or
+outdoors activities (e.g.~``camping''). A truly fair test of children's
+early vocabulary would contain a representative and balanced sample of
+words from all activities that children engage in, across demographic
+groups.
 

### 2022-02-01 paper/cogsci.Rmd

@@
                       fig.pos = "tb", fig.path='figs/',
-                      echo=F, warning=F, cache=F, 
+                      echo=F, warning=F, cache=TRUE, 
                       message=F, sanitize = T)
@@
 
+CDIs are a useful tool for measuring demographic effects because of their wide use across languages. 
+But their use in this context poses a chicken and egg problem. 
+While it is possible that there is an underlying difference in mean language ability between demographic groups, CDIs could also have bias for or against particular groups based on the words they include.
+
+The idea that some items on a test may show bias, favoring one group over another, is known in psychometrics as Differential Item Function (DIF; @holland1993differential]. 
+DIF can decrease the validity of a test, as the test may be overestimating the ability of test-takers in one group and underestimating the ability of another group -- not because of any underlying mean difference in ability between the groups, but simply because the test is unfair [@camilli2013ongoing].
+For example, if a vocabulary test is composed almost entirely of the names of farm equipment, the test will underestimate the knowledge of urban children, who have experience of other contexts, despite being less familiar with farm equipment than their rural peers.
+Of course, there may be an ability difference between demographic groups, as is likely the case for the female language advantage, which has been documented in many languages [@eriksson2012differences; @frank2021].
+And both observations could be true -- a test could be biased to inflate a true demographic difference.
+
+Our goal in the current study is to use DIF analyses to assess these possibilities. 
+Specifically, we test the 680 vocabulary items on the American CDI:WS for DIF along three dichotomized demographic dimensions: sex (male vs.
+female), maternal education (no more than secondary vs.
+at least some college), and race (white vs.
+non-white).
+
+The outline of this paper is as follows.
+First, we introduce the Wordbank data and the IRT model, and examine the overall size of demographic differences in early word learning based on the full CDI:WS.
+We then fit the IRT model to both groups for each demographic factor, and look at the item parameters for evidence of DIF, noting in particular how many items are significantly biased in favor of each group, and describing them qualitatively.
+Next, we use these item parameters to systematically prune biased items from the CDI:WS, and examine how the estimated size of demographic differences change as we prune more items.
+<!-- We identify a set of suspect items to eliminate (or in the future, replace), and provide updated estimates of the effect size of these demographic variables, were the biased items to be eliminated.-->
+Finally, using a corpus of child-directed speech, we examine the extent to which variation in word frequencies heard by boys vs. girls predicts sex-based DIF.
+Based on these analyses, we provide recommendations for next steps to be taken to identify and potentially replace items on the CDI showing demographic bias.
+
+# Methods
+
+## Vocabulary Data
+
 ```{r import-data}
@@
 ```
+
+### Participants
+
+```{r, echo=F}
+sex_tab <- table(d_demo$sex)
+ses_tab <- table(d_demo$ses_group)
+eth_tab <- table(d_demo$ethnicity)
+eth2_tab <- table(d_demo$eth_group)
+momed_tab <- table(d_demo$mom_ed)
+```
+
+We analyzed parent-reported Wordbank data from `r nrow(d_demo)` American English CDI: Words & Sentences administrations for children `r min(d_demo$age)` to `r max(d_demo$age)` months of age [@frank2017; @frank2021].
+Full demographic data are not reported in some datasets contributed to Wordbank: sex was available for `r nrow(d_demo)-sum(is.na(d_demo$sex))` children, race/ethnicity was available for `r nrow(d_demo)-sum(is.na(d_demo$ethnicity))` children, and maternal education (a proxy for socioeconomic status; SES) was available for `r nrow(d_demo)-sum(is.na(d_demo$ses_group))` children.
+
+The analysis of sex-based differences included CDI administrations from `r sex_tab["Female"]` female and `r sex_tab["Male"]` male children.
+The analysis of race-based differences included data from `r eth_tab["White"]` white, `r eth_tab["Asian"]` Asian, `r eth_tab["Black"]` Black, `r eth_tab["Hispanic"]` Hispanic, and `r eth_tab["Other"]` "Other" children.
+Due to sparse data for many categories, we binarized participants' race andethnicity as White (`r eth2_tab["White"]`) or Non-white (`r eth2_tab["Nonwhite"]`), recognizing that there is important variation between groups that this will fail to capture.
+Data for the maternal education analysis included CDI data from children whose mothers had the following levels of education: `r momed_tab["Primary"]` with no more than primary school education, `r momed_tab["Some Secondary"]` with some secondary school, `r momed_tab["Secondary"]` with no more than secondary school, `r momed_tab["Some College"]` with some college, `r momed_tab["College"]` with no more than a college degree, `r momed_tab["Some Graduate"]` with no more than some graduate school, and `r momed_tab["Graduate"]` with a graduate degree.
+Again due to data sparsity, we binarized the `r ses_tab["high"]` children whose mothers had at least some college or more as high maternal education (high-ME), and those whose mothers had at most high school (`r ses_tab["low"]` children) as low maternal education (low-ME).
+
+## Rasch Model
+
+The Rasch model, also known as the 1-parameter logistic (1PL) model, is the simplest Item Response Theory model, and is thus the easiest to use to investigate potential differences in item function across different groups of participants.
+The Rasch model jointly estimates for each child $j$ a latent ability $\theta_j$, and for each item $i$ a difficulty parameter $b_i$.
+In the model, the probability of child $j$ knowing (i.e., producing or understanding) a given item $i$ is
+
+$$P_{i}(x_i = 1 | b_{i},\theta_j ) = \frac{1}{1 + e^{-D(\theta_j - b_i )}}$$
+
+\noindent where $D$ is a constant scaling parameter ($D=1.702$) which makes the logistic closely match the ogive function in traditional factor analysis [@R-mirt; @reckase2009].
+Child ability ($\theta$) and item difficulty ($b$) distributions are standardized (i.e., mean of 0), and expected to be normally-distributed. 
+Children with high latent ability ($\theta$) will be more likely to produce any given item than children with lower latent ability, and more difficult items will be produced by fewer children (at any given $\theta$) than easier items.
+
+In the multigroup Rasch model, an item's difficulty is allowed to vary by group. 
+For example, in the sex-based multigroup model, item $i$'s difficulty is $b_{i}^{female}$ for females, and $b_{i}^{male}$ for males.
+To identify DIF, a multigroup Rasch model will be fitted for each demographic dimension of interest (sex, maternal education, and ethnicity), and we will examine the between-group difficulty difference for each item (e.g., $d_i = b_{i}^{female} - b_{i}^{male}$).
+If there is no DIF for a given item, then $d_i \approx 0$ as the two groups find the item equally difficult.
+
+# Results
+
+First, we examine the size of demographic effects on language ability in a baseline Rasch model fitted without regard to demographic group.
+Then we fit a multigroup Rasch model for each demographic factor, and characterize the between-group differences in item difficulties.
+Next, we re-examine the size of demographic effects after pruning biased items from the CDI:WS using different criteria, and recommend a threshold.
+Finally, we measure the strength of association between sex-related differences in language input and the degree of sex-related DIF.
 
@@
 ```
-CDIs are a useful tool for measuring demographic effects because of their wide use across languages. 
-<!-- While many languages show similar demographic effects on early word learning [e.g., the female advantage: @eriksson2012differences; and the first-born advantage: @frank2021], the size of these demographic effects vary across languages [see @frank2021, Ch.
-6].
--->
-<!-- Why might we see such variation? -->
-But their use in this context poses a chicken and egg problem. 
-While it is possible that there is an underlying difference in mean language ability between demographic groups, CDIs could also have bias for or against particular groups based on the words they include.
-<!-- If a hypothetical vocabulary test only consisted of words for cold-weather gear, it would naturally show bias against children who grew up in hot climates (or who were born at certain times of year).
--->
-<!-- , and that cross-linguistic variation is driven by differences in the language environments of different cultures, could it also be possible that the  -->
 
-The idea that some items on a test may show bias, favoring one group over another, is known in psychometrics as Differential Item Function (DIF; @holland1993differential]. 
-DIF can decrease the validity of a test, as the test may be overestimating the ability of test-takers in one group and underestimating the ability of another group -- not because of any underlying mean difference in ability between the groups, but simply because the test is unfair [@camilli2013ongoing].
-For example, if a vocabulary test is composed almost entirely of the names of farm equipment, the test will underestimate the knowledge of urban children, who have experience of other contexts, despite being less familiar with farm equipment than their rural peers.
-Of course, there may be an ability difference between demographic groups, as is likely the case for the female language advantage, which has been documented in many languages [@eriksson2012differences; @frank2021].
-And both observations could be true -- a test could be biased to inflate a true demographic difference.
-
-Our goal in the current study is to use DIF analyses to assess these possibilities. 
-Specifically, we test the 680 vocabulary items on the American CDI:WS for DIF along three dichotomized demographic dimensions: sex (male vs.
-female), maternal education (no more than secondary vs.
-at least some college), and race (white vs.
-non-white).
-
-<!-- A variety of statistical methods for detecting DIF have been proposed, and investigations have in several instances identified DIF for many items on tests favoring one group over another. -->
-
-The outline of this paper is as follows.
-First, we introduce the Wordbank data and the IRT model, and examine the overall size of demographic differences in early word learning based on the full CDI:WS.
-We then fit the IRT model to both groups for each demographic factor, and look at the item parameters for evidence of DIF, noting in particular how many items are significantly biased in favor of each group, and describing them qualitatively.
-Next, we use these item parameters to systematically prune biased items from the CDI:WS, and examine how the estimated size of demographic differences change as we prune more items.
-<!-- We identify a set of suspect items to eliminate (or in the future, replace), and provide updated estimates of the effect size of these demographic variables, were the biased items to be eliminated.-->
-Finally, using a corpus of child-directed speech, we examine the extent to which variation in word frequencies heard by boys vs. girls predicts sex-based DIF.
-Based on these analyses, we provide recommendations for next steps to be taken to identify and potentially replace items on the CDI showing demographic bias.
-
-# Methods
-
-## Vocabulary Data
-
-### Participants
-
-```{r, echo=F}
-sex_tab <- table(d_demo$sex)
-ses_tab <- table(d_demo$ses_group)
-eth_tab <- table(d_demo$ethnicity)
-eth2_tab <- table(d_demo$eth_group)
-momed_tab <- table(d_demo$mom_ed)
-```
-
-We analyze parent-reported Wordbank data from `r nrow(d_demo)` American English CDI: Words & Sentences administrations for children `r min(d_demo$age)` to `r max(d_demo$age)` months of age [@frank2017; @frank2021].
-Full demographic data are not reported in some datasets contributed to Wordbank: sex was available for `r nrow(d_demo)-sum(is.na(d_demo$sex))` children, race/ethnicity was available for `r nrow(d_demo)-sum(is.na(d_demo$ethnicity))` children, and maternal education (a proxy for socioeconomic status; SES) was available for `r nrow(d_demo)-sum(is.na(d_demo$ses_group))` children.
-
-The analysis of sex-based differences included CDI administrations from `r sex_tab["Female"]` female and `r sex_tab["Male"]` male children.
-The analysis of race-based differences included data from `r eth_tab["White"]` white, `r eth_tab["Asian"]` Asian, `r eth_tab["Black"]` Black, `r eth_tab["Hispanic"]` Hispanic, and `r eth_tab["Other"]` "Other" children.
-Due to sparse data for many categories, we binarized participants' race/ethnicity as White (`r eth2_tab["White"]`) or Non-white (`r eth2_tab["Nonwhite"]`), recognizing that there may be important variation between groups that this will fail to capture.
-Data for the maternal education analysis included CDI data from children whose mother's had the following levels of education: `r momed_tab["Primary"]` with no more than primary school education, `r momed_tab["Some Secondary"]` with some secondary school, `r momed_tab["Secondary"]` with no more than secondary school, `r momed_tab["Some College"]` with some college, `r momed_tab["College"]` with no more than a college degree, `r momed_tab["Some Graduate"]` with no more than some graduate school, and `r momed_tab["Graduate"]` with a graduate degree.
-Again due to data sparsity, we binarized the `r ses_tab["high"]` children whose mothers had at least some college or more as high maternal education (high-ME), and those whose mothers had at most high school (`r ses_tab["low"]` children) as low maternal education (low-ME).
-
-## Rasch Model
-
-The Rasch model, also known as the 1-parameter logistic (1PL) model, is the simplest Item Response Theory model, and is thus the easiest to use to investigate potential differences in item function across different groups of participants.
-The Rasch model jointly estimates for each child $j$ a latent ability $\theta_j$, and for each item $i$ a difficulty parameter $b_i$.
-In the model, the probability of child $j$ knowing (i.e., producing or understanding) a given item $i$ is
-
-$$P_{i}(x_i = 1 | b_{i},\theta_j ) = \frac{1}{1 + e^{-D(\theta_j - b_i )}}$$
-
-\noindent where $D$ is a constant scaling parameter ($D=1.702$) which makes the logistic closely match the ogive function in traditional factor analysis [@R-mirt; @reckase2009].
-Child ability ($\theta$) and item difficulty ($b$) distributions are standardized (i.e., mean of 0), and expected to be normally-distributed. 
-Children with high latent ability ($\theta$) will be more likely to produce any given item than children with lower latent ability, and more difficult items will be produced by fewer children (at any given $\theta$) than easier items.
-
-In the multigroup Rasch model, an item's difficulty is allowed to vary by group. 
-For example, in the sex-based multigroup model, item $i$'s difficulty is $b_{i}^{female}$ for females, and $b_{i}^{male}$ for males.
-To identify DIF, a multigroup Rasch model will be fitted for each demographic dimension of interest (sex, maternal education, and ethnicity), and we will examine the between-group difficulty difference for each item (e.g., $d_i = b_{i}^{female} - b_{i}^{male}$).
-If there is no DIF for a given item, then $d_i \approx 0$ as the two groups find the item equally difficult.
-
-# Results
-
-First, we will examine the size of demographic effects on language ability in a baseline Rasch model fitted without regard to demographic group.
-Then we will fit a multigroup Rasch model for each demographic factor, and characterize the between-group differences in item difficulties.
-Next, we will re-examine the size of demographic effects after pruning biased items from the CDI:WS using a varying threshold.
-Finally, we measure the strength of association between sex-related differences in language input and the degree of sex-related DIF.
-
-## Demographic effects in baseline Rasch model
+## Demographic effects in the baseline Rasch model
 We fit a baseline Rasch model to the entire dataset, without demographic information.
@@
 For example, there is some clustering at the top and bottom of Figure 2: at the top, "vagina"^[Parents using the CDI are instructed to check genital items if the child produces whatever word the family uses.], "tights", "dress (object)", and "doll" form a cluster of items that are much more well-known for females, while at the bottom, "penis" stands out as much more well-known by males.
-For the maternal education and race GLIMMERs (Figures 3 and 4)--and in the rest of the full sex GLIMMER, there are not clusters, but rather a continuum of smoothly varying differences with overlap. 
-This makes identifying items with DIF quite difficult, as different methods are likely to yield inconsistent results [@stenhaug2021treading].
+For the maternal education and race GLIMMERs (Figures 3 and 4) -- and in the rest of the full sex GLIMMER, there are not clusters, but rather a continuum of smoothly varying differences with overlap. 
+This pattern makes identifying items with DIF quite difficult, as different methods are likely to yield inconsistent results [@stenhaug2021treading].
 Hence, we next characterized the distributions of the item-level group difficulty differences, and measured the influence of pruning a varying number of items on the demographic effect sizes.
@@
 For sex, the median difficulty difference (male-female) was `r round(median(mm_sex$d_diff),2)` (M=`r round(mean(mm_sex$d_diff),2)`, sd=`r round(sd(mm_sex$d_diff),2)`), with `r length(which(mm_sex$d_diff>0))`/680 items being easier for females than males.
-The fact that the bulk of this distribution favors females shows that the female language advantage is pervasive across the CDI:WS, and suggests that it is likely to be a real ability advantage, unless the bulk of these items are actually spoken more often to female children than to male children.
+The fact that the bulk of this distribution favors females shows that the female language advantage is pervasive across the CDI:WS, and suggests that it is likely to reflect more than the effect of a small number of biased words [cf.
+@frank2021].
+<!-- be a real ability advantage, unless the bulk of these items are actually spoken more often to female children than to male children. -->
 
@@
 For maternal education (ME), the median difficulty difference (low-high) was `r round(median(mm_ses$d_diff),2)` (M=`r round(mean(mm_ses$d_diff),2)`, sd=`r round(sd(mm_ses$d_diff),2)`), with `r length(which(mm_ses$d_diff>0))`/680 items being easier for high-ME than low-ME children.
-With the mean and median difficulty differences close to 0, and roughly half of the words favoring each ME group, it is tempting to conclude that the CDI:WS items are somewhat balanced with respect to ME (and thus SES, by proxy).
+With the mean and median difficulty differences close to 0, and roughly half of the words favoring each ME group, it seems that the CDI:WS items are somewhat balanced with respect to ME (and thus SES, by proxy).
 
@@
 
-Given the smoothly-varying distributions of DIF shown by each demographic factor, we chose to evaluate multiple thresholds for pruning the extreme-valued items from each distribution.
-For each demographic factor, we pruned items with difficulty difference from the mean at varying thresholds, from >.25 SD to >3 SD, in increments of .25 SD. 
+If only a small number of items show DIF, pruning decisions are simple; this situation was not the case for any of the three characteristics we examined. 
+Instead, distributions of DIF varied smoothly by each demographic factor. 
+Thus, we chose to evaluate multiple thresholds for pruning the extreme-valued items from each distribution.
+For each demographic factor, we pruned items with a between-group difficulty difference from the mean difference at varying thresholds, from >.25 SD to >3 SD, in increments of .25 SD. 
 At each SD threshold, a potentially different subset of items are excluded for each model, but with the goal of creating a single CDI that is less biased on all dimensions, we pruned the union of the subsets excluded from each model.
@@
 Where is the exclusion threshold that jointly minimizes the effect sizes of these demographic variables? 
-The strict exclusion of all items with >.25 SD difficulty difference -- 661/680 (97%) of the CDI:WS -- best optimized this, but seems too extreme of a culling.
+The strict exclusion of all items with >.25 SD difficulty difference -- 661/680 (97%) of the CDI:WS -- best optimized this, but seems far too extreme of a culling.
 The next best thresholds are >.5 SD -- again, too extreme -- or 3 SD, a remarkably lax criterion that only excludes 19 items in total (11 due to sex bias).
-Only slightly worse than these is >2.25 SD, which excludes 59 items (9% of the CDI:WS), and shows a modest effect of both race ($\beta=0.48$) and maternal education ($\beta=.19$), and a near-median effect size for sex ($\beta=-.59$).
+Only slightly worse than these is >2.25 SD, which excludes 59 items (9%), and shows a modest effect of both race ($\beta=0.48$) and maternal education ($\beta=.19$), and a near-median effect size for sex ($\beta=-.59$).
 We characterize these 59 potentially biased items below, and consider whether we might recommend pruning them from the CDI:WS.
@@
 a male child and the size of the female (vs.
-male) advantage for that CDI word was modest, but significant ($r=0.18$, $p<.001$), suggesting that some of the sex bias seen in CDI items is explained by differences in language input received by girls vs.
-boys.
+male) advantage for that CDI word was modest, but significant ($r=0.18$, $p<.001$).
+Some of the sex bias seen in CDI items is related to differences in language input received by girls vs. boys -- though of course this differential input could be elicited due to different interests.
 
@@
 
-We investigated the CDI:WS, a popular parent-report measure of children's early vocabulary, for potential demographic bias, examining the distribution of words' estimated difficulties for children of mothers with high- vs.
-low-education children (a proxy for family SES), females vs.
-males, and for white vs.
+We investigated the CDI:WS, a popular parent-report measure of children's early vocabulary, for potential demographic bias, examining the distribution of word difficulties for children of mothers with high vs.
+low education (a proxy for family SES), for female vs.
+male children, and for white vs.
 non-white children.
-The IRT-based analysis revealed differential item functioning (DIF) for many items along each demographic dimension, but only in the case of sex were clear clusters of items that were more well-known to females (including feminine clothing and genitalia), and a clear item that was more well-known to males (male genitalia).
-For the rest of the items, and for SES- and race-based analysis, there was a smooth continuum of DIF, making the boundary of true DIF subjective, as this would rely on knowing the true difference in language ability between groups--which we reciprocally estimate from instruments like the CDI.
-To move forward, we identified candidate DIF items systematically pruning the extremes of each distribution, excluding outliers for each demographic across a range of thresholds, and seeking to minimize the size of demographic effects.
+Our IRT-based analysis revealed differential item functioning (DIF) for many items along each demographic dimension, but only in the case of sex were clear clusters of items that were more well-known to females (including feminine clothing and genitalia), and a clear item that was more well-known to males (male genitalia).
+For the rest of the items, and for SES- and race-based analysis, there was a smooth continuum of DIF, making the boundary of true DIF subjective.
+To move forward, we identified candidate DIF items by systematically pruning the extremes of each distribution, excluding outliers for each demographic across a range of thresholds, and seeking to minimize the size of demographic effects.
 Although many exclusion thresholds decreased the size of maternal education and race effects, the female advantage actually grew under most prunings, with the majority of excluded extrema being stereotypically-male nouns (e.g, "truck", "fireman").
@@
 This analysis highlights a fundamental difficulty of identifying DIF: an analyst must either know the expected magnitude of an ability difference between the two demographic groups, or know a set of "anchor" items that are equally difficult (i.e., unbiased) for both groups [for an overview, see @stenhaug2021treading].
-But in the universe of children's early language, there is only a finite set of early-learned words to choose from--and we may expect many of them to be biased for various environmental reasons (e.g., boys and girls know their respective genitalia; children in Florida may not use mittens or skis).
-Hence, the presence of DIF on a set of items could indicate a degree of varying linguistic input, rather than an actual advantage for one demographic over another.
+But in the universe of children's early language, there is only a finite set of early-learned words to choose from -- and we may expect many of them to be biased for various environmental reasons (e.g., boys and girls know the names for their respective anatomy; children in Florida may not use mittens or skis).
+Hence, the presence of DIF on a set of items could indicate varying linguistic and environmental input, rather than an actual advantage for one demographic over another.
 
-Choosing an exclusion threshold that struck a balance of reducing race- and ME-based bias and not eliminating too many items, we identified 59 suspect items--the majority of which favored white/high-ME groups.
+Choosing an exclusion threshold that struck a balance of reducing race- and ME-based bias and not eliminating too many items, we identified 59 items -- the majority of which favored white/high-ME groups.
 These extrema were predominantly nouns, including many animals (esp. for maternal education), vehicles and professions (esp. for sex). 
-Notably, no verbs and few syntactically complex items were identified as outliers, which may suggest that these lexical classes are less prone to environmental influence, and perhaps more generally unbiased.
+Notably, no verbs and few syntactically complex items were identified as outliers, which may suggest that these lexical classes are less prone to bias via the internal or external mechanisms leading to the effects we observed.
 <!--or something..seems important, as nouns are more referential/context-dependent-->
-Pruning these 59 outliers would decrease the demographic effects of maternal education (unpruned $\beta=.23$; pruned $\beta=.19$) and race (unpruned $\beta=.50$; pruned $\beta=.48$).
-In contrast, the majority of the sex-biased outliers were easier for the disadvantaged group (males), meaning that the removal of these extrema increased the female language advantage on the CDI:WS (unpruned $\beta=0.56$; pruned $\beta=.59$).
-In aggregate, these analyses seem to suggest that the CDI:WS may slightly overestimate the magnitude of differences in language ability along the dimensions of maternal education and race, and perhaps underestimate the female advantage.
+Pruning these 59 outliers would slightly decrease the demographic effects of maternal education (unpruned $\beta=.23$; pruned $\beta=.19$) and race (unpruned $\beta=.50$; pruned $\beta=.48$).
+In contrast, the majority of the sex-biased outliers were easier for the disadvantaged group (males), meaning that the removal of these extrema would increase the female language advantage on the CDI:WS (unpruned $\beta=0.56$; pruned $\beta=.59$).
+In aggregate, these analyses seem to suggest that the CDI:WS may very slightly overestimate the magnitude of differences in language ability along the dimensions of maternal education and race, and perhaps underestimate the female advantage.
 
 This investigation is only a first step in measuring demographic bias in the items on the CDI:WS.
-More research is needed to determine which CDI items that show DIF are more subject to environmental influence, as we found only a modest association between the relative frequency of words in child-directed speech to boys vs.
+More research is needed to determine whether CDI items that show DIF are more related to environmental frequencies, as we found only a modest association between the relative frequency of words in child-directed speech to boys vs.
 girls and the amount of sex-based DIF.
@@
 If certain activities are driving early word learning for high-ME children, what are the activities (and associated vocabulary) that low-SES households are instead engaging in?
-Note, too, that SES, race, and sex are only three of many possible demographic dimensions of interest.
+Critically this enterprise will require a greater diversity of publicly-available corpora of child language; we view the lack of such data as a substantial issue.
+
+SES, race, and sex are only three of many possible demographic dimensions of interest.
 For example, geographic region is likely predictive of children's early experience with (and thus knowledge of) CDI items related to winter weather (e.g., "snow", "mittens") or outdoors activities (e.g.

### 2022-02-01 paper/cogsci.Rmd
minor edits, fig tweaks

@@
 
-One challenge in comprehensive assessment of these differences is that is difficult to get a complete measure of young children's language skill.
+One challenge in comprehensive assessment of these differences is that it is difficult to get a complete measure of young children's language skill.
 Long recordings are prohibitively difficult to collect and transcribe, and yet any short recording (e.g., a 1-hour play session) will elicit only a small proportion of the words and constructions that children know.
@@
 The analysis of race-based differences included data from `r eth_tab["White"]` white, `r eth_tab["Asian"]` Asian, `r eth_tab["Black"]` Black, `r eth_tab["Hispanic"]` Hispanic, and `r eth_tab["Other"]` "Other" children.
-Due to sparse data for many categories, we binarized participants' race andethnicity as White (`r eth2_tab["White"]`) or Non-white (`r eth2_tab["Nonwhite"]`), recognizing that there is important variation between groups that this will fail to capture.
+Due to sparse data for many categories, we binarized participants' race and ethnicity as White (`r eth2_tab["White"]`) or Non-white (`r eth2_tab["Nonwhite"]`), recognizing that there is important variation between groups that this will fail to capture.
 Data for the maternal education analysis included CDI data from children whose mothers had the following levels of education: `r momed_tab["Primary"]` with no more than primary school education, `r momed_tab["Some Secondary"]` with some secondary school, `r momed_tab["Secondary"]` with no more than secondary school, `r momed_tab["Some College"]` with some college, `r momed_tab["College"]` with no more than a college degree, `r momed_tab["Some Graduate"]` with no more than some graduate school, and `r momed_tab["Graduate"]` with a graduate degree.
@@
 Finally, we measure the strength of association between sex-related differences in language input and the degree of sex-related DIF.
+The scripts and models needed to reproduce our analyses are available on [OSF](https://osf.io/57rsw/?view_only=2b6ecb61fe08458293af7421d276932a).
 
@@
 It is important, however, to inspect the full plots, for if there is a cluster of items in a GLIMMER, the analyst may conclude that these items are strong candidates for DIF on that dimension.
-For example, there is some clustering at the top and bottom of Figure 2: at the top, "vagina"^[Parents using the CDI are instructed to check genital items if the child produces whatever word the family uses.], "tights", "dress (object)", and "doll" form a cluster of items that are much more well-known for females, while at the bottom, "penis" stands out as much more well-known by males.
-For the maternal education and race GLIMMERs (Figures 3 and 4) -- and in the rest of the full sex GLIMMER, there are not clusters, but rather a continuum of smoothly varying differences with overlap. 
+For example, there is some clustering at the top and bottom of the sex GLIMMER (Fig.
+2, left): at the top, "vagina"^[Parents using the CDI are instructed to check genital items if the child produces whatever word the family uses.], "tights", "dress (object)", and "doll" form a cluster of items that are much more well-known for females, while at the bottom, "penis" stands out as much more well-known by males.
+For the maternal education and race GLIMMERs -- and in the rest of the full sex GLIMMER, there are not clusters, but rather a continuum of smoothly varying differences with overlap. 
 This pattern makes identifying items with DIF quite difficult, as different methods are likely to yield inconsistent results [@stenhaug2021treading].
@@
 
-```{r glimmer-combo, fig.env = "figure*", out.width="\\linewidth", fig.pos = "h", fig.width=6.5, fig.align = "center", set.cap.width=T, num.cols.cap=2, fig.cap = "GLIMMER plot of a sample of CDI:WS words from the sex bias model.
-Words at the top are more well-known by females, while those at the bottom are more known for males."}
+```{r glimmer-combo, fig.env = "figure*", out.width="\\linewidth", fig.pos = "h", fig.width=6.5, fig.align = "center", set.cap.width=T, num.cols.cap=2, fig.cap = "GLIMMER plots of a sample of CDI:WS words from the sex bias model (left), the maternal education model (center), and the race model (right).
+Words at the top are more well-known by one group (e.g., females, for sex), while those at the bottom are more known by the other group (e.g., males), and the selection in the middle show roughly the median difficulty difference for those groups."}
 knitr::include_graphics("figs/smGLIMMER_combo.pdf")
@@
   ggplot(aes(x=SDthreshold, y=beta, color=extreme)) + 
-  geom_point() + #theme(legend.position = "bottom") + 
+  geom_point() + theme_bw(base_size=9) +
   labs(color="Excluded items") + ylim(-.7, -.4) + ggtitle("Sex") +
@@
   ggplot(aes(x=SDthreshold, y=beta, color=extreme)) + 
-  geom_point() + 
+  geom_point() + theme_bw(base_size=9) +
   labs(color="Excluded items") + ylim(-.3, -.1) + ggtitle("Maternal Education") +
@@
   ggplot(aes(x=SDthreshold, y=beta, color=extreme)) + 
-  geom_point() + 
+  geom_point() + theme_bw(base_size=9) +
   labs(color="Excluded items") + ylim(.35, .55) + ggtitle("Race") +
@@
 #white=.48
-#highSES=0.20
+#highME=0.20
 #female=0.57
@@
 
-```{r biased-words, include=T, out.width="\\linewidth", fig.env = "figure", fig.pos = "H", fig.align='center', fig.width=3.5, fig.height=3.2, set.cap.width=T, num.cols.cap=1, fig.cap = "The 59 items showing extreme bias (difficulty difference $>2.25$ SD) for one or more demographic."}
-knitr::include_graphics("figs/extrema2p25sd.pdf")
+```{r biased-words, include=T, out.width="\\linewidth", fig.env = "figure", fig.pos = "h", fig.align='center', fig.width=3.5, fig.height=3.2, set.cap.width=T, num.cols.cap=1, fig.cap = "The 59 items showing extreme bias (difficulty difference $>2.25$ SD) for one or more demographic.
+The 19 items in bold are those showing a difference $>3$ SD."}
+knitr::include_graphics("figs/extrema2p25sd_3sd_bold.pdf")
 ```

### 2022-02-01 paper/cogsci.tex
minor edits, fig tweaks

@@
 
-\title{Estimating demographic bias on tests of children's early
-vocabulary}
+\title{Estimating demographic bias on tests of children's early vocabulary}
 
@@
 Frank (mcfrank@stanford.edu)} \\ Department of Psychology, Stanford Unviersity \\ Stanford, CA 94305 USA }
-
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newenvironment{CSLReferences}%
-  {}%
-  {\par}
 
@@
 One challenge in comprehensive assessment of these differences is that
-is difficult to get a complete measure of young children's language
+it is difficult to get a complete measure of young children's language
 skill. Long recordings are prohibitively difficult to collect and
@@
 We focus here on the MacArthur-Bates Communicative Development
-Inventories {[}CDIs; Larry Fenson et al. (1994); L. Fenson et al.
-(2007){]}, a set of parent-reported measures of children's productive
-and receptive language skills, which offer a low-cost and reliable way
-to estimate children's early language skills (Larry Fenson et al.,
-1994). CDIs have shown good predictive validity (Bornstein \& Putnick,
-2012; Duff, Reen, Plunkett, \& Nation, 2015; e.g., Larry Fenson et al.,
-1994). Our primary focus is the vocabulary checklist portion of the CDI
-Words \& Sentences (CDI:WS) form, comprised of 680 early-learned words
-across 22 categories (e.g., animals, vehicles, action words, pronouns)
-selected to assess the productive vocabulary of children 16 to 30 months
-of age. For each item on the CDI:WS, caregivers are asked to respond
-whether the target child has been heard to say (i.e.~produce) the given
-item. Children's total vocabulary score on the CDI:WS is tightly
-correlated with other facets of early language (e.g., grammatical
-competence and gesture), suggesting that the language system is
-``tightly woven'' (Frank, Braginsky, Yurovsky, \& Marchman, 2021). Due
+Inventories (CDIs; Fenson et al., 1994, 2007), a set of parent-reported
+measures of children's productive and receptive language skills, which
+offer a low-cost and reliable way to estimate children's early language
+skills (Fenson et al., 1994). CDIs have shown good predictive validity
+(e.g., Fenson et al., 1994; Bornstein \& Putnick, 2012; Duff, Reen,
+Plunkett, \& Nation, 2015). Our primary focus is the vocabulary
+checklist portion of the CDI Words \& Sentences (CDI:WS) form, comprised
+of 680 early-learned words across 22 categories (e.g., animals,
+vehicles, action words, pronouns) selected to assess the productive
+vocabulary of children 16 to 30 months of age. For each item on the
+CDI:WS, caregivers are asked to respond whether the target child has
+been heard to say (i.e.~produce) the given item. Children's total
+vocabulary score on the CDI:WS is tightly correlated with other facets
+of early language (e.g., grammatical competence and gesture), suggesting
+that the language system is ``tightly woven'' (Frank et al., 2021). Due
 to these desirable properties, CDIs have been adapted to dozens of
 languages, and a central repository of CDI data contributed from all
-over the world has been created {[}Wordbank; Frank, Braginsky, Yurovsky,
-\& Marchman (2017); Frank, Braginsky, Yurovsky, \& Marchman (2021){]}.
+over the world has been created (Wordbank; Frank, Braginsky, Yurovsky,
+\& Marchman, 2017; Frank et al., 2021).
 
@@
 the female language advantage, which has been documented in many
-languages (Eriksson et al., 2012; Frank, Braginsky, Yurovsky, \&
-Marchman, 2021). And both observations could be true -- a test could be
-biased to inflate a true demographic difference.
+languages (Eriksson et al., 2012; Frank et al., 2021). And both
+observations could be true -- a test could be biased to inflate a true
+demographic difference.
 
@@
 CDI: Words \& Sentences administrations for children 16 to 30 months of
-age (Frank, Braginsky, Yurovsky, \& Marchman, 2017, 2021). Full
-demographic data are not reported in some datasets contributed to
-Wordbank: sex was available for 4094 children, race/ethnicity was
-available for 2715 children, and maternal education (a proxy for
-socioeconomic status; SES) was available for 5520 children.
+age (Frank et al., 2017, 2021). Full demographic data are not reported
+in some datasets contributed to Wordbank: sex was available for 4094
+children, race/ethnicity was available for 2715 children, and maternal
+education (a proxy for socioeconomic status; SES) was available for 5520
+children.
 
@@
 Hispanic, and 93 ``Other'' children. Due to sparse data for many
-categories, we binarized participants' race andethnicity as White (2202)
-or Non-white (513), recognizing that there is important variation
+categories, we binarized participants' race and ethnicity as White
+(2202) or Non-white (513), recognizing that there is important variation
 between groups that this will fail to capture. Data for the maternal
@@
 Finally, we measure the strength of association between sex-related
-differences in language input and the degree of sex-related DIF.
+differences in language input and the degree of sex-related DIF. The
+scripts and models needed to reproduce our analyses are available on
+\href{https://osf.io/57rsw/?view_only=2b6ecb61fe08458293af7421d276932a}{OSF}.
 
@@
 that dimension. For example, there is some clustering at the top and
-bottom of Figure 2: at the top, ``vagina''\footnote{Parents using the
-  CDI are instructed to check genital items if the child produces
-  whatever word the family uses.}, ``tights,'' ``dress (object),'' and
-``doll'' form a cluster of items that are much more well-known for
-females, while at the bottom, ``penis'' stands out as much more
-well-known by males. For the maternal education and race GLIMMERs
-(Figures 3 and 4) -- and in the rest of the full sex GLIMMER, there are
-not clusters, but rather a continuum of smoothly varying differences
-with overlap. This pattern makes identifying items with DIF quite
-difficult, as different methods are likely to yield inconsistent results
-(Stenhaug, Frank, \& Domingue, 2021). Hence, we next characterized the
+bottom of the sex GLIMMER (Fig. 2, left): at the top,
+``vagina''\footnote{Parents using the CDI are instructed to check
+  genital items if the child produces whatever word the family uses.},
+``tights'', ``dress (object)'', and ``doll'' form a cluster of items
+that are much more well-known for females, while at the bottom,
+``penis'' stands out as much more well-known by males. For the maternal
+education and race GLIMMERs -- and in the rest of the full sex GLIMMER,
+there are not clusters, but rather a continuum of smoothly varying
+differences with overlap. This pattern makes identifying items with DIF
+quite difficult, as different methods are likely to yield inconsistent
+results (Stenhaug et al., 2021). Hence, we next characterized the
 distributions of the item-level group difficulty differences, and
@@
 
-\caption[GLIMMER plot of a sample of CDI:WS words from the sex bias model]{GLIMMER plot of a sample of CDI:WS words from the sex bias model.
-Words at the top are more well-known by females, while those at the bottom are more known for males.}\label{fig:glimmer-combo}
+\caption[GLIMMER plots of a sample of CDI:WS words from the sex bias model (left), the maternal education model (center), and the race model (right)]{GLIMMER plots of a sample of CDI:WS words from the sex bias model (left), the maternal education model (center), and the race model (right).
+Words at the top are more well-known by one group (e.g., females, for sex), while those at the bottom are more known by the other group (e.g., males), and the selection in the middle show roughly the median difficulty difference for those groups.}\label{fig:glimmer-combo}
 \end{figure*}
@@
 suggests that it is likely to reflect more than the effect of a small
-number of biased words (cf. Frank, Braginsky, Yurovsky, \& Marchman,
-2021).
+number of biased words (cf.~Frank et al., 2021).
 
@@
 ME-biased items are nouns (15/27). Notably, many of the items more known
-by high-ME children were animals (``zebra,'' ``owl'') and animal sounds
-(``baa baa,'' ``moo''). Items more known by low-ME children include a
-few sweet treats: ``candy,'' ``soda/pop,'' and ``gum.'' For race, 21
+by high-ME children were animals (``zebra'', ``owl'') and animal sounds
+(``baa baa'', ``moo''). Items more known by low-ME children include a
+few sweet treats: ``candy'', ``soda/pop'', and ``gum''. For race, 21
 extrema were identified, only 8 of which were easier for non-white
@@
 were nouns, but many others were early-learned words and phrases (e.g.,
-``up,'' ``all gone,'' ``uh-oh,'' ``bye,'' ``mommy''). Only 10 of the 59
+``up'', ``all gone'', ``uh-oh'', ``bye'', ``mommy''). Only 10 of the 59
 items were extreme in more than one model, with most of the overlap
@@
 \begin{CodeChunk}
-\begin{figure}[H]
+\begin{figure}[h]
 
-{\centering \includegraphics[width=\linewidth]{figs/extrema2p25sd}
+{\centering \includegraphics[width=\linewidth]{figs/extrema2p25sd_3sd_bold}
 
@@
 
-\caption[The 59 items showing extreme bias (difficulty difference $>2.25$ SD) for one or more demographic]{The 59 items showing extreme bias (difficulty difference $>2.25$ SD) for one or more demographic.}\label{fig:biased-words}
+\caption[The 59 items showing extreme bias (difficulty difference $>2.25$ SD) for one or more demographic]{The 59 items showing extreme bias (difficulty difference $>2.25$ SD) for one or more demographic.
+The 19 items in bold are those showing a difference $>3$ SD.}\label{fig:biased-words}
 \end{figure}
@@
 most prunings, with the majority of excluded extrema being
-stereotypically-male nouns (e.g, ``truck,'' ``fireman'').
+stereotypically-male nouns (e.g, ``truck'', ``fireman'').
 
@@
 that are equally difficult (i.e., unbiased) for both groups (for an
-overview, see Stenhaug, Frank, \& Domingue, 2021). But in the universe
-of children's early language, there is only a finite set of
-early-learned words to choose from -- and we may expect many of them to
-be biased for various environmental reasons (e.g., boys and girls know
-the names for their respective anatomy; children in Florida may not use
-mittens or skis). Hence, the presence of DIF on a set of items could
-indicate varying linguistic and environmental input, rather than an
-actual advantage for one demographic over another.
+overview, see Stenhaug et al., 2021). But in the universe of children's
+early language, there is only a finite set of early-learned words to
+choose from -- and we may expect many of them to be biased for various
+environmental reasons (e.g., boys and girls know the names for their
+respective anatomy; children in Florida may not use mittens or skis).
+Hence, the presence of DIF on a set of items could indicate varying
+linguistic and environmental input, rather than an actual advantage for
+one demographic over another.
 
@@
 predictive. For example, many of the extrema favoring high-ME children
-are animals and animal sounds (e.g., ``quack quack,'' ``woof woof,''
-``baa baa,'' ``duck,'' ``sheep,'' ``giraffe,'' ``zebra''): do high-SES
+are animals and animal sounds (e.g., ``quack quack'', ``woof woof'',
+``baa baa'', ``duck'', ``sheep'', ``giraffe'', ``zebra''): do high-SES
 households visit the zoo more often, or perhaps engage in other
@@
 predictive of children's early experience with (and thus knowledge of)
-CDI items related to winter weather (e.g., ``snow,'' ``mittens'') or
+CDI items related to winter weather (e.g., ``snow'', ``mittens'') or
 outdoors activities (e.g.~``camping''). A truly fair test of children's
@@
 \hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
 \leavevmode\hypertarget{ref-bornstein2016stability}{}%
@@
 \leavevmode\hypertarget{ref-R-mirt}{}%
-Chalmers, R. P. (2012). {mirt}: A multidimensional item response theory
-package for the {R} environment. \emph{Journal of Statistical Software},
-\emph{48}(6), 1--29.
-http://doi.org/\href{https://doi.org/10.18637/jss.v048.i06}{10.18637/jss.v048.i06}
+Chalmers, R. P. (2012). mirt: A multidimensional item response theory
+package for the R environment. \emph{Journal of Statistical Software},
+\emph{48}(6), 1--29. \url{http://doi.org/10.18637/jss.v048.i06}
 
@@
 \leavevmode\hypertarget{ref-fenson1994}{}%
-Fenson, Larry, Dale, P. S., Reznick, J. S., Bates, E., Thal, D. J.,
+Fenson, L., Dale, P. S., Reznick, J. S., Bates, E., Thal, D. J.,
 Pethick, S. J., \ldots{} Stiles, J. (1994). Variability in early
@@
 \leavevmode\hypertarget{ref-macwhinney2000childes}{}%
-MacWhinney, B. (2000). \emph{The CHILDES project: Tools for analyzing
+MacWhinney, B. (2000). \emph{The childes project: Tools for analyzing
 talk. Transcription format and programs} (Vol. 1). Psychology Press.
@@
 Rowe, M. L., \& Goldin-Meadow, S. (2009). Differences in early gesture
-explain SES disparities in child vocabulary size at school entry.
+explain ses disparities in child vocabulary size at school entry.
 \emph{Science}, \emph{323}(5916), 951--953.

### 2022-02-03 GLIMMER_production.R
look at overlap of biased items

@@ -262,9 +262,28 @@ AOAA_OAT <- function(d_mat, group, fname=c()) {
   return(list(model=constrained, items_removed=items_removed, dif_stats=difm))
 }
 
-ses_aoaa <- AOAA_OAT(d_mat, ses_group, "ses")
-sex_aoaa <- AOAA_OAT(d_mat, sex_group, "sex")
-race_aoaa <- AOAA_OAT(d_mat, eth_group, "race")
+ses_aoaa <- AOAA_OAT(d_mat, ses_group, "ses") # removed 314 items
+sex_aoaa <- AOAA_OAT(d_mat, sex_group, "sex") # removed 170 items
+race_aoaa <- AOAA_OAT(d_mat, eth_group, "race") # removed 318 items
+
+intersect(ses_aoaa$items_removed$word, sex_aoaa$items_removed$word) # 90
+
+require(VennDiagram)
+library(RColorBrewer)
+
+venn.diagram(
+  x = list(ses_aoaa$items_removed$word, 
+           sex_aoaa$items_removed$word, 
+           race_aoaa$items_removed$word),
+  fill=brewer.pal(3, "Pastel2"),
+  category.names = c("Maternal Ed" , "Sex" , "Race"),
+  cat.cex = 1.1,
+  cat.fontface = "bold",
+  cex = 1.2,
+  fontface = "bold",
+  filename = 'AOAA-OAT_venn_diagram.png',
+  output=TRUE
+)
 
 # drop uses a constrained model (e.g. same group means) and then tests whether dropping
 #dif_itemN = DIF(mod_intuitive_sex, which.par=c('d'), scheme='drop', items2test=1:2)

### 2022-02-24 paper/cogsci_word.Rmd
added funding statement

--- 2022-02-01 paper/cogsci.Rmd
@@
 
-output: cogsci2016::cogsci_paper
+output: word_document
 #final-submission: \cogscifinalcopy
@@
 [Redacted for anonymous review.]
-<!--We thank members of the Language and Cognition lab for their feedback.-->
+<!--We thank members of the Language and Cognition lab for their feedback.
+This research was supported by the Stanford Maternal and Child Health Research Institute.  
+-->
 

### 2022-05-05 paper/cogsci_final.Rmd
CogSci revisions

--- 2022-02-01 paper/cogsci.Rmd
@@
 author-information: > 
-    \author{{\large \bf George Kachergis (kachergis@stanford.edu)}
-    \AND {\large \bf Nathan Francis (nathan99@stanford.edu)}
-    \AND {\large \bf Michael C. Frank (mcfrank@stanford.edu)} \\ Department of Psychology, Stanford Unviersity \\ Stanford, CA 94305 USA
+    \author{{\large \bf George Kachergis, Nathan Francis, and Michael C. Frank}
+    \AND {\large \tt  \{kachergis, nathan99, mcfrank\} @stanford.edu} \\ Department of Psychology, Stanford Unviersity \\ Stanford, CA 94305 USA
     }
@@
     (high vs. low). For each dimension, we identified dozens of strongly biased items, and showed that 
-    eliminating these items reduced the expected ability difference between groups. Additionally,
-    we investigated how well the relative frequency of words spoken to young girls vs. boys predicted 
-    sex-based word learning bias, and discuss possible sources of demographic differences in early word learning.
+    eliminating even some of these items can reduce the magnitude of differences between groups. 
+    Additionally, we investigated how well the relative frequency of words spoken to young girls vs.
+    boys predicted sex-based word learning bias, and discuss possible sources of demographic
+    differences in early word learning.
 
@@
 output: cogsci2016::cogsci_paper
-#final-submission: \cogscifinalcopy
+final-submission: \cogscifinalcopy
 ---
@@
 
-CDIs are a useful tool for measuring demographic effects because of their wide use across languages. 
+CDIs are a useful tool for measuring demographic effects because of their ease of use and wide distribution to different groups. 
 But their use in this context poses a chicken and egg problem. 
-While it is possible that there is an underlying difference in mean language ability between demographic groups, CDIs could also have bias for or against particular groups based on the words they include.
+While it is possible that there is an underlying difference in mean language ability (or vocabulary size) between demographic groups, CDIs could also have bias for or against particular groups based on the particular words they include, which different groups may have a varying amount of exposure to.
 
@@
 a male child and the size of the female (vs.
-male) advantage for that CDI word was modest, but significant ($r=0.18$, $p<.001$).
+male) advantage for that CDI word was weak, but significant ($r=0.18$, $p<.001$).
 Some of the sex bias seen in CDI items is related to differences in language input received by girls vs. boys -- though of course this differential input could be elicited due to different interests.
@@
 This investigation is only a first step in measuring demographic bias in the items on the CDI:WS.
-More research is needed to determine whether CDI items that show DIF are more related to environmental frequencies, as we found only a modest association between the relative frequency of words in child-directed speech to boys vs.
+More research is needed to determine whether CDI items that show DIF are more related to environmental frequencies, as we found a weak though reliable association between the relative frequency of words in child-directed speech to boys vs.
 girls and the amount of sex-based DIF.
@@
 
-[Redacted for anonymous review.]
-<!--We thank members of the Language and Cognition lab for their feedback.-->
+We thank members of the Language and Cognition lab and three anonymous reviewers for their feedback.
 

### 2022-05-27 paper/cogsci_final.Rmd
minor edit
final render

@@
 tokens) have often been predictive of language learning outcomes in  demographic studies [@rowe2009differences; @huttenlocher1991early].
-We used a corpus-based analysis to investigate the extent to which word frequency in child-directed speech to male vs.
+In an attempt to investigate this, we analyzed the extent to which word frequency in a corpus of child-directed speech to male vs.
 female children was predictive of the amount of sex-based bias shown by CDI items.
@@
 
-We thank members of the Language and Cognition lab and three anonymous reviewers for their feedback.
+This research was supported by the Stanford Maternal and Child Health Research Institute. We thank members of the Language and Cognition lab and three anonymous reviewers for their feedback.
 

### 2022-09-07 paper/manuscript.Rmd
adding APA manuscript for topiCS submission

--- 2022-05-27 paper/cogsci_final.Rmd
@@
 ---
-title: "Estimating demographic bias on tests of children's early vocabulary"
-bibliography: library.bib
-csl: apa6.csl
-document-params: "10pt, letterpaper"
+title             : "Estimating demographic bias on tests of children's early vocabulary"
+shorttitle        : "Demographic bias on children's vocabulary tests"
 
-author-information: > 
-    \author{{\large \bf George Kachergis, Nathan Francis, and Michael C. Frank}
-    \AND {\large \tt  \{kachergis, nathan99, mcfrank\} @stanford.edu} \\ Department of Psychology, Stanford Unviersity \\ Stanford, CA 94305 USA
-    }
+author: 
+  - name          : "George Kachergis"
+    affiliation   : "1"
+    corresponding : yes    # Define only one corresponding author
+    address       : "450 Jane Stanford Way"
+    email         : "gkacherg@stanford.edu"
+    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
+      - "Conceptualization"
+      - "Writing - Original Draft Preparation"
+      - "Writing - Review & Editing"
+      - "Supervision"
+  - name          : "Nathan Francis"
+    affiliation   : "1"
+    role:
+      - "Writing - Review & Editing"
+  - name          : "Michael C. Frank"
+    affiliation   : "1"
+    role:
+      - "Conceptualization"
+      - "Writing - Original Draft Preparation"
+      - "Writing - Review & Editing"
+      - "Supervision"
 
-abstract: >
+affiliation:
+  - id            : "1"
+    institution   : "Stanford University"
+#  - id            : "2"
+#    institution   : "Konstanz Business School"
+
+authornote: |
+  Department of Psychology
+
+abstract: |
   Children's early language skill has been linked to later educational outcomes, making
@@
 
-keywords: >
-    language acquisition; word learning; measuring instrument bias; demographic bias
+keywords          : "language acquisition; word learning; measuring instrument bias; demographic bias"
+wordcount         : "4008"
 
-output: cogsci2016::cogsci_paper
-final-submission: \cogscifinalcopy
+bibliography      : "library.bib"
+
+floatsintext      : no
+linenumbers       : yes
+draft             : no
+mask              : no
+
+figurelist        : no
+tablelist         : no
+footnotelist      : no
+
+classoption       : "man"
+output            : papaja::apa6_pdf
 ---
 
-```{r global_options, include=FALSE}
-knitr::opts_chunk$set(fig.width=3, fig.height=3, fig.crop = F, 
-                      fig.pos = "tb", fig.path='figs/',
-                      echo=F, warning=F, cache=TRUE, 
-                      message=F, sanitize = T)
+```{r setup, include = FALSE}
+library("papaja")
+r_refs("library.bib")
 ```
@@
 
-# Introduction
+```{r analysis-preferences}
+# Seed for random number generation
+set.seed(42)
+knitr::opts_chunk$set(cache.extra = knitr::rand_seed)
+```
 
@@
 CDIs have shown good predictive validity [e.g., @fenson1994; @bornstein2012stability; @duff2015infant].
-Our primary focus is the vocabulary checklist portion of the CDI Words & Sentences (CDI:WS) form, comprised of 680 early-learned words across 22 categories (e.g., animals, vehicles, action words, pronouns) selected to assess the productive vocabulary of children 16 to 30 months of age.
+Our primary focus is the vocabulary checklist portion of the CDI Words & Sentences (CDI:WS) form, comprised of 680 early-learned words across 22 categories (e.g., animals, vehicles, action words, pronouns) selected to assess the productive vocabulary of children 16 to 30 months of age.[^1]
 For each item on the CDI:WS, caregivers are asked to respond whether the target child has been heard to say (i.e. produce) the given item.
@@
 
-<!-- [^1]: The CDI: Words and Gestures form includes a subset of the CDI:WS items and targets children 12 to 18 months of age, measuring both comprehension and production. -->
+[^1]: The CDI: Words and Gestures form includes a subset of the CDI:WS items and targets children 12 to 18 months of age, measuring both comprehension and production.
 
@@
 Next, we use these item parameters to systematically prune biased items from the CDI:WS, and examine how the estimated size of demographic differences change as we prune more items.
-<!-- We identify a set of suspect items to eliminate (or in the future, replace), and provide updated estimates of the effect size of these demographic variables, were the biased items to be eliminated.-->
 Finally, using a corpus of child-directed speech, we examine the extent to which variation in word frequencies heard by boys vs. girls predicts sex-based DIF.
@@
 @frank2021].
-<!-- be a real ability advantage, unless the bulk of these items are actually spoken more often to female children than to male children. -->
 
@@
 
+\newpage
+
 # References
 
-```{r}
-# References will be generated automatically by Pandoc and included here.
-# The following code is some latex to format the bibliography. Do not remove it.
-```
-
-\setlength{\parindent}{-0.1in} 
-\setlength{\leftskip}{0.125in}
-\noindent
+::: {#refs custom-style="Bibliography"}
+:::

### 2022-09-08 paper/manuscript.Rmd
readying topiCS submission

@@
     affiliation   : "1"
-    role:
-      - "Conceptualization"
-      - "Writing - Original Draft Preparation"
-      - "Writing - Review & Editing"
-      - "Supervision"
+#    role:
+#      - "Conceptualization"
+#      - "Writing - Original Draft Preparation"
+#      - "Writing - Review & Editing"
+#      - "Supervision"
 
@@
     institution   : "Stanford University"
-#  - id            : "2"
-#    institution   : "Konstanz Business School"
 
@@
 
-We focus here on the MacArthur-Bates Communicative Development Inventories [CDIs; @fenson1994; @Fenson2007], a set of parent-reported measures of children's productive and receptive language skills, which offer a low-cost and reliable way to estimate children's early language skills [@fenson1994].
+We focus here on the MacArthur-Bates Communicative Development Inventories (CDIs) [@fenson1994; @Fenson2007], a set of parent-reported measures of children's productive and receptive language skills, which offer a low-cost and reliable way to estimate children's early language skills [@fenson1994].
 CDIs have shown good predictive validity [e.g., @fenson1994; @bornstein2012stability; @duff2015infant].
@@
 
-```{r baseline-ability-vs-age-hor, fig.env = "figure*", out.width="\\linewidth", fig.pos = "t", fig.align='center', fig.width=6.5, fig.height=2.8, set.cap.width=T, num.cols.cap=2, fig.cap = "Language ability vs.
+```{r baseline-ability-vs-age-hor, fig.env = "figure*", out.width="\\linewidth", fig.pos = "t", fig.align='center', fig.width=6.5, fig.height=3, set.cap.width=T, num.cols.cap=2, fig.cap = "Language ability vs.
 age by demographic group, from the baseline Rasch model."}
@@
   ggplot(aes(x=jitter(age), y=ability, color=sex)) + 
-  geom_point(alpha=.1) + geom_smooth() + theme_classic(base_size=9) + 
+  geom_point(alpha=.1) + geom_smooth() + theme_classic(base_size=10) + 
   xlab("Age (months)") + ylab("Language Ability") + 
@@
   ggplot(aes(x=jitter(age), y=ability, color=ses_group)) + 
-  geom_point(alpha=.1) + geom_smooth() + theme_classic(base_size=9) + 
+  geom_point(alpha=.1) + geom_smooth() + theme_classic(base_size=10) + 
   xlab("Age (months)") + ylab("Language Ability") +
@@
   ggplot(aes(x=jitter(age), y=ability, color=eth_group)) + 
-  geom_point(alpha=.1) + geom_smooth() + theme_classic(base_size=9) + 
+  geom_point(alpha=.1) + geom_smooth() + theme_classic(base_size=10) + 
   xlab("Age (months)") + ylab("Language Ability") + 
@@
 
-<!--
-## Recommended Pruning Threshold: 2SD
-Considering just those items whose difficulty was >2SD from the mean difficulty difference identified 25 items with extreme sex-based difficulty differences, only 7 of which were easier for females, with the other 18 items favoring males.
-For SES, 39 extrema were identified, only 13 of which were easier for low-SES children.
-For race, 31 extrema were identified, only 11 of which were easier for non-white children.
-Thus, while if anything the bulk of the extrema in the sex-based model advantaged the lower language ability group (males), the extrema in the other two models mostly favored the advantaged groups: 26/39 extreme items in the SES model favored high-SES children, and 20/31 items in the race model favored white children.
-17 CDI items were identified as extrema in more than one model: "tractor" and "vroom" were extreme in all three models, "choo choo" was extreme in both SES and sex models, "give me five!" was extreme in both race and sex models, and 13 other items were extreme in both SES and race models ("grrr", "moo", "quack quack", "uh oh", "duck", "owl", "candy", "gum", "walker", "daddy", "pet's name", "up", and "so").
-In our final analysis, we explore the effect of pruning the 15 items that were extreme in both SES and race models.
--->
-
 ```{r items-to-remove}
@@
 We characterize these 59 potentially biased items below, and consider whether we might recommend pruning them from the CDI:WS.
+
+<!--
+## Recommended Pruning Threshold: 2SD
+Considering just those items whose difficulty was >2SD from the mean difficulty difference identified 25 items with extreme sex-based difficulty differences, only 7 of which were easier for females, with the other 18 items favoring males.
+For SES, 39 extrema were identified, only 13 of which were easier for low-SES children.
+For race, 31 extrema were identified, only 11 of which were easier for non-white children.
+Thus, while if anything the bulk of the extrema in the sex-based model advantaged the lower language ability group (males), the extrema in the other two models mostly favored the advantaged groups: 26/39 extreme items in the SES model favored high-SES children, and 20/31 items in the race model favored white children.
+17 CDI items were identified as extrema in more than one model: "tractor" and "vroom" were extreme in all three models, "choo choo" was extreme in both SES and sex models, "give me five!" was extreme in both race and sex models, and 13 other items were extreme in both SES and race models ("grrr", "moo", "quack quack", "uh oh", "duck", "owl", "candy", "gum", "walker", "daddy", "pet's name", "up", and "so").
+In our final analysis, we explore the effect of pruning the 15 items that were extreme in both SES and race models.
+-->
 
@@
 If certain activities are driving early word learning for high-ME children, what are the activities (and associated vocabulary) that low-SES households are instead engaging in?
-Critically this enterprise will require a greater diversity of publicly-available corpora of child language; we view the lack of such data as a substantial issue.
+Critically, this enterprise will require a greater diversity of publicly-available corpora of child language; we view the lack of such data as a substantial issue.
 

### 2022-09-08 paper/manuscript.Rmd
handle warnings, figures in text

@@
 
-floatsintext      : no
+floatsintext      : yes
 linenumbers       : yes
@@
 Children's total vocabulary score on the CDI:WS is tightly correlated with other facets of early language (e.g., grammatical competence and gesture), suggesting that the language system is "tightly woven" [@frank2021].
-Due to these desirable properties, CDIs have been adapted to dozens of languages, and a central repository of CDI data contributed from all over the world has been created [Wordbank; @frank2017; @frank2021].
+Due to these desirable properties, CDIs have been adapted to dozens of languages, and a central repository of CDI data contributed from all over the world has been created (Wordbank) [@frank2017; @frank2021].
 
@@
 
-```{r baseline-ability-vs-age-hor, fig.env = "figure*", out.width="\\linewidth", fig.pos = "t", fig.align='center', fig.width=6.5, fig.height=3, set.cap.width=T, num.cols.cap=2, fig.cap = "Language ability vs.
+```{r baseline-ability-vs-age-hor, warning=F, out.width="\\linewidth", fig.align='center', fig.width=6.5, fig.height=3, fig.cap = "Language ability vs.
 age by demographic group, from the baseline Rasch model."}
+# fig.env = "figure*", fig.pos = "t", set.cap.width=T, num.cols.cap=2,
 p1 <- d_demo %>% filter(!is.na(sex)) %>%
@@
 
-```{r glimmer-combo, fig.env = "figure*", out.width="\\linewidth", fig.pos = "h", fig.width=6.5, fig.align = "center", set.cap.width=T, num.cols.cap=2, fig.cap = "GLIMMER plots of a sample of CDI:WS words from the sex bias model (left), the maternal education model (center), and the race model (right).
+```{r glimmer-combo, out.width="\\linewidth", fig.width=6.5, fig.align = "center", fig.cap = "GLIMMER plots of a sample of CDI:WS words from the sex bias model (left), the maternal education model (center), and the race model (right).
 Words at the top are more well-known by one group (e.g., females, for sex), while those at the bottom are more known by the other group (e.g., males), and the selection in the middle show roughly the median difficulty difference for those groups."}
@@
 
-```{r pruned-coefs-hor, fig.env = "figure*", out.width="\\linewidth", fig.pos = "h", fig.align='center', fig.width=6.5, fig.height=2.8, set.cap.width=T, num.cols.cap=2, fig.cap = "Size of demographic effects (regression coefficients) with different pruning thresholds."}
+```{r pruned-coefs-hor, out.width="\\linewidth", fig.width=6.5, fig.height=3, fig.cap = "Size of demographic effects (regression coefficients) with different pruning thresholds."}
 load(here("data/pruned_demo_coefs.Rdata"))
@@
 
-```{r biased-words, include=T, out.width="\\linewidth", fig.env = "figure", fig.pos = "h", fig.align='center', fig.width=3.5, fig.height=3.2, set.cap.width=T, num.cols.cap=1, fig.cap = "The 59 items showing extreme bias (difficulty difference $>2.25$ SD) for one or more demographic.
+```{r biased-words, include=T, out.width=".7\\linewidth", fig.align='center', fig.width=3.5, fig.cap = "The 59 items showing extreme bias (difficulty difference $>2.25$ SD) for one or more demographic.
 The 19 items in bold are those showing a difference $>3$ SD."}

### 2022-09-13 paper/manuscript.Rmd
Mike's edits incorporated

@@
     role:
+      - "Writing - Original Draft Preparation"
       - "Writing - Review & Editing"
@@
     affiliation   : "1"
-#    role:
-#      - "Conceptualization"
-#      - "Writing - Original Draft Preparation"
-#      - "Writing - Review & Editing"
-#      - "Supervision"
+    role:
+      - "Conceptualization"
+      - "Writing - Original Draft Preparation"
+      - "Writing - Review & Editing"
+      - "Supervision"
 
@@
 keywords          : "language acquisition; word learning; measuring instrument bias; demographic bias"
-wordcount         : "4008"
+wordcount         : "3988"
 
@@
 One challenge in comprehensive assessment of these differences is that it is difficult to get a complete measure of young children's language skill.
-Long recordings are prohibitively difficult to collect and transcribe, and yet any short recording (e.g., a 1-hour play session) will elicit only a small proportion of the words and constructions that children know.
+Long recordings are prohibitively difficult to collect and transcribe, yet any short recording (e.g., a 1-hour play session) will elicit only a small proportion of the words and constructions that children know.
 Thus, researchers of early word learning have constructed tests with hundreds of words, intentionally oversampling words that are more likely to be known by young children.
@@
 
-<!--Inspired by the utility and widespread use of the CDI, researchers have recently been using psychometric models on CDI data to construct short, adaptive tests to reliably assess language ability using only a small subset of the CDI items [e.g., @mayor2019; @kachergis2021cat].
+<!--Inspired by the utility and widespread use of the CDI, researchers have recently been using psychometric models on CDI data to construct short, adaptive tests to reliably assess language ability using only a small subset of the CDI items [e.g., @mayor2019; @kachergis2022cat].
 These psychometric models typically come from the Item-Response Theory (IRT) framework [@Baker2001], which assumes that not only test-takers (here, children being reported on) have normally-distributed ability, but that items (words on the CDI) have normally-distributed difficulty.
@@
 
-The idea that some items on a test may show bias, favoring one group over another, is known in psychometrics as Differential Item Function (DIF; @holland1993differential]. 
+The idea that some items on a test may show bias, favoring one group over another, is known in psychometrics as Differential Item Function (DIF) [@holland1993differential]. 
 DIF can decrease the validity of a test, as the test may be overestimating the ability of test-takers in one group and underestimating the ability of another group -- not because of any underlying mean difference in ability between the groups, but simply because the test is unfair [@camilli2013ongoing].
 For example, if a vocabulary test is composed almost entirely of the names of farm equipment, the test will underestimate the knowledge of urban children, who have experience of other contexts, despite being less familiar with farm equipment than their rural peers.
-Of course, there may be an ability difference between demographic groups, as is likely the case for the female language advantage, which has been documented in many languages [@eriksson2012differences; @frank2021].
-And both observations could be true -- a test could be biased to inflate a true demographic difference.
+Of course, there may be a true ability difference between demographic groups, as is likely the case for the female language advantage, which has been documented in many languages [@eriksson2012differences; @frank2021].
+And both observations could be true -- a test could be biased to inflate *or* deflate a true demographic difference.
 
 Our goal in the current study is to use DIF analyses to assess these possibilities. 
-Specifically, we test the 680 vocabulary items on the American CDI:WS for DIF along three dichotomized demographic dimensions: sex (male vs.
+We focus on the American English data from Wordbank due to its large size and due to the presence of demographic data for our variables of interest.
+Specifically, we test the 680 vocabulary items on the American English CDI:WS for DIF along three dichotomized demographic dimensions: sex (male vs.
 female), maternal education (no more than secondary vs.
@@
 First, we introduce the Wordbank data and the IRT model, and examine the overall size of demographic differences in early word learning based on the full CDI:WS.
-We then fit the IRT model to both groups for each demographic factor, and look at the item parameters for evidence of DIF, noting in particular how many items are significantly biased in favor of each group, and describing them qualitatively.
+We then fit the IRT model to both groups for each demographic factor and examine the item parameters for evidence of DIF.
 Next, we use these item parameters to systematically prune biased items from the CDI:WS, and examine how the estimated size of demographic differences change as we prune more items.
 Finally, using a corpus of child-directed speech, we examine the extent to which variation in word frequencies heard by boys vs. girls predicts sex-based DIF.
-Based on these analyses, we provide recommendations for next steps to be taken to identify and potentially replace items on the CDI showing demographic bias.
+Based on these analyses, we provide recommendations for next steps to be taken to identify and potentially replace items on the CDI that show demographic bias.
 
@@
 
-The Rasch model, also known as the 1-parameter logistic (1PL) model, is the simplest Item Response Theory model, and is thus the easiest to use to investigate potential differences in item function across different groups of participants.
+The Rasch model, also known as the 1-parameter logistic (1PL) model, is the simplest Item Response Theory model.
+Although we have elsewhere shown that the 2-parameter logistic model often fits CDI data better [@kachergis2022cat], the simpler Rasch model is easier to use to evaluate potential differences in item function across different groups of participants.
 The Rasch model jointly estimates for each child $j$ a latent ability $\theta_j$, and for each item $i$ a difficulty parameter $b_i$.
@@
 Figure 2 shows GLIMMERs for a selection of CDI items for sex (left), maternal education (middle), and race (right).
-The full GLIMMERs, with all 680 CDI:WS items, are available on [OSF](https://osf.io/57rsw/?view_only=2b6ecb61fe08458293af7421d276932a), but were too large to include here.
-It is important, however, to inspect the full plots, for if there is a cluster of items in a GLIMMER, the analyst may conclude that these items are strong candidates for DIF on that dimension.
-For example, there is some clustering at the top and bottom of the sex GLIMMER (Fig.
-2, left): at the top, "vagina"^[Parents using the CDI are instructed to check genital items if the child produces whatever word the family uses.], "tights", "dress (object)", and "doll" form a cluster of items that are much more well-known for females, while at the bottom, "penis" stands out as much more well-known by males.
+The full GLIMMERs, with all 680 CDI:WS items, are large but can be found at OSF^[1], but were too large to include here.
+Inspecting GLIMMER plots can reveal clusters of items, indicating candidates for DIF on that dimension.
+For example, there is some clustering at the top and bottom of the sex GLIMMER (Figure 2, left): at the top, "vagina"^[Parents using the CDI are instructed to check genital items if the child produces whatever word the family uses.], "tights", "dress (object)", and "doll" form a cluster of items that are much more well-known for females, while at the bottom, "penis" stands out as much more well-known by males.
 For the maternal education and race GLIMMERs -- and in the rest of the full sex GLIMMER, there are not clusters, but rather a continuum of smoothly varying differences with overlap. 
@@
 Hence, we next characterized the distributions of the item-level group difficulty differences, and measured the influence of pruning a varying number of items on the demographic effect sizes.
+
+^[1]: [https://osf.io/57rsw/](https://osf.io/57rsw/?view_only=2b6ecb61fe08458293af7421d276932a)
 
@@
 Similar to the approach taken by @braginsky2016gender, we used the CHILDES corpus of transcripts from dyadic play sessions [@macwhinney2000childes], which are labeled with the sex of the target child, but not other demographic variables.
-We found a total of 5213750 female-directed tokens and 6091950 male-directed tokens that matched 662 of the 680 CDI:WS items, normalized the word frequencies to tokens per million for each target sex, and calculated the log-odds of each word being spoken to a female (i.e., $log(p_{f} / p_{m}$), meaning unbiased words will have log-odds of 0, while those spoken more often to females will have log-odds $>0$, and those spoken more often males will have log-odds $<0$.
+We found a total of 5,213,750 female-directed tokens and 6,091,950 male-directed tokens that matched 662 of the 680 CDI:WS items, normalized the word frequencies to tokens per million for each target sex, and calculated the log-odds of each word being spoken to a female (i.e., $log(p_{f} / p_{m}$), meaning unbiased words will have log-odds of 0, while those spoken more often to females will have log-odds $>0$, and those spoken more often males will have log-odds $<0$.
 For example "doll" was spoken 156 times to females, and 138 times to males, thus $p_{f} = 156/(156+138) = .53$, $p_{m} = 1-p_{f} = 0.47$, and $log(p_{f}/p_{m})$ = 0.12), while "police" was spoken 138 times to females, and 225 times to males, and thus has log-odds = -0.49).

### 2022-09-14 paper/manuscript.Rmd
credit and abstract edits

@@
       - "Conceptualization"
+      - "Methodology"
+      - "Software"
+      - "Investigation"
+      - "Data Curation"
       - "Writing - Original Draft Preparation"
       - "Writing - Review & Editing"
+      - "Visualization"
       - "Supervision"
@@
     role:
+      - "Software"
+      - "Investigation"
       - "Writing - Original Draft Preparation"
@@
       - "Conceptualization"
+      - "Methodology"
       - "Writing - Original Draft Preparation"
@@
       - "Supervision"
+      - "Funding acquisition"
 
@@
   dimensions of sex (male vs. female), race (white vs. non-white), and maternal education 
-  (high vs. low). For each dimension, we identified dozens of strongly biased items, and showed that 
-  eliminating even some of these items can reduce the magnitude of differences between groups. 
+  (high vs. low). For each dimension, many items showed bias; removing these items slightly
+  reduced the magnitude of group differences, but did not eliminate them. 
   Additionally, we investigated how well the relative frequency of words spoken to young girls vs.

### 2022-09-14 paper/manuscript.Rmd
fixed citation bracket issues

@@
 Children's total vocabulary score on the CDI:WS is tightly correlated with other facets of early language (e.g., grammatical competence and gesture), suggesting that the language system is "tightly woven" [@frank2021].
-Due to these desirable properties, CDIs have been adapted to dozens of languages, and a central repository of CDI data contributed from all over the world has been created (Wordbank) [@frank2017; @frank2021].
+Due to these desirable properties, CDIs have been adapted to dozens of languages, and a central repository of CDI data contributed from all over the world has been created [Wordbank\; @frank2017].
 
@@
 
-The idea that some items on a test may show bias, favoring one group over another, is known in psychometrics as Differential Item Function (DIF) [@holland1993differential].
+The idea that some items on a test may show bias, favoring one group over another, is known in psychometrics as Differential Item Function [DIF\; @holland1993differential].
 DIF can decrease the validity of a test, as the test may be overestimating the ability of test-takers in one group and underestimating the ability of another group -- not because of any underlying mean difference in ability between the groups, but simply because the test is unfair [@camilli2013ongoing].
